### 2022-11-15 docs/papers/adjoint/adjoint.tex
Add adjoint method paper source and scripts

\documentclass[modern, trackchanges, dvipsnames]{aastex631}
  \urlstyle{sf}

\usepackage{CJK}
\usepackage{microtype}
\usepackage{savesym}
  \savesymbol{lambdabar}
\usepackage{newpxtext, eulerpx}
  \restoresymbol{newpx}{lambdabar}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
  %\allowdisplaybreaks
\usepackage{bm}
\usepackage{nicefrac}
\usepackage[caption=false]{subfig}
\usepackage{tikz}
  \usetikzlibrary{positioning, fit, calc, arrows.meta}
\usepackage{booktabs}

\newcommand{\pmwd}{{\usefont{T1}{nova}{m}{sl}pmwd}}

\newcommand{\gkai}[1]{\begin{CJK*}{UTF8}{gkai}\raisebox{.1em}{(}#1\raisebox{.1em}{)}\end{CJK*}}

\renewcommand{\sectionautorefname}{Sec.}
\renewcommand{\subsectionautorefname}{Sec.}
\renewcommand{\appendixautorefname}{App.}
\renewcommand{\figureautorefname}{Fig.}
\newcommand{\subfigureautorefname}{\figureautorefname}

\newcommand{\deltaD}{\delta^\textsc{d}}
\newcommand{\deltaK}{\delta^\textsc{k}}
\renewcommand{\d}{d}
\newcommand{\p}{\partial}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cH}{\mathcal{H}}
\bmdefine{\vzero}{0}
\bmdefine{\vI}{I}
\bmdefine{\vnabla}{\nabla}
\bmdefine{\vtheta}{\theta}  % parameters
\bmdefine{\vomega}{\omega}  % white noise modes
\bmdefine{\vk}{k}  % wavevectors
\bmdefine{\vx}{x}  % comoving and canonical coordinates
\bmdefine{\vq}{q}  % Lagrangian coordinates
\bmdefine{\vs}{s}  % displacements
\bmdefine{\vp}{p}  % canonical momenta
\bmdefine{\va}{a}  % accelerations
\bmdefine{\vz}{z}  % states
\bmdefine{\vf}{f}
\bmdefine{\vF}{F}
\bmdefine{\vDelta}{\Delta}
\bmdefine{\vLambda}{\Lambda}
\bmdefine{\vlambda}{\lambda}
\bmdefine{\vvarphi}{\varphi}
\bmdefine{\vxi}{\xi}  % x adjoint
\bmdefine{\vpi}{\pi}  % p adjoint
\bmdefine{\valpha}{\alpha}  % force vjp x gradient
\bmdefine{\vzeta}{\zeta}  % force vjp theta gradient
\newcommand{\half}{\nicefrac12}
\newcommand{\As}{A_\mathrm{s}}
\newcommand{\ns}{n_\mathrm{s}}
\newcommand{\Omegam}{\Omega_\mathrm{m}}
\newcommand{\Omegab}{\Omega_\mathrm{b}}
\newcommand{\Mpc}{\mathrm{Mpc}}
\newcommand{\ic}{\mathrm{i}}
\newcommand{\Plin}{P_\mathrm{lin}}

\newcommand{\GPU}{NVIDIA A100 80GB SXM4}

\newcommand{\HL}[1]{\textcolor{Bittersweet}{#1}}

\begin{document}

\title{\large Differentiable Cosmological Simulation with Adjoint Method
\vspace{0.3em}}

\author[0000-0002-0701-1410]{\normalsize Yin Li \gkai{李寅}}
\affiliation{Department of Mathematics and Theory, Peng Cheng
Laboratory, Shenzhen, Guangdong 518066, China}
\affiliation{Center for Computational Mathematics, Flatiron Institute,
New York, New York 10010, USA}
\affiliation{Center for Computational Astrophysics, Flatiron Institute,
New York, New York 10010, USA}

\author[0000-0002-1670-2248]{\normalsize Chirag Modi}
\affiliation{Center for Computational Mathematics, Flatiron Institute,
New York, New York 10010, USA}
\affiliation{Center for Computational Astrophysics, Flatiron Institute,
New York, New York 10010, USA}

\author[0000-0001-5044-7204]{\normalsize Drew Jamieson}
\affiliation{Max Planck Institute for Astrophysics, 85748 Garching bei
M\"unchen, Germany}

\author[0000-0002-9300-2632]{\normalsize Yucheng Zhang \gkai{张宇澄}}
\affiliation{Department of Mathematics and Theory, Peng Cheng
Laboratory, Shenzhen, Guangdong 518066, China}
\affiliation{Department of Physics, New York University, New York, New
York 10003, USA}

\author[0000-0003-0745-9431]{\normalsize Libin Lu \gkai{陆利彬}}
\affiliation{Center for Computational Mathematics, Flatiron Institute,
New York, New York 10010, USA}

\author[0000-0001-5590-0581]{\normalsize Yu Feng \gkai{冯雨}}
\affiliation{Berkeley Center for Cosmological Physics, University of
California, Berkeley, California 94720, USA}

\author[0000-0001-7956-0542]{\normalsize Fran\c{c}ois Lanusse}
\affiliation{AIM, CEA, CNRS, Universit\'e Paris-Saclay, Universit\'e
Paris Diderot,\\ Sorbonne Paris Cit\'e, F-91191 Gif-sur-Yvette, France}

\author[0000-0003-2895-8715]{\normalsize Leslie Greengard}
\affiliation{Center for Computational Mathematics, Flatiron Institute,
New York, New York 10010, USA}
\affiliation{Courant Institute, New York University, New York, New York
10012, USA}

\shorttitle{Differentiable Simulation with Adjoint Method}
\shortauthors{Li et al.}

\correspondingauthor{Yin Li}
\email{eelregit@gmail.com}

\begin{abstract}

Rapid advances in deep learning have brought not only myriad powerful
neural networks, but also breakthroughs that benefit established
scientific research.
In particular, automatic differentiation (AD) tools and computational
accelerators like GPUs have facilitated forward modeling of the Universe
with differentiable simulations.
Current differentiable cosmological simulations are limited by memory,
thus are subject to a trade-off between time and space/mass resolution.
They typically integrate for only tens of time steps, unlike the
standard non-differentiable simulations.
We present a new approach free of such constraints, using the adjoint
method and reverse time integration.
It enables larger and more accurate forward modeling, and will improve
gradient based optimization and inference.
We implement it in a particle-mesh (PM) $N$-body library \pmwd\
(particle-mesh with derivatives).
Based on the powerful AD system \texttt{JAX}, \pmwd\ is fully
differentiable, and is highly performant on GPUs.

\end{abstract}

\vspace{1em}
\section{Introduction}

Current established workflows of statistical inference from cosmological
datasets involve reducing cleaned data to summary statistics like the
power spectrum, and predicting these statistics using perturbation
theories, semi-analytic models, or simulation-calibrated emulators.
These can be suboptimal due to the limited model fidelity and the
risk of information loss in data compression.
Cosmological simulations \citep{HockneyEastwood1988, AnguloHahn2022} can
accurately predict structure formation even in the nonlinear regime at
the level of the fields.
Using simulations as forward models also naturally accounts for the
cross-correlation of different observables, and can easily incorporate
systematic errors.
This approach has been intractable due to the large computational costs
on conventional CPU clusters, but rapid advances in accelerator
technology like GPUs opens the possibility of simulation-based modeling
and inference \citep{CranmerEtAl2020}.
Furthermore, model differentiability enabled by AD libraries can
accelerate parameter constraint with gradient-based optimization and
inference.
A differentiable field-level forward model combining these two features
is able to constrain physical parameters together with the initial
conditions of the Universe.

The first differentiable cosmological simulations, such as BORG and
ELUCID \citep{BORG, ELUCID}, were developed before the advent of modern
AD systems, and were based on implementations of analytic derivatives
\citep[see e.g.,][App.~D]{SeljakEtAl2017}).
Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
vmad, SeljakEtAl2017, FlowPM} compute gradients using AD engines, namely
\texttt{vmad} (written by the same authors) and \texttt{TensorFlow},
respectively.
Both analytic differentiation and AD backpropagate the gradients through
the whole history, thus requires saving the states at all time steps in
memory.
Therefore, they are subject to a trade-off between time and space/mass
resolution, and typically can integrate for only tens of time steps,
much smaller than the standard non-differentiable simulations.
As a result, they lose accuracy on small scales and in dense regions
where the time resolution is important, e.g., in weak lensing
\citep{MADLens}.

Alternatively, the adjoint method provides systematic ways of deriving
model gradients under constraints \citep{Pontryagin1962}, such as those
imposed by the $N$-body equations of motion in a simulated Universe.
It identifies a set of adjoint variables, dual to the state variables in
the simulation.
For time-dependent problems, these variables evolves backward in time by
a dual set of equations, known as the adjoint equations.
Their evolution depends on the states in the forward run, which can be
re-simulated with reverse time integration if the dynamics is
reversible, thereby dramatically reducing the memory cost
\citep{NeuralODE}.
Furthermore, we take the discretize-then-optimize approach
\citep[e.g.,][]{ANODE} to ensure gradients propagate backward along the
\emph{same} discrete trajectory as taken by the forward time
integration.
In brief, to compute the gradient we only need to evolve a simulation
forward, and then backward jointly with its dual adjoint equations.
We introduce the adjoint method for generic time-dependent problems in
both continuous and discrete cases in \autoref{app:adj}.

We implement the adjoint method with reverse time integration in a new
differentiable PM library \pmwd\ using \texttt{JAX} \citep{pmwd}.
\pmwd\ is memory efficient at gradient computation with a space
complexity independent of the number of time steps, and is computation
efficient when running on GPUs.

\begin{figure*}[t]
\centering
\tikzstyle{node style}=[
  node distance=2em, minimum size=2em, text centered,
  to/.style={-to, gray, thick},
  toto/.style={double distance=1pt, -Implies, gray, thick},
  from/.style={to-, gray, thick},
]
\tikzstyle{box style}=[
  rectangle, inner sep=0.4em, rounded corners, draw=gray, thick,
]
\subfloat[model structure]{
\label{fig:model}
\begin{tikzpicture}[node style]
  \node (boltz) {Boltz};
  \node (ic) [right=of boltz] {IC} edge [from] (boltz);
  \node (integ) [right=of ic] {Integ} edge [from] (ic);
  \node (obs) [below=of integ] {Obs}
        edge [from] (ic)
        edge [from] (integ);
  \node (obj) [below=of obs] {Obj} edge [from] (obs);
  \node (param) [below=of ic] {$\vtheta, \vomega$}
        edge [to] (boltz)
        edge [toto] (ic)
        edge [to] (integ)
        edge [to] (obs)
        edge [toto] (obj);
  \node (box) [box style, fit=(integ) (obs)] {};
  \useasboundingbox ($ (box.north) + (0, 1em) $);
\end{tikzpicture}
}
\hfill
\subfloat[integration-observation loop]{
\label{fig:loop}
\begin{tikzpicture}[node style]
  \node (ic) {IC};
  \node (integ_1) [right=of ic] {$I_1$} edge [from] (ic);
  \node (obs_1) [below=of integ_1] {$O_1$}
        edge [from] (ic)
        edge [from] (integ_1);
  \node (integ_1i) [right=of integ_1] {$\cdots$} edge [from] (integ_1);
  \node (obs_1i) [below=of integ_1i] {$\cdots$}
        edge [from] (integ_1i)
        edge [from] (obs_1);
  \node (integ_i) [right=of integ_1i] {$I_i$} edge [from] (integ_1i);
  \node (obs_i) [below=of integ_i] {$O_i$}
        edge [from] (integ_i)
        edge [from] (obs_1i);
  \node (integ_in) [right=of integ_i] {$\cdots$} edge [from] (integ_i);
  \node (obs_in) [below=of integ_in] {$\cdots$}
        edge [from] (integ_in)
        edge [from] (obs_i);
  \node (integ_n) [right=of integ_in] {$I_n$} edge [from] (integ_in);
  \node (obs_n) [below=of integ_n] {$O_n$}
        edge [from] (integ_n)
        edge [from] (obs_in);
  \node (obj) [below=of obs_n] {Obj} edge [from] (obs_n);
  \node (box) [box style, fit=(integ_1) (obs_n)] {};
  \node [box style, inner sep=0em, dashed, fit=(integ_i) (obs_i)] {};
  \useasboundingbox ($ (box.north) + (0, 1em) $);
\end{tikzpicture}
}
\\
\hfill
\subfloat[$i$-th (leapfrog) integration-observation step]{
\label{fig:step}
\begin{tikzpicture}[node style, node distance=2em and 1.5em]
  \node (kick_a) {$K_{i-1}^{i-\half}$};
  \node (drift) [right=of kick_a] {$D_{i-1}^i$} edge [from] (kick_a);
  \node (force) [right=of drift] {$F_i$} edge [from] (drift);
  \node (kick_b) [right=of force] {$K_{i-\half}^i$}
        edge [from] (force);
  \node (obs) [below=of kick_b] {$O_{i-1}^i$} edge [from] (kick_b);
  \node (integ_im1) [left=of kick_a, minimum width=0em] {}
        edge [to] (kick_a);
  \node (integ_ip1) [right=of kick_b, minimum width=0em] {}
        edge [from] (kick_b);
  \node (obs_im1) [below=of integ_im1, minimum width=0em] {}
        edge [to] (obs);
  \node (obs_ip1) [below=of integ_ip1, minimum width=0em] {}
        edge [from] (obs);
  \node (box) [box style, dashed, fit=(kick_a) (kick_b) (obs)] {};
  \useasboundingbox ($ (box.south) - (0, 1em) $) --
                    ($ (box.north) + (0, 1em) $);
\end{tikzpicture}
}
\caption{Simulation-based forward model of the Universe.
\subref{fig:model} shows the overall model structure.
Single arrows from the cosmological parameters $\vtheta$ and white noise
field $\vomega$ indicate dependence on $\vtheta$ only, while double
arrows imply dependence on both.
The time integration loop in \subref{fig:loop} expands the solid box in
\subref{fig:model}, and a single time step in \subref{fig:step} further
expands the dashed box in \subref{fig:loop}.
We describe different operators in \autoref{sec:method}: Boltzmann
solver (``Boltz'') and initial condition generator (``IC'') in
\autoref{sec:ic}; force solver ($F$) in \autoref{sec:force}; time
integration (``Integ''), kick ($K$), and drift ($D$) in
\autoref{sec:integ}; observation (``Obs'' and $O$) and objective
(``Obj'') in \autoref{sec:obsobj}.
Gradients flow backward with all arrows reversed (\autoref{sec:adj}).
}
\end{figure*}

\vspace{1em}
\section{Methods}
\label{sec:method}

We first review all components of the $N$-body simulation-based forward
model of the cosmological structure formation, and then introduce the
adjoint method for each part, and how they can be combined to
differentiate a whole simulation at low memory cost.

\vspace{1em}
\subsection{Initial Conditions \& Perturbation Theories}
\label{sec:ic}

$N$-body particles discretize the uniform distribution of matter at the
beginning of cosmic history (scale factor $a(t) \to 0$) at their
Lagrangian positions $\vq$, typically on a Cartesian grid, from which
they then evolve by displacements $\vs$ to their later positions, i.e.,
%
\begin{equation}
\vx = \vq + \vs(\vq).
\end{equation}
%
To account for the cosmic background that expands globally, while $\vx$
is the comoving position relative to this background, the physical
position grows with the expansion scale factor, i.e., $a \vx$.
The expansion rate is described by the Hubble parameter $H \triangleq \d
\ln a / \d t$.

The initial conditions of particles can be set perturbatively when
$|\vnabla \cdot \vs|$, linear approximation of the density fluctuation,
is much smaller than 1.
We compute the initial displacements and momenta using the second order
Lagrangian perturbation theory \citep[2LPT,][]{BouchetEtAl1995}:
%
\begin{align}
\vs &= D_1 \vs^{(1)} + D_2 \vs^{(2)}, \nonumber\\
\vp &= a^2 \dot\vx = a^2 \dot\vs
  = a^3 H \bigl( D_1' \vs^{(1)} + D_2' \vs^{(2)} \bigr),
\label{ic}
\end{align}
%
where $\vp$ is the canonical momentum\footnote{We omit the particle mass
$m$ in the canonical momentum $\vp = m a^2 \dot\vx$ throughout for
brevity.} for canonical coordinate $\vx$.
The temporal and spatial dependences separate at each order: the $i$-th
order growth factor $D_i$ is only a function of scale factor $a$ (or
time $t$), and the $i$-th order displacement field $\vs^{(i)}$ depends
only on $\vq$.
We have also used two types of time derivatives, $\dot\Box \triangleq
\d\Box / \d t$ and $\Box' \triangleq \d\Box / \d \ln a$, related by
$\dot\Box = H \Box'$.

Both the first and second order displacements are potential flows,
%
\begin{equation}
\vs^{(1)} = - \vnabla_\vq \phi_\vs^{(1)},
\quad
\vs^{(2)} = - \vnabla_\vq \phi_\vs^{(2)},
\end{equation}
%
with the scalar potentials sourced by\footnote{Note the difference in
the sign of the source of $\phi_\vs^{(2)}$ from the usual LPT
convention.}
%
\begin{align}
\nabla_\vq^2 \phi_\vs^{(1)} &= \delta^{(1)}(\vq), \nonumber\\
%
\nabla_\vq^2 \phi_\vs^{(2)} &= {\textstyle \sum_{i<j}} \bigl(
  \phi_{\vs,ii}^{(1)}\phi_{\vs,jj}^{(1)}
  - \phi_{\vs,ij}^{(1)}\phi_{\vs,ji}^{(1)}
\bigr),
\end{align}
%
where $\phi_{\vs,ij} \triangleq \p^2\phi_\vs / \p q_i \p q_j$,
and $\delta^{(1)}$ is the linear order of the overdensity field,
$\delta$, related to the density field $\rho$ and mean matter density
$\bar\rho$ by $\delta \triangleq \rho / \bar\rho - 1$.

The linear overdensity $\delta^{(1)}$, which sources the 2LPT particle
initial conditions, is a homogeneous and isotropic Gaussian random field
in the consensus cosmology, with its Fourier transform
$\delta^{(1)}(\vk) = \int\!\d\vq\, \delta^{(1)}(\vq) e^{- i \vk \cdot
\vq}$ characterized by the linear matter power spectrum $\Plin$,
%
\begin{equation}
\bigl\langle \delta^{(1)}(\vk) \delta^{(1)}(\vk') \bigr\rangle
= (2\pi)^3 \deltaD(\vk + \vk') \Plin(k)
\simeq \deltaK(\vk + \vk') V \Plin(k).
\end{equation}
%
The angle bracket takes the ensemble average of all possible
realizations.
Homogeneity demands that different wavevectors are uncorrelated, thus
the Dirac delta $\deltaD$ in the first equality.
And with isotropy, $\Plin$ does not depend on the direction of the
wavevector $\vk$, but only on its magnitude, the wavenumber $k
\triangleq |\vk|$.
In a periodic box of volume $V$, where $\vk$ is discrete, $\deltaD$ is
replaced by the Kronecker delta $\deltaK$ in the second equality.
Numerically, we can easily generate a $\delta^{(1)}$ realization by
sampling each Fourier mode independently,
%
\begin{equation}
\delta^{(1)}(\vk) = \sqrt{V \Plin(k)} \, \omega(\vk),
\end{equation}
%
with $\omega(\vk)$ being any Hermitian white noise, i.e., Fourier
transform of a real white noise field $\omega(\vq)$.

Cosmological perturbation theory gives the linear power spectrum
as\footnote{Here the time dependence of the linear power, $\Plin(k, a) =
\Plin(k) D_1^2(a)$, has been left to \eqref{ic}}
%
\begin{equation}
\frac{k^3}{2\pi^2} \Plin(k)
= \frac{4}{25} \As
  \Bigl( \frac{k}{k_\mathrm{pivot}} \Bigr)^{\ns - 1}
  T^2(k) \frac{c^4 k^4}{\Omegam^2 H_0^4},
\label{Plin}
\end{equation}
%
where the shape of $\Plin$ is determined by the transfer function $T$,
solution to the linearized Einstein-Boltzmann equations \citep{CAMB,
CLASS}.
$T$ depends on the cosmological parameters,
%
\begin{equation*}
\vtheta = (
  \As, \ns, \Omegam, \Omegab, h, \cdots),
\end{equation*}
%
some of which already appear in \eqref{Plin}: $\As$ is the amplitude of
the primordial power spectrum defined at some fixed scale
$k_\mathrm{pivot}$; $\ns$ describes the shape of the primordial power
spectrum; $\Omegam$ is the total matter density parameter; $\Omegab$ is
the baryonic matter density parameter; and $H_0$ is the Hubble constant,
often parameterized by the dimensionless $h$ as $H_0 = 100 h \,
\mathrm{km}/\mathrm{s}/\Mpc$.
Other parameters may enter in extensions of the standard $\Lambda$ cold
dark matter ($\Lambda$CDM) cosmology.

In summary, other than the discretized white noise modes $\vomega$, to
generate initial conditions we need the growth functions $D$ and the
transfer function $T$, both of which depend on the cosmological
parameters $\vtheta$.
We compute $D$ by solving the ordinary differential equations (ODEs)
given in \autoref{app:growth}, and employ the fitting formula for $T$
from \citet{EisensteinHu1998}.
We illustrate these dependencies in the upper left triangle of
\autoref{fig:model}.

At early times and/or lower space/mass resolution, LPT can be accurate
enough to directly compare to the observational data.
However, more expensive integration of the $N$-body dynamics is
necessary in the nonlinear regime.
During LPT and the time integration we can ``observe'' the simulation
predictions by interpolating on the past light cone of a chosen
observer.
These form the upper right square of \autoref{fig:model}.

\vspace{1em}
\subsection{Force Evaluation}
\label{sec:force}

The core of gravitational $N$-body simulation is the gravity solver.
The gravitational potential sourced by matter density fluctuation
follows the Poisson equation
%
\begin{equation}
\nabla^2 \phi(\vx) = \frac32 \frac{\Omegam H_0^2}{a} \delta(\vx),
\end{equation}
%
where $\nabla^2$ is the Laplacian with respect to $\vx$.
We separate the time dependence by defining $\varphi \triangleq a \phi$,
so $\varphi$ satisfies
%
\begin{equation}
\nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \, \delta(\vx),
\end{equation}
%
that only depends on the matter overdensity $\delta$.

While our adjoint method is general, we employ the PM solver in \pmwd\
for efficiency, and leave the implementation of short-range forces to
future development.
With the PM method, we evaluate $\delta(\vx)$ on an auxiliary mesh by
\emph{scattering} particle masses to the nearest grid points.
We use the usual cloud-in-cell (CIC), or trilinear, interpolation
\citep{HockneyEastwood1988}, to compute the fractions of a particle at
$\vx'$ going to a grid point at $\vx$,
%
\begin{equation}
W(\vx, \vx') = \prod_{i=1}^3
  \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_\mathrm{cell}}, 0 \Bigr),
\label{cic}
\end{equation}
%
where $l_\mathrm{cell}$ is the mesh cell size.

The gravitational field, $- \vnabla \varphi$, can then be readily
computed on the mesh with the fast Fourier transform (FFT), as the above
partial differential equation becomes an algebraic one in Fourier space:
%
\begin{equation}
- k^2 \varphi(\vk) = \frac32 \Omegam H_0^2 \, \delta(\vk).
\end{equation}
%
And in Fourier space, $- \vnabla \varphi(\vx)$ is just $- i \vk
\varphi(\vk)$, each component of which can be transformed back and
combined into the force field.
With 4 (1 forward and 3 inverse) FFTs, we can obtain $- \vnabla
\varphi(\vx)$ from $\delta(\vx)$, with both on the mesh, efficiently.

Finally, we interpolate the particle accelerations by \emph{gathering}
$- \vnabla \varphi$ from the same grid points with the same weights, as
given in \eqref{cic}.

\vspace{1em}
\subsection{Time Integration}
\label{sec:integ}

$N$-body particles move by the following equations of motion
%
\begin{align}
\dot\vx &= \frac{\vp}{a^2}, \nonumber\\
\dot\vp &= -\vnabla\phi = - \frac{\vnabla\varphi}a.
\label{eom}
\end{align}
%
We use the \texttt{FastPM} time stepping \citep{FastPM}, designed to
reproduce the first order Lagrangian perturbation theory, also known as
the Zel'dovich approximation (hereafter ZA), in the linear regime.
We present a simplified derivation below.

$N$-body simulations integrate \eqref{eom} in discrete steps
(\autoref{fig:loop}), typically with a symplectic integrator that
updates $\vx$ and $\vp$ alternately.
From $t_a$ to $t_b$,
%
\begin{align}
\vx_b
  &= \vx_a + \int_{t_a}^{t_b} \frac{\vp}{a^2} \,\d t
  \approx \vx_a + \frac{\vp(t_c)}{G_D(t_c)}
    \int_{t_a}^{t_b} \frac{G_D}{a^2} \,\d t
  \triangleq \vx_a + \vp_c D(t_c, t_a, t_b), \nonumber\\
%
\vp_b
  &= \vp_a - \int_{t_a}^{t_b} \frac{\vnabla\varphi}a \,\d t
  \approx \vp_a - \frac{\vnabla\varphi(t_c)}{G_K(t_c)}
    \int_{t_a}^{t_b} \frac{G_K}a \,\d t
  \triangleq \vp_a - \vnabla\varphi_c K(t_c, t_a, t_b),
\label{eom_integ}
\end{align}
%
which are named the drift and kick operators, respectively.
In the second equalities of each equation we have introduced two
time-dependent functions $G_D$ and $G_K$.
As approximations, they have been taken out of the integrals together
with $\vp$ and $\vnabla\varphi$ at some intermediate representative time
$t_c$.
We can make the approximation more accurate by choosing $G_D$ to have a
time dependence closer to that of $\vp$.
Likewise for $\vnabla\varphi$ and $G_K$.
However, in most codes $G_D$ and $G_K$ are simply set to 1
\citep{QuinnEtAl1997}, lowering the accuracy when the number of time
steps are limited.

\texttt{FastPM} chooses $G_D$ and $G_K$ according to the ZA growth
history, thereby improves the accuracy on large scales and at early
times.
In ZA, the displacements are proportional to the linear growth factor,
$\vs \propto D_1$, which determines the time dependences of the momenta
and the accelerations by \eqref{eom}.
Therefore, we can set $G_D$ and $G_K$ in \eqref{eom_integ} to
%
\begin{align}
G_D &:= a^2 \dot D_1, \nonumber\\
G_K &:= a \dot G_D.
\label{G_ZA}
\end{align}
%
They are functions of $D_1$ and its derivatives, as given by
\eqref{G_eval}.
With these choices, the drift and kick factors, defined in
\eqref{eom_integ}, then become
%
\begin{align}
D(t_c, t_a, t_b) &=
  \frac{D_1(t_b) - D_1(t_a)}{G_D(t_c)}, \nonumber\\
%
K(t_c, t_a, t_b) &=
  \frac{G_D(t_b) - G_D(t_a)}{G_K(t_c)}.
\end{align}
%

While these operators are generally applicable in any symplectic
integrator, we use them in the second order kick-drift-kick leapfrog, or
velocity Verlet, integration scheme \citep{QuinnEtAl1997}.
From $t_{i-1}$ to $t_i$, the particles' state $(\vx, \vp)$ is updated in
the following order:
%
\begin{alignat}{2}
&K_{i-1}^{i-\half}: &\qquad \vp_{i-\half} &= \vp_{i-1}
  + \va_{i-1} K(t_{i-1}, t_{i-1}, t_{i-\half}), \nonumber\\
%
&D_{i-1}^i: & \vx_i &= \vx_{i-1}
  + \vp_{i-\half} D(t_{i-\half}, t_{i-1}, t_i), \nonumber\\
%
&F_i: & \va_i &\triangleq -\vnabla\varphi(\vx_i), \nonumber\\
%
&K_{i-\half}^i: & \vp_i &= \vp_{i-\half}
  + \va_i K(t_i, t_{i-\half}, t_i).
\label{KDFK}
\end{alignat}
%
The left column names the operators as shown in \autoref{fig:step}.
The force operator $F$ on the third line computes the accelerations as
described in \autoref{sec:force}.
It caches the results in $\va$, so that they can be used again by the
first $K$ in the next step.
Note that we need to initialize $\va_0$ with $F_0$ before the first time
step.

\vspace{1em}
\subsection{Observation \& Objective}
\label{sec:obsobj}

Because all observables live on our past light cone, we model
observations on the fly by interpolating the $j$-th particle's state
$\hat\vz_j = (\hat\vx_j, \hat\vp_j)$ when they cross the light cone at
$\hat t_j$.
Given $\vz = (\vx, \vp)$ at $t_{i-1}$ and $t_i$, we can parametrize
intermediate particle trajectories with cubic Hermite splines.
Combined with the $\vtheta$-dependent propagation of the light front, we
can solve for the intersections\footnote{Even analytically if the light
propagation is also approximated cubically.}, at which we can record the
observed\footnote{Note that we focus on the dark matter phase space
here, and leave more generic observable modeling to a future work.}
$\hat\vz$.
\autoref{fig:step} illustrates the observation operator $O$ and its
dependencies on the previous and the current time step.

We can compare the simulated observables to the observational data,
either at level of the fields or the summary statistics, by some
objective function in case of optimization, or by a posterior
probability for Bayesian inference.
We refer to both cases by objective and denote it by $\cJ$ throughout.
Note that in general it can also depend on $\vtheta$ and $\vomega$, in
the form of regularization or prior probability.

Formally, we can combine the observation and objective operators as
%
\begin{equation}
\cJ(\hat\vz, \vtheta, \vomega)
  = \cJ \bigl(
    \hat\vz(\vz_0, \cdots, \vz_n, \vtheta), \vtheta, \vomega
  \bigr),
\label{O}
\end{equation}
%
as illustrated in the lower right triangle in \autoref{fig:model}.
Note this form also captures the conventional simulation snapshots at
the end of a time step or those interpolated between two consecutive
steps, so we model all these cases as observations in \pmwd.

\vspace{1em}
\subsection{Adjoint Method}
\label{sec:adj}

The adjoint method provides systematic ways of deriving the objective
gradient under constraints \citep{Pontryagin1962}, here imposed by the
$N$-body equations of motion.
In \autoref{app:adj} we introduce the adjoint method for generic
time-dependent problems in both continuous and discrete cases.
The continuous case is easier to understand and has pedagogical values,
while the discrete case is the useful one in our application, for we
want to propagate the gradients backward along the \emph{same} path as
that of the forward time integration.

For the $N$-body particles, the state variable\footnote{State and
adjoint vectors in the adjoint equations include enumeration of
particles, e.g., $\vnabla\vvarphi$ includes the $\vnabla\varphi$ of each
particle.} is $\vz = (\vx, \vp)^\intercal$.
Their adjoint variables help to accumulate the objective gradient while
evolving backward in time by the adjoint equation.
Let's denote them by $\vlambda = (\vxi, \vpi)$.
We can compare each step of \eqref{KDFK} and \eqref{O} to
\eqref{explicit}, and write down its adjoint equation following
\eqref{explicit_adj}.
Taking $D_{i-1}^i$ for example, we can write it explicitly as
%
\begin{equation*}
\begin{pmatrix} \vx_i \\ \vp_{i-\half} \end{pmatrix}
= \begin{pmatrix} \vx_{i-1} \\ \vp_{i-\half} \end{pmatrix}
+ \begin{pmatrix}
    \vp_{i-\half} D(t_{i-\half}, t_{i-1}, t_i) \\ \vzero
  \end{pmatrix},
\end{equation*}
%
in the form of \eqref{explicit}.
By \eqref{explicit_adj}, its adjoint equation is
%
\begin{align*}
\begin{pmatrix} \vxi_{i-\half} & \vpi_{i-1} \end{pmatrix}
&= \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
+ \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
  \begin{pmatrix}
    \vzero & \vI \, D(t_{i-\half}, t_{i-1}, t_i) \\
    \vzero & \vzero \\
  \end{pmatrix} \\
%
&= \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
- \begin{pmatrix}
    \vzero & \vxi_{i-\half} D(t_{i-\half}, t_i, t_{i-1})
  \end{pmatrix},
\end{align*}
%
where we have used the fact that $D(t_c, t_a, t_b) = - D(t_c, t_b,
t_a)$, and left the $\p\cJ / \p\vz_i$ term to the $O$ operator.
This also naturally determines the subscripts of $\xi$ and $\pi$.

Repeating the derivation for $K$ and $O$, and flipping the arrow of
time, we present the adjoint equation time stepping for \eqref{KDFK}
from $t_i$ to $t_{i-1}$:
%
\begin{alignat}{2}
&K_i^{i-\half}: &\qquad \vp_{i-\half} &= \vp_i
  + \va_i K(t_i, t_i, t_{i-\half}), \nonumber\\
&& \vxi_{i-\half} &= \vxi_i
  - \valpha_i K(t_i, t_i, t_{i-\half}), \nonumber\\
%
&D_i^{i-1}: & \vx_{i-1} &= \vx_i
  + \vp_{i-\half} D(t_{i-\half}, t_i, t_{i-1}), \nonumber\\
&& \vpi_{i-1} &= \vpi_i
  - \vxi_{i-\half} D(t_{i-\half}, t_i, t_{i-1}), \nonumber\\
%
&F_{i-1}: & \va_{i-1} &= -\vnabla\vvarphi(\vx_{i-1}), \nonumber\\
&& \valpha_{i-1} &\triangleq - \vpi_{i-1}
  \cdot \frac{\p\vnabla\vvarphi_{i-1}}{\p\vx_{i-1}}, \nonumber\\
&& \vzeta_{i-1} &\triangleq - \vpi_{i-1}
  \cdot \frac{\p\vnabla\vvarphi_{i-1}}{\p\vtheta}, \nonumber\\
%
&K_{i-\half}^{i-1}: & \vp_{i-1} &= \vp_{i-\half}
  + \va_{i-1} K(t_{i-1}, t_{i-\half}, t_{i-1}), \nonumber\\
&& \vxi_{i-1} &= \vxi_{i-\half}
  - \valpha_{i-1} K(t_{i-1}, t_{i-\half}, t_{i-1}), \nonumber\\
%
&O_i^{i-1}: &\qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
  + \frac{\p\cJ}{\p\vx_{i-1}}, \nonumber\\
&& \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1} + \frac{\p\cJ}{\p\vp_{i-1}}.
\label{KDFKO_adj}
\end{alignat}
%
Like $\va$, we have introduced $\valpha$ and $\vzeta$ to cache the
vector-Jacobian products (VJPs) on their right-hand sides, for the next
time step in the kick operator and the objective gradient (see below),
respectively.
Note that in the reverse order, the $F$ operator is at $t_{i-1}$ instead
of $t_i$ as in \eqref{KDFK}, and we need to initialize $\va_n$,
$\valpha_n$, and $\vzeta_n$ with $F_n$ before stepping from $t_n$ to
$t_{n-1}$.
Likewise, the gradient of $O_n^{n-1}$ at $t_n$ is absent in
\eqref{KDFKO_adj} but enter via the initial conditions following
\eqref{explicit_adj}.
Explicitly, the initial conditions of \eqref{KDFKO_adj} are
%
\begin{equation}
\vxi_n = \frac{\p\cJ}{\p\vx_n},
\vpi_n = \frac{\p\cJ}{\p\vp_n},
\va_n = -\vnabla\vvarphi(\vx_n),
\valpha_n = - \vpi_n \cdot \frac{\p\vnabla\vvarphi_n}{\p\vx_n},
\vzeta_n = - \vpi_n \cdot \frac{\p\vnabla\vvarphi_n}{\p\vtheta}.
\end{equation}

The VJPs in $F$ and the $\p\cJ / \p\vz$'s in $O$ can be computed by AD
if the whole forward integration and observation history of \eqref{KDFK}
and \eqref{O} has been saved.
However, this can be too costly spatially for GPUs, whose memories are
much smaller than those of CPUs.
Alternatively, we take advantage of the reversibility of the $N$-body
dynamics and the leapfrog integrator, and recover the history by reverse
time integration, which we have already included on the first lines of
the $K$ and $D$ operators in \eqref{KDFKO_adj}.
We can integrate the leapfrog and the adjoint equations jointly backward
in time, and still benefit from the convenience of AD in computing VJPs
and $\p\cJ / \p\vz$'s.
In practice, the numerical reversibility can be challenged by the
chaotic $N$-body dynamics, which we find is generally not a concern for
our applications in the result section.

Finally, during the reverse time integration, we can accumulate
objective gradient following \eqref{explicit_grad}:
%
\begin{multline}
\frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
%
+ \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vtheta}
+ \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vtheta}
%
+ \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
+ \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
%
- \sum_{i=1}^n \Bigl[
  \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
    \bigr) K(t_i, t_i, t_{i-\half}) \\
  + \vxi_i \cdot \vp_{i-\half}
    \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \\
  + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta} + \vzeta_{i-1}
    \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
  \Bigr],
\label{KDFKO_grad}
\end{multline}
%
in which the derivatives and VJPs of the regularization/prior, the
observation, the initial conditions, the kick and drift factors, and the
growth and transfer functions can all be conveniently computed by AD.
Also note that $\vtheta$ can stand for both itself and $\vomega$, which
we have omitted for brevity.
To implement \eqref{KDFKO_adj}-\eqref{KDFKO_grad} in \pmwd\ with
\texttt{JAX}, we only need to write custom VJP operators for the
high-level $N$-body integration-observation loop.

Other than that, we also implement custom VJPs for the scatter and
gather operations in \autoref{sec:force} following \citet{Feng2017}, and
these further save memory in gradient computations of those nonlinear
functions.

\begin{figure*}[t]
\centering
\includegraphics[width=0.85\linewidth]{slab.pdf}
\caption{Relative matter density field, $1+\delta$, at $a=1$, projected
from an $8\,\Mpc/h$ thick slab in a \pmwd\ simulation, that has evolved
$512^3$ particles with single precision and a $1024^3$ mesh in a
$(512\,\Mpc/h)^3$ box for 63 time steps.
The simulation takes only 20 seconds to finish on a \GPU\ GPU.
}
\label{fig:slab}
\end{figure*}

\vspace{1em}
\section{Results}

We implement the forward simulation and backward adjoint method in the
PM code \pmwd.
All results assume a simple $\Lambda$CDM cosmology: $\vtheta (\As =
2\times10^{-9}, \ns = 0.96, \Omegam = 0.3, \Omegab = 0.05, h = 0.7)$,
and use a \GPU\ GPU.

As in \texttt{FastPM}, the choice of time steps is flexible, and in fact
with \pmwd\ it can even be optimized in non-parametric ways to improve
simulation accuracy at fixed computation cost.
Here we use time steps linearly spaced in scale factor $a$, and leave
such optimization in a follow-up work.

\vspace{1em}
\subsection{Simulation}

We first test the forward simulations of \pmwd.
\autoref{fig:slab} shows the cosmic web in the final snapshot of a
fairly large simulation for the size of GPU memories.

Because GPUs are inherently parallel devices, they can output different
results for identical inputs.
To test the reproducibility, in \autoref{tab:rep} we compare the
root-mean-square deviations (RMSDs) of particle displacements and
velocities between two runs, relative to their respective standard
deviations, with different floating-point precisions, mesh sizes,
particle masses, and numbers of time steps.
Other than the precision, mesh size is the most important factor,
because a finer mesh can better resolve the most nonlinear and dense
structures, which can affect reproducibility as the order of many
operations can change easily.
The particle mass plays a similar role and less massive particles
generally take part in nonlinear motions at earlier times.
The number of time steps has very small impact except in the most
nonlinear cases.
And interestingly, more time steps improves the reproducibility most of
the time.

\begin{table*}[t]
\centering
\caption{\pmwd\ reproducibility on GPU.
GPUs can output different results for identical inputs.
We simulate $384^3$ particles from $a=1/64$ to $a=1$, with two
floating-point precisions, two mesh sizes, two particle masses, and two
time step sizes.
We take the root-mean-square deviations (RMSDs) of particle
displacements and velocities between two runs at $a=1$, and quote their
ratios to the respective standard deviations, about $6 \, \Mpc / h$ and
$3\times10^2 \, \mathrm{km} / \mathrm{s}$.
In general, the factors on the left of the left four columns affect the
reproducibility more than the those on the right, and lower rows are
more reproducible than the upper ones.
}
\label{tab:rep}
\begin{tabular}{cccccc}
  \toprule
  precision & cell/ptcl & ptcl mass $[10^{10} M_\odot]$ & time steps
    & disp rel diff & vel rel diff \\
  \midrule
  single & 8 & 1 &  63 & $1.6\times10^{-6}$  & $4.8\times10^{-5}$  \\
  single & 8 & 1 & 126 & $5.1\times10^{-7}$  & $1.3\times10^{-5}$  \\
  single & 8 & 8 &  63 & $3.1\times10^{-7}$  & $2.4\times10^{-6}$  \\
  single & 8 & 8 & 126 & $3.1\times10^{-7}$  & $2.3\times10^{-6}$  \\
  single & 1 & 1 &  63 & $1.4\times10^{-7}$  & $1.5\times10^{-6}$  \\
  single & 1 & 1 & 126 & $1.3\times10^{-7}$  & $1.4\times10^{-6}$  \\
  single & 1 & 8 &  63 & $1.1\times10^{-7}$  & $4.5\times10^{-7}$  \\
  single & 1 & 8 & 126 & $9.9\times10^{-8}$  & $4.4\times10^{-7}$  \\
  double & 8 & 1 &  63 & $1.4\times10^{-15}$ & $3.6\times10^{-14}$ \\
  double & 8 & 1 & 126 & $8.3\times10^{-16}$ & $1.7\times10^{-14}$ \\
  double & 8 & 8 &  63 & $5.4\times10^{-16}$ & $4.1\times10^{-15}$ \\
  double & 8 & 8 & 126 & $5.9\times10^{-16}$ & $4.3\times10^{-15}$ \\
  double & 1 & 1 &  63 & $2.4\times10^{-16}$ & $2.3\times10^{-15}$ \\
  double & 1 & 1 & 126 & $2.3\times10^{-16}$ & $2.3\times10^{-15}$ \\
  double & 1 & 8 &  63 & $1.9\times10^{-16}$ & $7.9\times10^{-16}$ \\
  double & 1 & 8 & 126 & $1.8\times10^{-16}$ & $7.8\times10^{-16}$ \\
  \bottomrule
\end{tabular}
\end{table*}

\vspace{1em}
\subsection{Differentiation}

Model differentiation evolves the adjoint equations backward in time.
To save memory, trajectory of the model state in the forward run is not
saved, but re-simulated together with the adjoint equations.
Even though in principle the $N$-body systems are reversible, in
practice the reconstructed trajectory can differ from the forward one
due to loss of small-scale information.
To test the reversibility, in \autoref{tab:rev} we compare the RMSDs of
the forward-then-reversed particle displacements and velocities from the
LPT initial conditions, relative to their respective standard
deviations, that are very small at the \emph{initial} time.
As before, we vary the floating-point precision, the mesh size, the
particle mass, and the number of time steps.
The order of factor importance and their effects are the same as in the
reproducibility test.
This is because more nonlinear structures are more difficult to reverse.
One way to improve reversibility is use higher order LPT to initialize
the $N$-body simulations at later times, when the displacements and
velocities are not as small.
We leave this for future development.

\begin{table*}[t]
\centering
\caption{\pmwd\ reversibility on GPU.
Our adjoint method reduces memory cost by reconstructing the forward
evolution with reverse time integration.
We test the numerical reversibility by comparing the displacements and
velocities of particles that have evolved to $a=1$ and then reversed to
$a=1/64$, to those of the LPT initial conditions at $a=1/64$, in RMSDs.
We take their ratios to the respective standard deviations, about $0.1
\, \Mpc / h$ and $0.7 \, \mathrm{km} / \mathrm{s}$.
Their smallness contributes partly to the fact that the quoted relative
differences here being much greater than those in \autoref{tab:rep}.
With the same setup as that in \autoref{tab:rep}, we see the same
general trend---left factors are more important than the right ones,
and lower rows are more reversible than the upper ones.
}
\label{tab:rev}
\begin{tabular}{cccccc}
  \toprule
  precision & cell/ptcl & ptcl mass $[10^{10} M_\odot]$ & time steps
    & disp rel diff & vel rel diff \\
  \midrule
  single & 8 & 1 &  63 & $5.2\times10^{-2}$  & $7.1\times10^{-2}$  \\
  single & 8 & 1 & 126 & $2.1\times10^{-2}$  & $3.6\times10^{-2}$  \\
  single & 8 & 8 &  63 & $3.3\times10^{-3}$  & $7.6\times10^{-3}$  \\
  single & 8 & 8 & 126 & $3.7\times10^{-3}$  & $7.0\times10^{-3}$  \\
  single & 1 & 1 &  63 & $1.4\times10^{-3}$  & $2.2\times10^{-3}$  \\
  single & 1 & 1 & 126 & $1.3\times10^{-3}$  & $1.7\times10^{-3}$  \\
  single & 1 & 8 &  63 & $4.3\times10^{-4}$  & $7.3\times10^{-4}$  \\
  single & 1 & 8 & 126 & $4.4\times10^{-4}$  & $6.3\times10^{-4}$  \\
  double & 8 & 1 &  63 & $5.4\times10^{-11}$ & $1.3\times10^{-10}$ \\
  double & 8 & 1 & 126 & $3.5\times10^{-11}$ & $7.0\times10^{-11}$ \\
  double & 8 & 8 &  63 & $5.8\times10^{-12}$ & $1.4\times10^{-11}$ \\
  double & 8 & 8 & 126 & $6.4\times10^{-12}$ & $1.2\times10^{-11}$ \\
  double & 1 & 1 &  63 & $2.2\times10^{-12}$ & $3.4\times10^{-12}$ \\
  double & 1 & 1 & 126 & $2.1\times10^{-12}$ & $2.9\times10^{-12}$ \\
  double & 1 & 8 &  63 & $7.2\times10^{-13}$ & $1.2\times10^{-12}$ \\
  double & 1 & 8 & 126 & $6.9\times10^{-13}$ & $9.4\times10^{-13}$ \\
  \bottomrule
\end{tabular}
\end{table*}

Next, we want to verify that our adjoint method yields the same
gradients as those computed by AD.
As explained in \autoref{sec:adj}, \pmwd\ already utilizes AD on most of
the differentiation tasks.
To get the AD gradients we disable our custom VJP implementations on the
$N$-body time integration, and the scatter and gather operations.
In \autoref{fig:grads}, we compare the adjoint and AD gradients on a
smaller problem, because AD already runs out of memory if we double the
number of time steps or increase the space/mass resolution by
$2^3\times$.
To compensate for good statistics, we perform both adjoint and AD runs
for 8 times, and compare their results by an asymmetric difference of
$X_i - Y_j$ only for $j<i$, where $i$ and $j$ index the 8 runs.
We can set $X$ and $Y$ to adjoint and AD respectively for comparison, or
set both to either adjoint or AD to check their respective
reproducibility.
The adjoint and AD gradients on the real white noise agree very well,
and so are their gradients on cosmological parameters.
We find both gradients are consistent among different runs, with AD
being more reproducible without the uncertainty from reverse time
integration.
Interestingly, the asymmetric difference of AD gradients is not
symmetric about 0, implying some difference between the early and late
AD runs, i.e., possible breaking of order- or time-invariance in AD.

\begin{figure*}[t]
\centering
\includegraphics[width=0.4\linewidth]{grads.pdf}
\\
\vspace{1em}
\includegraphics[height=0.45\linewidth]{grads_cmp.pdf}
\hfill
\includegraphics[height=0.45\linewidth]{grads_diff.pdf}
\caption{Adjoint gradients of a $128\times128$ slice of the real white
noise fields, in a \pmwd\ simulation of $128^3$ particles in a
$(128\,\Mpc/h)^3$ box, with a $256^3$ mesh, 15 time steps, and single
precision.
We compare the adjoint gradients to those by AD, for which we have
disabled the custom gradient implementation on the scatter, gather, and
$N$-body time stepping operators.
The adjoint and AD gradients agree as expected, with a RMSD of $\approx
4\times10^{-5}$, 3 orders of magnitude smaller than the standard
deviation of the gradients itself, $\approx 0.015$.
It is also comparable to the difference between two different adjoint
gradients, with a RMSD $\approx 5\times10^{-5}$.
Different AD gradients are more consistent, with a tighter RMSD of
$\approx 1\times10^{-5}$ due to the absence of uncertainty from reverse
time integration.
}
\label{fig:grads}
\end{figure*}

Our last test on the adjoint gradients uses them in a toy optimization
problem in \autoref{fig:optim}.
We use a simple mean squared error loss between the projected density at
$a=1$ and a target image, and the Adam optimizer \citep{Adam} with a
learning rate of $0.1$ and other default hyperparameters.
Holding the cosmological parameters fixed, we can optimize the real
white noise field to source initial particles to evolve into our target
pattern, in hundreds to thousands of iterations.

\begin{figure*}[t]
\centering
\subfloat[initial conditions at $a=1/64$]{
  \includegraphics[width=0.45\linewidth]{optim_ini.pdf}
}
\hfill
\subfloat[final snapshot at $a=1$]{
  \includegraphics[width=0.45\linewidth]{optim_0.pdf}
}
\\
\subfloat[reverse evolution back to $a=1/64$]{
  \includegraphics[width=0.45\linewidth]{optim_rev.pdf}
}
\hfill
\subfloat[optimization for 10 iterations at $a=1$]{
  \includegraphics[width=0.45\linewidth]{optim_10.pdf}
}
\\
\subfloat[optimization for 100 iterations at $a=1$]{
  \includegraphics[width=0.45\linewidth]{optim_100.pdf}
}
\hfill
\subfloat[optimization for 1000 iterations at $a=1$]{
  \includegraphics[width=0.45\linewidth]{optim_1000.pdf}
}
\\
\caption{A toy problem where we optimize the initial conditions by
gradient descent to make some interesting pattern after projection.
The particles originally fill a $16\times27\times16$ grid, and then
evolve from $a=1/64$ to $a=1$ for 63 time steps with \emph{single
precision} and a $32\times54\times32$ mesh in a $160\times270\times160
\, \Mpc^3/h^3$ box.
We use the adjoint method and reverse time integration, assuming the
latter can reconstruct the forward evolution history accurately.
We validate this by demonstrating that the particles evolve backward to
align on the initial grid.
The optimized initial conditions successfully evolve into the target
pattern, which improves with more iterations.
}
\label{fig:optim}
\end{figure*}

\vspace{1em}
\subsection{Performance}

\pmwd\ benefits from GPU accelerations and efficient CUDA
implementations of scatter and gather operations.
In \autoref{fig:scaling}, we present weak scaling test of \pmwd, and
find both the LPT and the $N$-body parts scale well except for very
small problems.
The growth function solution has a constant cost, and generally does not
affect problem of moderate sizes.
However, for a small number of particles and few time steps, it can
dominate the computation, in which case one can accelerate the growth
computation with an emulator \citep{KwanModiEtAl2022}.

\begin{figure*}[t]
\centering
\includegraphics[width=0.7\linewidth]{scaling.pdf}
\caption{Weak scaling of \pmwd. Both the 2LPT and $N$-body components
scale well from $512^3$ particles to $64^3$ particles, below which they
become overhead dominated.
Solving the growth ODEs takes a constant time, and can dominate the cost
for small numbers of particles and few time steps, but generally does
not affect problems with more than $128^3$ particles.
}
\label{fig:scaling}
\end{figure*}

\vspace{1em}
\section{Conclusions}

In this work, we develop the adjoint method for memory efficient
differentiable cosmological simulations, exploiting the reversible
nature of $N$-body Hamiltonian dynamics, and implement it with
\texttt{JAX} in a new PM library, \pmwd{}.
We have validated the numerical reversibility and the accuracy of the
adjoint gradients.

Being both computation and memory efficient, \pmwd\ enables larger and
more accurate cosmological dark matter simulations.
The next step involves modeling cosmological observables such as
galaxies.
One can achieve this with analytic, semi-analytic, and deep learning
components running based on or in parallel with \pmwd.
In the future, it can also facilitate the simultaneous modeling of
multiple observables and the understanding of the astrophysics at play.

\pmwd\ will benefit all the forward modeling approaches in cosmology,
and will improve gradient based optimization and field-level inference
to simultaneously constrain the cosmological parameters and the initial
conditions of the Universe.
The efficiency of \pmwd{} also makes it a promising route to generate
the large amount of training data needed by the likelihood-free
inference frameworks \citep{CranmerEtAl2020, DELFI}.

Currently, these applications require more development, including
distributed parallel capability on multiple GPUs \citep{FlowPM},
optimization of spatiotemporal resolution of the PM solvers \citep{LDL,
LanzieriLanusseEtAl2022, ZhangLiEtAl}, short-range supplement by direct
summation of the particle-particle (PP) forces on GPUs \citep{HACC,
PKDGRAV3, ABACUS}, differentiable models for observables, etc.
We plan to pursue these in the future.

\vspace{1em}
\textit{\large Acknowledgements:}
YL is grateful to all organizers of the 2013 KICP Summer School on
Computational Cosmology \citep{KICPsummer2013}, especially Andrey
Kravtsov and Nick Gnedin.
We thank Alex Barnett, Biwei Dai, Boris Leistedt, ChangHoon Hahn,
Daisuke Nagai, Dan Foreman-Mackey, Dan Fortunato, David Duvenaud, David
Spergel, Dhairya Malhotra, Erik Thiede, Francisco Villaescusa-Navarro,
Fruzsina Agocs, Giulio Fabbian, Jens St\"ucker, Kaze Wong, Kazuyuki
Akitsu, Lehman Garrison, Manas Rachh, Ra\'ul Angulo, Shirley Ho, Simon
White, Teppei Okumura, Tomomi Sunayama, Ulrich Steinwandel, Uro\v{s}
Seljak, Wen Yan, Wenda Zhou, William Coulton, Xiangchong Li, and Yueying
Ni for interesting discussions.
The Flatiron Institute is supported by the Simons Foundation.

\vspace{1em}
\textit{\large Code Availability:}
\pmwd\ is open-source on
\href{https://github.com/eelregit/pmwd}{GitHub}, including the
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint}{source
and scripts of this paper}.
Also see the companion code paper on
\href{https://arxiv.org/abs/2211.THIS+1}{arXiv} and
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/pmwd}{GitHub}.

\vspace{1em}
\textit{\large Software:}
\texttt{JAX} \citep{JAX}, \texttt{NumPy} \citep{NumPy}, \texttt{SciPy}
\citep{SciPy}, and \texttt{matplotlib} \citep{matplotlib}.

\vspace{1em}
\appendix

\vspace{1em}
\section{Growth equations}
\label{app:growth}

The 2LPT growth functions follow the following ODEs:
\begin{align}
  D_1'' + \Bigl( 2 + \frac{H'}H \Bigr) D_1' - \frac32 \Omegam(a) D_1
  &= 0, \nonumber\\
  D_2'' + \Bigl( 2 + \frac{H'}H \Bigr) D_2' - \frac32 \Omegam(a) D_2
  &= \frac32 \Omegam(a) D_1^2,
\end{align}
where
\begin{equation}
    \Omegam(a) = \frac{\Omegam H_0^2}{a^3 H^2}
\end{equation}
If the universe has been sufficiently matter dominated at $a_\ic$, with
$\Omegam(a_\ic) \to 1$ and $H'/H \to -3/2$, the initial conditions of
the ODEs can be set as the growing mode in this limiting case
\begin{align}
    D_1(a_\ic) &= a_\ic, \nonumber\\
    D_1'(a_\ic) &= a_\ic, \nonumber\\
    D_2(a_\ic) &= \frac37 a_\ic^2, \nonumber\\
    D_2'(a_\ic) &= \frac67 a_\ic^2.
\end{align}
$D_1 \approx a$ in the matter dominated era, before being suppressed by
dark energy.

The above growth equations can be written in such suppression factors,
\begin{equation}
  G_m \triangleq D_m / a^m,
\end{equation}
as
\begin{align}
  G_1'' + \Bigl( 4 + \frac{H'}H \Bigr) G_1'
  + \Bigl( 3 + \frac{H'}H - \frac32 \Omegam(a) \Bigr) G_1
  &= 0, \nonumber\\
  G_2'' + \Bigl( 6 + \frac{H'}H \Bigr) G_2'
  + \Bigl( 8 + 2\frac{H'}H - \frac32 \Omegam(a) \Bigr) G_2
  &= \frac32 \Omegam(a) G_1^2,
\end{align}
with initial conditions
\begin{align}
    G_1(a_\ic) &= 1, \nonumber\\
    G_1'(a_\ic) &= 0, \nonumber\\
    G_2(a_\ic) &= \frac37, \nonumber\\
    G_2'(a_\ic) &= 0.
\end{align}
We solve the growth equations in $G_n$ instead of $D_n$, with the
\texttt{JAX} adaptive ODE integrator implementing the adjoint
method\footnote{The growth equation adjoint method aims to solve the
ODEs in their continuous limit, this is different from our
discretize-then-optimize $N$-body adjoint method.}.
This is because the former can be integrated backward in time more
accurately for early times, which can improve the adjoint gradients.

We can then evaluate the \texttt{FastPM} time integration factors in
\eqref{G_ZA} by
\begin{align}
  G_D &= a^2 \dot D_1 = a^2 H D_1', \nonumber\\
  G_K &= a \dot G_D = a^3 H^2 \Bigl[
    D_1'' + \Bigl(2 + \frac{H'}H \Bigr) D_1'
  \Bigr],
  \label{G_eval}
\end{align}
and
\begin{align}
  D_m &= a^m G_m \nonumber\\
  D_m' &= a^m \bigl( m G_m + G_1' \Bigr), \nonumber\\
  D_m'' &= a^m \bigl( m^2 G_1 + 2m G_1' + G_1'' \Bigr).
\end{align}

\vspace{1em}
\section{Adjoint Method}
\label{app:adj}

We introduce the adjoint method for generic time-dependent ODEs, and
derive the adjoint equations following a pedagogical tutorial by
\citet{Bradley2019}.
We then adopt the discretize-then-optimize approach, and derive the
discrete adjoint equations, that is more suitable for the $N$-body
symplectic time integration.

\vspace{1em}
\subsection{Variational (Tangent) and Adjoint Equations}

Consider a vector state $\vz(t)$ subject to the following ODEs and
initial conditions
%
\begin{equation}
\dot\vz = \vf(\vz, t, \vtheta),
\qquad
\vz(t_0) = \vz_0(\vtheta),
\label{ode}
\end{equation}
%
for $t \in [t_0, t_1]$.
Here the initial conditions can depend on the parameters $\vtheta$.

A perturbation in the initial conditions propagates forward in time.
For $\vz(t, t_0, \vz_0)$, the Jacobian of state variables describing
this,
%
\begin{equation}
\vDelta = \frac{\p\vz}{\p\vz_0},
\end{equation}
%
evolves from identity $\vDelta_0 = \vI$ by
%
\begin{equation}
\dot\vDelta = \frac{\p\vf}{\p\vz} \cdot \vDelta,
\label{variational}
\end{equation}
%
following from \eqref{ode}.
\eqref{variational} is known as the \emph{variational or tangent
equation}.

The backward version of $\vDelta$,
%
\begin{equation}
\vLambda = \frac{\p\vz_1}{\p\vz},
\end{equation}
%
evolves towards identity $\vLambda_1 = \vI$ by
%
\begin{equation}
\dot\vLambda = - \vLambda \cdot \frac{\p\vf}{\p\vz}.
\label{adjoint}
\end{equation}
%
\eqref{adjoint} is called the \emph{adjoint equation}, for the
right-hand side is $(\p\vf / \p\vz)^\intercal \vLambda$.
It can be derived from the time-invariance of $\vLambda \cdot \vDelta$:
%
\begin{equation}
\vLambda \cdot \dot\vDelta + \dot\vLambda \cdot \vDelta
= \frac\d{\d t} (\vLambda \cdot \vDelta)
= \frac\d{\d t} \Bigl( \frac{\p\vz_1}{\p\vz}
  \cdot \frac{\p\vz}{\p\vz_0} \Bigr)
= \frac\d{\d t} \frac{\p\vz_1}{\p\vz_0}
= \vzero.
\end{equation}

Alternatively, the adjoint equation can be derived from the variational
equation, using the facts that $\d (\bm{M}^{-1}) = - \bm{M}^{-1} \, \d
\bm{M} \, \bm{M}^{-1}$ and $(\p\vz / \p\tilde\vz)^{-1} = \p\tilde\vz /
\p\vz$.
Like \eqref{variational} and \eqref{adjoint}
%
\begin{equation}
\frac\d{\d t} \frac{\p\vz}{\p\tilde\vz}
= \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\tilde\vz},
\qquad
\frac\d{\d t} \frac{\p\tilde\vz}{\p\vz}
= - \frac{\p\tilde\vz}{\p\vz} \cdot \frac{\p\vf}{\p\vz}.
\end{equation}
As we can see next, the adjoint equation takes a similar form when one
optimize an objective function of the state.

\vspace{1em}
\subsection{Objective on the Final State}

In the simplest case, the objective function depends only on the final
state, e.g., the last snapshot of a simulation, and possibly the
parameters too in the form of regularization or prior information, i.e.,
$\cJ(\vz_1, \vtheta)$.
To optimize the objective under the constraint given by the ODEs, we can
introduce a time-dependent function $\vlambda(t)$ as the Lagrange
multiplier:
%
\begin{equation}
\cL = \cJ(\vz_1, \vtheta)
- \int_{t_0}^{t_1} \vlambda(t)
  \cdot \bigl[ \dot\vz - \vf(\vz, t, \vtheta) \bigr] \, \d t.
\end{equation}
%
Note the minus sign we have introduced in front of $\vlambda$ for later
convenience.

Its total derivative with respect to $\vtheta$ is
%
\begin{equation}
\frac{\d\cL}{\d\vtheta}
= \frac{\p\cJ}{\p\vtheta}
+ \frac{\p\cJ}{\p\vz_1} \cdot \frac{\p\vz_1}{\p\vtheta}
- \int_{t_0}^{t_1} \vlambda
  \cdot \Bigl( \frac{\p\dot\vz}{\p\vtheta}
    - \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
    - \frac{\p\vf}{\p\vtheta} \Bigr) \, \d t.
\end{equation}
%
Integrating the first term of the integrand by parts:
%
\begin{equation}
\int_{t_0}^{t_1} \vlambda \cdot \frac{\p\dot\vz}{\p\vtheta} \, \d t
= \int_{t_0}^{t_1} \vlambda \cdot \frac{\p^2\vz}{\p\vtheta\p t} \, \d t
= \Bigl[ \vlambda \cdot \frac{\p\vz}{\p\vtheta} \Bigr]_{t_0}^{t_1}
- \int_{t_0}^{t_1} \dot \vlambda \cdot \frac{\p\vz}{\p\vtheta} \, \d t,
\end{equation}
%
and plugging it back:
%
\begin{equation}
\frac{\d\cL}{\d\vtheta}
= \frac{\p\cJ}{\p\vtheta}
+ \Bigl( \frac{\p\cJ}{\p\vz_1} - \vlambda_1 \Bigr)
  \cdot \frac{\p\vz_1}{\p\vtheta}
+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \int_{t_0}^{t_1} \Bigl[
  \Bigl( \vlambda \cdot \frac{\p\vf}{\p\vz} + \dot\vlambda \Bigr)
  \cdot \frac{\p\vz}{\p\vtheta}
  + \vlambda \cdot \frac{\p\vf}{\p\vtheta} \Bigr] \, \d t.
\end{equation}

Now we are free to choose
%
\begin{equation}
\dot\vlambda = - \vlambda \cdot \frac{\p\vf}{\p\vz},
\qquad
\vlambda_1 = \frac{\p\cJ}{\p\vz_1},
\label{adjoint1}
\end{equation}
%
that allow us to avoid all $\p\vz/\p\vtheta$ terms in the final
objective gradient:
%
\begin{equation}
\frac{\d\cJ}{\d\vtheta}
= \frac{\d\cL}{\d\vtheta}
= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \int_{t_0}^{t_1} \vlambda \cdot \frac{\p\vf}{\p\vtheta} \, \d t,
\end{equation}
%
in which the first two terms come from the regularization and initial
conditions, respectively.
\eqref{adjoint1} is the adjoint equation for the objective $\cJ(\vz_1,
\vtheta)$.
With the initial conditions set at the final time, we can integrate it
backward in time to obtain $\vlambda(t)$, which enters the above
equation and yields $\d\cJ / \d\vtheta$.

Note that \eqref{adjoint1} has the same form as \eqref{adjoint}, and
their solutions are related by
%
\begin{equation}
\vlambda = \vlambda_1 \cdot \vLambda
= \frac{\p\cJ}{\p\vz_1} \cdot \frac{\p\vz_1}{\p\vz}
= \frac{\p\cJ}{\p\vz}.
\end{equation}
And like $\vLambda \cdot \vDelta$, $\vlambda \cdot \vDelta$ is
time-invariant:
%
\begin{equation}
\vlambda_0 \cdot \vDelta_0
= \vlambda_1 \cdot \vDelta_1
= \vlambda \cdot \vDelta
= \frac{\p\cJ}{\p\vz} \cdot \frac{\p\vz}{\p\vz_0}
= \frac{\p\cJ}{\p\vz_0}
= \vlambda_0.
\end{equation}
%
Computing $\vlambda_1 \cdot \vDelta_1$ directly is expensive, but
solving \eqref{adjoint1} backward for $\vlambda_0$ is cheap.
This is related to the fact that the reverse-mode AD or backpropagation
is cheaper than the forward mode for optimization.

\vspace{1em}
\subsection{Objective on the State History}

The adjoint method applies to more complex cases too.
Let's consider an objective as a functional of the evolution history
with some regularization $\cR$ on $\vtheta$
%
\begin{equation}
\cJ = \cR(\vtheta) + \int_{t_0}^{t_1} g(\vz, t, \vtheta) \, \d t.
\end{equation}
%
The derivation is similar:
%
\begin{equation}
\cL = \cR(\vtheta)
+ \int_{t_0}^{t_1} \bigl\{ g(\vz, t, \vtheta) -
  \vlambda(t) \cdot \bigl[ \dot\vz - \vf(\vz, t, \vtheta) \bigr]
  \bigr\} \, \d t
\end{equation}
%
has gradient
%
\begin{align}
\frac{\d\cL}{\d\vtheta}
&= \frac{\d\cR}{\d\vtheta}
+ \int_{t_0}^{t_1} \Bigl[
    \frac{\p g}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
    + \frac{\p g}{\p\vtheta}
    - \vlambda \cdot \Bigl( \frac{\p\dot\vz}{\p\vtheta}
    - \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
    - \frac{\p\vf}{\p\vtheta} \Bigr) \Bigr] \, \d t \\
%
&= \frac{\d\cR}{\d\vtheta}
- \vlambda_1 \cdot \frac{\p\vz_1}{\p\vtheta}
+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \int_{t_0}^{t_1} \Bigl[
  \Bigl( \frac{\p g}{\p\vz} + \vlambda \cdot \frac{\p\vf}{\p\vz}
    + \dot\vlambda \Bigr) \cdot \frac{\p\vz}{\p\vtheta}
  + \frac{\p g}{\p\vtheta} + \vlambda \cdot \frac{\p\vf}{\p\vtheta}
\Bigr] \, \d t. \nonumber
\end{align}

The adjoint equation becomes
%
\begin{equation}
\dot\vlambda = - \frac{\p g}{\p\vz}
  - \vlambda \cdot \frac{\p\vf}{\p\vz},
\qquad
\vlambda_1 = \vzero,
\end{equation}
%
with the objective gradient
%
\begin{equation}
\frac{\d\cJ}{\d\vtheta}
= \frac{\d\cL}{\d\vtheta}
= \frac{\d\cR}{\d\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \int_{t_0}^{t_1} \Bigl( \frac{\p g}{\p\vtheta}
  + \vlambda \cdot \frac{\p\vf}{\p\vtheta} \Bigr) \, \d t.
\end{equation}

The time-invariant is now
%
\begin{equation}
\vlambda_1 \cdot \vDelta_1
= \vlambda_0 \cdot \vDelta_0
  + \int_{t_0}^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t
= \vlambda \cdot \vDelta
  + \int_t^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t'
= \vzero,
\end{equation}
%
so
%
\begin{equation}
\vlambda \cdot \vDelta
= - \int_t^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t'
= - \int_t^{t_1} \frac{\p g}{\p\vz_0} \, \d t'.
\end{equation}

\vspace{1em}
\subsection{Objective on the Observables}
\label{app:lightcone}

Now let's consider an objective that depends on different state
components at different times, e.g., in the Universe where further
objects intersected our past light cone earlier.
It falls between the previous scenarios, and we can derive its adjoint
equation similarly.

We denote the observables by $\hat\vz$, with different components $\hat
z_j$ affecting the objective $\cJ(\hat\vz, \vtheta)$ at different times
$\hat t_j$, i.e., $\hat z_i \triangleq z_i(\hat t_i)$.
The Lagrangian becomes
%
\begin{equation}
\cL = \cJ(\hat\vz, \vtheta)
- \sum_j \int_{t_0}^{\hat t_j} \lambda_j(t)
  \bigl[ \dot z_j - f_j(\vz, t, \vtheta) \bigr] \, \d t,
\end{equation}
%
constraining only the parts of trajectories inside the light cone.
Its gradient is
%
\begin{align}
\frac{\d\cL}{\d\vtheta}
&= \frac{\p\cJ}{\p\vtheta}
+ \frac{\p\cJ}{\p\hat\vz} \cdot \frac{\p\hat\vz}{\p\vtheta}
- \sum_j \int_{t_0}^{\hat t_j}
  \Bigl( \lambda_j \frac{\p\dot z_j}{\p\vtheta}
    - \vlambda \cdot \frac{\p\vf}{\p z_j} \frac{\p z_j}{\p\vtheta}
    - \lambda_j \frac{\p f_j}{\p\vtheta} \Bigr) \, \d t \\
%
&= \frac{\p\cJ}{\p\vtheta}
+ \sum_j \Bigl( \frac{\p\cJ}{\p\hat\vz} - \hat\vlambda \Bigr)
  \cdot \frac{\p\hat\vz}{\p\vtheta}
+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \sum_j \int_{t_0}^{\hat t_j} \Bigl[
  \Bigl( \vlambda \cdot \frac{\p\vf}{\p z_j} + \dot\lambda_j \Bigr)
  \frac{\p z_j}{\p\vtheta}
  + \lambda_j \frac{\p f_j}{\p\vtheta} \Bigr] \, \d t, \nonumber
\end{align}
%
where we have defined $\hat\vlambda$ similarly with components
$\hat\lambda_j \triangleq \lambda_j(\hat t_j)$.

Now we find the adjoint equation
%
\begin{equation}
\dot\vlambda = - \vlambda \cdot \frac{\p\vf}{\p\vz},
\qquad
\hat\vlambda = \frac{\p\cJ}{\p\hat\vz},
\end{equation}
%
that has the same form as \eqref{adjoint1}, with a slightly different
initial condition given at the respective observation time of each
component.
The objective gradient is also similar to the previous cases:
%
\begin{equation}
\frac{\d\cJ}{\d\vtheta}
= \frac{\d\cL}{\d\vtheta}
= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \sum_j \int_{t_0}^{\hat t_j} \lambda_j \frac{\p f_j}{\p\vtheta} \, \d t.
\end{equation}

Note that even though $\lambda_j(t)$ for $t > \hat t_j$ does not affect
the final gradient, they can enter the right-hand side of the adjoint
equation, and affect those $\lambda_k$ with $t < \hat t_k$, i.e., inside
the light cone.
Physically however, $\p f_j / \p z_k$ should vanish for spacelike
separated pairs of $z_j$ and $z_k$, even though the Newtonian
approximation we adopt introduces some small deviation.
Therefore, we can set $\lambda_j(t)$ to 0 for $t > \hat t_j$, and bump
it to $\p\cJ / \p\hat z_j$ at $\hat t_j$.

\vspace{1em}
\subsection{Discretize Then Optimize}

In practice, the time integration of \eqref{ode} is discrete.
Consider the explicit methods,
%
\begin{equation}
\vz_{i+1} = \vz_i + \vF_i(\vz_i, \vtheta), \qquad i = 0, \cdots, n-1,
\label{explicit}
\end{equation}
%
which include the leapfrog integrator commonly used for Hamiltonian
dynamics.
We want to propagate the gradients backward along the \emph{same}
discrete trajectory as taken by the forward integration.
Therefore, instead of the continuous adjoint equations derived above, we
need the adjoint method for the discrete integrator.

Without loss of generality, we derive the adjoint equation for an
objective depending on the state at all time steps, which can be easily
specialized to the 3 cases discussed above with only slight
modifications.
The discretized Lagrangian is now
%
\begin{equation}
\cL = \cJ(\vz_0, \cdots, \vz_n, \vtheta)
- \sum_{i=0}^{n-1} \vlambda_{i+1}
  \cdot \bigl[ \vz_{i+1} - \vz_i - \vF_i(\vz_i, \vtheta) \bigr],
\end{equation}
%
whose gradient is
%
\begin{align}
\frac{\d\cL}{\d\vtheta}
&= \frac{\p\cJ}{\p\vtheta}
+ \sum_{i=0}^n \frac{\p\cJ}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
- \sum_{i=0}^{n-1} \vlambda_{i+1}
  \cdot \Bigl( \frac{\p\vz_{i+1}}{\p\vtheta} - \frac{\p\vz_i}{\p\vtheta}
    - \frac{\p\vF_i}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
    - \frac{\p\vF_i}{\p\vtheta} \Bigr) \nonumber\\
%
&= \frac{\p\cJ}{\p\vtheta}
+ \sum_{i=0}^n \frac{\p\cJ}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
- \sum_{i=1}^n \vlambda_i \cdot \frac{\p\vz_i}{\p\vtheta}
+ \sum_{i=0}^{n-1} \vlambda_{i+1}
  \cdot \Bigl( \frac{\p\vz_i}{\p\vtheta}
    + \frac{\p\vF_i}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
    + \frac{\p\vF_i}{\p\vtheta} \Bigr) \nonumber\\
&= \frac{\p\cJ}{\p\vtheta}
+ \Bigl( \frac{\p\cJ}{\p\vz_n} - \vlambda_n \Bigr)
  \cdot \frac{\p\vz_n}{\p\vtheta}
+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta} \nonumber\\
&\hspace{4em} + \sum_{i=0}^{n-1} \Bigl[
  \Bigl( \frac{\p\cJ}{\p\vz_i} - \vlambda_i + \vlambda_{i+1}
    + \vlambda_{i+1} \cdot \frac{\p\vF_i}{\p\vz_i} \Bigr)
  \cdot \frac{\p\vz_i}{\p\vtheta}
  + \vlambda_{i+1} \cdot \frac{\p\vF_i}{\p\vtheta} \Bigr].
\end{align}

So the discrete adjoint equation is
%
\begin{equation}
\vlambda_{i-1} = \vlambda_i
  + \vlambda_i \cdot \frac{\p\vF_{i-1}}{\p\vz_{i-1}}
  + \frac{\p\cJ}{\p\vz_{i-1}},
\quad
i = n, \cdots, 1;
\qquad
\vlambda_n = \frac{\p\cJ}{\p\vz_n}.
\label{explicit_adj}
\end{equation}
%
We can iterate it backward in time to compute the final objective
gradient:
%
\begin{equation}
\frac{\d\cJ}{\d\vtheta}
= \frac{\d\cL}{\d\vtheta}
= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
+ \sum_{i=1}^n \vlambda_i \cdot \frac{\p\vF_{i-1}}{\p\vtheta}.
\label{explicit_grad}
\end{equation}
These equations are readily adaptable to simulated observables.
For snapshots at $t_n$ or interpolated between $t_{n-1}$ and $t_n$,
all $\p\cJ / \p\vz_i$ vanish except for the last one or two,
respectively.
For light cones, as discussed in \autoref{app:lightcone}, each component
of $\hat\vz$ is interpolated at different times, thus all $\p\cJ /
\p\vz_i$ vanish except for those times relevant for its interpolation,
and the corresponding $\lambda_i$ can be set to zero for $i$ greater
than the intersection time.

At the $i$-th iteration the adjoint variable requires the
vector-Jacobian product (VJP) $\vlambda_i \cdot \p\vF_{i-1} /
\p\vz_{i-1}$ and the partial objective derivative $\p\cJ / \p\vz_{i-1}$
at the next time step, which can be easily computed by AD if the whole
forward history of \eqref{explicit} has been saved.
However, this can be extremely costly in memory, which can be alleviated
by checkpointing algorithms such as Revolve and its successors
\citep{Revolve}.
Alternatively, if a solution to \eqref{ode} is unique, we can integrate
it backward and recover the history, which is easy for reversible
Hamiltonian dynamics and with reversible integrators such as leapfrog.
When the $N$-body dynamics become too chaotic, one can use more precise
floating-point numbers and/or save multiple checkpoints\footnote{This is
different from the checkpointing in the Revolve algorithm, which needs
to rerun the forward iterations.} during the forward evolution, from
which the backward evolution can be resumed piecewise.

%\vspace{1em}
%\section{Hamiltonian Dynamics}
%
%Consider a Hamiltonian system, with $\vz = (\vx, \vp)$ where $\vx$ and
%$\vp$ are the canonical coordinates and momenta.
%Other variables split likewise: $\vDelta = (\vDelta_\vx, \vDelta_\vp)$,
%$\vLambda = (\vLambda_\vx, \vLambda_\vp)$, and $\vlambda =
%(\vlambda_\vx, \vlambda_\vp)$.
%Hamilton's equation of motion is
%%
%\begin{equation}
%\vf = \begin{pmatrix}
%  \hphantom{-} \p\cH/\p \vp \\ - \p\cH/\p \vx
%\end{pmatrix}.
%\label{Hamilton}
%\end{equation}
%%
%The dynamics is time-reversible: flipping the arrow of time $\d/\d t \to
%- \d/\d t$ and the direction of momenta $\vp \to -\vp$ preserve the form
%of \eqref{ode}.
%It can be reversible even after discretization if appropriate integrator
%is used.
%
%The variational equation \eqref{variational} and the adjoint equation
%\eqref{adjoint} become
%%
%\begin{align}
%\begin{pmatrix}
%  \dot\vDelta_{\vx\vx} & \dot\vDelta_{\vx\vp} \\
%  \dot\vDelta_{\vp\vx} & \dot\vDelta_{\vp\vp}
%\end{pmatrix}
%&= \begin{pmatrix}
%  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vx}
%  & \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp} \\
%  - \dfrac{\p^2\cH}{\p\vx\p\vx}
%  & - \dfrac{\p^2\cH}{\p\vx\p\vp} \\
%\end{pmatrix}
%\cdot
%\begin{pmatrix}
%  \vDelta_{\vx\vx} & \vDelta_{\vx\vp} \\
%  \vDelta_{\vp\vx} & \vDelta_{\vp\vp}
%\end{pmatrix}, \nonumber \\
%%
%\qquad
%- \begin{pmatrix}
%  \dot\vLambda_{\vx\vx} & \dot\vLambda_{\vx\vp} \\
%  \dot\vLambda_{\vp\vx} & \dot\vLambda_{\vp\vp}
%\end{pmatrix}
%&= \begin{pmatrix}
%  \hphantom{-} \dfrac{\p^2\cH}{\p\vx\p\vp}
%  & - \dfrac{\p^2\cH}{\p\vx\p\vx} \\
%  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp}
%  & - \dfrac{\p^2\cH}{\p\vp\p\vx} \\
%\end{pmatrix}
%\cdot
%\begin{pmatrix}
%  \vLambda_{\vx\vx} & \vLambda_{\vx\vp} \\
%  \vLambda_{\vp\vx} & \vLambda_{\vp\vp}
%\end{pmatrix}.
%\end{align}
%%
%Interestingly, their forms are identical once we swap rows of
%$\vLambda$, and flip the directions of time and momenta.
%
%As discussed before, $\vlambda$ follows adjoint equations similar to
%that of $\vLambda$ above.
%One can integrate it and \eqref{Hamilton} backward together to compute
%$\d\cJ / \d\vtheta$.
%
%
%\vspace{1em}
%\section{\texorpdfstring{$N$}{N}-body Dynamics}
%
%We model self-gravitating collisionless dark matter with $N$ particles
%each of mass $m$ in a periodic box of comoving volume $V$, whose
%physical length expands proportional to a scale factor $a(t)$.
%The canonical coordinates are the comoving positions $\vx = (\vx_1,
%\cdots, \vx_N)$, and their canonical momenta are $\vp = (\vp_1, \cdots,
%\vp_N)$ with $\vp_i = a^2 m \dot\vx_i$ for each particle.
%
%The $N$-body Hamiltonian is
%%
%\begin{equation}
%\cH = \sum_i \frac{\vp_i^2}{2 m a^2}
%+ \sum_{i<j} \frac{m^2 \varphi(\vx_i - \vx_j)}{a},
%\end{equation}
%%
%where $\varphi$ is the potential of a unit-mass particle at $a=1$ under
%periodic boundary condition
%%
%\begin{equation}
%\nabla^2 \varphi
%= 4\pi G \bigl[ \deltaD(\vx - \vx') - V^{-1} \bigr].
%\end{equation}
%%
%So the only nonzero second $\cH$ derivatives are
%%
%\begin{equation}
%\frac{\p^2\cH}{\p\vp_i\p\vp_j} = \frac{\vI_3}{2 m a^2},
%\qquad
%\frac{\p^2\cH}{\p\vx_i\p\vx_j}
%= \deltaK_{ij} \sum_{k \neq i} \frac{m^2}a
%  \frac{\p\varphi(\vx_i-\vx_k)}{\p\vx_i\p\vx_i}
%+ \frac{m^2}a \frac{\p\varphi(\vx_i-\vx_j)}{\p\vx_i\p\vx_j},
%\end{equation}
%%
%where $\vI_3$ is the $3\times3$ identity matrix, the sum is the tidal
%tensor at $\vx_i$, and the second term vanishes when $i=j$.

\bibliographystyle{aasjournal}
\bibliography{adjoint}

%\listofchanges

\end{document}

### 2022-11-15 docs/papers/adjoint/grads.py

@@ -1,0 +1,139 @@
+import os
+
+import numpy as np
+import jax
+import jax.numpy as jnp
+import matplotlib.pyplot as plt
+
+from pmwd import (
+    Configuration,
+    SimpleLCDM,
+    boltzmann,
+    white_noise, linear_modes,
+    lpt,
+    nbody,
+    scatter,
+)
+from pmwd.vis_util import simshow
+
+
+def model(modes, cosmo, conf):
+    cosmo = boltzmann(cosmo, conf)
+    modes = linear_modes(modes, cosmo, conf)
+    ptcl, obsvbl = lpt(modes, cosmo, conf)
+    ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
+    dens = scatter(ptcl, conf)
+    return dens
+
+
+def obj(tgt_dens, modes, cosmo, conf):
+    cosmo = boltzmann(cosmo, conf)
+    modes = linear_modes(modes, cosmo, conf)
+    ptcl, obsvbl = lpt(modes, cosmo, conf)
+    ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
+    dens = scatter(ptcl, conf)
+    return (dens - tgt_dens).var()
+
+obj_grad = jax.grad(obj, argnums=(1, 2))
+
+
+ptcl_spacing = 1.
+ptcl_grid_shape = (128,) * 3
+conf = Configuration(ptcl_spacing, ptcl_grid_shape, mesh_shape=2,
+                     a_start=1/16, a_nbody_maxstep=1/16)
+
+cosmo = SimpleLCDM(conf)
+
+
+# control the target dens variation
+fname = 'dens.npy'
+if not os.path.exists(fname):
+    seed = 0  # seed for target
+    modes = white_noise(seed, conf)
+
+    dens = model(modes, cosmo, conf)  # target density
+    jnp.save(fname, dens)
+dens = jnp.load(fname)
+
+
+# control the input modes variation
+fname = 'modes.npy'
+if not os.path.exists(fname):
+    seed = 1
+    modes = white_noise(seed, conf, real=True)
+    jnp.save(fname, modes)
+modes = jnp.load(fname)
+
+
+n = 8
+fname_am = 'grads_am{}.npy'  # adjoint mode gradients
+fname_ad = 'grads_ad{}.npy'  # AD mode gradients
+
+if not os.path.exists(fname_am.format(0)):  # adjoint gradients
[…]

### 2022-11-15 docs/papers/adjoint/optim.py

@@ -1,0 +1,117 @@
+from functools import partial
+
+from PIL import Image, ImageFont, ImageDraw
+import jax
+import jax.numpy as jnp
+from jax.example_libraries.optimizers import adam
+import matplotlib.pyplot as plt
+
+from pmwd import (
+    Configuration,
+    SimpleLCDM,
+    boltzmann,
+    white_noise, linear_modes,
+    lpt,
+    nbody,
+    scatter,
+)
+from pmwd.vis_util import simshow
+
+
+
+def model(modes, cosmo, conf):
+    modes = linear_modes(modes, cosmo, conf)
+    ptcl, obsvbl = lpt(modes, cosmo, conf)
+    ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
+    return ptcl
+
+
+def scatter_2d(ptcl, conf):
+    dens = jnp.zeros(tuple(2*s for s in conf.mesh_shape), dtype=conf.float_dtype)
+    dens = scatter(ptcl, conf, mesh=dens, val=1, cell_size=conf.cell_size / 2)
+    return dens.sum(axis=2)
+
+
+pmwdshow = partial(simshow, figsize=(27/2, 8), cmap='cividis', colorbar=False)
+
+
+def obj(tgt, modes, cosmo, conf):
+    ptcl = model(modes, cosmo, conf)
+    dens = scatter_2d(ptcl, conf)
+    return (dens - tgt).var() / tgt.var()
+
+obj_valgrad = jax.value_and_grad(obj, argnums=1)
+
+
+def optim(tgt, modes, cosmo, conf, iters=100, lr=0.1):
+    init, update, get_params = adam(lr)
+    state = init(modes)
+
+    def step(step, state, tgt, cosmo, conf):
+        modes = get_params(state)
+        value, grads = obj_valgrad(tgt, modes, cosmo, conf)
+        state = update(step, grads, state)
+        return value, state
+
+    tgt = jnp.asarray(tgt)
+    for i in range(iters):
+        value, state = step(i, state, tgt, cosmo, conf)
+
+    modes = get_params(state)
+    return value, modes
+
+
+text = 'pmwd'
[…]

### 2022-11-15 docs/papers/adjoint/scaling.py

@@ -1,0 +1,87 @@
+import os
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+mesh_shapes = [1, 2]
+ptcl_grid_sizes = [2**n for n in range(5, 10)]
+
+
+number = 1
+repeat = 8
+exec_times = np.empty((2 + len(mesh_shapes), len(ptcl_grid_sizes)))
+fname = 'scaling.txt'
+
+if not os.path.exists(fname):
+    import timeit
+
+    from pmwd import (
+        Configuration,
+        SimpleLCDM,
+        boltzmann,
+        white_noise, linear_modes,
+        lpt,
+        nbody,
+    )
+
+
+    for i, mesh_shape in enumerate(mesh_shapes):
+        for j, ptcl_grid_size in enumerate(ptcl_grid_sizes):
+            ptcl_spacing = 1.
+            ptcl_grid_shape = (ptcl_grid_size,) * 3
+            conf = Configuration(ptcl_spacing, ptcl_grid_shape, mesh_shape=mesh_shape)
+            print(f'ptcl_grid_size={ptcl_grid_size}, mesh_shape={mesh_shape}:')
+
+            cosmo = SimpleLCDM(conf)
+
+            seed = 0
+            modes = white_noise(seed, conf)
+
+
+            boltzmann(cosmo, conf)  # do not overwrite cosmo
+            t = timeit.repeat('boltzmann(cosmo, conf).h.block_until_ready()',
+                              repeat=repeat, number=number, globals=globals())
+            t = min(t) / number
+            exec_times[0, j] = t
+            print(f'boltz {t}')
+            cosmo = boltzmann(cosmo, conf)  # overwrite cosmo
+
+
+            ptcl, obsvbl = lpt(linear_modes(modes, cosmo, conf), cosmo, conf)  # do not overwrite modes
+            t = timeit.repeat('lpt(linear_modes(modes, cosmo, conf), cosmo, conf)[0].vel.block_until_ready()',
+                              repeat=repeat, number=number, globals=globals())
+            t = min(t) / number
+            print(f'  lpt {t}')
+            exec_times[1, j] = t
+
+
+            nbody(ptcl, obsvbl, cosmo, conf)  # do not overwrite ptcl and obsvbl
[…]

### 2022-11-15 docs/papers/adjoint/slab.py

@@ -1,0 +1,38 @@
+import matplotlib.pyplot as plt
+plt.style.use('font.mplstyle')
+
+from pmwd import (
+    Configuration,
+    SimpleLCDM,
+    boltzmann,
+    white_noise, linear_modes,
+    lpt,
+    nbody,
+    scatter,
+)
+from pmwd.vis_util import simshow
+
+
+def model(modes, cosmo, conf):
+    cosmo = boltzmann(cosmo, conf)
+    modes = linear_modes(modes, cosmo, conf)
+    ptcl, obsvbl = lpt(modes, cosmo, conf)
+    ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
+    return ptcl, obsvbl
+
+
+ptcl_spacing = 1.
+ptcl_grid_shape = (512,) * 3
+conf = Configuration(ptcl_spacing, ptcl_grid_shape, mesh_shape=2)
+
+cosmo = SimpleLCDM(conf)
+
+seed = 0
+modes = white_noise(seed, conf)
+
+ptcl, obsvbl = model(modes, cosmo, conf)
+
+dens = scatter(ptcl, conf)
+fig, _ = simshow(dens[:16].mean(axis=0), norm='CosmicWebNorm')
+fig.savefig('slab.pdf')
+plt.close(fig)

### 2022-11-15 docs/papers/pmwd/make.sh
Add pmwd code paper source

@@ -1,0 +1,17 @@
+#!/usr/bin/env bash
+
+# adapted from https://github.com/dfm/emcee/blob/main/joss/make_latex.sh
+# and https://github.com/openjournals/whedon/blob/master/resources/docker-entrypoint.sh
+
+#echo Downloading...
+#rm -rf latex.template apa.csl joss-logo.png aas-logo.png
+#url=https://raw.githubusercontent.com/openjournals/whedon/ca417d169c382b02ee5dc858e6651aa080c5c40a/resources/joss
+## use chicago style if w/o apa.csl
+#wget -q $url/{latex.template,aas-logo.png}
+#wget -q $url/logo.png -O joss-logo.png
+#echo Done
+
+paper=pmwd
+
+pandoc $paper.md -Cso $paper.tex --template latex.template
+pandoc $paper.md -Co $paper.pdf --template latex.template

### 2022-11-18 docs/papers/sto/sto.tex
Add sto paper source

\documentclass[usenatbib]{mnras}

% Allow "Thomas van Noord" and "Simon de Laguarde" and alike to be sorted by "N" and "L" etc. in the bibliography.
% Write the name in the bibliography as "\VAN{Noord}{Van}{van} Noord, Thomas"
\DeclareRobustCommand{\VAN}[3]{#2}
\let\VANthebibliography\thebibliography
\def\thebibliography{\DeclareRobustCommand{\VAN}[3]{##3}\VANthebibliography}

\usepackage{microtype}
\usepackage{newpxtext, eulerpx}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
%\usepackage{booktabs}

\newcommand{\pmwd}{{\usefont{T1}{nova}{m}{sl}pmwd}}

\renewcommand{\sectionautorefname}{Sec.}
\renewcommand{\subsectionautorefname}{Sec.}
\renewcommand{\appendixautorefname}{App.}
\renewcommand{\figureautorefname}{Fig.}
%\newcommand{\subfigureautorefname}{\figureautorefname}

%FIXME copied from ../adjoint/adjoint.tex
\newcommand{\deltaD}{\delta^\textsc{d}}
\newcommand{\deltaK}{\delta^\textsc{k}}
\renewcommand{\d}{d}
\newcommand{\p}{\partial}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cH}{\mathcal{H}}
\bmdefine{\vzero}{0}
\bmdefine{\vI}{I}
\bmdefine{\vnabla}{\nabla}
\bmdefine{\vtheta}{\theta}  % parameters
\bmdefine{\vomega}{\omega}  % white noise modes
\bmdefine{\vk}{k}  % wavevectors
\bmdefine{\vx}{x}  % comoving and canonical coordinates
\bmdefine{\vq}{q}  % Lagrangian coordinates
\bmdefine{\vs}{s}  % displacements
\bmdefine{\vp}{p}  % canonical momenta
\bmdefine{\va}{a}  % accelerations
\bmdefine{\vz}{z}  % states
\bmdefine{\vf}{f}
\bmdefine{\vF}{F}
\bmdefine{\vDelta}{\Delta}
\bmdefine{\vLambda}{\Lambda}
\bmdefine{\vlambda}{\lambda}
\bmdefine{\vvarphi}{\varphi}
\bmdefine{\vxi}{\xi}  % x adjoint
\bmdefine{\vpi}{\pi}  % p adjoint
\bmdefine{\valpha}{\alpha}  % force vjp x gradient
\bmdefine{\vzeta}{\zeta}  % force vjp theta gradient
\newcommand{\half}{\nicefrac12}
\newcommand{\As}{A_\mathrm{s}}
\newcommand{\ns}{n_\mathrm{s}}
\newcommand{\Omegam}{\Omega_\mathrm{m}}
\newcommand{\Omegab}{\Omega_\mathrm{b}}
\newcommand{\Mpc}{\mathrm{Mpc}}
\newcommand{\ic}{\mathrm{i}}
\newcommand{\Plin}{P_\mathrm{lin}}

\newcommand{\YL}[1]{\textcolor{Bittersweet}{#1}}

\title[Spatiotemporally Optimized Simulation]
{Spatiotemporal Optimization of Cosmological Particel-Mesh Simulation}

\author[Zhang, Li, Jamieson, et al.]{
%
Yucheng Zhang,$^{1, 2}$
%
Yin Li,$^{1, 3}$\thanks{Email: XYZ, eelregit@gmail.com, and XYZ}
%
and Drew Jamieson$^{4}$
%
\\$^1$Department of Mathematics and Theory, Peng Cheng Laboratory,
Shenzhen, Guangdong 518066, China
%
\\$^2$Center for Cosmology and Particle Physics, Department of Physics,
New York University, New York, NY 10003, USA
%
\\$^3$Center for Computational Astrophysics \& Center for Computational
Mathematics, Flatiron Institute, New York, NY 10010, USA
%
\\$^4$Max Planck Institute for Astrophysics, 85748 Garching bei
M\"unchen, Germany
}

\date{Accepted XXX. Received YYY; in original form ZZZ}
\pubyear{2023}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\begin{abstract}
We sharpen PM force and optimize integration time steps.
\end{abstract}

\begin{keywords}
cosmology: large-scale structure of Universe
-- methods: numerical
-- software: development
\end{keywords}

\section{Introduction}

\section{Methods}

\subsection{Spatial Optimization}

\subsection{Temporal Optimization}

\begin{table*}
  \centering
  \caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
    parameters.
    Note that the grid ratio need next fast len to determine the mesh
    shape for fast FFT.
    Given the box size, the mesh shape determines the cell size.
    The softening parameter gives the ratio of the comving softening
    length to the mean particle spacing.
    The curvature $\Omega_k$ is related to the separate universe
    simulation.
    We sample parameters applicable to GADGET-4 during data generation,
    and those applicable to \pmwd\ during training.
    \textsuperscript\dag The box size is determined jointly by the
    \pmwd\ mesh shape and mesh cell size below, which are assumed to be
    sampled independently.
    In practice, we sample one conditioned on the other and the box
    size.
    * Number of time steps from beginning to end ...
  }
  \label{tab:param}
  \begin{tabular}{lcccr}
    \toprule
    applicability & parameter & distribution & range & sampling \\
    \midrule
    & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
    \cmidrule(lr){2-5}
    & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
    \cmidrule(lr){2-5}
    & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
    \cmidrule(lr){2-4}
    & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
    \cmidrule(lr){2-4}
    GADGET-4 & $A_\mathrm{s} \times 10^9$ & log-uniform & [1, 4) \\
    \cmidrule(lr){2-4}
    \& \pmwd\ & $n_\mathrm{s}$ & log-uniform & [0.75, 1.25) & RQMC \\
    \cmidrule(lr){2-4}
    & $\Omega_\mathrm{m}$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
    \cmidrule(lr){2-4}
    & $\Omega_\mathrm{b} / \Omega_\mathrm{m}$ & log-uniform & [1/16, 1/2) \\
    \cmidrule(lr){2-4}
    & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
    \cmidrule(lr){2-4}
    & $h$ & log-uniform & [0.5, 1) \\
    \cmidrule(lr){1-4}
    GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
    \cmidrule(lr){1-5}
    & mesh shape & log-uniform & [128, 512] \\
    \cmidrule(lr){2-4}
    \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
    \cmidrule(lr){2-4}
    & number of time steps* & log-uniform & [10, 1000] \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsection{Parameters and Configurations}

We run 512 GADGET-4 simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
Each simulation has $128^3$ particles and writes 121 snapshots linearly
spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
spacing $1/128$.
$\Delta a$ is an offset, common for all snapshots in each simulation,
that varies as in \autoref{tab:param} and helps to sample different
alignment between snapshot and start/stop times.

The curvature $\Omega_k$ is related to the separate universe simulation
\citep{LiEtAl2014a, WagnerEtAl2015}.

\begin{figure*}
  \centering
  \includegraphics[width=0.9\linewidth]{sobol.pdf}
  \caption{Randomized Quasi-Monte Carlo (RQMC) configuration with
    scrambled Sobol sequence of 512 points in 9D.
    Lower triangular panels show the 2D projections and the diagonal
    panels are the 1D cumulative histograms.
    From left to right (top to bottom), we use each dimension of the
    sample to scale the parameters as ordered in \autoref{tab:param}.
    We use the \texttt{scipy.stats.qmc} package \citep{SciPy} to
    generate the Sobol sequence \citep{Sobol1967}, which uses the
    direction number from \citet{JoeKuo2008} and the Owen scrambling
    \citep{Owen1998}.
    We search among 65536 scrambling seeds to minimize the mixture
    discrepancy (a uniformity measure) proposed in \citet{Zhou2013MD}.
  }
  \label{fig:sobol}
\end{figure*}

\begin{figure*}
  \centering
  %\includegraphics[width=0.8\linewidth]{name.ext}
  \caption{1D histogram of the parameters sampled with RQMC and scaled
  as in \autoref{tab:param}}.
  \label{fig:hist}
\end{figure*}

\subsection{Simulations}

\citet{GADGET-4}

We configure GADGET-4 with high force and time integration accuracy
settings.
For gravitational force, we choose pure FMM with $p=5$ and opening angle
0.4.
%FIXME list other from Config.sh and param.txt

\section{Results}

\section{Conclusions}

\section*{Acknowledgements}

The Flatiron Institute is supported by the Simons Foundation.

\section*{Code Availability}

\pmwd\ is open-source on
\href{https://github.com/eelregit/pmwd}{GitHub}, including the
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{source
and scripts of this paper}.

\bibliographystyle{mnras}
\bibliography{sto}

\bsp  % typesetting comment
\label{lastpage}
\end{document}

### 2022-11-18 docs/papers/adjoint/perf.py
Edit drafts, ApJS submission

@@ -11,7 +11,7 @@ ptcl_grid_sizes = [2**n for n in range(5, 10)]
 number = 1
 repeat = 8
 exec_times = np.empty((2 + len(mesh_shapes), len(ptcl_grid_sizes)))
-fname = 'scaling.txt'
+fname = 'perf.txt'
 
 if not os.path.exists(fname):
     import timeit
@@ -83,5 +83,5 @@ ax.set_xlabel('ptcl grid size')
 ax.set_ylabel('wall time [sec]')
 ax.legend()
 
-fig.savefig('scaling.pdf')
+fig.savefig('perf.pdf')
 plt.close(fig)

### 2022-11-25 docs/papers/sto/sto.tex
sto paper: add cubic Hermite equations

@@
 \usepackage{graphicx}
-%\usepackage{booktabs}
+\usepackage{booktabs}
 
@@
 \section{Methods}
+
+Given two snapshots at $a_i$ and $a_{i+1}$, we get the particle displacements
+in between at $a$ using interpolation with the cubic Hermite spline,
+\begin{align}
+  \bm{x}(a) =\ &h_{00}(\alpha)\bm{x}_i + h_{10}(\alpha)(a_{i+1} - a_i)\bm{v}_i + \nonumber\\
+           &h_{01}(\alpha)\bm{x}_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\bm{v}_{i+1} \,,
+\end{align}
+where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis functions read
+\begin{align}
+  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1 \,,\nonumber\\
+  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha \,,\nonumber\\
+  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2 \,,\nonumber\\
+  h_{11}(\alpha) &= \alpha^3 - \alpha^2 \,.
+\end{align}
+The particle velocity is then given by the derivative $d\bm{x}/da$.
 

### 2022-12-01 docs/papers/sto/sto.tex
FIXUP draft: add some obj text

@@
 \bmdefine{\vp}{p}  % canonical momenta
+\bmdefine{\vv}{v}  % FIXME some velocities
 \bmdefine{\va}{a}  % accelerations
@@
 \begin{align}
-  \bm{x}(a) =\ &h_{00}(\alpha)\bm{x}_i + h_{10}(\alpha)(a_{i+1} - a_i)\bm{v}_i + \nonumber\\
-           &h_{01}(\alpha)\bm{x}_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\bm{v}_{i+1} \,,
+  \vx(a) =\ &h_{00}(\alpha)\vx_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
+           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1} \,,
 \end{align}
@@
 \end{align}
-The particle velocity is then given by the derivative $d\bm{x}/da$.
+The particle velocity is then given by the derivative $\d\vx / \d a$.
+
+\subsection{Objective Function}
+
+Again combine Lagrangian and Eulernian fields.
+
+A mean squared error (MSE) loss on the fields would focus too much on
+the small scales. In this work we choose uniform weights in the scale
+space, by weighing each mode of wavenumber $k$ by $k^{-3}$ in the MSE
+loss
+\begin{equation}
+  \cJ = \mathbb{E} \sum_\vk k^{-3}
+    \bigl| \hat f(\vk) - f(\vk) \bigr|^2.
+\end{equation}
 

### 2022-12-01 pmwd/sto_util.py
FIXUP add an option for output contraint in MLP

@@ -13,6 +13,7 @@ from pmwd.pm_util import rfftnfreq
 class MLP(nn.Module):
     features: Sequence[int]
     activator: Callable[[jnp.ndarray], jnp.ndarray] = nn.softplus
+    outivator: Callable[[jnp.ndarray], jnp.ndarray] = nn.softplus
 
     def setup(self):
         self.layers = [nn.Dense(f) for f in self.features]
@@ -23,6 +24,10 @@ class MLP(nn.Module):
             x = lyr(x)
             if i != len(self.layers)-1:
                 x = self.activator(x)
+            else:
+                if self.outivator is not None:
+                    x = self.outivator(x)
+
         return x
 
 
@@ -100,9 +105,9 @@ def sharpening(pot, cosmo, conf, a):
     # neural nets
     nets = [MLP(features=n) for n in conf.so_nodes]
     # modification factors to the laplace potential
-    g_k = 1 + nets[0].apply(cosmo.so_params[0], fts[0])
+    g_k = nets[0].apply(cosmo.so_params[0], fts[0])
     g_k = g_k.reshape(g_k.shape[:-1])  # remove the trailing axis of dim one
-    f_kvec = [1 + nets[1].apply(cosmo.so_params[1], ft) for ft in fts[1]]
+    f_kvec = [nets[1].apply(cosmo.so_params[1], ft) for ft in fts[1]]
     f_kvec = [f.reshape(f.shape[:-1]) for f in f_kvec]
     pot *= g_k * math.prod(f_kvec)
     return pot

### 2022-12-02 docs/papers/sto/sto.tex
FIXUP draft: add so methods

@@
 \bmdefine{\vtheta}{\theta}  % parameters
+\bmdefine{\vvartheta}{\vartheta}  % NN parameters
 \bmdefine{\vomega}{\omega}  % white noise modes
@@
 \newcommand{\ic}{\mathrm{i}}
-\newcommand{\Plin}{P_\mathrm{lin}}
+\newcommand{\linear}{\mathrm{lin}}
+\newcommand{\tophat}{\mathrm{TH}}
+\newcommand{\gauss}{\mathrm{G}}
 
@@
 
-Again combine Lagrangian and Eulernian fields.
+Again combine Lagrangian and Eulerian fields.
 
@@
 
+An alternative loss function can be the Eulernian density weighted
+particle displacement MSE.
+
 \subsection{Spatial Optimization}
+
+The dimensionless power spectrum
+%
+\begin{equation}
+\Delta^2(k) \triangleq \frac{k^3 P(k)}{2 \pi^2}.
+\end{equation}
+
+Variance of linear matter overdensity in a top-hat window
+%
+\begin{equation}
+\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
+  \Delta_\linear^2(k) W_\tophat^2(kR),
+\end{equation}
+where $W_\tophat(kR) = \cdots$ is the Fourier transform of the top-hat
+window.
+Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
+$W_\gauss(kR) = \cdots$.
+
+Various nonlinear scales are
+\begin{itemize}
+\item $k_P^{-1}$: determined by $\Delta_\linear^2(k_P) \triangleq 1$;
+\item $R_\tophat$: determined by $\sigma_\tophat^2(R_\tophat) \triangleq
+  1$;
+\item $R_\gauss$: determined by $\sigma_\gauss^2(R_\tophat) \triangleq
+  1$;
+\item $R_\mathrm{d}$: rms linear theory displacement;
+\item $\d R / \d\ln a$: time derivatives of the above scales.
+\end{itemize}
+
+Other relevant scales are
+\begin{itemize}
+\item $l_\mathrm{cell}$: PM cell size.
+\item particle spacing;
+\item softening length.
+\end{itemize}
+
+Relevant scale-independent dimensionless factors are
+\begin{itemize}
+\item $D(a)$: linear growth factor;
+\item $f(a)$: linear growth rate $f \triangleq \d\ln D / \d\ln a$;
+\item $\Omegam(a)$: matter density parameter at $a$;
+\item $\d\ln H / \d\ln a$: time derivative of Hubble expansion;
+\item $a$: scale factor;
+\item $\Delta a$: time step size in $a$.
+\end{itemize}
+
+We combine the above list of parameters in $\vvartheta$.
+
+Force sharpening
+%
+\begin{equation}
+\frac{k_i}{k^2} \to \frac{k_i}{k^2}
+  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
+\end{equation}
+%
+where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
+function of the dimensionless features, e.g.,
+%
+\begin{equation}
+g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots, D, f, \cdots).
+\end{equation}
 
@@
 parameters.
-    Note that the grid ratio need next fast len to determine the mesh
-    shape for fast FFT.
+Note that the grid ratio need next fast len to determine the mesh shape
+for fast FFT.
 Given the box size, the mesh shape determines the cell size.
-    The softening parameter gives the ratio of the comving softening
-    length to the mean particle spacing.
-    The curvature $\Omega_k$ is related to the separate universe
-    simulation.
-    We sample parameters applicable to GADGET-4 during data generation,
-    and those applicable to \pmwd\ during training.
-    \textsuperscript\dag The box size is determined jointly by the
-    \pmwd\ mesh shape and mesh cell size below, which are assumed to be
-    sampled independently.
-    In practice, we sample one conditioned on the other and the box
-    size.
-    * Number of time steps from beginning to end ...
+The softening parameter gives the ratio of the comving softening length
+to the mean particle spacing.
+The curvature $\Omega_k$ is related to the separate universe simulation.
+We sample parameters applicable to GADGET-4 during data generation, and
+those applicable to \pmwd\ during training.
+\textsuperscript\dag The box size is determined jointly by the \pmwd\
+mesh shape and mesh cell size below, which are assumed to be sampled
+independently.
+In practice, we sample one conditioned on the other and the box size.
+* Number of time steps from $a=1/16$ to $a=1$.
 }
@@
   \cmidrule(lr){2-4}
-    GADGET-4 & $A_\mathrm{s} \times 10^9$ & log-uniform & [1, 4) \\
+  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
   \cmidrule(lr){2-4}
-    \& \pmwd\ & $n_\mathrm{s}$ & log-uniform & [0.75, 1.25) & RQMC \\
+  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
   \cmidrule(lr){2-4}
-    & $\Omega_\mathrm{m}$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
+  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
   \cmidrule(lr){2-4}
-    & $\Omega_\mathrm{b} / \Omega_\mathrm{m}$ & log-uniform & [1/16, 1/2) \\
+  & $\Omegab / \Omegam$ & log-uniform & [1/16, 1/2) \\
   \cmidrule(lr){2-4}

### 2022-12-05 docs/papers/sto/sto.tex
FIXUP draft: fill in top-hat and Gaussian windows

@@
 \end{equation}
-where $W_\tophat(kR) = \cdots$ is the Fourier transform of the top-hat
-window.
+where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
+Fourier transform of the top-hat window.
 Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
-$W_\gauss(kR) = \cdots$.
+$W_\gauss(kR) = e^{-(kR)^2/2}$.
 

### 2022-12-07 docs/papers/pp/pp.tex
Add 4PM draft

\documentclass[a4paper]{article}

\usepackage{CJK}
\usepackage{microtype}
%\usepackage{savesym}
%  \savesymbol{lambdabar}
\usepackage{newpxtext,eulerpx}
%  \restoresymbol{newpx}{lambdabar}
\usepackage[T1]{fontenc}
\usepackage[margin=3cm]{geometry}
%\usepackage{setspace}
\usepackage{amsmath}
%  \allowdisplaybreaks
\usepackage{bm}
%\usepackage{mathrsfs}
%\usepackage{soul}
%\usepackage{tensor}
\usepackage{graphicx}
%\usepackage{booktabs}
%\usepackage{placeins}
%\usepackage{subcaption}
%\usepackage[caption=false]{subfig}  % revtex hates caption
\usepackage{hyperref}
  \hypersetup{
    colorlinks=true,
    linkcolor=BrickRed,
    filecolor=BrickRed,
    citecolor=LimeGreen,
    urlcolor=Cerulean,
  }
\usepackage{natbib}
\usepackage[dvipsnames,svgnames,x11names]{xcolor}
\usepackage{comment}
  \newif\ifshow
  %\showtrue  % show
  \showfalse  % hide
  %
  \ifshow
    \includecomment{hide}
  \else
    \excludecomment{hide}
  \fi
\usepackage{tikz}
  \usetikzlibrary{positioning, fit, calc, arrows.meta}

\newcommand{\pmwd}{{\usefont{T1}{nova}{m}{sl}pmwd}}

\newcommand{\gkai}[1]{\begin{CJK*}{UTF8}{gkai}\raisebox{.1em}{(}#1\raisebox{.1em}{)}\end{CJK*}}

\renewcommand{\sectionautorefname}{Sec.}
\renewcommand{\subsectionautorefname}{Sec.}
\renewcommand{\appendixautorefname}{App.}
\renewcommand{\figureautorefname}{Fig.}
\newcommand{\subfigureautorefname}{\figureautorefname}

\newcommand{\order}{\mathcal{O}}
\newcommand{\deltaD}{\delta^\textsc{d}}
\newcommand{\deltaK}{\delta^\textsc{k}}
\renewcommand{\d}{d}
\newcommand{\p}{\partial}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cH}{\mathcal{H}}
\bmdefine{\vzero}{0}
\bmdefine{\vI}{I}
\bmdefine{\vnabla}{\nabla}
\bmdefine{\vtheta}{\theta}  % parameters
\bmdefine{\vomega}{\omega}  % white noise modes
\bmdefine{\vk}{k}  % wavevectors
\bmdefine{\vn}{n}  % wavevector indices
\bmdefine{\vx}{x}  % comoving and canonical coordinates
\bmdefine{\vq}{q}  % Lagrangian coordinates
\bmdefine{\vs}{s}  % displacements
\bmdefine{\vp}{p}  % canonical momenta
\bmdefine{\va}{a}  % accelerations
\bmdefine{\vz}{z}  % states
\bmdefine{\vf}{f}
\bmdefine{\vF}{F}
\bmdefine{\vvarphi}{\varphi}
\newcommand{\As}{A_\mathrm{s}}
\newcommand{\ns}{n_\mathrm{s}}
\newcommand{\Omegam}{\Omega_\mathrm{m}}
\newcommand{\Omegab}{\Omega_\mathrm{b}}
\newcommand{\Mpc}{\mathrm{Mpc}}
\newcommand{\ic}{\mathrm{i}}
\newcommand{\knyq}{k_\mathrm{Nyq}}
\newcommand{\kfun}{k_\mathrm{fun}}

\newcommand{\GPU}{NVIDIA A100 80GB SXM4}

\newcommand{\YL}[1]{\textcolor{Bittersweet}{#1}}

\begin{document}

\title{Patching Particle-Particle Particle-Mesh Method}

\author{Yin Li \gkai{李寅}}

\date{2022-12}

\maketitle

\begin{abstract}
Adding the PP compensation to the PM force.
\end{abstract}

\section{Introduction}

\section{Methods}

\subsection{Particle-Mesh Force}

\YL{Alex suggested starting from 1D.}

\YL{Describe briefly normal PM here. Also introduce possible filter
below. Explain that we are not doing splitting the usual way.}

\subsubsection{Partial Sum}

In a periodic box of size $L_1 \times \cdots \times L_d$ in
$\mathbb{R}^d$, Poisson's equation,
%
\begin{equation}
\nabla^2 u = f,
\end{equation}
%
becomes
%
\begin{equation}
- k_\vn^2 u(\vk_\vn) = f(\vk_\vn),
\end{equation}
%
in Fourier space, with discretized wavevectors $\vk_\vn \triangleq (2\pi
n_1 / L_1, \cdots, 2\pi n_d / L_d )^\intercal$ and integrer $n_i \in
\mathbb{Z}/N_i\mathbb{Z}$.
And the force $- \vnabla u$ becomes $i \vk_\vn f / k_\vn^2$.

At finite resolution of the mesh cell size $l$, the bandwidth is
truncated at the Nyquist frequency $\knyq = \pi / l$.
In a 1D periodic length $L$ with a mesh of size $N = L / l$, the
partial sum is equivalent to a convolution by (the Dirichlet kernel)
%
\begin{align}
\frac1{2 L} \Bigg(
  \sum_{n=-\lfloor \frac{N}2\rfloor}^{\lfloor \frac{N}2\rfloor}
  + \sum_{n=-\lfloor \frac{N-1}2\rfloor}^{\lfloor \frac{N-1}2\rfloor}
\Bigg)
\exp\bigl( i n\kfun x \bigr)
%
&= \frac{
  \sin \bigl[
    \bigl( \bigl\lfloor\frac{N}2\bigr\rfloor + \frac12 \bigr) \kfun x
  \bigr]
  + \sin \bigl[
    \bigl( \bigl\lfloor\frac{N-1}2\bigr\rfloor + \frac12 \bigr) \kfun x
  \bigr]
}{2 \sin \bigl( \frac12 \kfun x \bigr) L} \nonumber\\
%
&= \frac{\sin \bigl( \knyq x \bigr)}
        {\sin \bigl( \frac12 \kfun x \bigr) L}
  \cdot \begin{cases}
    \cos \bigl( \frac12 \kfun x \bigr) & \text{if } N \text{ is even}, \\
    1 & \text{if } N \text{ is odd}.
  \end{cases}
\end{align}
%
$\kfun = 2\pi / L$ is the fundamental frequency, and the first sum
includes the symmetrized Nyquist mode for even $N$.

To preserve Hermiticity, for even $N$ we truncate the force even before
$\kfun$, due to the sign ambiguity of the wavevector at the Nyquist
frequency.
%
\begin{equation}
\frac1L
  \sum_{n=-\lfloor \frac{N-1}2\rfloor}^{\lfloor \frac{N-1}2\rfloor}
  \exp\bigl( i n\kfun x \bigr)
%
= \frac{\sin \bigl[
    \bigl( \bigl\lfloor\frac{N-1}2\bigr\rfloor + \frac12 \bigr) \kfun x
  \bigr]
}{\sin \bigl( \frac12 \kfun x \bigr) L}
\end{equation}

\YL{TODO: verify that in 1D the partial sum explains the force error.}

\YL{Minimize force error above a certain radius, e.g. beyond 27 cells,
(in a way similar to choosing NUFFT kernel?):
\begin{itemize}
\item \url{https://arxiv.org/abs/1712.04732},
\item \url{https://www.frontiersin.org/articles/10.3389/fphy.2016.00028/full},
\item \url{https://journals.aps.org/pre/pdf/10.1103/PhysRevE.95.063303}.
\end{itemize}
}

\YL{Filter/window design: Jupyter notebook with interactive sliders,
\href{https://docs.scipy.org/doc/scipy/reference/signal.windows.html}{scipy.signal.windows},
\href{https://docs.scipy.org/doc/scipy/reference/signal.html}{scipy.signal}.
}

\YL{2 possible ways to optimize:
suppress the oscillation outside of the PP patch size;
shift the phase of the oscillation so that there are both positive and
negative parts between two nearby mesh points.
}

\subsubsection{Equivalent Point Source}

simple cubic (CIC), bcc, interlocking/interlacing cubic.
\YL{See \citet{HockneyEastwood1988} for the interlacing one.
Other references:
\begin{itemize}
\item \url{https://aip.scitation.org/doi/10.1063/1.3430521},
\item \url{https://aip.scitation.org/doi/pdf/10.1063/1.3430521},
\item \url{https://aip.scitation.org/doi/pdf/10.1063/1.3657407}.
\end{itemize}
}

%* Suppose the two interlacing meshes are A and B, is the following a way
%  to have 16 equivalent grid points (up to octopole)?
%  1. src: (A + shifted B) at A, tgt: A
%  2. src: (B + shifted A) at B, tgt: B

\subsection{Particle-Particle Force}

\subsection{Force Patching}

\YL{Quantify force error including (position-dependent) magnitude and
volume.}

For cosmological applications, we want to optimize for point clouds that
are uniform (and Gaussian with known power spectrum) on large scales.
\YL{Complexity analysis given variance of particle density.}

\YL{Optimal mesh cell size given variance of particle density for fixed
error budget, cosmology and time dependence.}

\section{Results}

\section{Conclusions}

\bibliographystyle{plainnat}
\bibliography{pp}

\end{document}

### 2022-12-27 docs/papers/sto/sto.tex
FIXUP draft: nothing
Draft: add some observable snapshot text

@@
 \bmdefine{\vtheta}{\theta}  % parameters
-\bmdefine{\vvartheta}{\vartheta}  % NN parameters
+\bmdefine{\vvartheta}{\vartheta}  % relevant scales and factors to form
+                                  % dimensionless features as NN inputs
 \bmdefine{\vomega}{\omega}  % white noise modes
@@
 
+\subsection{Observable}
+
 Given two snapshots at $a_i$ and $a_{i+1}$, we get the particle displacements
@@
 The particle velocity is then given by the derivative $\d\vx / \d a$.
+
+Our observable is a snapshot $\hat \vz = (\hat\vx, \hat\vp)^\intercal$ at any
+given $a$, which in general can be given by the interpolation of two snapshots
+$\vz_{k-1}$ and $\vz_k$.
+Following the discussion in \citet{Li2022a}, the observe operator in the adjoint
+equation time stepping from $a_i$ to $a_{i-1}$ now reads
+\begin{align}
+  O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
+  + \frac{\p\cJ}{\p\hat\vx}\frac{\p\hat\vx}{\p\vx_{i-1}} \,, \nonumber\\
+                    \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
+  + \frac{\p\cJ}{\p\hat\vp}\frac{\p\hat\vp}{\p\vp_{i-1}} \,,
+\end{align}
+where $(\vxi, \vpi)$ are the adjoint variables of $(\vx, \vp)$, and
+${\p\hat\vz}/{\p\vz_i}$ is zero except for the relevant $\vz_{k-1}$ and $\vz_k$.
+Similarly we have in the initial conditions of the adjoint equations
+\begin{align}
+  \vxi_n &= \frac{\p\cJ}{\p\hat\vx}\frac{\p\hat\vx}{\p\vx_n} \,, \nonumber\\
+  \vpi_n &= \frac{\p\cJ}{\p\hat\vp}\frac{\p\hat\vp}{\p\vp_n} \,.
+\end{align}
 
@@
 \begin{equation}
-g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots, D, f, \cdots).
+g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots; D, f, \cdots).
 \end{equation}

### 2022-12-27 pmwd/gravity.py
FIXUP rename sto_util to so_util

@@ -4,7 +4,7 @@ from jax import custom_vjp
 from pmwd.scatter import scatter
 from pmwd.gather import gather
 from pmwd.pm_util import fftfreq, fftfwd, fftinv
-from pmwd.sto_util import sharpening
+from pmwd.so_util import sharpening
 
 
 @custom_vjp

### 2022-12-28 docs/papers/sto/sto.tex
Draft: add obj gradient equation

@@
 \usepackage{booktabs}
+\usepackage{nicefrac}
 
@@
   O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
-  + \frac{\p\cJ}{\p\hat\vx}\frac{\p\hat\vx}{\p\vx_{i-1}} \,, \nonumber\\
+  + \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vx_{i-1}} \,, \nonumber\\
                     \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
-  + \frac{\p\cJ}{\p\hat\vp}\frac{\p\hat\vp}{\p\vp_{i-1}} \,,
+  + \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vp_{i-1}} \,,
 \end{align}
-where $(\vxi, \vpi)$ are the adjoint variables of $(\vx, \vp)$, and
-${\p\hat\vz}/{\p\vz_i}$ is zero except for the relevant $\vz_{k-1}$ and $\vz_k$.
+where $\vlambda = (\vxi, \vpi)$ are the adjoint variables of $\vz = (\vx,
+\vp)^\intercal$, and ${\p\hat\vz}/{\p\vz_i}$ is zero except for the relevant
+$\vz_{k-1}$ and $\vz_k$.
 Similarly we have in the initial conditions of the adjoint equations
 \begin{align}
-  \vxi_n &= \frac{\p\cJ}{\p\hat\vx}\frac{\p\hat\vx}{\p\vx_n} \,, \nonumber\\
-  \vpi_n &= \frac{\p\cJ}{\p\hat\vp}\frac{\p\hat\vp}{\p\vp_n} \,.
+  \vxi_n &= \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vx_n} \,, \nonumber\\
+  \vpi_n &= \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vp_n} \,.
 \end{align}
+In the reverse time evolution of the adjoint equations, these adjoint varibales
+remain zero until the time stepping from $a_{k+1}$ to $a_k$.
 
@@
 particle displacement MSE.
+
+As derived in \citet{Li2022a}, the gradient of the objective function
+$\cJ(\hat\vz(\vz_{k-1}, \vz_k, \vtheta), \theta)$ is accumulated through
+\begin{multline}
+  \frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
+  %
+  + \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vtheta}
+  + \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vtheta}
+  %
+  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
+  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
+  %
+  - \sum_{i=1}^n \Bigl[
+    \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
+      \bigr) K(t_i, t_i, t_{i-\half}) \\
+    + \vxi_i \cdot \vp_{i-\half}
+      \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \\
+    + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta} + \vzeta_{i-1}
+      \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
+    \Bigr],
+\label{KDFKO_grad}
+\end{multline}
+where $\vtheta$ could include cosmological parameters, initial conditions, as
+well as STO parameters in this work.
 
@@
 \subsection{Temporal Optimization}
+
+We use a B-spline to model the function $a(t)$ that maps uniformly sampled $t$
+values to optimally distributed $a$ time steps.
+The spline is parameterized with a few control points that are updated during
+training.
 
@@
 
-\pmwd\ is open-source on
+\pmwd\ \citep{Li2022b} is open-source on
 \href{https://github.com/eelregit/pmwd}{GitHub}, including the

### 2023-01-11 docs/papers/sto/sto.tex
Draft: symplectic integration and its adjoint

--- 2022-12-05 docs/papers/sto/sto.tex
@@
 \bmdefine{\vz}{z}  % states
+\bmdefine{\vo}{o}  % observable
 \bmdefine{\vf}{f}
@@
 
+\subsection{Simulation and Differentiation}
+
+We briefly recap the discussion in \citet{Li2022a} and introduce the notations.
+We denote the particles' state with $\vz = (\vx, \vp)^\intercal$ and their
+adjoint variables with $\vlambda = (\vxi, \vpi)$.
+The forward simulation is evolved using a symplectic integrator.
+From $a_{i-1}$ to $a_i$, the particles' state is updated alternately for $\vx$
+and $\vp$ through a loop over the symplectic integration indexed with $j$
+\begin{alignat}{2}
+&D_{j-1}^j: &\qquad \vx_j &= \vx_{j-1} +
+  \vp_{j-1} D(a^p_{j-1}, a^x_{j-1}, a^x_j), \nonumber\\
+%
+&F_j: & \va_j &\triangleq -\vnabla\varphi(\vx_j), \nonumber\\
+%
+&K_{j-1}^j: & \vp_j &= \vp_{j-1} + \va_j K(a^x_j, a^p_{j-1}, a^p_j).
+\label{DFK}
+\end{alignat}
+The intermediate times are given by
+\begin{align}
+a^x_j &= (1 - C^x_j) a_{i-1} + C^x_j a_i, \nonumber\\
+a^p_j &= (1 - C^p_j) a_{i-1} + C^p_j a_i,
+\end{align}
+where $C^x$ and $C^p$ are the cumulative coefficients of the symplectic
+integrator, e.g. $C^x = (0, 0, 1)$ and $C^p = (0, 0.5, 1)$ for leapfrog.
+The reverse time integration from $a_i$ to $a_{i-1}$ and adjoint equations
+\begin{alignat}{2}
+&K_j^{j-1}: &\qquad \vp_{j-1} &= \vp_j +
+  \va_j K(a^x_j, a^p_j, a^p_{j-1}), \nonumber\\
+%
+& & \vxi_{j-1} &= \vxi_j - \valpha_j K(a^x_j, a^p_j, a^p_{j-1}) \nonumber\\
+%
+&D_j^{j-1}: & \vx_{j-1} &= \vx_j +
+  \vp_{j-1} D(a^p_{j-1}, a^x_j, a^x_{j-1}), \nonumber\\
+%
+& & \vpi_{j-1} &= \vpi_j - \vxi_{j-1} D(a^p_{j-1}, a^x_j, a^x_{j-1}), \nonumber\\
+%
+&F_{j-1}: & \va_{j-1} &\triangleq -\vnabla\vvarphi(\vx_{j-1}), \nonumber\\
+%
+& & \valpha_{j-1} &\triangleq -\vpi_{j-1} \cdot
+  \frac{\p\vnabla\vvarphi_{j-1}}{\p\vx_{j-1}}, \nonumber\\
+& & \vzeta_{j-1} &\triangleq - \vpi_{j-1} \cdot
+  \frac{\p\vnabla\vvarphi_{j-1}}{\p\vtheta}.
+\label{KDF_adj}
+\end{alignat}
+
+We consider an objective function $\cJ(\vo, \theta)$ defined on the observable
+$\vo(\vz_0,...,\vz_n,\vtheta)$.
+The observe operator from $a_i$ to $a_{i-1}$ reads
+\begin{align}
+  O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
+  + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vx_{i-1}}, \nonumber\\
+                    \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
+  + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vp_{i-1}}.
+\end{align}
+
+% % TODO update the final gradient for loop over j
+% The final gradient is accumulated through
+% \begin{multline}
+%   \frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
+%   %
+%   + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vtheta}
+%   %
+%   + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
+%   + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
+%   %
+%   - \sum_{i=1}^n \Bigl[
+%     \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
+%       \bigr) K(t_i, t_i, t_{i-\half}) \\
+%     + \vxi_i \cdot \vp_{i-\half}
+%       \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta}
+%     \Bigr],
+% \label{KDFO_grad}
+% \end{multline}
+% where $\vtheta$ could include cosmological parameters, initial conditions, as
+% well as STO parameters in this work.
+
+\subsection{Snapshot Observable}
+
 Given two snapshots at $a_i$ and $a_{i+1}$, we get the particle displacements
@@
   \vx(a) =\ &h_{00}(\alpha)\vx_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
-           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1} \,,
+           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
 \end{align}
@@
 \begin{align}
-  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1 \,,\nonumber\\
-  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha \,,\nonumber\\
-  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2 \,,\nonumber\\
-  h_{11}(\alpha) &= \alpha^3 - \alpha^2 \,.
+  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
+  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
+  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
+  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
 \end{align}
 The particle velocity is then given by the derivative $\d\vx / \d a$.
+
+Our observable $\vo$ is a snapshot $\hat\vz = (\hat\vx, \hat\vp)^\intercal$ at a
+given time $a_o$, which in general can be given by the interpolation of two
+snapshots $\vz_{k-1}$ and $\vz_k$ where $a_{k-1} \leq a_o < a_k$.
+Then ${\p\vo}/{\p\vz_i}$ would be zero except for the relevant $\vz_{k-1}$ and
+$\vz_k$.
+We have in the initial conditions of the adjoint equations
+\begin{align}
+  \vxi_n &= \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vx_n}, \nonumber\\
+  \vpi_n &= \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vp_n}.
+\end{align}
+In the reverse time evolution of the adjoint equations, these adjoint varibales
+remain zero until the time stepping from $a_{k+1}$ to $a_k$.
 
@@
 \subsection{Temporal Optimization}
+
+We use a spline to model the function $a(t)$ that maps uniformly sampled $t$
+values to optimally distributed $a$ time steps.
+The spline is parameterized with a number of control points that are updated
+during training.
+The gradients (i.e. cotangents) are back propagated from an array of integration
+time steps to the control points through a VJP transformation of $a(t)$.
 

### 2023-01-11 docs/papers/sto/sto.tex
Draft: a few sentences on TO

@@
 values to optimally distributed $a$ time steps.
-The spline is parameterized with a number of control points that are updated
+The spline is parameterized with a number of control points that are optimized
 during training.
-The gradients (i.e. cotangents) are back propagated from an array of integration
-time steps to the control points through a VJP transformation of $a(t)$.
+The gradients (i.e. cotangents) are back propagated from the array of
+integration time steps to the control points through a VJP transformation of
+$a(t)$.
+
+The time-dependence of the discrete symplectic integrator mainly enters through
+the drift and kick factors in \eqref{DFK}.
+Besides, the input features of the SO NNs could also depend on time.
+Therefore when we perform TO together with SO, the force evaluation in
+\eqref{DFK} which contains SO also becomes time-dependent and needs to be
+optimized.
 

### 2023-01-11 pmwd/configuration.py
Change symplectic splitting back to a interface

@@ -1,20 +1,20 @@
-import os
 from functools import partial
 from math import ceil
 from typing import ClassVar, Optional, Tuple, Union
 
 from numpy.typing import DTypeLike
 import jax
-jax.config.update("jax_enable_x64", True)
 from jax import ensure_compile_time_eval
 import jax.numpy as jnp
-jnp.set_printoptions(precision=3, edgeitems=2, linewidth=128)
 from jax.tree_util import tree_map
 
 from pmwd.tree_util import pytree_dataclass
 
 
-os.environ['NVIDIA_TF32_OVERRIDE'] = '1'  # enable TF32
+jax.config.update("jax_enable_x64", True)
+
+
+jnp.set_printoptions(precision=3, edgeitems=2, linewidth=128)
 
 
 @partial(pytree_dataclass, aux_fields=Ellipsis, frozen=True)
@@ -75,6 +75,10 @@ class Configuration:
         Maximum scale factor N-body time integration step size. It determines the
         number of steps ``a_nbody_num``, the actual step size ``a_nbody_step``, and the
         steps ``a_nbody``.
+    symp_splits : tuple of float 2-tuples, optional
+        Symplectic splitting method composition, with each 2-tuples being drift and then
+        kick coefficients. Its adjoint has the same splits in reverse nested orders,
+        i.e., kick and then drift. Default is the Newton-Störmer-Verlet-leapfrog method.
     chunk_size : int, optional
         Chunk size to split particles in batches in scatter and gather to save memory.
 
@@ -124,11 +128,9 @@ class Configuration:
     a_lpt_maxstep: float = 1/128
     a_nbody_maxstep: float = 1/64
 
-    chunk_size: int = 2**24
+    symp_splits: Tuple[Tuple[float, float], ...] = ((0, 0.5), (1, 0.5))
 
-    # padded and cumsumed symplectic coefs
-    symp_cd: jnp.ndarray = jnp.array([[0, 0, 1], [0, 0.5, 1]], dtype=float_dtype)
-    symp_order: Optional[int] = None
+    chunk_size: int = 2**24
 
     def __post_init__(self):
         if self._is_transforming():
@@ -163,8 +165,11 @@ class Configuration:
         if self.growth_atol is None:
             object.__setattr__(self, 'growth_atol', growth_tol)
 
-        if self.symp_order is None:
-            object.__setattr__(self, 'symp_order', self.symp_cd.shape[1] - 1)
+        if any(len(s) != 2 for s in self.symp_splits):
[…]

### 2023-01-11 pmwd/nbody.py

@@ -1,6 +1,5 @@
 from functools import partial
 
-import numpy as np
 from jax import value_and_grad, jit, vjp, custom_vjp
 import jax.numpy as jnp
 from jax.tree_util import tree_map
@@ -43,7 +42,6 @@ def drift(a_vel, a_prev, a_next, ptcl, cosmo, conf):
     return ptcl.replace(disp=disp)
 
 
-# TODO deriv wrt a
 def drift_adj(a_vel, a_prev, a_next, ptcl, ptcl_cot, cosmo, cosmo_cot, conf):
     """Drift, and particle and cosmology adjoints."""
     factor_valgrad = value_and_grad(drift_factor, argnums=3)
@@ -70,7 +68,6 @@ def kick(a_acc, a_prev, a_next, ptcl, cosmo, conf):
     return ptcl.replace(vel=vel)
 
 
-# TODO deriv wrt a
 def kick_adj(a_acc, a_prev, a_next, ptcl, ptcl_cot, cosmo, cosmo_cot, cosmo_cot_force, conf):
     """Kick, and particle and cosmology adjoints."""
     factor_valgrad = value_and_grad(kick_factor, argnums=3)
@@ -105,7 +102,7 @@ def force_adj(a, ptcl, ptcl_cot, cosmo, conf):
     ptcl = ptcl.replace(acc=acc)
 
     # particle and cosmology vjp
-    a_cot, ptcl_cot_force, cosmo_cot_force, conf_cot = gravity_vjp(ptcl_cot.vel)
+    _, ptcl_cot_force, cosmo_cot_force, _ = gravity_vjp(ptcl_cot.vel)
     ptcl_cot = ptcl_cot.replace(acc=ptcl_cot_force.disp)
 
     return ptcl, ptcl_cot, cosmo_cot_force
@@ -141,29 +138,38 @@ def observe_init(a, ptcl, obsvbl, cosmo, conf):
 
 @jit
 def nbody_init(a, ptcl, obsvbl, cosmo, conf):
-    # ptcl = force(a, ptcl, cosmo, conf)
+    ptcl = force(a, ptcl, cosmo, conf)
 
-    # ptcl = coevolve_init(a, ptcl, cosmo, conf)
+    ptcl = coevolve_init(a, ptcl, cosmo, conf)
 
-    # obsvbl = observe_init(a, ptcl, obsvbl, cosmo, conf)
+    obsvbl = observe_init(a, ptcl, obsvbl, cosmo, conf)
 
[…]

### 2023-01-12 docs/papers/sto/sto.tex
Draft: just a few sentences

@@
 \subsection{Snapshot Observable}
+\label{sec:snapobs}
 
@@
 \subsection{Spatial Optimization}
+\label{sec:so}
 
@@
 
+In Fourier space, the gravitational force field reads
+\begin{equation}
+- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\end{equation}
 Force sharpening
@@
 
+Parameters in these neural networks are optimized during training.
+
 \subsection{Temporal Optimization}
+\label{sec:to}
 
@@
 the drift and kick factors in \eqref{DFK}.
-Besides, the input features of the SO NNs could also depend on time.
+Besides, the input features of the SO NNs introduced in \autoref{sec:so} could
+also depend on time.
 Therefore when we perform TO together with SO, the force evaluation in
@@
 Given the box size, the mesh shape determines the cell size.
-The softening parameter gives the ratio of the comving softening length
+The softening parameter gives the ratio of the comoving softening length
 to the mean particle spacing.
@@
 
+\subsection{Training}
+
+Given 512 Gadget simulations and 121 output snapshots each, in total we have
+61952 snapshots for training.
+Following \autoref{tab:param}, given a snapshot from Gadget simulations, we
+configure \pmwd\ with the same initial coniditions and cosmological parameters,
+and randomly sample the mesh shape (or the mesh cell size) and the number of
+time steps.
+
 \section{Results}

### 2023-02-01 docs/papers/sto/sto.tex
Draft: just a few sentences

@@
 
+We perform parallel training using a number of nodes that each hosts multiple
+GPU devices.
+The training on a given snapshot is performed on a single GPU device.
+The gradients are collected and averaged to update the STO parameters.
+To make sure that each individual training in a parallel run finishes in similar
+time, they share the same number of time steps.
+
 \section{Results}
+
+\subsection{SO function}
+
+We use the \texttt{PySR} \citep{PySR, Cranmer2020} symbolic regression code to
+fit for an analytic function expression of the SO neural network.
 

### 2023-02-07 docs/papers/sto/sto.tex
FIXUP draft: change title, add TODO

@@
 \title[Spatiotemporally Optimized Simulation]
-{Spatiotemporal Optimization of Cosmological Particel-Mesh Simulation}
+{Neural and Symbolic Spatiotemporal Optimization of Cosmological
+Particel-Mesh Simulation}
 
@@
 \end{keywords}
+
+\section*{TODO}
+\begin{itemize}
+\item Non-integer multiple mesh shapes: for uniform particle grid, test
+  (with double precision on CPUs) whether they start moving by artifact
+  forces on cell scales, in 3 cases: no deconvolution, with CIC
+  deconvolution, with SO.
+\item For TO, probably necessary to first incorporate Oliver Hahn's
+  recent improvement on FastPM integrator arXiv:2301:09655.
+\end{itemize}
 

### 2023-02-07 docs/papers/sto/sto.tex
FIXUP draft: add TODO

@@
 \begin{itemize}
+\item Be careful with the non-deterministic ICs generated on GPUs.
+  \begin{enumerate}
+  \item Saving them on disks and loading them during training solves
+    this but may introduce IO or memory problem;
+  \item Generating them on CPUs for training data and also training
+    might suffice while requireing such data loader implementation;
+  \item Generating them on GPUs for training data and training will
+    introduce additional source of initial deviations to be nonlinearly
+    amplified. There are flags that makes certain operations
+    deterministic but slow on GPUs, but those are enviroment variables
+    that function globally and cannot be applied locally only on the IC
+    generation during training.
+  \end{enumerate}
 \item Non-integer multiple mesh shapes: for uniform particle grid, test

### 2023-02-07 docs/papers/sto/sto.tex
FIXUP draft: add TODO

@@
     generation during training.
+  \item Generating ICs and then train on GPUs with double precision and
+    then compute the loss at single precision (Gadget4 should be run at
+    double precision) should work, and likely be the simplest solution.
+    But for this to work the final training loss should be much larger
+    than the level of reproducibility.
   \end{enumerate}

### 2023-02-07 pmwd/train_util.py
Add a func to scale Sobol samples to sim params

@@ -1,0 +1,51 @@
+import jax
+import jax.numpy as jnp
+import numpy as np
+import optax
+from functools import partial
+
+from pmwd import Configuration, Cosmology, nbody
+
+
+def scale_Sobol(fn='sobol.txt'):
+    """Scale the Sobol sequence samples, refer to the Table in the paper."""
+    sobol = np.loadtxt(fn).T
+    # functions mapping uniform random samples in [0, 1] to a desired one
+    f_uni = lambda x, a, b : a + x * (b - a)
+    f_log_uni = lambda x, a, b : np.exp(f_uni(x, np.log(a), np.log(b)))
+    def f_log_trap(x, a, b, c):
+        # a, b, c are the locations of the first 3 points of the symmetric trapezoid
+        h = 1 / (c - a)
+        x1 = (b - a) * h / 2
+        x2 = (2*c - b - a) * h / 2
+        y = np.zeros_like(x)
+        m = (x < x1)
+        y[m] = a + np.sqrt(2 * (b - a) * x[m] / h)
+        m = (x1 <= x) & (x < x2)
+        y[m] = x[m] / h + (a + b) / 2
+        m = (x2 <= x)
+        y[m] = c + b - a - np.sqrt((1 - x[m]) * 2 * (b - a) / h)
+        return np.exp(y)
+
+    # 0: box size, log-trapezoidal
+    sobol[0] = f_log_trap(sobol[0], np.log(128)+np.log(0.2),
+                          np.log(512)+np.log(0.2), np.log(128)+np.log(5)
+                          )
+    # 1: snapshot offset, uniform
+    sobol[1] = f_uni(sobol[1], 0, 1/128)
+    # 2: A_s_1e9, log-uniform
+    sobol[2] = f_log_uni(sobol[2], 1, 4)
+    # 3: n_s, log-uniform
+    sobol[3] = f_log_uni(sobol[3], 0.75, 1.25)
+    # 4: Omega_m, log-uniform
+    sobol[4] = f_log_uni(sobol[4], 1/5, 1/2)
+    # 5: Omega_b / Omega_m, log-uniform
+    sobol[5] = f_log_uni(sobol[5], 1/16, 1/2)
+    # 6: Omega_k / (1 - Omega_k), uniform
+    sobol[6] = f_uni(sobol[6], -0.5, 0.5)
+    # 7: h, log-uniform
[…]

### 2023-02-07 docs/papers/sto/sto.tex
FIXUP draft: rm scaled histogram placeholder
FIXUP draft: add TODO

@@
 \begin{itemize}
+\item Compare one pair of Gadget4 and \pmwd\ (with say 100 to 1000 time
+    steps) results and make sure that they agree perfectly on the large
+    scales. For this purpose the box size should be large enough (in
+    linear regime), say 1 Gpc/$h$.
 \item Be careful with the non-deterministic ICs generated on GPUs.
@@
 
-\begin{figure*}
-  \centering
-  %\includegraphics[width=0.8\linewidth]{name.ext}
-  \caption{1D histogram of the parameters sampled with RQMC and scaled
-  as in \autoref{tab:param}}.
-  \label{fig:hist}
-\end{figure*}
-
 \subsection{Simulations}

### 2023-02-08 docs/examples/sto/g4gen/sobol.py
FIXUP sobol sequence: gen and plot util

@@ -1,11 +1,32 @@
 import os
-import sys
 
 import numpy as np
 import matplotlib.pyplot as plt
 
+from pmwd.train_util import scale_Sobol
 
-def plt_proj(filename='sobol.txt', max_rows=None):
+
+def gen_sobol(filename, d=9, m=9):
+    from scipy.stats.qmc import Sobol, discrepancy
+
+    disc_min = np.inf
+    for seed in range(65536):
+        sampler = Sobol(d, scramble=True, seed=seed)  # d is the dimensionality
+        sample = sampler.random_base2(m)  # m is the log2 of the number of samples
+        disc = discrepancy(sample, method='MD')
+        if disc < disc_min:
+            seed_min = seed
+            disc_min = disc
+    print(f'0 <= seed = {seed_min} < 65536, minimizes mixture discrepancy = '
+            f'{disc_min}')
+    # seed_min = 55868, mixture discrepancy = 0.016109347957680598
+
+    sampler = Sobol(d, scramble=True, seed=seed_min)
+    sample = sampler.random_base2(m)
+    np.savetxt(filename, sample)
+
+
+def plt_proj(filename, max_rows=None):
     sample = np.loadtxt(filename, max_rows=max_rows)
 
     n, d = sample.shape
@@ -40,7 +61,7 @@ def plt_proj(filename='sobol.txt', max_rows=None):
             axes[i, j].scatter(* sample.T[[j, i]],
                                s=2, marker='o', alpha=0.75, linewidth=0)
 
-        axes[i, i].hist(sample[:, i], bins='sqrt',
+        axes[i, i].hist(sample[:, i], bins='sqrt', range=(0, 1),
                         density=True, cumulative=True, histtype='step')
 
         for j in range(i+1, d):
@@ -51,7 +72,58 @@ def plt_proj(filename='sobol.txt', max_rows=None):
     plt.close(fig)
 
 
+def plt_scaled(filename):
+    sample = scale_Sobol(filename).T
+    assert sample.shape[0] == 9
+
+    fig, axes = plt.subplots(nrows=3, ncols=3, sharey=True, figsize=(9, 9),
+                             subplot_kw={'box_aspect': 1})
+
+    params = np.array([
+        ('box size in Mpc', 'log', 25.6, 2560),
[…]

### 2023-02-08 pmwd/train_util.py
FIXUP scale_Sobol: shrink Ob/Om and Ok/(1-Ok) ranges

@@ -40,9 +40,11 @@ def scale_Sobol(fn='sobol.txt'):
     # 4: Omega_m, log-uniform
     sobol[4] = f_log_uni(sobol[4], 1/5, 1/2)
     # 5: Omega_b / Omega_m, log-uniform
-    sobol[5] = f_log_uni(sobol[5], 1/16, 1/2)
+    sobol[5] = f_log_uni(sobol[5], 1/8, 1/4)
+    sobol[5] *= sobol[4]  # get Omega_b
     # 6: Omega_k / (1 - Omega_k), uniform
-    sobol[6] = f_uni(sobol[6], -0.5, 0.5)
+    sobol[6] = f_uni(sobol[6], -1/3, 1/3)
+    sobol[6] = sobol[6] / (1 + sobol[6])  # get Omega_k
     # 7: h, log-uniform
     sobol[7] = f_log_uni(sobol[7], 0.5, 1)
     # 8: softening ratio

### 2023-02-23 docs/examples/sto/g4gen/setup_g4sims.py
FIXUP generate IC for g4 on GPU

@@ -1,6 +1,7 @@
 """A script to setup directories and files etc for Gadget-4 simulations."""
 import os
 import sys
+import jax
 
 import numpy as np
 
@@ -25,8 +26,9 @@ def gen_g4files(sim_dir, i, fn_sobol='sobol.txt',
     tpl_job : str
         The template for generating the job.sh file.
     """
-    # initial condition
-    ptcl, cosmo, conf, sobol = gen_ic(i, fn_sobol, re_sobol=True)
+    # generate initial condition on GPU, to be consistent with training
+    with jax.default_device(jax.devices('gpu')[0]):
+        ptcl, cosmo, conf, sobol = gen_ic(i, fn_sobol, re_sobol=True)
     write_gadget_hdf5(os.path.join(sim_dir, 'ic'), 1, conf.a_start, ptcl, cosmo, conf)
 
     with (open(tpl_config, 'r') as f,

### 2023-02-23 pmwdlib/kernels.cc.cu
fix bin stride

@@ -38,8 +38,8 @@ namespace jax_pmwd {
 
 #define DIM 3
 #define blk 1
-#define BINSIZE 1
-#define SCATTER_DEV_TIME
+#define BINSIZE 16
+//#define SCATTER_DEV_TIME
 
 // Macro to catch CUDA errors in CUDA runtime calls
 #define CUDA_SAFE_CALL(call)                                                   \
@@ -396,9 +396,9 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
 
     // parameters for shared mem using bins to group cells
     uint32_t bin_size = BINSIZE;
-    uint32_t nbinx = stride[0]/bin_size+1;
-    uint32_t nbiny = stride[1]/bin_size+1;
-    uint32_t nbinz = stride[2]/bin_size+1;
+    uint32_t nbinx = static_cast<uint32_t>(std::ceil(1.0*stride[0]/bin_size));
+    uint32_t nbiny = static_cast<uint32_t>(std::ceil(1.0*stride[1]/bin_size));
+    uint32_t nbinz = static_cast<uint32_t>(std::ceil(1.0*stride[2]/bin_size));
 
     uint32_t npts_mem_size = sizeof(uint32_t) * n_particle;
     uint32_t nbins_mem_size = sizeof(uint32_t) * nbinx*nbiny*nbinz;

### 2023-02-23 pmwdlib/kernels.cc.cu
adding offset and ptcl spacinng

@@ -88,7 +88,7 @@ cal_cellid(T_int2 n_particle, T_int1* pmid, T_float* disp, T_float cell_size, T_
 
 template <typename T_int1, typename T_int2, typename T_float>
 __global__ void
-cal_binid(T_int1 bin_size_x, T_int1 bin_size_y, T_int1 bin_size_z, T_int1 nbinx, T_int1 nbiny, T_int1 nbinz, T_int2 n_particle, T_int1* pmid, T_float* disp, T_float cell_size, T_int1 stridex, T_int1 stridey, T_int1 stridez, T_int1* binid, T_int1* sortidx){
+cal_binid(T_int1 bin_size_x, T_int1 bin_size_y, T_int1 bin_size_z, T_int1 nbinx, T_int1 nbiny, T_int1 nbinz, T_int2 n_particle, T_int1* pmid, T_float* disp, T_float cell_size, T_float ptcl_spacing, T_int1 stridex, T_int1 stridey, T_int1 stridez, T_float offsetx, T_float offsety, T_float offsetz, T_int1* binid, T_int1* sortidx){
 
     for(uint32_t tid = blockIdx.x * blockDim.x + threadIdx.x; tid < n_particle; tid+=gridDim.x*blockDim.x){
         // read particle data from global memory
@@ -97,11 +97,14 @@ cal_binid(T_int1 bin_size_x, T_int1 bin_size_y, T_int1 bin_size_z, T_int1 nbinx,
 
         // strides
         T_int1 g_stride[3] = {stridex, stridey, stridez};
+        //double g_offset[3] = {offsetx, offsety, offsetz};
+        T_float g_offset[3] = {offsetx, offsety, offsetz};
 
         // cell index for each dimension
         T_int1  c_index[DIM];
         for(int idim=0; idim<3; idim++){
-            c_index[idim] = (static_cast<int>(std::floor(p_disp[idim]/cell_size)+p_pmid[idim])%g_stride[idim]+g_stride[idim]) % g_stride[idim];
+            c_index[idim] = (static_cast<int>(std::floor((p_disp[idim]+g_offset[idim])/cell_size)+p_pmid[idim])%g_stride[idim]+g_stride[idim]) % g_stride[idim];
+            //c_index[idim] = (static_cast<int>(std::floor(p_disp[idim]/cell_size)+p_pmid[idim])%g_stride[idim]+g_stride[idim]) % g_stride[idim];
         }
         c_index[0] = c_index[0]/bin_size_x;
[…]

### 2023-02-28 pmwd/train_util.py
Add mesh_shape input in gen_ic

@@ -62,21 +62,20 @@ def scale_Sobol(fn='sobol.txt', ind=slice(None)):
     return sobol.T
 
 
-def gen_ic(i, fn_sobol='sobol.txt', re_sobol=False):
-    """Generate the initial condition for nbody.
+def gen_ic(i, fn_sobol='sobol.txt', re_sobol=False, mesh_shape=1):
+    """Generate the initial condition with lpt for nbody.
     The seed for white noise is simply the Sobol index i.
     """
     sobol = scale_Sobol(fn_sobol, i)  # scaled Sobol parameters at i
 
-    # initialize cosmo and conf based on the Sobol parameters
-    # Fields related to mesh shape and number of time steps in conf
-    # need to be further sampled and replaced for pmwd during training.
     conf = Configuration(
         ptcl_spacing = sobol[0] / 128,
         ptcl_grid_shape = (128,) * 3,
         a_start = 1 / 16,
         float_dtype = jnp.float64,
         growth_dt0 = 1,
+        mesh_shape = mesh_shape,
+        # TODO number of time steps
     )
 
     cosmo = Cosmology(

### 2023-02-28 docs/papers/sto/sto.tex
FIXUP draft: shrink Ob/Om and Ok/(1-Ok) ranges
FIXUP generate IC for g4 on GPU
update TODO

@@
     linear regime), say 1 Gpc/$h$.
+    \textbf{Update: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
+    consistent at the field level. To further check power spectrum.}
 \item Be careful with the non-deterministic ICs generated on GPUs.
+      \textbf{Update: we have shown that ICs generated on GPUs are deterministic
+      (but different from those on CPUs), generating ICs for both training data
+      and training on GPUs might be the simplest solution.}
   \begin{enumerate}
@@
   \cmidrule(lr){2-4}
-  & $\Omegab / \Omegam$ & log-uniform & [1/16, 1/2) \\
+  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
   \cmidrule(lr){2-4}
-  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-0.5, 0.5) \\
+  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
   \cmidrule(lr){2-4}

### 2023-02-28 pmwd/train_util.py
Set a_stop=1+1/128 in conf

@@ -72,6 +72,7 @@ def gen_ic(i, fn_sobol='sobol.txt', re_sobol=False, mesh_shape=1):
         ptcl_spacing = sobol[0] / 128,
         ptcl_grid_shape = (128,) * 3,
         a_start = 1 / 16,
+        a_stop = 1 + 1 / 128,
         float_dtype = jnp.float64,
         growth_dt0 = 1,
         mesh_shape = mesh_shape,

### 2023-03-05 pmwd/train_util.py
Add optional a inputs in gen_ic

@@ -62,7 +62,8 @@ def scale_Sobol(fn='sobol.txt', ind=slice(None)):
     return sobol.T
 
 
-def gen_ic(i, fn_sobol='sobol.txt', re_sobol=False, mesh_shape=1):
+def gen_ic(i, fn_sobol='sobol.txt', re_sobol=False,
+           a_start=1/16, a_stop=1+1/128, a_nbody_maxstep=1/64, mesh_shape=1):
     """Generate the initial condition with lpt for nbody.
     The seed for white noise is simply the Sobol index i.
     """
@@ -71,12 +72,13 @@ def gen_ic(i, fn_sobol='sobol.txt', re_sobol=False, mesh_shape=1):
     conf = Configuration(
         ptcl_spacing = sobol[0] / 128,
         ptcl_grid_shape = (128,) * 3,
-        a_start = 1 / 16,
-        a_stop = 1 + 1 / 128,
+        a_start = a_start,
+        a_stop = a_stop,
         float_dtype = jnp.float64,
         growth_dt0 = 1,
         mesh_shape = mesh_shape,
         # TODO number of time steps
+        a_nbody_maxstep=a_nbody_maxstep,
     )
 
     cosmo = Cosmology(

### 2023-03-05 docs/papers/sto/sto.tex
FIXUP draft: add xxx acknowledgement
FIXUP draft: typo

@@
 {Neural and Symbolic Spatiotemporal Optimization of Cosmological
-Particel-Mesh Simulation}
+Particle-Mesh Simulation}
 
@@
 
+YZ and YL were supported by The Major Key Project of PCL.
 The Flatiron Institute is supported by the Simons Foundation.

### 2023-03-09 docs/papers/adjoint/adjoint.tex

@@
 \usepackage[T1]{fontenc}
+\usepackage{fontawesome}
 \usepackage{amsmath}
-  %\allowdisplaybreaks
+%  \allowdisplaybreaks
 \usepackage{bm}
@@
 \usepackage[caption=false]{subfig}
+\usepackage[figure,figure*]{hypcap}
 \usepackage{tikz}
@@
 
-\title{\large Differentiable Cosmological Simulation with Adjoint Method
+\title{\large Differentiable Cosmological Simulation with the Adjoint Method
 \vspace{0.3em}}
@@
 Laboratory, Shenzhen, Guangdong 518066, China}
-\affiliation{Department of Physics, New York University, New York, New
-York 10003, USA}
+\affiliation{Center for Cosmology and Particle Physics, Department of
+Physics, New York University, New York, New York 10003, USA}
 
@@
 
-The first differentiable cosmological simulations, such as BORG and
-ELUCID \citep{BORG, ELUCID}, were developed before the advent of modern
-AD systems, and were based on implementations of analytic derivatives
-\citep[see e.g.,][App.~D]{SeljakEtAl2017}).
+The first differentiable cosmological simulations, such as BORG, ELUCID,
+and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
+advent of modern AD systems, and were based on implementations of
+analytic derivatives \citep[see e.g.,][App.~D]{SeljakEtAl2017}).
 Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
@@
 The initial conditions of particles can be set perturbatively when
-$|\vnabla \cdot \vs|$, linear approximation of the density fluctuation,
-is much smaller than 1.
+$|\vnabla \cdot \vs|$, the linear approximation of the density
+fluctuation, is much smaller than 1.
 We compute the initial displacements and momenta using the second order
@@
 \vp &= a^2 \dot\vx = a^2 \dot\vs
-  = a^3 H \bigl( D_1' \vs^{(1)} + D_2' \vs^{(2)} \bigr),
+  = a^2 H \bigl( D_1' \vs^{(1)} + D_2' \vs^{(2)} \bigr),
 \label{ic}
@@
 The gravitational potential sourced by matter density fluctuation
-follows the Poisson equation
+satisfies the Poisson equation
 %
@@
 %
-And in Fourier space, $- \vnabla \varphi(\vx)$ is just $- i \vk
-\varphi(\vk)$, each component of which can be transformed back and
-combined into the force field.
+In Fourier space, $- \vnabla \varphi(\vx)$ is just $- i \vk
+\varphi(\vk)$, each component of which can be transformed back to obtain
+the force field.
 With 4 (1 forward and 3 inverse) FFTs, we can obtain $- \vnabla
@@
 Combined with the $\vtheta$-dependent propagation of the light front, we
-can solve for the intersections\footnote{Even analytically if the light
-propagation is also approximated cubically.}, at which we can record the
-observed\footnote{Note that we focus on the dark matter phase space
-here, and leave more generic observable modeling to a future work.}
+can solve for the intersections, at which we can record the observed
 $\hat\vz$.
+The solution can even be analytic if the light propagation is also
+approximated cubically.
+Note that we only ``observe'' the dark matter phase space here, and
+leave more realistic observables to a future work, including the forward
+modeling of real observational effects.
 \autoref{fig:step} illustrates the observation operator $O$ and its
@@
 pattern, which improves with more iterations.
+Also see the animated reverse time evolution
+\href{https://youtu.be/Epsgh6vr0qs}{\faYoutubePlay} and initial
+condition optimization
+\href{https://youtu.be/vD6lbjHP3SY}{\faYoutubePlay} on YouTube.
 }
@@
 implementations of scatter and gather operations.
-In \autoref{fig:scaling}, we present weak scaling test of \pmwd, and
-find both the LPT and the $N$-body parts scale well except for very
-small problems.
+In \autoref{fig:perf}, we present performance test of \pmwd, and find
+both the LPT and the $N$-body parts scale well except for very small
+problems.
 The growth function solution has a constant cost, and generally does not
@@
 \centering
-\includegraphics[width=0.7\linewidth]{scaling.pdf}
-\caption{Weak scaling of \pmwd. Both the 2LPT and $N$-body components
+\includegraphics[width=0.7\linewidth]{perf.pdf}
+\caption{Performance of \pmwd. Both the 2LPT and $N$-body components
 scale well from $512^3$ particles to $64^3$ particles, below which they
@@
 }
-\label{fig:scaling}
+\label{fig:perf}
 \end{figure*}
@@
 Ni for interesting discussions.
+YL and YZ were supported by The Major Key Project of PCL.
 The Flatiron Institute is supported by the Simons Foundation.
@@
 \textit{\large Code Availability:}
-\pmwd\ is open-source on
-\href{https://github.com/eelregit/pmwd}{GitHub}, including the
-\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint}{source
-and scripts of this paper}.
-Also see the companion code paper on
-\href{https://arxiv.org/abs/2211.THIS+1}{arXiv} and
-\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/pmwd}{GitHub}.
+\pmwd\ is open-source on GitHub
+\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
+files and scripts of this paper
+\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint}{\faFile}.
+Also see the companion code paper \citep{pmwd}
+\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/pmwd}{\faFile}.
 
@@
 \end{equation}
-as
+for $m \in \{1, 2\}$, as
 \begin{align}
@@
 \end{align}
-We solve the growth equations in $G_n$ instead of $D_n$, with the
+We solve the growth equations in $G_m$ instead of $D_m$, with the
 \texttt{JAX} adaptive ODE integrator implementing the adjoint
 method\footnote{The growth equation adjoint method aims to solve the
-ODEs in their continuous limit, this is different from our
-discretize-then-optimize $N$-body adjoint method.}.
+ODEs in their continuous limit, i.e., with the optimize-then-discretize
+approach, this is different from our discretize-then-optimize $N$-body
+adjoint method.}.
 This is because the former can be integrated backward in time more
@@
 %
-evolves towards identity $\vLambda_1 = \vI$ by
+evolves backward in time from identity $\vLambda_1 = \vI$ by
 %
@@
 z_j$ affecting the objective $\cJ(\hat\vz, \vtheta)$ at different times
-$\hat t_j$, i.e., $\hat z_i \triangleq z_i(\hat t_i)$.
+$\hat t_j$, i.e., $\hat z_j \triangleq z_j(\hat t_j)$.
 The Lagrangian becomes
@@
 &= \frac{\p\cJ}{\p\vtheta}
-+ \sum_j \Bigl( \frac{\p\cJ}{\p\hat\vz} - \hat\vlambda \Bigr)
++ \Bigl( \frac{\p\cJ}{\p\hat\vz} - \hat\vlambda \Bigr)
   \cdot \frac{\p\hat\vz}{\p\vtheta}
@@
 $\hat\lambda_j \triangleq \lambda_j(\hat t_j)$.
+In the first equality, we have also dropped a vanishing term $\sum_j
+\hat\lambda_j \bigl[ \dot z_j - f_j(\hat t_j) \bigr] \p\hat t_j /
+\p\vtheta$, i.e., $\p\hat t_j / \p\vtheta$ does not directly enter the
+gradient.
 

### 2023-03-10 tests/scatter_consistency.py
test scatter two branches consistency

@@ -1,0 +1,53 @@
+import numpy as np
+import jax.numpy as jnp
+from jax import device_put
+import pmwd
+import time
+from pmwd import (
+    Configuration,
+    Cosmology, SimpleLCDM,
+    boltzmann,
+    white_noise, linear_modes,
+    lpt,
+    nbody,
+    scatter,
+    Particles,
+)
+from pmwd.scatter import _scatter
+from jax import random
+import os
+os.environ['XLA_PYTHON_CLIENT_MEM_FRACTION'] = '.99'
+key = random.PRNGKey(101)
+ngrid = 512
+print("loading")
+ptcl_spacing = 1.7  # Lagrangian space Cartesian particle grid spacing, in Mpc/h by default
+ptcl_grid_shape = (ngrid,) * 3
+conf = Configuration(ptcl_spacing, ptcl_grid_shape, mesh_shape=1)  # 1x mesh shape
+ptcl = Particles.gen_grid(conf)
+pmid = ptcl.pmid.astype(jnp.uint32)
+#pmid = ptcl.pmid
+mesh = random.uniform(key, shape=(ngrid,ngrid,ngrid),dtype=jnp.float32,minval=0.0,maxval=2.0)
+val = random.uniform(key, shape=(ngrid*ngrid*ngrid,),dtype=jnp.float32,minval=0.0,maxval=2.0)
+disp = random.uniform(key, shape=(ngrid*ngrid*ngrid,3),dtype=jnp.float32,minval=0.0,maxval=2.0)
+cell_size = ptcl_spacing
+offset = tuple((0.,0.,0.))
+
+print("jax simple")
+mesh0 = mesh*0;
+mesh0 = _scatter(pmid, disp, conf, mesh0, val, offset, None).block_until_ready()
+start = time.time()
+for ii in range(10):
+    mesh0 = _scatter(pmid, disp, conf, mesh0, val, offset, None).block_until_ready()
+print(time.time() - start)
+
+print("jax complex")
+mesh1 = mesh*0
+mesh1 = _scatter(pmid, disp, conf, mesh1, val, offset, cell_size).block_until_ready()
+start = time.time()
+for ii in range(10):
+    mesh1 = _scatter(pmid, disp, conf, mesh1, val, offset, cell_size).block_until_ready()
+print(time.time() - start)
+
+print("mesh1 max:",mesh1.max())
+print("mesh1 min:",mesh1.min())
+print("diff abs max:", jnp.abs(mesh0-mesh1).max())

### 2023-03-10 tests/scatter_consistency.py
update consistency test

@@ -48,6 +48,8 @@ for ii in range(10):
     mesh1 = _scatter(pmid, disp, conf, mesh1, val, offset, cell_size).block_until_ready()
 print(time.time() - start)
 
+print("mesh0 max:",mesh0.max())
+print("mesh0 min:",mesh0.min())
 print("mesh1 max:",mesh1.max())
 print("mesh1 min:",mesh1.min())
 print("diff abs max:", jnp.abs(mesh0-mesh1).max())

### 2023-03-10 tests/scatter_consistency.py
update test

@@ -24,8 +24,7 @@ ptcl_spacing = 1.7  # Lagrangian space Cartesian particle grid spacing, in Mpc/h
 ptcl_grid_shape = (ngrid,) * 3
 conf = Configuration(ptcl_spacing, ptcl_grid_shape, mesh_shape=1)  # 1x mesh shape
 ptcl = Particles.gen_grid(conf)
-pmid = ptcl.pmid.astype(jnp.uint32)
-#pmid = ptcl.pmid
+pmid = ptcl.pmid
 mesh = random.uniform(key, shape=(ngrid,ngrid,ngrid),dtype=jnp.float32,minval=0.0,maxval=2.0)
 val = random.uniform(key, shape=(ngrid*ngrid*ngrid,),dtype=jnp.float32,minval=0.0,maxval=2.0)
 disp = random.uniform(key, shape=(ngrid*ngrid*ngrid,3),dtype=jnp.float32,minval=0.0,maxval=2.0)
@@ -53,3 +52,4 @@ print("mesh0 min:",mesh0.min())
 print("mesh1 max:",mesh1.max())
 print("mesh1 min:",mesh1.min())
 print("diff abs max:", jnp.abs(mesh0-mesh1).max())
+print("diff std:", jnp.std(mesh0-mesh1))

### 2023-03-11 docs/papers/sto/sto.tex
FIXUP draft: edit TODOs

@@
 \begin{itemize}
+\item Moving some of the following ongoing and completed tasks to the
+    main text and add preliminary figures.
+\item Implement the rest of the nonlinear scales, including the rms
+    displacement with np.trapz of $P$ over $\ln k$, the $\ln a$ time
+    derivatives, etc.
 \item Compare one pair of Gadget4 and \pmwd\ (with say 100 to 1000 time
@@
       and training on GPUs might be the simplest solution.}
-  \begin{enumerate}
-  \item Saving them on disks and loading them during training solves
-    this but may introduce IO or memory problem;
-  \item Generating them on CPUs for training data and also training
-    might suffice while requireing such data loader implementation;
-  \item Generating them on GPUs for training data and training will
-    introduce additional source of initial deviations to be nonlinearly
-    amplified. There are flags that makes certain operations
-    deterministic but slow on GPUs, but those are enviroment variables
-    that function globally and cannot be applied locally only on the IC
-    generation during training.
-  \item Generating ICs and then train on GPUs with double precision and
-    then compute the loss at single precision (Gadget4 should be run at
-    double precision) should work, and likely be the simplest solution.
-    But for this to work the final training loss should be much larger
-    than the level of reproducibility.
-  \end{enumerate}
 \item Non-integer multiple mesh shapes: for uniform particle grid, test

### 2023-03-13 pmwdlib/kernels.cc.cu
test round(x) and floor(x+0.5)

@@ -295,7 +295,8 @@ scatter_kernel_sm(T_int1* pmid, T_float* disp, T_float cell_size, T_float ptcl_s
                 t_disp[idim] -= floor(t_disp[idim]/L1[idim])*L1[idim];
                 c_index[idim] = static_cast<T_int1>(std::floor(t_disp[idim]/cell_size));
                 t_disp[idim]  = g_disp[idim] - c_index[idim]*cell_size;
-                t_disp[idim] -= round(t_disp[idim]/L1)*L1;
+                //t_disp[idim] -= round(t_disp[idim]/L1)*L1;
+                t_disp[idim] -= floor(t_disp[idim]/L1+0.5)*L1;
                 t_disp[idim] /= cell_size;
             }
 

### 2023-03-13 pmwdlib/kernels.cc.cu
minor comments

@@ -96,7 +96,7 @@ cal_binid(T_int1 bin_size_x, T_int1 bin_size_y, T_int1 bin_size_z, T_int1 nbinx,
         T_float p_disp[DIM] = {disp[tid*DIM + 0], disp[tid*DIM + 1], disp[tid*DIM + 2]};
 
         // strides
-        T_int1 p_stride[3] = {ptcl_gridx, ptcl_gridy, ptcl_gridz};
+        //T_int1 p_stride[3] = {ptcl_gridx, ptcl_gridy, ptcl_gridz};
         T_int1 g_stride[3] = {stridex, stridey, stridez};
         T_float g_offset[3] = {offsetx, offsety, offsetz};
 
@@ -477,6 +477,7 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
     cudaEventElapsedTime(&milliseconds, start, stop);
     printf("cuda kernel SortPairs cellids: %f milliseconds\n", milliseconds);
 
+    // TODO: change to sparse histogram, upper_bound is slow
     cudaEventRecord(start);
     thrust::counting_iterator<uint32_t> search_begin(0);
     thrust::upper_bound(thrust::device_ptr<uint32_t>(d_index), thrust::device_ptr<uint32_t>(d_index)+uint32_t(n_particle),

### 2023-03-13 docs/papers/sto/sto.tex
FIXUP draft: edit TODOs
FIXUP draft: edit TODO

@@
     scales. For this purpose the box size should be large enough (in
-    linear regime), say 1 Gpc/$h$.
-    \textbf{Update: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
-    consistent at the field level. To further check power spectrum.}
-\item Be careful with the non-deterministic ICs generated on GPUs.
-      \textbf{Update: we have shown that ICs generated on GPUs are deterministic
-      (but different from those on CPUs), generating ICs for both training data
-      and training on GPUs might be the simplest solution.}
+    linear regime), say 1 Gpc/$h$.\newline
+    Update: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
+    consistent at the field level. To further check power spectrum. For odd mesh
+    shapes (e.g. 1, 3, 5), pmwd and Gadget are perfectly consistent on large
+    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
+    slightly different. To figure out this weird issue.}
+\item Be careful with the non-deterministic ICs generated on GPUs.\newline
+      Update: we have shown that ICs generated on GPUs are deterministic (but
+      different from those on CPUs), generating ICs for both training data and
+      training on GPUs might be the simplest solution.
+\item Incorporate 3lpt branch; testing; find a\_start for different mass
+  resolutions
+\item For TO, probably necessary to first incorporate Oliver Hahn's
+  recent improvement on FastPM integrator arXiv:2301:09655.
+\item Compare various field and halo summary stats to compare sto with
+  raw pmwd, deconvolved pmwd, and gadget test cases.
 \item Non-integer multiple mesh shapes: for uniform particle grid, test
@@
   deconvolution, with SO.
-\item For TO, probably necessary to first incorporate Oliver Hahn's
-  recent improvement on FastPM integrator arXiv:2301:09655.
 \end{itemize}

### 2023-03-13 pmwdlib/kernels.cc.cu
fix typo

@@ -295,8 +295,8 @@ scatter_kernel_sm(T_int1* pmid, T_float* disp, T_float cell_size, T_float ptcl_s
                 t_disp[idim] -= floor(t_disp[idim]/L1[idim])*L1[idim];
                 c_index[idim] = static_cast<T_int1>(std::floor(t_disp[idim]/cell_size));
                 t_disp[idim]  = g_disp[idim] - c_index[idim]*cell_size;
-                //t_disp[idim] -= round(t_disp[idim]/L1)*L1;
-                t_disp[idim] -= floor(t_disp[idim]/L1+0.5)*L1;
+                //t_disp[idim] -= round(t_disp[idim]/L1[idim])*L1[idim];
+                t_disp[idim] -= floor(t_disp[idim]/L1[idim]+0.5)*L1[idim];
                 t_disp[idim] /= cell_size;
             }
 

### 2023-03-13 pmwd/configuration.py
Add nonlinear scales based on Plin, TopHat, Gauss

@@ -7,7 +7,7 @@ from jax import ensure_compile_time_eval
 from jax.typing import DTypeLike
 import jax.numpy as jnp
 from jax.tree_util import tree_map
-from mcfit import TophatVar
+from mcfit import TophatVar, GaussVar
 
 from pmwd.tree_util import pytree_dataclass
 
@@ -179,6 +179,11 @@ class Configuration:
                 self,
                 'var_tophat',
                 TophatVar(self.transfer_k[1:], lowring=True, backend='jax'),
+            )
+            object.__setattr__(
+                self,
+                'var_gauss',
+                GaussVar(self.transfer_k[1:], lowring=True, backend='jax'),
             )
 
         # ~ 1.5e-8 for float64, 3.5e-4 for float32

### 2023-03-14 pmwd/so_util.py
Add nonlinear scale based on rms linear disp

@@ -63,25 +63,31 @@ def init_mlp_params(n_input, nodes, zero_params=None):
 def nonlinear_scales(cosmo, conf, a):
     k = conf.transfer_k[1:]
     D = growth(a, cosmo, conf)
-    Dfac = -2 * D**(-3) * growth(a, cosmo, conf, deriv=1)  # d(1/D^2) / dlna
+    dD = growth(a, cosmo, conf, deriv=1)
+    dD2i = -2 * D**(-3) * dD  # d(1/D^2) / dlna
     interp_valgrad = jax.value_and_grad(jnp.interp, argnums=0)
 
     # dimensionless linear power
     Plin = linear_power(k, None, cosmo, conf)  # no a dependence
     k_P, dk_P = interp_valgrad(1 / D**2, k**3 * Plin / (2 * jnp.pi**2), k)
-    dk_P *= Dfac
+    dk_P *= dD2i
 
     # TopHat variance, var is decreasing with R
     # but for jnp.interp, xp must be increasing, thus the reverse [::-1]
     R_TH, dR_TH = interp_valgrad(1 / D**2, cosmo.varlin[::-1], conf.varlin_R[::-1])
-    dR_TH *= Dfac
+    dR_TH *= dD2i
 
     # Gaussian variance
     R, varlin_G = conf.var_gauss(Plin, extrap=True)
     R_G, dR_G = interp_valgrad(1 / D**2, varlin_G[::-1], R[::-1])
-    dR_G *= Dfac
+    dR_G *= dD2i
 
-    return (1/k_P, R_TH, R_G, -dk_P/k_P**2, dR_TH, dR_G)
+    # rms linear theory displacement
+    Rd = jnp.sqrt(jnp.trapz(k * Plin, x=jnp.log(k)) / (2 * jnp.pi**2))
+    dRd = Rd * dD
+    Rd *= D
+
+    return (1/k_P, R_TH, R_G, Rd, -dk_P/k_P**2, dR_TH, dR_G, dRd)
 
 
 # TODO add more relevant factors

### 2023-03-14 docs/papers/sto/sto.tex
FIXUP draft: update TODO

@@
     displacement with np.trapz of $P$ over $\ln k$, the $\ln a$ time
-    derivatives, etc.
+    derivatives, etc.\newline
+    \textbf{Update}: code has been added.
 \item Compare one pair of Gadget4 and \pmwd\ (with say 100 to 1000 time
@@
     linear regime), say 1 Gpc/$h$.\newline
-    Update: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
+    \textbf{Update}: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
     consistent at the field level. To further check power spectrum. For odd mesh
@@
 \item Be careful with the non-deterministic ICs generated on GPUs.\newline
-      Update: we have shown that ICs generated on GPUs are deterministic (but
-      different from those on CPUs), generating ICs for both training data and
-      training on GPUs might be the simplest solution.
+      \textbf{Update}: we have shown that ICs generated on GPUs are
+      deterministic (but different from those on CPUs), generating ICs for both
+      training data and training on GPUs might be the simplest solution.
 \item Incorporate 3lpt branch; testing; find a\_start for different mass

### 2023-03-14 pmwd/so_util.py
FIXUP draft: update TODO

@@ -94,9 +94,10 @@ def nonlinear_scales(cosmo, conf, a):
 @jax.jit
 def sotheta(cosmo, conf, a):
     theta_l = jnp.asarray([  # quantities of dim L
+        *nonlinear_scales(cosmo, conf, a),
         conf.ptcl_spacing,
         conf.cell_size,
-        *nonlinear_scales(cosmo, conf, a),
+        # softening length?
     ])
     theta_o = jnp.asarray([  # dimensionless quantities
         a,
@@ -104,6 +105,7 @@ def sotheta(cosmo, conf, a):
         growth(a, cosmo, conf),
         growth(a, cosmo, conf, deriv=1),
         H_deriv(a, cosmo),
+        # time step size?
     ])
     return (theta_l, theta_o)
 

### 2023-03-14 pmwd/boltzmann.py
FIXUP move Gauss var evaluation to varlin_integ

@@ -287,8 +287,9 @@ def varlin_integ(cosmo, conf):
     Plin = linear_power(conf.var_tophat.x, None, cosmo, conf)
 
     _, varlin = conf.var_tophat(Plin, extrap=True)
+    _, varlin_g = conf.var_gauss(Plin, extrap=True)
 
-    return cosmo.replace(varlin=varlin)
+    return cosmo.replace(varlin=varlin, varlin_g=varlin_g)
 
 
 def varlin(R, a, cosmo, conf):

### 2023-03-14 pmwd/cosmology.py

@@ -73,6 +73,8 @@ class Cosmology:
 
     varlin: Optional[Array] = field(default=None, compare=False)
 
+    varlin_g: Optional[jnp.ndarray] = field(default=None, compare=False)
+
     # list of parameters of SO neural nets, order: f, g, h
     so_params: Optional[list[FrozenDict]] = None
 

### 2023-03-14 pmwd/so_util.py

@@ -78,8 +78,7 @@ def nonlinear_scales(cosmo, conf, a):
     dR_TH *= dD2i
 
     # Gaussian variance
-    R, varlin_G = conf.var_gauss(Plin, extrap=True)
-    R_G, dR_G = interp_valgrad(1 / D**2, varlin_G[::-1], R[::-1])
+    R_G, dR_G = interp_valgrad(1 / D**2, cosmo.varlin_g[::-1], conf.varlin_R_g[::-1])
     dR_G *= dD2i
 
     # rms linear theory displacement

### 2023-03-14 pmwd/configuration.py
FIXUP move Gauss var evaluation to varlin_integ

@@ -337,3 +337,8 @@ class Configuration:
     def varlin_R(self):
         """Linear matter overdensity variance in a top-hat window of radius R in [L], of ``cosmo_dtype``."""
         return self.var_tophat.y
+
+    @property
+    def varlin_R_g(self):
+        """Linear matter overdensity variance in a Gauss window of radius R in [L], of ``cosmo_dtype``."""
+        return self.var_gauss.y

### 2023-03-21 docs/examples/sto/g4gen/sobol.py
FIXUP sobol sequence: add 9 extra for testing

@@ -6,7 +6,7 @@ import matplotlib.pyplot as plt
 from pmwd.train_util import scale_Sobol
 
 
-def gen_sobol(filename, d=9, m=9):
+def gen_sobol(filename, d=9, m=9, extra=9):
     from scipy.stats.qmc import Sobol, discrepancy
 
     disc_min = np.inf
@@ -22,7 +22,7 @@ def gen_sobol(filename, d=9, m=9):
     # seed_min = 55868, mixture discrepancy = 0.016109347957680598
 
     sampler = Sobol(d, scramble=True, seed=seed_min)
-    sample = sampler.random_base2(m)
+    sample = sampler.random(n=2**m + extra)
     np.savetxt(filename, sample)
 
 

### 2023-03-21 docs/papers/sto/sto.tex
FIXUP draft: edit dimensionless NN input factors
FIXUP draft: edit TODOs

@@
   resolutions
-\item For TO, probably necessary to first incorporate Oliver Hahn's
-  recent improvement on FastPM integrator arXiv:2301:09655.
-\item Compare various field and halo summary stats to compare sto with
-  raw pmwd, deconvolved pmwd, and gadget test cases.
+\item Probably necessary to first incorporate List \& Hahn's recent
+  improvement on FastPM integrator arXiv:2301:09655.
+\item Compare sto DM and halo fields and summary stats with those of raw
+  pmwd, deconvolved pmwd, and gadget, on the last 9 test cases.
+\item Demonstrate capability of generalization, on some cases with
+  extrapolating cosmological parameters, etc.
 \item Non-integer multiple mesh shapes: for uniform particle grid, test
@@
 \begin{itemize}
-\item $D(a)$: linear growth factor;
-\item $f(a)$: linear growth rate $f \triangleq \d\ln D / \d\ln a$;
+\item $G_m(a) \triangleq D_m(a) / a^m$: linear growth given in
+  suppression factors for $m \in \{1, 2\}$;
+\item $\d\ln G_m / \d\ln a$: with $m=1$ case related to the growth rate
+  $f \triangleq \d\ln D_1 / \d\ln a$;
 \item $\Omegam(a)$: matter density parameter at $a$;
 \item $\d\ln H / \d\ln a$: time derivative of Hubble expansion;
-\item $a$: scale factor;
-\item $\Delta a$: time step size in $a$.
+\item $\Delta\ln a$: time step size in $\ln a$.
 \end{itemize}

### 2023-03-25 pmwdlib/kernels.cc.cu
fix memory bug

@@ -540,7 +540,7 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
     T *particle_values = reinterpret_cast<T *>(buffers[2]);
     void *work_d = buffers[4];
     T *grid_values = reinterpret_cast<T *>(buffers[5]);
-    uint32_t *work_i_d = static_cast<uint32_t *>(work_d);
+    char *work_i_d = static_cast<char *>(work_d);
 
     // parameters for shared mem using bins to group cells
     uint32_t bin_size = BINSIZE;
@@ -550,12 +550,12 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
 
     uint32_t npts_mem_size = sizeof(uint32_t) * n_particle;
     uint32_t nbins_mem_size = sizeof(uint32_t) * nbinx*nbiny*nbinz;
-    uint32_t* d_sortidx = work_i_d;
-    uint32_t* d_sortidx_buff = &work_i_d[npts_mem_size];
-    uint32_t* d_index = &work_i_d[2*npts_mem_size];
-    uint32_t* d_index_buff = &work_i_d[3*npts_mem_size];
-    uint32_t* d_bin_count = &work_i_d[4*npts_mem_size];
-    uint32_t* d_bin_start = &work_i_d[4*npts_mem_size + nbins_mem_size];
+    uint32_t* d_sortidx = (uint32_t*)work_i_d;
+    uint32_t* d_sortidx_buff = (uint32_t*)&work_i_d[npts_mem_size];
+    uint32_t* d_index = (uint32_t*)&work_i_d[2*npts_mem_size];
+    uint32_t* d_index_buff = (uint32_t*)&work_i_d[3*npts_mem_size];
+    uint32_t* d_bin_count = (uint32_t*)&work_i_d[4*npts_mem_size];
+    uint32_t* d_bin_start = (uint32_t*)&work_i_d[4*npts_mem_size + nbins_mem_size];
     void     *d_temp_storage = (void*)&work_i_d[4*npts_mem_size + 2*nbins_mem_size + sizeof(uint32_t)];
     int block_size = 1024;
     int grid_size = ((n_particle + block_size) / block_size);

### 2023-03-25 pmwdlib/jax_pmwd_common.h
adding sort ops cuda side

@@ -17,6 +17,11 @@ struct PmwdDescriptor {
   size_t tmp_storage_size;
 };
 
+struct SortKeysDescriptor {
+    int64_t n_keys;
+    size_t tmp_storage_size;
+};
+
 }
 
 #endif

### 2023-03-25 pmwdlib/kernels.cc.cu

@@ -519,6 +519,41 @@ gather_kernel_sm(T_int1* pmid, T_float* disp, T_float cell_size, T_int1* stride,
 }
 
 template <typename T>
+void sort_keys_kernel(cudaStream_t stream, void** buffers, const char* opaque, std::size_t opaque_len){
+#ifdef SCATTER_TIME
+    cudaEvent_t start, stop;
+    cudaEventCreate(&start);
+    cudaEventCreate(&stop);
+#endif
+
+    // inputs/outputs
+    const SortKeysDescriptor<T> *descriptor = unpack_descriptor<SortKeysDescriptor>(opaque, opaque_len);
+    int64_t n_keys = descriptor->n_keys;
+    size_t temp_storage_bytes = descriptor->tmp_storage_size;
+    T *d_keys = reinterpret_cast<T *>(buffers[0]);
+    void *work_d = buffers[1];
+    char *work_i_d = static_cast<char *>(work_d);
+
+    uint32_t keys_mem_size = sizeof(T) * n_keys;
+    T* d_keys_buff = (T*)work_i_d;
+    void *d_temp_storage = (void*)&work_i_d[keys_mem_size];
+
+#ifdef SCATTER_TIME
+    cudaEventRecord(start);
+#endif
+    cub::DoubleBuffer<T> d_keys_dbuff(d_keys, d_keys_buff);
+    cub::DeviceRadixSort::SortKeys(d_temp_storage, temp_storage_bytes, d_keys_dbuff, n_keys);
+    d_keys = d_keys_dbuff.Current();
+#ifdef SCATTER_TIME
+    cudaEventRecord(stop);
+    cudaEventSynchronize(stop);
+    milliseconds = 0;
+    cudaEventElapsedTime(&milliseconds, start, stop);
+    printf("cuda kernel Sort: %f milliseconds\n", milliseconds);
+#endif
+}
+
+template <typename T>
 void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::size_t opaque_len){
 #ifdef SCATTER_TIME
     cudaEvent_t start, stop;

### 2023-03-25 pmwdlib/kernels.h

@@ -16,6 +16,11 @@ void scatterf(cudaStream_t stream, void** buffers, const char* opaque, std::size
 void gatherf(cudaStream_t stream, void** buffers, const char* opaque, std::size_t opaque_len);
 
 int64_t get_workspace_size(int64_t n_ptcls, uint32_t stride_x, uint32_t stride_y, uint32_t stride_z, size_t& temp_storage_bytes);
+
+int64_t get_sort_keys_workspace_size_f32(int64_t n_keys, size_t& temp_storage_bytes);
+int64_t get_sort_keys_workspace_size_f64(int64_t n_keys, size_t& temp_storage_bytes);
+int64_t get_sort_keys_workspace_size_i32(int64_t n_keys, size_t& temp_storage_bytes);
+int64_t get_sort_keys_workspace_size_i64(int64_t n_keys, size_t& temp_storage_bytes);
 }  // namespace jax_pmwd
 
 #endif

### 2023-03-25 pmwdlib/pybind11_kernel_helpers.h

@@ -31,6 +31,10 @@ pybind11::bytes build_descriptor(int64_t n_particle, T ptcl_spacing, T cell_size
       PmwdDescriptor<T>{n_particle, ptcl_spacing, cell_size, {offset_1, offset_2, offset_3}, {ptcl_grid_1, ptcl_grid_2, ptcl_grid_3}, {stride_1, stride_2, stride_3}, tmp_storage_size});
 }
 
+pybind11::bytes build_sort_keys_descriptor(int64_t n_keys, size_t tmp_storage_size) {
+  return pack_descriptor(SortKeysDescriptor{n_keys, tmp_storage_size});
+}
+
 }  // namespace jax_pmwd
 
 #endif

### 2023-04-06 docs/examples/sto/train/test_jax_dist.py
Add the job script and a test of jax distributed

@@ -1,0 +1,19 @@
+import os
+import jax
+from pprint import pprint
+
+procid = int(os.getenv('SLURM_PROCID'))
+n_local_gpus = int(os.getenv('SLURM_GPUS_PER_TASK'))
+print(f'\n ## process id: {procid}')
+
+jax.distributed.initialize(local_device_ids=[i for i in range(n_local_gpus)])
+
+print(f'local device count: {jax.local_device_count()}')
+pprint(jax.local_devices())
+
+if procid == 0:
+    print(f'global device count: {jax.device_count()}')
+    pprint(jax.devices())
+
+xs = jax.numpy.ones(jax.local_device_count())
+print(jax.pmap(lambda x: jax.lax.psum(x, 'i'), axis_name='i')(xs))

### 2023-04-06 docs/examples/sto/train/job.sh
Add the job script and a test of jax distributed

@@ -1,0 +1,14 @@
+#!/usr/bin/bash
+
+#SBATCH --partition=gpu
+#SBATCH --nodes=2
+#SBATCH --ntasks-per-node=1
+#SBATCH --gpus-per-task=4
+#SBATCH --constraint=a100-80gb,ib
+#SBATCH --exclusive
+#SBATCH --mem=0
+
+module purge
+
+source ~/mambaforge/bin/activate
+srun python test_jax_dist.py

### 2023-04-13 pmwd/train_util.py
FIXUP parallel loading of data from disk

@@ -7,6 +7,7 @@ import optax
 from functools import partial
 from torch.utils.data import Dataset, DataLoader
 import os
+from joblib import Parallel, delayed
 
 from pmwd import (
     Configuration,
@@ -105,17 +106,18 @@ def gen_ic(seed, conf, cosmo):
     return ptcl, cosmo
 
 
-# TODO make this parallel?
 def read_g4data(sobol_ids, sims_dir, snaps_per_sim, fn_sobol):
     data = {}
-    for i, sidx in enumerate(sobol_ids):
+    def load_sobol(i, sidx):
         data[i] = {}
         sobol = scale_Sobol(fn_sobol, sidx)
         for j in range(snaps_per_sim):
             snap_file = os.path.join(sims_dir, f'{sidx:03}',
-                                        'output', f'snapshot_{j:03}')
+                                     'output', f'snapshot_{j:03}')
             pos, vel, a = read_gadget_hdf5(snap_file)
             data[i][j] = (pos, vel, a, sidx, sobol)
+    Parallel(n_jobs=min(16, len(sobol_ids)), prefer='threads', require='sharedmem')(
+        delayed(load_sobol)(i, sidx) for i, sidx in enumerate(sobol_ids))
     return data
 
 

### 2023-04-13 docs/papers/adjoint/demo_data_time_evo.py
Add demo animation scripts

@@ -1,0 +1,59 @@
+import numpy as np
+import jax.numpy as jnp
+from tqdm.notebook import tqdm
+import h5py
+
+import jax
+
+from pmwd import (
+    Configuration,
+    SimpleLCDM,
+    boltzmann,
+    white_noise, linear_modes,
+    lpt,
+    nbody,
+    scatter,
+)
+from pmwd.nbody import nbody_init, nbody_step
+
+
+ptcl_spacing = 10.
+ptcl_grid_shape = (16, 27, 16)
+mesh_shape = (32, 54, 32)
+
+conf = Configuration(ptcl_spacing, ptcl_grid_shape, mesh_shape=2,
+                     a_nbody_maxstep=1/512)
+a_nbody = conf.a_nbody
+cosmo = SimpleLCDM(conf)
+seed = 0
+modes = white_noise(seed, conf)
+
+cosmo = boltzmann(cosmo, conf)
+modes = linear_modes(modes, cosmo, conf)
+ptcl, obsvbl = lpt(modes, cosmo, conf)
+
+f = h5py.File('data/time_evo.h5', 'w')
+
+def save_snap(f, i, a, ptcl, conf, prefix):
+    pos = np.asarray(ptcl.pos())
+    dset = f.create_dataset(f'{prefix}_{i}', data=pos)
+    dset.attrs.create('a', a)
+
+
+# forward simulation
+ptcl, obsvbl = nbody_init(a_nbody[0], ptcl, obsvbl, cosmo, conf)
+save_snap(f, 0, a_nbody[0], ptcl, conf, 'fw')
+
+for i, (a_prev, a_next) in enumerate(tqdm(zip(a_nbody[:-1], a_nbody[1:]), total=len(a_nbody)-1)):
+    ptcl, obsvbl = nbody_step(a_prev, a_next, ptcl, obsvbl, cosmo, conf)
+    save_snap(f, i+1, a_next, ptcl, conf, 'fw')
+
+# reverse simulation
+a_nbody_r = a_nbody[::-1]
+save_snap(f, 0, a_nbody_r[0], ptcl, conf, 're')
+
+for i, (a_prev, a_next) in enumerate(tqdm(zip(a_nbody_r[:-1], a_nbody_r[1:]), total=len(a_nbody_r)-1)):
+    ptcl, obsvbl = nbody_step(a_prev, a_next, ptcl, obsvbl, cosmo, conf)
+    save_snap(f, i+1, a_next, ptcl, conf, 're')
+
+f.close()

### 2023-04-13 docs/papers/pp/pp.tex
FIXUP draft: a question
FIXUP draft: add complexity analysis

@@
 \usepackage{bm}
+\usepackage{dsfont}
 %\usepackage{mathrsfs}
@@
 
+\DeclareMathOperator*{\E}{\mathds{E}}
+\DeclareMathOperator{\tsum}{{\textstyle\sum}}
+\newcommand{\1}{\mathds{1}}
 \newcommand{\order}{\mathcal{O}}
@@
 
-\YL{TODO: verify that in 1D the partial sum explains the force error.}
+\YL{TODO: verify that in 1D the partial sum explains the force error.
+Two ways: mass integral vs force partial sum, off by the Nyquist?}
 
@@
 
+\subsection{Complexity}
+
+$N$ particles, $M$ cells, volume $V$, mean matter density
+$\bar\rho$, particle mass $m = \bar\rho V / N$, density field $\rho(\vx)
+= \bar\rho [1 + \delta(\vx)]$, $C_i$ is the $i$-th cell, and $D_i$ are
+the cells whose particles directly interact with those in $C_i$ (so $C_i
+\subsetneq D_i$).
+Each cell $C$ has volume $V / M$, and each direct interacting region $D$
+has volume $d V / M$ with $d$ being an integer greater than 1.
+The total number of pairs of interactions is
+%
+\begin{align}
+\tsum_{i=1}^M \E \bigl[ \tsum_{p\in C_i} \1_{p'\in D_i} \bigr]
+%
+&\simeq \frac{N^2}{V^2} \tsum_{i=1}^M \E \biggl[
+  \int_{C_i} \!\!\!\d\vx \int_{D_i} \!\!\!\d\vx'\,
+  [1 + \delta(\vx)] [1 + \delta(\vx')] \biggr] \nonumber\\
+%
+&= \frac{M N^2}{V^2} \int_C \!\!\d\vx \int_D \!\!\d\vx'\,
+  \bigl[ 1 + \xi(\vx' - \vx) \bigr]
+= \frac{d N^2}M  \bigl( 1 + \sigma^2_{CD} \bigr),
+\end{align}
+%
+where $\xi$ is the matter density 2-point correlation function, and we
+have replaced the sums over particles with the integrals over the
+density fields, $\sum_p \to m^{-1} \int \!\d\vx\, \rho(\vx) = N V^{-1}
+\int \!\d\vx\, [1 + \delta(\vx)]$, in deriving the first equality.
+We have also used the density covariance between $C$ and $D$, defined as
+%
+\begin{equation}
+\sigma^2_{CD} \triangleq \int \!\!\frac{\d\vk}{(2\pi)^3}\,
+  P(k) W_C^*(\vk) W_D(\vk)
+= \frac{M^2}{d V^2} \int_C \!\!\d\vx \int_D \!\!\d\vx'\,
+  \xi(\vx' - \vx),
+\end{equation}
+%
+where $P(k)$ is the matter density power spectrum and the Fourier
+transform of $\xi$.
+$\sigma^2_{CD}$ depends on the geometry of $C$ and $D$, through their
+Fourier space windows:
+%
+\begin{align}
+W_C(\vk) &\triangleq \frac{M}V \int_C \!\!\d\vx\, e^{i\vk\cdot\vx},
+  \nonumber\\
+W_D(\vk) &\triangleq \frac{M}{dV} \int_D \!\!\d\vx\, e^{i\vk\cdot\vx},
+\end{align}
+%
+both normalized to 1 at $\vk=\vzero$.
+
 \section{Results}

### 2023-04-14 docs/examples/sto/train/train.py
FIXUP number of input features of SO NNs

@@ -52,12 +52,8 @@ if __name__ == "__main__":
                           num_workers=0, collate_fn=lambda x: x)
 
     # structure of the so neural nets
-    # dummy conf and cosmo for getting the number of so input features
     printinfo('initializing SO parameters')
-    conf = Configuration(1., (128,)*3)
-    cosmo = SimpleLCDM(conf)
-    cosmo = boltzmann(cosmo, conf)
-    n_input = [soft_len(cosmo, conf)] * 3  # three nets
+    n_input = [soft_len()] * 3  # three nets
     nodes = [[2*n, n//2, 1] for n in n_input]
     so_params = init_mlp_params(n_input, nodes)
 

### 2023-04-14 pmwd/so_util.py

@@ -8,6 +8,9 @@ from typing import Sequence, Callable
 import math
 
 from pmwd import (
+    Configuration,
+    SimpleLCDM,
+    boltzmann,
     H_deriv,
     Omega_m_a,
     growth,
@@ -119,7 +122,11 @@ def sotheta(cosmo, conf, a):
     return (theta_l, theta_o)
 
 
-def soft_len(cosmo, conf):
+def soft_len():
+    # get the length of SO input features with dummy conf and cosmo
+    conf = Configuration(1., (128,)*3)
+    cosmo = SimpleLCDM(conf)
+    cosmo = boltzmann(cosmo, conf)
     tl, to = sotheta(cosmo, conf, conf.a_start)
     return len(tl) + len(to)
 

### 2023-04-24 docs/papers/sto/sto.tex
FIXUP draft: edit TODOs
FIXUP draft: add comments

@@
 \begin{itemize}
+\item Separate sto to SO and TO.
 \item Moving some of the following ongoing and completed tasks to the
@@
   deconvolution, with SO.
+\item In TO, time step spacing can be important for light cone
+  generation. We can train on light cone target data to learn TO for
+  light cones. The result may be related to $c \d t/H = c \d a / a^2 H$.
 \end{itemize}
@@
 \begin{equation}
-\frac{k_i}{k^2} \to \frac{k_i}{k^2}
+\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
   h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
@@
 Parameters in these neural networks are optimized during training.
+
+\YL{While $g$ is spherically symmetric by assumption, $f$ and $h$ are
+expected to be even functions, i.e., real valued in Fourier space.
+It is interesting to symbolically regress both $h(k_i)$ and $k_i h(k_i)$
+to see which one gives simpler expression, given the possible different
+rankings by complexity.
+For example, $k_i h(k_i)$ may have a simpler form as a nonlinear odd
+function.}
 

### 2023-05-04 pmwd/train_util.py
FIXUP add tensorboard writer & misc

@@ -7,6 +7,7 @@ from functools import partial
 from torch.utils.data import Dataset
 import os
 from joblib import Parallel, delayed
+import matplotlib.pyplot as plt
 
 from pmwd import (
     Configuration,
@@ -21,6 +22,7 @@ from pmwd import (
 )
 from pmwd.pm_util import rfftnfreq
 from pmwd.io_util import read_gadget_hdf5
+from pmwd.spec_util import powspec
 
 
 def scale_Sobol(fn='sobol.txt', ind=slice(None)):
@@ -118,7 +120,7 @@ def read_g4data(sobol_ids, sims_dir, snaps_per_sim, fn_sobol):
                                      'output', f'snapshot_{j:03}')
             pos, vel, a = read_gadget_hdf5(snap_file)
             data[i][j] = (pos, vel, a, sidx, sobol)
-    Parallel(n_jobs=min(16, len(sobol_ids)), prefer='threads', require='sharedmem')(
+    Parallel(n_jobs=min(8, len(sobol_ids)), prefer='threads', require='sharedmem')(
         delayed(load_sobol)(i, sidx) for i, sidx in enumerate(sobol_ids))
     return data
 
@@ -147,15 +149,13 @@ class G4snapDataset(Dataset):
 
 
 def _loss_dens_mse(dens, dens_t):
-    return jnp.sum((dens - dens_t)**2)
-
+    return (dens - dens_t).var()
 
 def _loss_disp_mse(ptcl, ptcl_t):
-    return jnp.sum((ptcl.disp - ptcl_t.disp)**2)
-
+    return (ptcl.disp - ptcl_t.disp).var()
 
 def _loss_vel_mse(ptcl, ptcl_t):
-    return jnp.sum((ptcl.vel - ptcl_t.vel)**2)
+    return (ptcl.vel - ptcl_t.vel).var()
 
 
 @jax.custom_vjp
@@ -163,25 +163,27 @@ def _loss_scale_wmse(kvec, dens_k, dens_t_k):
     k2 = sum(k**2 for k in kvec)
     diff = dens_k - dens_t_k
     loss = jnp.sum(jnp.where(k2 != 0, jnp.abs(diff)**2 / k2**1.5, 0))
-    return loss, (k2, diff)
+    return jnp.log(loss), (loss, k2, diff)
 
 def _loss_scale_wmse_fwd(kvec, dens_k, dens_t_k):
[…]

### 2023-05-04 docs/examples/sto/train/train.py
FIXUP add tensorboard writer & misc

@@ -7,6 +7,8 @@ procid = int(os.getenv('SLURM_PROCID'))
 n_tasks_per_node = int(os.getenv('SLURM_NTASKS_PER_NODE'))
 os.environ['CUDA_VISIBLE_DEVICES'] = str(procid % n_tasks_per_node)
 
+os.environ['XLA_PYTHON_CLIENT_MEM_FRACTION'] = '.95'
+
 import jax
 import jax.numpy as jnp
 import numpy as np
@@ -19,7 +21,7 @@ import time
 import pickle
 
 from pmwd.so_util import soft_len, init_mlp_params
-from pmwd.train_util import G4snapDataset, train_step
+from pmwd.train_util import G4snapDataset, train_step, visins
 
 
 def printinfo(s, flush=False):
@@ -33,8 +35,8 @@ if __name__ == "__main__":
     jax.distributed.initialize(local_device_ids=[0])
 
     # hyper parameters of training
-    n_epochs = 1
-    learning_rate = 0.01
+    n_epochs = 10
+    learning_rate = 0.1
     sobol_ids = np.arange(0, 8)
 
     # RNGs with fixed seeds, for same randomness across processes
@@ -48,14 +50,14 @@ if __name__ == "__main__":
     # load training data
     printinfo('preparing the data loader')
     g4data = G4snapDataset('g4sims', sobol_ids=sobol_ids)
-    g4loader = DataLoader(g4data, batch_size=None, shuffle=True, generator=tc_rng,
+    g4loader = DataLoader(g4data, batch_size=None, shuffle=False, generator=tc_rng,
                           num_workers=0, collate_fn=lambda x: x)
 
     # structure of the so neural nets
     printinfo('initializing SO parameters & optimizer')
     n_input = [soft_len()] * 3  # three nets
     so_nodes = [[2*n, n, 1] for n in n_input]
-    so_params = init_mlp_params(n_input, so_nodes)
+    so_params = init_mlp_params(n_input, so_nodes, scheme='last_w0_b1')
 
     optimizer = optax.adam(learning_rate=learning_rate)
     opt_state = optimizer.init(so_params)
@@ -67,27 +69,35 @@ if __name__ == "__main__":
 
     tic = time.perf_counter()
     for epoch in range(n_epochs):
-        for i, g4snap in enumerate(g4loader):
+        for step, g4snap in enumerate(g4loader):
             pos, vel, a, sidx, sobol = g4snap
 
[…]

### 2023-05-04 pmwd/train_util.py
FIXUP option to load given snap ids

@@ -110,42 +110,51 @@ def gen_ic(seed, conf, cosmo):
     return ptcl, cosmo
 
 
-def read_g4data(sobol_ids, sims_dir, snaps_per_sim, fn_sobol):
+def read_g4data(sims_dir, sobol_ids, snap_ids, fn_sobol):
     data = {}
-    def load_sobol(i, sidx):
-        data[i] = {}
-        sobol = scale_Sobol(fn_sobol, sidx)
-        for j in range(snaps_per_sim):
-            snap_file = os.path.join(sims_dir, f'{sidx:03}',
-                                     'output', f'snapshot_{j:03}')
+    def load_sobol(sobo):
+        data[sobo] = {}
+        sobol = scale_Sobol(fn_sobol, sobo)
+        for snap in snap_ids:
+            snap_file = os.path.join(sims_dir, f'{sobo:03}',
+                                     'output', f'snapshot_{snap:03}')
             pos, vel, a = read_gadget_hdf5(snap_file)
-            data[i][j] = (pos, vel, a, sidx, sobol)
+            data[sobo][snap] = (pos, vel, a, sobo, sobol)
     Parallel(n_jobs=min(8, len(sobol_ids)), prefer='threads', require='sharedmem')(
-        delayed(load_sobol)(i, sidx) for i, sidx in enumerate(sobol_ids))
+        delayed(load_sobol)(sobo) for sobo in sobol_ids)
     return data
 
 
 class G4snapDataset(Dataset):
 
     def __init__(self, sims_dir, sobol_ids=None, sobols_edge=None,
-                 snaps_per_sim=121, fn_sobol='sobol.txt'):
+                 snap_ids=None, snaps_per_sim=121, fn_sobol='sobol.txt'):
         self.sims_dir = sims_dir
+
         if sobol_ids is None:
             sobol_ids = np.arange(*sobols_edge)
+        self.sobol_ids = sobol_ids
+
+        if snap_ids is not None:
+            self.snap_ids = snap_ids
+            self.snaps_per_sim = len(snap_ids)
+        else:
+            self.snap_ids = np.arange(snaps_per_sim)
+            self.snaps_per_sim = snaps_per_sim
+
         self.n_sims = len(sobol_ids)
-        self.snaps_per_sim = snaps_per_sim
         self.n_snaps = self.n_sims * self.snaps_per_sim
 
[…]

### 2023-05-15 pmwd/sto/so.py
FIXUP add another init scheme w/ small weigths

@@ -42,13 +42,23 @@ def init_mlp_params(n_input, nodes, scheme=None):
     # by default in flax.linen.Dense, kernel: lecun_norm, bias: 0
     params = [nn.init(key, x) for nn, key, x in zip(nets, keys, xs)]
 
-    # for the last layer: set weights to zero and bias to one
-    # TODO instead of zero, use very small random values for weights?
+    # for the last layer: set bias to one & weights to zero
     if scheme == 'last_w0_b1':
         for i, p in enumerate(params):
             p = unfreeze(p)
             p['params'][f'layers_{len(nodes[i])-1}']['kernel'] = (
                 jnp.zeros((nodes[i][-2], nodes[i][-1])))
+            p['params'][f'layers_{len(nodes[i])-1}']['bias'] = (
+                jnp.ones(nodes[i][-1]))
+            params[i] = freeze(p)
+
+    # for the last layer: set bias to one & weights to small random values
+    if scheme == 'last_ws_b1':
+        keys = random.split(random.PRNGKey(1), len(params))
+        for i, p in enumerate(params):
+            p = unfreeze(p)
+            p['params'][f'layers_{len(nodes[i])-1}']['kernel'] = (
+                random.normal(keys[i], (nodes[i][-2], nodes[i][-1]))) * 1e-4
             p['params'][f'layers_{len(nodes[i])-1}']['bias'] = (
                 jnp.ones(nodes[i][-1]))
             params[i] = freeze(p)

### 2023-05-15 docs/papers/pp/pp.tex
FIXUP draft: add 1D force analysis

@@
 \usepackage[T1]{fontenc}
-\usepackage[margin=3cm]{geometry}
+\usepackage[margin=2cm]{geometry}
 %\usepackage{setspace}
@@
 %\usepackage{mathrsfs}
-%\usepackage{soul}
+\usepackage{soul}
 %\usepackage{tensor}
@@
     filecolor=BrickRed,
-    citecolor=LimeGreen,
-    urlcolor=Cerulean,
+    citecolor=BrickRed,
+    urlcolor=RoyalBlue,
   }
-\usepackage{natbib}
-\usepackage[dvipsnames,svgnames,x11names]{xcolor}
+\usepackage[numbers]{natbib}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{comment}
@@
 \renewcommand{\subsectionautorefname}{Sec.}
+\renewcommand{\subsubsectionautorefname}{Sec.}
 \renewcommand{\appendixautorefname}{App.}
@@
 \newcommand{\1}{\mathds{1}}
+\newcommand{\R}{\mathds{R}}
+\newcommand{\Z}{\mathds{Z}}
 \newcommand{\order}{\mathcal{O}}
@@
 \bmdefine{\vn}{n}  % wavevector indices
+\bmdefine{\vm}{m}  % grid point indices
+\bmdefine{\vj}{j}  % integer indices
 \bmdefine{\vx}{x}  % comoving and canonical coordinates
@@
 
-\YL{Alex suggested starting from 1D.}
-
-\YL{Describe briefly normal PM here. Also introduce possible filter
-below. Explain that we are not doing splitting the usual way.}
+\YL{Describe briefly PM (and our corrections) here. Also introduce
+possible filter below. Explain that we are not doing splitting the
+usual way.}
 
@@
 
-In a periodic box of size $L_1 \times \cdots \times L_d$ in
-$\mathbb{R}^d$, Poisson's equation,
+In a periodic box of size $L_1 \times \cdots \times L_d$ in $\R^d$,
+Poisson's equation,
 %
@@
 \nabla^2 u = f,
+\label{Poisson}
 \end{equation}
@@
 in Fourier space, with discretized wavevectors $\vk_\vn \triangleq (2\pi
-n_1 / L_1, \cdots, 2\pi n_d / L_d )^\intercal$ and integrer $n_i \in
-\mathbb{Z}/N_i\mathbb{Z}$.
+n_1 / L_1, \cdots, 2\pi n_d / L_d )^\intercal$ and $n_i \in \Z$.
 And the force $- \vnabla u$ becomes $i \vk_\vn f / k_\vn^2$.
+When \eqref{Poisson} is further discretized by a $N_1 \times \cdots
+\times N_d$ mesh of cell size $l$, with $\vx_\vm \triangleq (m_1 l,
+\cdots, m_d l)^\intercal$ and $m_i \in \Z/N_i\Z$, $\vk_\vn$ is also
+periodic with $n_i \in \Z/N_i\Z$.
 
-At finite resolution of the mesh cell size $l$, the bandwidth is
-truncated at the Nyquist frequency $\knyq = \pi / l$.
-In a 1D periodic length $L$ with a mesh of size $N = L / l$, the
-partial sum is equivalent to a convolution by (the Dirichlet kernel)
+At finite resolution, the bandwidth is truncated at the Nyquist
+frequency $\knyq = \pi / l$.
+In 1D, the partial sum is equivalent to a convolution by (the Dirichlet
+kernel)
 %
@@
 \frac1{2 L} \Bigg(
-  \sum_{n=-\lfloor \frac{N}2\rfloor}^{\lfloor \frac{N}2\rfloor}
-  + \sum_{n=-\lfloor \frac{N-1}2\rfloor}^{\lfloor \frac{N-1}2\rfloor}
-\Bigg)
-\exp\bigl( i n\kfun x \bigr)
+  \sum_{n=-\lfloor\frac{N}2\rfloor}^{\lfloor\frac{N}2\rfloor}
+  + \sum_{n=-\lfloor\frac{N-1}2\rfloor}^{\lfloor\frac{N-1}2\rfloor}
+\Bigg) e^{i k_n x}
 %
@@
 \begin{equation}
-\frac1L
-  \sum_{n=-\lfloor \frac{N-1}2\rfloor}^{\lfloor \frac{N-1}2\rfloor}
-  \exp\bigl( i n\kfun x \bigr)
+\frac1L \sum_{n=-\lfloor\frac{N-1}2\rfloor}^{\lfloor\frac{N-1}2\rfloor}
+  e^{i k_n x}
 %
@@
   \bigl( \bigl\lfloor\frac{N-1}2\bigr\rfloor + \frac12 \bigr) \kfun x
-  \bigr]
-}{\sin \bigl( \frac12 \kfun x \bigr) L}
+\bigr]}{\sin \bigl( \frac12 \kfun x \bigr) L}.
 \end{equation}
+%
 
-\YL{TODO: verify that in 1D the partial sum explains the force error.
-Two ways: mass integral vs force partial sum, off by the Nyquist?}
+In 1D, the partial summed force of a point unit source at $x=0$ is
+%
+\begin{equation}
+\frac1L \Bigg(
+  \sum_{n=-\lfloor\frac{N-1}2\rfloor}^{-1}
+  + \sum_{n=1}^{\lfloor\frac{N-1}2\rfloor}
+\Bigg) \frac{i k_n}{k_n^2} e^{i k_n x}
+= -2 \sum_{n=1}^{\lfloor\frac{N-1}2\rfloor} \frac{\sin k_n x}{k_n L},
+\end{equation}
+%
+which as $N\to\infty$ has the limit
+%
+\begin{equation}
+-2 \sum_{n=1}^\infty \frac{\sin k_n x}{k_n L} = \frac{x}L - \frac12,
+\qquad 0 < x < L,
+\end{equation}
+%
+following, e.g., (1.441.1) in \cite{TISP}.
+Therefore, the 1D periodic force is linear, in contrast to constant as
+in the open boundary case.
+And all frequencies with $n > \lfloor\frac{N-1}2\rfloor$ sum up to the
+error between the linear force and its partial sum.
 
-\YL{Minimize force error above a certain radius, e.g. beyond 27 cells,
-(in a way similar to choosing NUFFT kernel?):
+We can introduce a window $w(k_n)$ to correct this force error, so that
+at all $x_m \neq 0$ the corrected partial sum gives the linear force,
+i.e.,
+%
+\begin{equation}
+\sum_{n=1}^{N-1} \frac{i w(k_n)}{k_n L} e^{i k_n x_m}
+= \frac{x_m}L - \frac12,
+\qquad m \in \{1, \cdots, N-1\},
+\end{equation}
+%
+with vanishing self-force at $m=0$.
+We can discrete Fourier transform the right hand side to find
+%
+\begin{equation}
+w(k_n) = \frac{-i k_n L}N \sum_{m=1}^{N-1}
+\Bigl( \frac{x_m}L - \frac12 \Bigr) e^{-i k_n x_m}
+%
+= - k_n l \sum_{m=1}^{N-1}
+\Bigl( \frac{x_m}L - \frac12 \Bigr) \sin k_n x_m
+%
+= \frac{k_n l / 2}{\tan k_n l / 2}.
+\end{equation}
+%
+We have used (1.352.1) from \cite{TISP} in the last equality.
+\YL{Note the similarity to a deconvolution of the first order B-spline
+window.}
+Alternatively, combining eqX and eqY,
+%
+\begin{equation}
+w(k_n) = \sum_{j=-\infty}^\infty \frac{n}{n+jN}
+\end{equation}
+%
+
+In 3D:
+%
+\begin{equation}
+\frac{n_i w_i(\vk_\vn)}{\vn^2}
+= \sum_\vj \frac{(\vn + \vj N)_i}{(\vn + \vj N)^2}
+\end{equation}
+%
+
+For example, $i=3$:
+%
+\begin{equation}
+\frac{n_3 w_3(\vk_\vn)}{\vn^2}
+= \sum_\vj
+\frac{n_3 + j_3 N}{(n_1 + j_1 N)^2 + (n_2 + j_2 N)^2 + (n_3 + j_3 N)^2}
+\end{equation}
+%
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=0.7\linewidth]{force1D.pdf}
+\caption{1D periodic forces with grids of even and odd lengths, from a
+point source at $x=0$.}
+\label{fig:force1D}
+\end{figure*}
+
+\subsubsection{Windowed partial sum?}
+
+Consider the $p$-th order B-spline window in 1D,
+%
+\begin{equation}
+W_p(k_n) = \biggl( \frac{\sin k_n l / 2}{k_n l / 2} \biggr)^p.
+\end{equation}
+%
+When the particles are scattered to the mesh (see \autoref{sec:equiv})
+using such a kernel, one usual practice is to approximately ``undo'' the
+smearing by a deconvolution.
+This is done by dividing the force by $W_p^2(k_n)$, with $W_p$ squared
+because of the additional smearing when gathering the particle forces
+from the mesh.
+\st{
+If we take the deconvolutions into account in modeling the partial sum
+approximation, this will change the shape of the Dirichlet kernel.
+However, we are not able to find the analytic expression of the new
+kernel.
+}
+
+\YL{\st{Minimize force error above a certain radius, e.g. beyond 27 cells,
+(in a way similar to choosing NUFFT kernel?)}:
 \begin{itemize}
@@
 
-\YL{2 possible ways to optimize:
+\YL{\st{2 possible ways to optimize:
 suppress the oscillation outside of the PP patch size;
 shift the phase of the oscillation so that there are both positive and
-negative parts between two nearby mesh points.
+negative parts between two nearby mesh points.}
 }
@@
 \subsubsection{Equivalent Point Source}
+\label{sec:equiv}
 
 simple cubic (CIC), bcc, interlocking/interlacing cubic.
-\YL{See \citet{HockneyEastwood1988} for the interlacing one.
+\YL{See \cite{HockneyEastwood1988} for the interlacing one.
 Other references:
@@
 are uniform (and Gaussian with known power spectrum) on large scales.
-\YL{Complexity analysis given variance of particle density.}
-
-\YL{Optimal mesh cell size given variance of particle density for fixed
-error budget, cosmology and time dependence.}
 
@@
 %
-&\simeq \frac{N^2}{V^2} \tsum_{i=1}^M \E \biggl[
+&\simeq \tsum_{i=1}^M \frac{N^2}{V^2}
   \int_{C_i} \!\!\!\d\vx \int_{D_i} \!\!\!\d\vx'\,
-  [1 + \delta(\vx)] [1 + \delta(\vx')] \biggr] \nonumber\\
+  \E \bigl\{ [1 + \delta(\vx)] [1 + \delta(\vx')] \bigr\} \nonumber\\
 %
@@
 
+\YL{Optimal mesh cell size given $\sigma^2_{CD}$ for fixed error budget,
+cosmology and time dependence.}
+
 \section{Results}
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{plain}
 \bibliography{pp}

### 2023-05-15 docs/papers/pp/force1D.py

@@ -1,0 +1,115 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+def force1D_sum(N, x=None, kernel=lambda k: 1, error=False):
+    """Partial summed 1D force kernel, with cell size l=1."""
+    n = np.arange(1, (N-1)//2 + 1)
+    n = n[:, None]
+    k = 2 * np.pi / N * n
+    if x is None:
+        x = np.arange(N)
+    force = -2 * (kernel(k) / (k * N) * np.sin(k * x)).sum(axis=0)
+    if error:
+        force -= force1D(N, x=x)
+    return force
+
+
+def force1D_pm(N, d=None, x_src=0.5, kernel=lambda k: 1, error=False):
+    """PM 1D force kernel, with cell size l=1."""
+    freq_period = 2 * np.pi
+    nyquist = np.pi
+    eps = nyquist * np.finfo(np.float64).eps
+
+    k = np.fft.rfftfreq(N) * freq_period
+
+    # 1D scatter
+    dens = np.zeros(N)
+    i_src = np.floor(x_src).astype(int)
+    t_src = x_src - i_src
+    dens[i_src % N] = 1 - t_src
+    dens[(i_src+1) % N] = t_src
+
+    dens = np.fft.rfft(dens)
+    pot = np.where(k != 0, - dens / k**2, 0)
+    grad = np.where(np.abs(np.abs(k) - nyquist) <= eps, 0, -1j * k * pot)
+    grad *= kernel(k)
+    grad = np.fft.irfft(grad, n=N)
+
+    # 1D gather
+    if d is None:
+        d = np.arange(N, dtype=float)
+    x = x_src + d
+    i = np.floor(x).astype(int)
+    t = x - i
+    grad = (1-t) * grad[i % N] + t * grad[(i+1) % N]
+
+    if error:
+        grad -= force1D(N, x=d)
+
+    return grad
+
+
+def force1D(N, x=None):
+    """Analytic 1D force kernel, with cell size l=1."""
+    if x is None:
+        x = np.arange(N)
+    return x / N - 0.5
+
+
+def force1D_corr(k):
+    """1D force correction kernel, with cell size l=1."""
+    return np.where(k != 0, 0.5 * k / np.tan(0.5 * k), 1)
+
+
+plt.style.use('font.mplstyle')
+
+rows = ['1D force', 'force error']
+cols = [16, 15] #, 126, 125]
+
+fig, axes = plt.subplots(
[…]

### 2023-05-16 docs/papers/sto/sto.tex
FIXUP draft: usepackage xcolor
FIXUP draft: change loss function

@@
 \usepackage{bm}
+\usepackage{dsfont}
 \usepackage{graphicx}
@@
 \usepackage{nicefrac}
+\usepackage[dvipsnames]{xcolor}
 
@@
 \item Compare sto DM and halo fields and summary stats with those of raw
-  pmwd, deconvolved pmwd, and gadget, on the last 9 test cases.
+  pmwd, deconvolved pmwd, and gadget, on the last 9 test cases and
+  noncubic box.
 \item Demonstrate capability of generalization, on some cases with
@@
 
-Again combine Lagrangian and Eulerian fields.
+A mean squared error (MSE) loss on the fields is dominated by the small
+scale modes.
+In this work we choose uniform weights in the scale space on the
+relative MSE.
+We bin the modes logarithmically by \nicefrac13 octave in scale, e.g.,
+in the $i$-th $k$-bin, $K_i \triangleq [k_i, k_{i+1})$, with $k_i =
+2^{1+i/3} \pi / L$ and $i = 0, 1, \cdots$.
+Our loss on $f$ sums up the MSE relative to the target field power in
+logarithmic bins:
+%
+\begin{equation}
+\cL_f = \sum_i
+\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |f(\vk')|^2}.
+\end{equation}
+%
+To understand the form of the loss function, let's introduce for each
+mode the square root ratio of power
+%
+\begin{equation}
+T(\vk) \triangleq
+\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk')|^2}},
+\end{equation}
+%
+and the cross correlation coefficient
+%
+\begin{equation}
+r(\vk) \triangleq
+\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
+     {\sqrt{|\hat f(\vk)|^2 |f(\vk')|^2}}.
+\end{equation}
+%
+Now, under the mild assumptions that $T$ and $r$ are smooth and
+isotropic, we can show
+%
+\begin{equation}
+\cL_f \simeq \sum_\vk \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
+\end{equation}
+%
+that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
+when $\hat f$ is identical to $f$.
+Above we choose the binned version instead for the robustness against
+the noisy modewise $|f(\vk)|^2$ in the denominator.
 
-A mean squared error (MSE) loss on the fields would focus too much on
-the small scales. In this work we choose uniform weights in the scale
-space, by weighing each mode of wavenumber $k$ by $k^{-3}$ in the MSE
-loss
+We combine Lagrangian and Eulerian fields to form the total loss:
+%
 \begin{equation}
-  \cJ = \mathbb{E} \sum_\vk k^{-3}
-    \bigl| \hat f(\vk) - f(\vk) \bigr|^2.
+\cJ = \cL_\vs + \cL_\delta
++ \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
++ \cdots
 \end{equation}
+%
+\YL{(a simple sum since we have already done careful normalization in
+$\cL_f$'s) and also try}
+%
+\begin{equation}
+\cJ = \ln \cL_\vs + \ln \cL_\delta
++ \ln \cL_{(1+\delta)\vv} + \ln \cL_{(1+\delta)\vv\vv}
++ \cdots
+\end{equation}
+%
+Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
+Lagrangian space, while the rest are in the Eulerian space:
+%
+\begin{align}
+1 + \delta(\vx) &= \int\!\d^3\vq\,
+\deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+%
+\bigl[(1+\delta)\vv\bigr]_i(\vx) &= \int\!\d^3\vq\,
+\deltaD[ \vx - \vq - \vs(\vq)] v_i, \nonumber\\
+%
+\bigl[(1+\delta)\vv\vv\bigr]_{ij}(\vx) &= \int\!\d^3\vq\,
+\deltaD[ \vx - \vq - \vs(\vq)] v_i v_j,
+\end{align}
+%
+which we compute by scattering particles and their velocities to the
+\YL{choice of mesh for loss}.
 

### 2023-05-17 pmwd/spec_util.py
Add log and arbitrary binning in powspec

@@ -1,3 +1,4 @@
+import math
 from functools import reduce
 from operator import mul
 
@@ -6,8 +7,9 @@ import jax.numpy as jnp
 from pmwd.pm_util import rfftnfreq
 
 
-def powspec(f, spacing, dk=None, g=None, deconv=0, int_dtype=jnp.uint32):
-    """Compute auto or cross power spectrum in 3D in spherical bins.
+def powspec(f, spacing, bins=1j/3, g=None, deconv=0, cut_zero=True, cut_nyq=True,
+            int_dtype=jnp.uint32):
+    """Compute auto or cross power spectrum in 3D averaged in spherical bins.
 
     Parameters
     ----------
@@ -15,34 +17,40 @@ def powspec(f, spacing, dk=None, g=None, deconv=0, int_dtype=jnp.uint32):
         The field, with the last 3 axes for FFT and the other summed over.
     spacing : float
         Field grid spacing.
-    dk : float, optional
-        Wavenumber bin width. Default is the largest fundamental frequency.
+    bins : float, complex, or 1D array_like, optional
+        Wavenumber bins. A real number sets the linear spaced bin width in unit of the
+        smallest fundamental in 3D (right edge inclusive starting from zero); an
+        imaginary number sets the log spaced bin width in octave (left edge inclusive
+        starting from the smallest fundamental in 3D); and an array sets the bin edges
+        directly (right edge inclusive and must starting from zero).
     g : array_like, optional
         Another field of the same shape for cross correlation.
     deconv : int, optional
         Power of sinc factors to deconvolve in the power spectrum.
+    cut_zero : bool, optional
+        Whether to discard the bin containing the zero or DC mode.
+    cut_nyq : bool, optional
+        Whether to discard the bins beyond the Nyquist.
     int_dtype : dtype_like, optional
         Integer dtype for the number of modes.
 
     Returns
     -------
     k : jax.numpy.ndarray
-        Wavenumber, excluding the DC mode in the first bin and stopping before or at the
-        Nyquist frequency in the last one.
[…]

### 2023-05-17 docs/papers/sto/sto.tex
FIXUP draft: correction and comments

@@
 A mean squared error (MSE) loss on the fields is dominated by the small
-scale modes.
+scale modes \citep{HeEtAl2019}.
 In this work we choose uniform weights in the scale space on the
 relative MSE.
-We bin the modes logarithmically by \nicefrac13 octave in scale, e.g.,
-in the $i$-th $k$-bin, $K_i \triangleq [k_i, k_{i+1})$, with $k_i =
-2^{1+i/3} \pi / L$ and $i = 0, 1, \cdots$.
+We bin the modes logarithmically by \nicefrac13 octave in scale, with
+the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
+2^{1+i/3} \pi / L$.
 Our loss on $f$ sums up the MSE relative to the target field power in
@@
 \begin{equation}
-\cL_f \simeq \sum_\vk \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+\cL_f \propto \int \!\d\ln\!k\, \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
   + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
@@
 the noisy modewise $|f(\vk)|^2$ in the denominator.
+The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
+loss, while our modification is important to balance among all scales so
+that the neural network does not focus too much on either high or low
+$k$'s.
 
@@
 The curvature $\Omega_k$ is related to the separate universe simulation
-\citep{LiEtAl2014a, WagnerEtAl2015}.
+\citep{LiEtAl2014}.
 

### 2023-05-18 docs/papers/sto/sto.tex
FIXUP draft: add coauthors

@@
 %
-Yin Li,$^{1, 3}$\thanks{Email: XYZ, eelregit@gmail.com, and XYZ}
+Yin Li,$^{1, 3, 4}$\thanks{Email: XYZ, eelregit@gmail.com, and XYZ}
 %
-and Drew Jamieson$^{4}$
+Drew Jamieson,$^{5}$
+%
+Libin Lu,$^{3}$
+%
+and XYZ
 %
@@
 %
-\\$^3$Center for Computational Astrophysics \& Center for Computational
-Mathematics, Flatiron Institute, New York, NY 10010, USA
+\\$^3$Center for Computational Mathematics, Flatiron Institute, New
+York, NY 10010, USA
 %
-\\$^4$Max Planck Institute for Astrophysics, 85748 Garching bei
+\\$^4$Center for Computational Astrophysics, Flatiron Institute, New
+York, NY 10010, USA
+%
+\\$^5$Max Planck Institute for Astrophysics, 85748 Garching bei
 M\"unchen, Germany

### 2023-05-18 docs/papers/sto/sto.tex
FIXUP draft: loss engineering

@@
 2^{1+i/3} \pi / L$.
-Our loss on $f$ sums up the MSE relative to the target field power in
-logarithmic bins:
+Our loss on a field $f$ sums up its MSE relative to the target field
+power in logarithmic bins:
 %
 \begin{equation}
-\cL_f = \sum_i
-\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |f(\vk')|^2}.
+\cL_f = \frac1{N_K} \sum_i
+\frac{\sum_{k \in K_i} w(\vk) \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |f(\vk')|^2},
 \end{equation}
 %
+\textbf{
+where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
+factor that suppresses the high $k$ contribution in Eulerian fields such
+as the density field (with $w=1$ for the Lagrangian displacement field).
+This is because a PM solver gradually loses small scale information step
+by step, and without $w$ setting the priority, the loss function could
+encourage the network to fit the small scales at the cost of the
+intermediate scales.
+As a result, we can set the loss mesh size the same as that of the PM
+force solver.
+We can try for $w$:
+%
+\begin{equation}
+\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{and its square},
+\prod_{j=1}^3 \cos (k_j l / 2) \textrm{and its square},
+\end{equation}
+%
+which vanish when any $k_j$ reaches the Nyquist.
+}
+\YL{Maybe add a figure of $w$ along axis, face diagonal and body
+digonal, if this works.}
+
 To understand the form of the loss function, let's introduce for each
@@
 \begin{equation}
-\cL_f \propto \int \!\d\ln\!k\, \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+\cL_f \propto \int \!\d\ln\!k\, w(\vk)
+\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
   + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
@@
 \begin{itemize}
-\item $l_\mathrm{cell}$: PM cell size.
+\item $l$: PM cell size.
 \item particle spacing;

### 2023-05-29 docs/papers/sto/sto.tex
FIXUP draft: loss engineering

@@
 \begin{equation}
+\cL_f = \frac1{N_K} \sum_i \ln \biggl[
+\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |f(\vk')|^2} \biggr],
+\end{equation}
+%
+\begin{equation}
 \cL_f = \frac1{N_K} \sum_i
@@
 %
-\textbf{
 where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
@@
 \begin{equation}
-\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{and its square},
-\prod_{j=1}^3 \cos (k_j l / 2) \textrm{and its square},
+\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{to some power},
+\prod_{j=1}^3 \cos (k_j l / 2) \textrm{to some power},
 \end{equation}
@@
 which vanish when any $k_j$ reaches the Nyquist.
-}
 \YL{Maybe add a figure of $w$ along axis, face diagonal and body
@@
 isotropic, we can show
+%
+\begin{equation}
+\cL_f \propto \int \!\d\ln\!k\, \ln
+\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty;
+\end{equation}
 %

### 2023-05-31 docs/papers/sto/sto.tex
FIXUP draft: simplify loss
FIXUP draft: add disp MSE loss

@@
 \newcommand{\pmwd}{{\usefont{T1}{nova}{m}{sl}pmwd}}
+\newcommand{\GADGET}{{{\fontsize{7.7pt}{9pt}\selectfont GADGET}-4}}
 
@@
     \textbf{Update}: code has been added.
-\item Compare one pair of Gadget4 and \pmwd\ (with say 100 to 1000 time
+\item Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
     steps) results and make sure that they agree perfectly on the large
@@
     linear regime), say 1 Gpc/$h$.\newline
-    \textbf{Update}: for Sobol=6, boxsize=983Mpc, Gadget4 and \pmwd\ are visually
+    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
     consistent at the field level. To further check power spectrum. For odd mesh
-    shapes (e.g. 1, 3, 5), pmwd and Gadget are perfectly consistent on large
+    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
     scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
@@
 \item Compare sto DM and halo fields and summary stats with those of raw
-  pmwd, deconvolved pmwd, and gadget, on the last 9 test cases and
+  pmwd, deconvolved pmwd, and \GADGET, on the last 9 test cases and
   noncubic box.
@@
 
-\subsection{Objective Function}
+\subsection{Loss Function}
 
-A mean squared error (MSE) loss on the fields is dominated by the small
-scale modes \citep{HeEtAl2019}.
+For the displacement field, we adopt the mean squared error (MSE) loss
+that proved effective in many previous works
+\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
+%
+\begin{equation}
+\cL_\vs = \ln \sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2,
+\end{equation}
+%
+in which a particle originating from its Lagrangian position $\vq$ is
+displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
+prediction of \pmwd\ model in training.
+Note that we have further taken the logarithm to combine it with the
+other loss(es) below.
+% Say that only s loss is not enough even though in principle it is, and cite DrewEtAl x2 too
+
+However, in the Eulerian space, an MSE loss is dominated by the small
+scale modes.
 In this work we choose uniform weights in the scale space on the
@@
 2^{1+i/3} \pi / L$.
-Our loss on a field $f$ sums up its MSE relative to the target field
-power in logarithmic bins:
+Our loss on the overdensity field $\delta$ sums up its MSE relative to
+the target field power in logarithmic bins:
 %
 \begin{equation}
-\cL_f = \frac1{N_K} \sum_i \ln \biggl[
-\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |f(\vk')|^2} \biggr],
+\cL_\delta = \frac1{N_K} \sum_i \,\ln\!\!
+\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2,
 \end{equation}
@@
 \begin{equation}
-\cL_f = \frac1{N_K} \sum_i
-\frac{\sum_{k \in K_i} w(\vk) \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |f(\vk')|^2},
+\cL_\delta = \frac1{N_K} \sum_i
+\frac{\sum_{k \in K_i} w(\vk)
+      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |\delta(\vk')|^2},
 \end{equation}
@@
 \begin{equation}
-\cL_f \propto \int \!\d\ln\!k\, \ln
+\cL_\delta \propto \int \!\d\ln\!k\, \ln
 \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
-  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty;
+  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
 \end{equation}
 %
+up to a constant;
+%
 \begin{equation}
-\cL_f \propto \int \!\d\ln\!k\, w(\vk)
+\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
 \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
@@
 that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
-when $\hat f$ is identical to $f$.
+when $\hat\delta$ is identical to $\delta$.
 Above we choose the binned version instead for the robustness against
-the noisy modewise $|f(\vk)|^2$ in the denominator.
+the noisy modewise $|\delta(\vk)|^2$ in the denominator.
 The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
@@
 \centering
-\caption{Ranges of GADGET-4 and \pmwd\ configuration and cosmological
+\caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
 parameters.
@@
 The curvature $\Omega_k$ is related to the separate universe simulation.
-We sample parameters applicable to GADGET-4 during data generation, and
+We sample parameters applicable to \GADGET\ during data generation, and
 those applicable to \pmwd\ during training.
-\textsuperscript\dag The box size is determined jointly by the \pmwd\
+\textsuperscript\dag{}The box size is determined jointly by the \pmwd\
 mesh shape and mesh cell size below, which are assumed to be sampled
@@
 In practice, we sample one conditioned on the other and the box size.
-* Number of time steps from $a=1/16$ to $a=1$.
+*Number of time steps from $a=1/16$ to $a=1$.
 }
@@
   \cmidrule(lr){2-4}
-  GADGET-4 & $\As \times 10^9$ & log-uniform & [1, 4) \\
+  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
   \cmidrule(lr){2-4}
@@
   \cmidrule(lr){1-4}
-  GADGET-4 & softening ratio & log-uniform & [1/50, 1/20) \\
+  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
   \cmidrule(lr){1-5}
@@
 
-We run 512 GADGET-4 simulations that cover a wide range of
+We run 512 \GADGET\ simulations that cover a wide range of
 configurations and cosmological parameters, as listed in
@@
 
-We configure GADGET-4 with high force and time integration accuracy
+We configure \GADGET\ with high force and time integration accuracy
 settings.
@@
 
-Given 512 Gadget simulations and 121 output snapshots each, in total we have
+Given 512 \GADGET\ simulations and 121 output snapshots each, in total we have
 61952 snapshots for training.
-Following \autoref{tab:param}, given a snapshot from Gadget simulations, we
+Following \autoref{tab:param}, given a snapshot from \GADGET\ simulations, we
 configure \pmwd\ with the same initial coniditions and cosmological parameters,

### 2023-06-06 docs/papers/sto/so.tex
FIXUP draft: edit and separating SO out of STO

--- 2023-05-31 docs/papers/sto/sto.tex
@@
 \bmdefine{\vzeta}{\zeta}  % force vjp theta gradient
+\newcommand{\lna}{\ln\!a}
+\newcommand{\lnk}{\ln\!k}
 \newcommand{\half}{\nicefrac12}
@@
 
-\title[Spatiotemporally Optimized Simulation]
-{Neural and Symbolic Spatiotemporal Optimization of Cosmological
-Particle-Mesh Simulation}
+\title[Optimized Cosmological Simulation]
+{Neural and Symbolic Optimization of Cosmological Particle-Mesh
+Simulation}
 
-\author[Zhang, Li, Jamieson, et al.]{
+\author[Zhang \& Li et al.]{
 %
@@
 \begin{abstract}
-We sharpen PM force and optimize integration time steps.
+We sharpen the PM force with first neural networks and then symbolic
+regression, taking symmetry and dimensional analysis in consideration.
 \end{abstract}
@@
 \begin{itemize}
-\item Separate sto to SO and TO.
-\item Moving some of the following ongoing and completed tasks to the
-    main text and add preliminary figures.
-\item Implement the rest of the nonlinear scales, including the rms
-    displacement with np.trapz of $P$ over $\ln k$, the $\ln a$ time
-    derivatives, etc.\newline
-    \textbf{Update}: code has been added.
-\item Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
+\item (paper) Moving some of the following ongoing and completed tasks
+  to the main text and add preliminary figures.
+\item (data; YL) Incorporate 3lpt branch; testing; find a\_start for
+  different mass resolutions; probably replace linear snapshot spacing
+  according to the nonlinearity (estimate (with PM till convergence) the
+  std acc over the std vel) of each realization.
+\item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
     steps) results and make sure that they agree perfectly on the large
@@
     slightly different. To figure out this weird issue.}
-\item Be careful with the non-deterministic ICs generated on GPUs.\newline
-      \textbf{Update}: we have shown that ICs generated on GPUs are
-      deterministic (but different from those on CPUs), generating ICs for both
-      training data and training on GPUs might be the simplest solution.
-\item Incorporate 3lpt branch; testing; find a\_start for different mass
-  resolutions
-\item Probably necessary to first incorporate List \& Hahn's recent
-  improvement on FastPM integrator arXiv:2301:09655.
-\item Compare sto DM and halo fields and summary stats with those of raw
-  pmwd, deconvolved pmwd, and \GADGET, on the last 9 test cases and
-  noncubic box.
-\item Demonstrate capability of generalization, on some cases with
-  extrapolating cosmological parameters, etc.
-\item Non-integer multiple mesh shapes: for uniform particle grid, test
-  (with double precision on CPUs) whether they start moving by artifact
-  forces on cell scales, in 3 cases: no deconvolution, with CIC
-  deconvolution, with SO.
-\item In TO, time step spacing can be important for light cone
-  generation. We can train on light cone target data to learn TO for
-  light cones. The result may be related to $c \d t/H = c \d a / a^2 H$.
+\item (model; YL) Probably necessary to first incorporate List \& Hahn's
+  recent improvement on FastPM integrator arXiv:2301:09655.
+\item (train, valid) Show $T - 1$ in symlog with 0.01 linthresh, and
+  $1-r^2$ in log at 3 different $a$'s; Show $f$ and $g$ (along axis and
+  face diagonal) at 3 different $a$'s; Track mean epoch loss
+\item (train) Add random offset in loss grid (see \autoref{tab:param})
+  to remove the position dependence in loss.
+\item (train) Randomize snapshots of each simulation instead of
+  sequential; alternatively accumulate their gradients
+\item (train) 3 sets of grid: ptcl, force mesh, and loss grid. Explore
+  if we can use a finer loss grid, from 1x to say 1.5x and even 2x.
+\item (train, valid) Non-integer multiple mesh shapes: for uniform
+  particle grid, test (with double precision on CPUs) whether they start
+  moving by artifact forces on cell scales, in 3 cases: no
+  deconvolution, with CIC deconvolution, with SO.
+\item (train) For generalizability and robustness, consider testing
+  L1/L2 and dropout. Also they're worth adding if they give lower losses
+  at convergence. Note that 512 realization are not too many, and we
+  have features that can be very correlated.
+\item (train) find the best loss function for production training
+\item (train) optuna hyperparam search on a subset of training data
+  before production training
+  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna example}).
+\item (train) Reduce learning rate on (mean epoch) loss plateau in
+  production
+\item (valid) Interpolation test: compare SO DM and halo fields and
+  summary stats with those of raw pmwd, deconvolved pmwd, and \GADGET,
+  on the last 9 test cases and noncubic box.
+\item (valid) Extrapolation test: demonstrate generalization capability,
+  compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
 \end{itemize}
@@
 \begin{equation}
-\cL_\vs = \ln \sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2,
+\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
+                   {\sum_{\vq'} s(\vq')^2},
 \end{equation}
@@
 % Say that only s loss is not enough even though in principle it is, and cite DrewEtAl x2 too
+
+\YL{Even though the log should be enough for adaptive optimizer and
+fixed sim config, this may create problem when training on all config at
+the same time. So normalizations inside the log is added above and
+below.}
 
@@
 \begin{equation}
-\cL_\delta = \frac1{N_K} \sum_i \,\ln\!\!
-\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2,
+\cL_\delta = \frac1{N_K} \sum_i \ln
+\frac{\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |\delta(\vk')|^2},
 \end{equation}
@@
 intermediate scales.
-As a result, we can set the loss mesh size the same as that of the PM
-force solver.
-We can try for $w$:
-%
-\begin{equation}
-\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{to some power},
-\prod_{j=1}^3 \cos (k_j l / 2) \textrm{to some power},
-\end{equation}
-%
-which vanish when any $k_j$ reaches the Nyquist.
-\YL{Maybe add a figure of $w$ along axis, face diagonal and body
-digonal, if this works.}
 
@@
 %
-up to a constant;
+replacing
 %
@@
 \begin{equation}
-\cJ = \cL_\vs + \cL_\delta
+\cJ = \cL_\vs + \cL_\vv + \cL_\delta
 + \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
@@
 \begin{equation}
-\cJ = \ln \cL_\vs + \ln \cL_\delta
-+ \ln \cL_{(1+\delta)\vv} + \ln \cL_{(1+\delta)\vv\vv}
+\cJ = \ln\cL_\vs + \ln\cL_\vv + \ln\cL_\delta
++ \ln\cL_{(1+\delta)\vv} + \ln\cL_{(1+\delta)\vv\vv}
 + \cdots
@@
 
-\subsection{Spatial Optimization}
+\subsection{Optimization of Dynamics}
 \label{sec:so}
+
+\begin{table}
+\centering
+\caption{Possible features that can affect force sharpening. TODO}
+\label{tab:feat}
+\begin{tabular}{rl}
+\toprule
+type & feature \\
+\midrule
+& $R_\mathrm{d}$: RMS linear displacement \\
+\cmidrule(lr){2-2}
+& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+& $\d R / \d\lna$: time derivatives of the above \\
+\midrule
+& particle spacing \\
+\cmidrule(lr){2-2}
+other scales & $l$: PM force mesh cell size \\
+\cmidrule(lr){2-2}
+& softening length \\
+\midrule
+& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
+\cmidrule(lr){2-2}
+& $\d\ln G_m / \d\lna$ \\
+\cmidrule(lr){2-2}
+dimensionless factors & $\Omegam(a)$ \\
+\cmidrule(lr){2-2}
+& $\d\ln\!H / \d\lna$ \\
+\cmidrule(lr){2-2}
+& $\Delta\lna$: time step size in $\lna$ \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-Various nonlinear scales are
-\begin{itemize}
-\item $k_P^{-1}$: determined by $\Delta_\linear^2(k_P) \triangleq 1$;
-\item $R_\tophat$: determined by $\sigma_\tophat^2(R_\tophat) \triangleq
-  1$;
-\item $R_\gauss$: determined by $\sigma_\gauss^2(R_\tophat) \triangleq
-  1$;
-\item $R_\mathrm{d}$: rms linear theory displacement;
-\item $\d R / \d\ln a$: time derivatives of the above scales.
-\end{itemize}
-
-Other relevant scales are
-\begin{itemize}
-\item $l$: PM cell size.
-\item particle spacing;
-\item softening length.
-\end{itemize}
+The RMS linear displacement
+%
+\begin{equation}
+R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
+\end{equation}
+%
 
@@
   suppression factors for $m \in \{1, 2\}$;
-\item $\d\ln G_m / \d\ln a$: with $m=1$ case related to the growth rate
-  $f \triangleq \d\ln D_1 / \d\ln a$;
+\item $\d\ln G_m / \d\lna$: with $m=1$ case related to the growth rate
+  $f \triangleq \d\ln D_1 / \d\lna$;
 \item $\Omegam(a)$: matter density parameter at $a$;
-\item $\d\ln H / \d\ln a$: time derivative of Hubble expansion;
-\item $\Delta\ln a$: time step size in $\ln a$.
+\item $\d\ln\!H / \d\lna$: time derivative of Hubble expansion;
+\item $\Delta\lna$: time step size in $\lna$.
 \end{itemize}
@@
 \frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
+  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
 \end{equation}
 %
-where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
-function of the dimensionless features, e.g.,
+where $f$, $g$ are neural networks, each approximating a nonlinear
+function of the wavevector components and all the features considered
+above in dimensionless combinations, e.g.,
 %
 \begin{equation}
-g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots; D, f, \cdots).
+f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
 \end{equation}
+%
+$f$ and $g$ are optimized during training to correct and sharpen the
+gravitational forces.
+Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
+broken by the PM solver to the O\textsubscript{h} discrete symmetry with
+48 elements involving permutations and reversals of the 3 principal axes
+of the simulation box.
+The PM evaluation of each force component can further break the symmetry
+along that axis, leaving the remaining 2 permutationally symmetric.
+Therefore, we implement $f$ and $g$ to be even functions of the
+wavevector components, and $g$ to be permutation equivariant in all 3
+components, so that $f g$ can be the most general way to modify the
+force kernel.
+An even and real kernel in configuration space is also even in Fourier
+space.
+We take the absolute value for the evenness, and sort the 3 wavevector
+compontents to achieve the permutation equivariance.
 
-Parameters in these neural networks are optimized during training.
-
-\YL{While $g$ is spherically symmetric by assumption, $f$ and $h$ are
-expected to be even functions, i.e., real valued in Fourier space.
-It is interesting to symbolically regress both $h(k_i)$ and $k_i h(k_i)$
-to see which one gives simpler expression, given the possible different
-rankings by complexity.
-For example, $k_i h(k_i)$ may have a simpler form as a nonlinear odd
+\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
+etc to see which one gives simpler expression, given the possible
+different rankings by complexity.
+For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
 function.}
@@
 mesh shape and mesh cell size below, which are assumed to be sampled
-independently.
+independently. \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
 In practice, we sample one conditioned on the other and the box size.
@@
 & number of time steps* & log-uniform & [10, 1000] \\
+\cmidrule(lr){1-4}
+training & loss grid offset in its own spacing & uniform & $[0, 1)^3$ \\
 \bottomrule
@@
 The curvature $\Omega_k$ is related to the separate universe simulation
-\citep{LiEtAl2014}.
+\citep{LiEtAl2014, WagnerEtAl2015}.
+
+We have verified IC generation on GPUs is deterministic though different
+from those on CPUs, so we generate ICs for both training data and
+training on GPUs.
 
@@
 The training on a given snapshot is performed on a single GPU device.
-The gradients are collected and averaged to update the STO parameters.
+The gradients are collected and averaged to update the SO parameters.
 To make sure that each individual training in a parallel run finishes in similar
@@
 
-We use the \texttt{PySR} \citep{PySR, Cranmer2020} symbolic regression code to
+We use the \texttt{PySR} \citep{Cranmer2023} symbolic regression code to
 fit for an analytic function expression of the SO neural network.

### 2023-06-07 docs/papers/sto/so.tex
FIXUP draft: edit TODOs and text

@@
   std acc over the std vel) of each realization.
+\item (model; YL) Change nbody for loop by lax scan and/or while loops.
+  while\_loop can help to save compilation when trained on different
+  snapshots. Performance difference between scan and while?
 \item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
@@
   recent improvement on FastPM integrator arXiv:2301:09655.
-\item (train, valid) Show $T - 1$ in symlog with 0.01 linthresh, and
-  $1-r^2$ in log at 3 different $a$'s; Show $f$ and $g$ (along axis and
-  face diagonal) at 3 different $a$'s; Track mean epoch loss
-\item (train) Add random offset in loss grid (see \autoref{tab:param})
-  to remove the position dependence in loss.
+\item (train, test) For $\vs(\vq)$, $\vv(\vq)$, $\delta(\vx)$,
+  $(1+\delta)\vv(\vx)$, $(1+\delta)\vv\vv(\vx)$, show $T - 1$ in symlog
+  with 0.01 linthresh, and $1-r^2$ in log at 3 different $a$'s; Show $f$
+  and $g$ (along axis and face diagonal) at 3 different $a$'s; Track
+  mean epoch loss
 \item (train) Randomize snapshots of each simulation instead of
-  sequential; alternatively accumulate their gradients
-\item (train) 3 sets of grid: ptcl, force mesh, and loss grid. Explore
-  if we can use a finer loss grid, from 1x to say 1.5x and even 2x.
-\item (train, valid) Non-integer multiple mesh shapes: for uniform
+  sequential; alternatively accumulate their gradients; Normalize loss
+  by number of time steps
+\item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add
+  random offset in loss grid (see \autoref{tab:param}) to remove the
+  position dependence in loss; likewise for force mesh? how about only
+  training (for lower loss), only ``inference'' (for higher accuracy),
+  or both? the random force mesh offset can also be replaced by a
+  structured scheme, e.g., alternating between 2 interlocking grids
+\item (train) Explore if we can use a finer loss grid, from 1x to say
+  1.5x and even 2x.
+\item (train, test) Non-integer multiple mesh shapes: for uniform
   particle grid, test (with double precision on CPUs) whether they start
@@
   deconvolution, with CIC deconvolution, with SO.
+\item (train) find the best loss function for production training
 \item (train) For generalizability and robustness, consider testing
-  L1/L2 and dropout. Also they're worth adding if they give lower losses
-  at convergence. Note that 512 realization are not too many, and we
-  have features that can be very correlated.
-\item (train) find the best loss function for production training
+  weight decay and dropout. Also they're worth adding if they give lower
+  losses at convergence. Note that 512 realization are not too many, and
+  we have features that can be very correlated.
 \item (train) optuna hyperparam search on a subset of training data
@@
   production
-\item (valid) Interpolation test: compare SO DM and halo fields and
+\item (test) Interpolation test: compare SO DM and halo fields and
   summary stats with those of raw pmwd, deconvolved pmwd, and \GADGET,
   on the last 9 test cases and noncubic box.
-\item (valid) Extrapolation test: demonstrate generalization capability,
+\item (test) Extrapolation test: demonstrate generalization capability,
   compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
@@
 
+\YL{
+For a field $f$ in either Lagrangian or Eulerian space, we have thought
+of 3 forms of MSE-inspired loss:
+%
+\begin{equation}
+\cL_f = \ln \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
+                   {\sum_{\vq'} f(\vq')^2},
+\end{equation}
+%
+(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
+their Fourier wavevector $\vk$ in both cases)
+%
+\begin{equation}
+\cL_f = \frac1{N_K} \sum_i \ln
+\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} | f(\vk')|^2},
+\end{equation}
+%
+and
+%
+\begin{equation}
+\cL_f = \ln \biggl[ \frac1{N_K} \sum_i
+\frac{\sum_{k \in K_i} w(\vk)
+      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
+\end{equation}
+%
+Then the question is which loss we should use for which $f$, to get
+stable and robust results, e.g., no bump for large or small boxes.
+To further limit our scope in this paper, we should probably only
+consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
+velocity $\vv(\vq)$.
+}
+
 For the displacement field, we adopt the mean squared error (MSE) loss
@@
 \begin{equation}
-\cJ = \cL_\vs + \cL_\vv + \cL_\delta
-+ \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
-+ \cdots
+\cJ = \frac1{N_a} (\cL_\vs + [\cL_\vv] + \cL_\delta),
 \end{equation}
 %
+where $N_a$ is the time steps it take to generate the snapshot for each
+iteration.
 \YL{(a simple sum since we have already done careful normalization in
@@
 \begin{equation}
-\cJ = \ln\cL_\vs + \ln\cL_\vv + \ln\cL_\delta
-+ \ln\cL_{(1+\delta)\vv} + \ln\cL_{(1+\delta)\vv\vv}
-+ \cdots
+\cJ = \frac1{N_a} (\ln\cL_\vs + [\ln\cL_\vv] + \ln\cL_\delta).
 \end{equation}
 %
+
+\YL{TODO move this above:}
 Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
-Lagrangian space, while the rest are in the Eulerian space:
+Lagrangian space, while the density field is in the Eulerian space:
 %
-\begin{align}
-1 + \delta(\vx) &= \int\!\d^3\vq\,
-\deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+\begin{equation}
+1 + \delta(\vx) = \int\!\d^3\vq\, \deltaD[ \vx - \vq - \vs(\vq)],
+\end{equation}
 %
-\bigl[(1+\delta)\vv\bigr]_i(\vx) &= \int\!\d^3\vq\,
-\deltaD[ \vx - \vq - \vs(\vq)] v_i, \nonumber\\
-%
-\bigl[(1+\delta)\vv\vv\bigr]_{ij}(\vx) &= \int\!\d^3\vq\,
-\deltaD[ \vx - \vq - \vs(\vq)] v_i v_j,
-\end{align}
-%
-which we compute by scattering particles and their velocities to the
-\YL{choice of mesh for loss}.
-
-An alternative loss function can be the Eulernian density weighted
-particle displacement MSE.
+which we compute by scattering particles to the choice of grid for loss.
 
@@
 wavevector components, and $g$ to be permutation equivariant in all 3
-components, so that $f g$ can be the most general way to modify the
-force kernel.
+components, so that $f g$ is the most general modification to the force
+kernel.
 An even and real kernel in configuration space is also even in Fourier
@@
 
-Given 512 \GADGET\ simulations and 121 output snapshots each, in total we have
-61952 snapshots for training.
-Following \autoref{tab:param}, given a snapshot from \GADGET\ simulations, we
-configure \pmwd\ with the same initial coniditions and cosmological parameters,
-and randomly sample the mesh shape (or the mesh cell size) and the number of
-time steps.
+Given 512 \GADGET\ simulations and 121 output snapshots each, in total
+we have 61952 snapshots for training.
+Following \autoref{tab:param}, in each iteration, we randomly sample the
+snapshots, the number of time steps, and the mesh shape (or its cell
+size), and configure \pmwd\ with the same initial coniditions,
+cosmological parameters and other common configurations.
+To balance the work of GPUs during training, in each iteration we use
+the same snapshot $a$, number of time steps, and mesh shape on all GPU.
 
-We perform parallel training using a number of nodes that each hosts multiple
-GPU devices.
-The training on a given snapshot is performed on a single GPU device.
+We perform parallel training using a number of nodes that each hosts
+multiple GPU devices.
+The training on each snapshot is performed on a single GPU device.
 The gradients are collected and averaged to update the SO parameters.
-To make sure that each individual training in a parallel run finishes in similar
-time, they share the same number of time steps.
+
+Hyperparameters:
 

### 2023-06-07 docs/papers/sto/sto.tex
FIXUP draft: loss norm caveat
FIXUP draft: adding loss norm inside log
FIXUP draft: edit TODOs and text

@@
 
-\author[Zhang, Li, Jamieson, et al.]{
+\author[Zhang \& Li et al.]{
 %
@@
 \begin{abstract}
-We sharpen PM force and optimize integration time steps.
+We optimize integration time steps with first neural networks and then
+symbolic regression, taking dimensional analysis in consideration.
 \end{abstract}
@@
 \begin{itemize}
-\item Separate sto to SO and TO.
-\item Moving some of the following ongoing and completed tasks to the
-    main text and add preliminary figures.
-\item Implement the rest of the nonlinear scales, including the rms
-    displacement with np.trapz of $P$ over $\ln k$, the $\ln a$ time
-    derivatives, etc.\newline
-    \textbf{Update}: code has been added.
-\item Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
-    steps) results and make sure that they agree perfectly on the large
-    scales. For this purpose the box size should be large enough (in
-    linear regime), say 1 Gpc/$h$.\newline
-    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
-    consistent at the field level. To further check power spectrum. For odd mesh
-    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
-    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
-    slightly different. To figure out this weird issue.}
-\item Be careful with the non-deterministic ICs generated on GPUs.\newline
-      \textbf{Update}: we have shown that ICs generated on GPUs are
-      deterministic (but different from those on CPUs), generating ICs for both
-      training data and training on GPUs might be the simplest solution.
-\item Incorporate 3lpt branch; testing; find a\_start for different mass
-  resolutions
-\item Probably necessary to first incorporate List \& Hahn's recent
-  improvement on FastPM integrator arXiv:2301:09655.
-\item Compare sto DM and halo fields and summary stats with those of raw
-  pmwd, deconvolved pmwd, and \GADGET, on the last 9 test cases and
-  noncubic box.
-\item Demonstrate capability of generalization, on some cases with
-  extrapolating cosmological parameters, etc.
-\item Non-integer multiple mesh shapes: for uniform particle grid, test
-  (with double precision on CPUs) whether they start moving by artifact
-  forces on cell scales, in 3 cases: no deconvolution, with CIC
-  deconvolution, with SO.
 \item In TO, time step spacing can be important for light cone
@@
 \begin{equation}
-\cJ = \cL_\vs + \cL_\delta
+\cJ = \frac1{N_a} (\cL_\vs + \cL_\vv + \cL_\delta
 + \cL_{(1+\delta)\vv} + \cL_{(1+\delta)\vv\vv}
-+ \cdots
++ \cdots),
 \end{equation}
 %
+where $N_a$ is the time steps it take to generate the snapshot for each
+iteration.
 \YL{(a simple sum since we have already done careful normalization in
@@
 \begin{equation}
-\cJ = \ln \cL_\vs + \ln \cL_\delta
-+ \ln \cL_{(1+\delta)\vv} + \ln \cL_{(1+\delta)\vv\vv}
-+ \cdots
+\cJ = \frac1{N_a} (\ln\cL_\vs + \ln\cL_\vv + \ln\cL_\delta
++ \ln\cL_{(1+\delta)\vv} + \ln\cL_{(1+\delta)\vv\vv}
++ \cdots).
 \end{equation}
@@
 which we compute by scattering particles and their velocities to the
-\YL{choice of mesh for loss}.
-
-An alternative loss function can be the Eulernian density weighted
-particle displacement MSE.
+choice of grid for loss.
 
@@
 \label{sec:so}
+
+\begin{table}
+\centering
+\caption{Possible features that can affect force sharpening. TODO}
+\label{tab:feat}
+\begin{tabular}{rl}
+\toprule
+type & feature \\
+\midrule
+& $R_\mathrm{d}$: RMS linear displacement \\
+\cmidrule(lr){2-2}
+& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+& $\d R / \d\lna$: time derivatives of the above \\
+\midrule
+& particle spacing \\
+\cmidrule(lr){2-2}
+other scales & $l$: PM force mesh cell size \\
+\cmidrule(lr){2-2}
+& softening length \\
+\midrule
+& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
+\cmidrule(lr){2-2}
+& $\d\ln G_m / \d\lna$ \\
+\cmidrule(lr){2-2}
+dimensionless factors & $\Omegam(a)$ \\
+\cmidrule(lr){2-2}
+& $\d\ln\!H / \d\lna$ \\
+\cmidrule(lr){2-2}
+& $\Delta\lna$: time step size in $\lna$ \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 \frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  h(k_i; \vvartheta) g(k; \vvartheta) \prod_{j=1}^3 f(k_j; \vvartheta),
+  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
 \end{equation}
 %
-where $f$, $g$, $h$ are neural networks, each approximating a nonlinear
-function of the dimensionless features, e.g.,
+where $f$, $g$ are neural networks, each approximating a nonlinear
+function of the wavevector components and all the features considered
+above in dimensionless combinations, e.g.,
 %
 \begin{equation}
-g(k; \vvartheta) = g(k/k_P, kR_\tophat, \cdots; D, f, \cdots).
+f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
 \end{equation}
+%
+$f$ and $g$ are optimized during training to correct and sharpen the
+gravitational forces.
+Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
+broken by the PM solver to the O\textsubscript{h} discrete symmetry with
+48 elements involving permutations and reversals of the 3 principal axes
+of the simulation box.
+The PM evaluation of each force component can further break the symmetry
+along that axis, leaving the remaining 2 permutationally symmetric.
+Therefore, we implement $f$ and $g$ to be even functions of the
+wavevector components, and $g$ to be permutation equivariant in all 3
+components, so that $f g$ is the most general modification to the force
+kernel.
+An even and real kernel in configuration space is also even in Fourier
+space.
+We take the absolute value for the evenness, and sort the 3 wavevector
+compontents to achieve the permutation equivariance.
 
-Parameters in these neural networks are optimized during training.
-
-\YL{While $g$ is spherically symmetric by assumption, $f$ and $h$ are
-expected to be even functions, i.e., real valued in Fourier space.
-It is interesting to symbolically regress both $h(k_i)$ and $k_i h(k_i)$
-to see which one gives simpler expression, given the possible different
-rankings by complexity.
-For example, $k_i h(k_i)$ may have a simpler form as a nonlinear odd
+\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
+etc to see which one gives simpler expression, given the possible
+different rankings by complexity.
+For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
 function.}
@@
 
-Given 512 \GADGET\ simulations and 121 output snapshots each, in total we have
-61952 snapshots for training.
-Following \autoref{tab:param}, given a snapshot from \GADGET\ simulations, we
-configure \pmwd\ with the same initial coniditions and cosmological parameters,
-and randomly sample the mesh shape (or the mesh cell size) and the number of
-time steps.
+Given 512 \GADGET\ simulations and 121 output snapshots each, in total
+we have 61952 snapshots for training.
+Following \autoref{tab:param}, in each iteration, we randomly sample the
+snapshots, the number of time steps, and the mesh shape (or its cell
+size), and configure \pmwd\ with the same initial coniditions,
+cosmological parameters and other common configurations.
+To balance the work of GPUs during training, in each iteration we use
+the same snapshot $a$, number of time steps, and mesh shape on all GPU.
 
-We perform parallel training using a number of nodes that each hosts multiple
-GPU devices.
-The training on a given snapshot is performed on a single GPU device.
-The gradients are collected and averaged to update the STO parameters.
-To make sure that each individual training in a parallel run finishes in similar
-time, they share the same number of time steps.
+We perform parallel training using a number of nodes that each hosts
+multiple GPU devices.
+The training on each snapshot is performed on a single GPU device.
+The gradients are collected and averaged to update the SO parameters.
+
+Hyperparameters:
 

### 2023-06-09 docs/papers/sto/so.tex
Draft: TODO update

@@
 \newcommand{\YL}[1]{\textcolor{Bittersweet}{#1}}
+\newcommand{\YZ}[1]{\textcolor{MidnightBlue}{#1}}
 
@@
   sequential; alternatively accumulate their gradients; Normalize loss
-  by number of time steps
+  by number of time steps\newline
+  \YZ{\textbf{Update}: Normalizing loss by number of time steps seems to halt
+  the training of late time snapshots, i.e. they stop improving.}
 \item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add

### 2023-06-16 docs/papers/sto/so.tex
FIXUP draft: abstract and loss
FIXUP SO draft: rm a paragraph on TO

@@
 regression, taking symmetry and dimensional analysis in consideration.
+
+We combine differentiable simulation, deep learning, [decision tree
+(feature selection)], and symbolic regression to extract the final kernel
+applicable to all other cosmological simulation code.
 \end{abstract}
@@
 \begin{equation}
-\cL_f = \ln \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
+\cL_f = \frac1{N_a} \ln
+  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
        {\sum_{\vq'} f(\vq')^2},
@@
 %
+where $N_a$ is the time steps it take to generate the snapshot for each
+iteration.
 (or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
@@
 \begin{equation}
-\cL_f = \frac1{N_K} \sum_i \ln
+\cL_f = \frac1{N_a N_K} \sum_i \ln
 \frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
@@
 \begin{equation}
-\cL_f = \ln \biggl[ \frac1{N_K} \sum_i
+\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
 \frac{\sum_{k \in K_i} w(\vk)
@@
 \begin{equation}
-\cJ = \frac1{N_a} (\cL_\vs + [\cL_\vv] + \cL_\delta),
-\end{equation}
-%
-where $N_a$ is the time steps it take to generate the snapshot for each
-iteration.
-\YL{(a simple sum since we have already done careful normalization in
-$\cL_f$'s) and also try}
-%
-\begin{equation}
-\cJ = \frac1{N_a} (\ln\cL_\vs + [\ln\cL_\vv] + \ln\cL_\delta).
+\cJ = \cL_\vs + [\cL_\vv] + \cL_\delta.
 \end{equation}
@@
 function.}
-
-\subsection{Temporal Optimization}
-\label{sec:to}
-
-We use a spline to model the function $a(t)$ that maps uniformly sampled $t$
-values to optimally distributed $a$ time steps.
-The spline is parameterized with a number of control points that are optimized
-during training.
-The gradients (i.e. cotangents) are back propagated from the array of
-integration time steps to the control points through a VJP transformation of
-$a(t)$.
-
-The time-dependence of the discrete symplectic integrator mainly enters through
-the drift and kick factors in \eqref{DFK}.
-Besides, the input features of the SO NNs introduced in \autoref{sec:so} could
-also depend on time.
-Therefore when we perform TO together with SO, the force evaluation in
-\eqref{DFK} which contains SO also becomes time-dependent and needs to be
-optimized.
 

### 2023-06-16 docs/papers/sto/so.tex
FIXUP draft: change to aastex template

@@
-\documentclass[usenatbib]{mnras}
+\documentclass[modern, trackchanges, dvipsnames]{aastex631}
+  \urlstyle{sf}
 
-% Allow "Thomas van Noord" and "Simon de Laguarde" and alike to be sorted by "N" and "L" etc. in the bibliography.
-% Write the name in the bibliography as "\VAN{Noord}{Van}{van} Noord, Thomas"
-\DeclareRobustCommand{\VAN}[3]{#2}
-\let\VANthebibliography\thebibliography
-\def\thebibliography{\DeclareRobustCommand{\VAN}[3]{##3}\VANthebibliography}
-
+\usepackage{CJK}
 \usepackage{microtype}
+\usepackage{savesym}
+  \savesymbol{lambdabar}
 \usepackage{newpxtext, eulerpx}
+  \restoresymbol{newpx}{lambdabar}
 \usepackage[T1]{fontenc}
+\usepackage{fontawesome}
 \usepackage{amsmath}
 \usepackage{bm}
-\usepackage{dsfont}
-\usepackage{graphicx}
+\usepackage{nicefrac}
+\usepackage[caption=false]{subfig}
+\usepackage[figure,figure*]{hypcap}
+%\usepackage{tikz}
+%  \usetikzlibrary{positioning, fit, calc, arrows.meta}
 \usepackage{booktabs}
-\usepackage{nicefrac}
-\usepackage[dvipsnames]{xcolor}
 
 \newcommand{\pmwd}{{\usefont{T1}{nova}{m}{sl}pmwd}}
-\newcommand{\GADGET}{{{\fontsize{7.7pt}{9pt}\selectfont GADGET}-4}}
+\newcommand{\GADGET}{{{\fontsize{10pt}{12pt}\selectfont GADGET}-4}}
 
-\renewcommand{\sectionautorefname}{Sec.}
-\renewcommand{\subsectionautorefname}{Sec.}
-\renewcommand{\appendixautorefname}{App.}
-\renewcommand{\figureautorefname}{Fig.}
+%\renewcommand{\sectionautorefname}{Sec.}
+%\renewcommand{\subsectionautorefname}{Sec.}
+%\renewcommand{\appendixautorefname}{App.}
+%\renewcommand{\figureautorefname}{Fig.}
 %\newcommand{\subfigureautorefname}{\figureautorefname}
@@
 
+\newcommand{\GPU}{NVIDIA H100 PCIe}  % 80GB memory
+
 \newcommand{\YL}[1]{\textcolor{Bittersweet}{#1}}
@@
 
-\title[Optimized Cosmological Simulation]
-{Neural and Symbolic Optimization of Cosmological Particle-Mesh
-Simulation}
+\begin{document}
 
-\author[Zhang \& Li et al.]{
-%
-Yucheng Zhang,$^{1, 2}$
-%
-Yin Li,$^{1, 3, 4}$\thanks{Email: XYZ, eelregit@gmail.com, and XYZ}
-%
-Drew Jamieson,$^{5}$
-%
-Libin Lu,$^{3}$
-%
-and XYZ
-%
-\\$^1$Department of Mathematics and Theory, Peng Cheng Laboratory,
-Shenzhen, Guangdong 518066, China
-%
-\\$^2$Center for Cosmology and Particle Physics, Department of Physics,
-New York University, New York, NY 10003, USA
-%
-\\$^3$Center for Computational Mathematics, Flatiron Institute, New
-York, NY 10010, USA
-%
-\\$^4$Center for Computational Astrophysics, Flatiron Institute, New
-York, NY 10010, USA
-%
-\\$^5$Max Planck Institute for Astrophysics, 85748 Garching bei
-M\"unchen, Germany
-}
+\title{\large Neural, Tree, and Symbolic Optimization of Cosmological
+Particle-Mesh Simulation
+\vspace{0.3em}}
 
-\date{Accepted XXX. Received YYY; in original form ZZZ}
-\pubyear{2023}
+\author[0000-0002-9300-2632]{\normalsize Yucheng Zhang}
+\affiliation{Department of Mathematics and Theory, Peng Cheng
+Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{Center for Cosmology and Particle Physics, New York
+University, New York, New York 10003, USA}
 
-\begin{document}
-\label{firstpage}
-\pagerange{\pageref{firstpage}--\pageref{lastpage}}
-\maketitle
+\author[0000-0002-0701-1410]{\normalsize Yin Li}
+\affiliation{Center for Computational Mathematics, Flatiron Institute,
+New York, New York 10010, USA}
+\affiliation{Department of Mathematics and Theory, Peng Cheng
+Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute,
+New York, New York 10010, USA}
+
+\author[0000-0001-5044-7204]{\normalsize Drew Jamieson}
+\affiliation{Max Planck Institute for Astrophysics, 85748 Garching bei
+M\"unchen, Germany}
+
+\author[0000-0003-0745-9431]{\normalsize Libin Lu}
+\affiliation{Center for Computational Mathematics, Flatiron Institute,
+New York, New York 10010, USA}
+
+\shorttitle{Optimized Cosmological Simulation}
+\shortauthors{Zhang \& Li et al.}
+
+\correspondingauthor{XYZ, Yin Li, and XYZ}
+\email{XYZ, eelregit@gmail.com, and XYZ}
 
 \begin{abstract}
+
 We sharpen the PM force with first neural networks and then symbolic
@@
 applicable to all other cosmological simulation code.
+
 \end{abstract}
-
-\begin{keywords}
-cosmology: large-scale structure of Universe
--- methods: numerical
--- software: development
-\end{keywords}
 
@@
   compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
+\item Fix cross references and other journal quirks as a patch
 \end{itemize}
 
+\vspace{1em}
 \section{Introduction}
 
+\vspace{1em}
+\section{Results}
+
+\vspace{1em}
+\subsection{SO function}
+
+We use the \texttt{PySR} \citep{Cranmer2023} symbolic regression code to
+fit for an analytic function expression of the SO neural network.
+
+\vspace{1em}
+\section{Discussion}
+
+\vspace{1em}
 \section{Methods}
 
+\vspace{1em}
 \subsection{Simulation and Differentiation}
@@
 
+\vspace{1em}
 \subsection{Snapshot Observable}
@@
 
+\vspace{1em}
 \subsection{Loss Function}
@@
 T(\vk) \triangleq
-\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk')|^2}},
+\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
 \end{equation}
@@
 \frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
-     {\sqrt{|\hat f(\vk)|^2 |f(\vk')|^2}}.
+     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
 \end{equation}
@@
 
+\vspace{1em}
 \subsection{Optimization of Dynamics}
@@
 
+\vspace{1em}
 \subsection{Parameters and Configurations}
@@
 
+\vspace{1em}
 \subsection{Simulations}
@@
 
+\vspace{1em}
 \subsection{Training}
@@
 
-\section{Results}
-
-\subsection{SO function}
-
-We use the \texttt{PySR} \citep{Cranmer2023} symbolic regression code to
-fit for an analytic function expression of the SO neural network.
-
-\section{Conclusions}
-
-\section*{Acknowledgements}
-
+\vspace{1em}
+\textit{\large Acknowledgements:}
 YZ and YL were supported by The Major Key Project of PCL.
@@
 
-\section*{Code Availability}
+\vspace{1em}
+\textit{\large Code Availability:}
+\pmwd\ \citep{Li2022b} is open-source on GitHub
+\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
+files and scripts of this paper
+\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
 
-\pmwd\ \citep{Li2022b} is open-source on
-\href{https://github.com/eelregit/pmwd}{GitHub}, including the
-\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{source
-and scripts of this paper}.
-
-\bibliographystyle{mnras}
+\bibliographystyle{aasjournal}
 \bibliography{sto}
 
-\bsp  % typesetting comment
-\label{lastpage}
+%\listofchanges
+
 \end{document}

### 2023-06-25 docs/papers/sto/so.tex
FIXUP draft methods: add SR subsection

@@
 \vspace{1em}
-\subsection{SO function}
-
-We use the \texttt{PySR} \citep{Cranmer2023} symbolic regression code to
-fit for an analytic function expression of the SO neural network.
-
-\vspace{1em}
 \section{Discussion}
@@
 \vspace{1em}
-\subsection{Simulation and Differentiation}
+\subsection{Differentiable Simulation}
 
@@
 
-Our observable $\vo$ is a snapshot $\hat\vz = (\hat\vx, \hat\vp)^\intercal$ at a
-given time $a_o$, which in general can be given by the interpolation of two
-snapshots $\vz_{k-1}$ and $\vz_k$ where $a_{k-1} \leq a_o < a_k$.
-Then ${\p\vo}/{\p\vz_i}$ would be zero except for the relevant $\vz_{k-1}$ and
-$\vz_k$.
-We have in the initial conditions of the adjoint equations
+We refer to the quantities that are directly related to the objective function
+as the observables. Our observable $\vo$ is a snapshot $\hat\vz = (\hat\vx,
+\hat\vp)^\intercal$ at a given time $a_o$, which in general can be given by the
+interpolation of two snapshots $\vz_{k-1}$ and $\vz_k$ where $a_{k-1} \leq a_o <
+a_k$. Then ${\p\vo}/{\p\vz_i}$ would be zero except for the relevant $\vz_{k-1}$
+and $\vz_k$. We have in the initial conditions of the adjoint equations
 \begin{align}
@@
 remain zero until the time stepping from $a_{k+1}$ to $a_k$.
+
+\vspace{1em}
+\subsection{Spatial Optimization of Dynamics}
+\label{sec:so}
+
+\begin{table}
+\centering
+\caption{Possible features that can affect force sharpening. TODO}
+\label{tab:feat}
+\begin{tabular}{rl}
+\toprule
+type & feature \\
+\midrule
+& $R_\mathrm{d}$: RMS linear displacement \\
+\cmidrule(lr){2-2}
+& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
+\cmidrule(lr){2-2}
+& $\d R / \d\lna$: time derivatives of the above \\
+\midrule
+& particle spacing \\
+\cmidrule(lr){2-2}
+other scales & $l$: PM force mesh cell size \\
+\cmidrule(lr){2-2}
+& softening length \\
+\midrule
+& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
+\cmidrule(lr){2-2}
+& $\d\ln G_m / \d\lna$ \\
+\cmidrule(lr){2-2}
+dimensionless factors & $\Omegam(a)$ \\
+\cmidrule(lr){2-2}
+& $\d\ln\!H / \d\lna$ \\
+\cmidrule(lr){2-2}
+& $\Delta\lna$: time step size in $\lna$ \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+The dimensionless power spectrum
+%
+\begin{equation}
+\Delta^2(k) \triangleq \frac{k^3 P(k)}{2 \pi^2}.
+\end{equation}
+
+Variance of linear matter overdensity in a top-hat window
+%
+\begin{equation}
+\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
+  \Delta_\linear^2(k) W_\tophat^2(kR),
+\end{equation}
+where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
+Fourier transform of the top-hat window.
+Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
+$W_\gauss(kR) = e^{-(kR)^2/2}$.
+
+The RMS linear displacement
+%
+\begin{equation}
+R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
+\end{equation}
+%
+
+Relevant scale-independent dimensionless factors are
+\begin{itemize}
+\item $G_m(a) \triangleq D_m(a) / a^m$: linear growth given in
+  suppression factors for $m \in \{1, 2\}$;
+\item $\d\ln G_m / \d\lna$: with $m=1$ case related to the growth rate
+  $f \triangleq \d\ln D_1 / \d\lna$;
+\item $\Omegam(a)$: matter density parameter at $a$;
+\item $\d\ln\!H / \d\lna$: time derivative of Hubble expansion;
+\item $\Delta\lna$: time step size in $\lna$.
+\end{itemize}
+
+We combine the above list of parameters in $\vvartheta$.
+
+In Fourier space, the gravitational force field reads
+\begin{equation}
+- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\end{equation}
+Force sharpening
+%
+\begin{equation}
+\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
+  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
+\end{equation}
+%
+where $f$, $g$ are neural networks, each approximating a nonlinear
+function of the wavevector components and all the features considered
+above in dimensionless combinations, e.g.,
+%
+\begin{equation}
+f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
+\end{equation}
+%
+$f$ and $g$ are optimized during training to correct and sharpen the
+gravitational forces.
+Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
+broken by the PM solver to the O\textsubscript{h} discrete symmetry with
+48 elements involving permutations and reversals of the 3 principal axes
+of the simulation box.
+The PM evaluation of each force component can further break the symmetry
+along that axis, leaving the remaining 2 permutationally symmetric.
+Therefore, we implement $f$ and $g$ to be even functions of the
+wavevector components, and $g$ to be permutation equivariant in all 3
+components, so that $f g$ is the most general modification to the force
+kernel.
+An even and real kernel in configuration space is also even in Fourier
+space.
+We take the absolute value for the evenness, and sort the 3 wavevector
+compontents to achieve the permutation equivariance.
 
@@
 
-\vspace{1em}
-\subsection{Optimization of Dynamics}
-\label{sec:so}
-
-\begin{table}
-\centering
-\caption{Possible features that can affect force sharpening. TODO}
-\label{tab:feat}
-\begin{tabular}{rl}
-\toprule
-type & feature \\
-\midrule
-& $R_\mathrm{d}$: RMS linear displacement \\
-\cmidrule(lr){2-2}
-& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
-\cmidrule(lr){2-2}
-nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
-\cmidrule(lr){2-2}
-& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
-\cmidrule(lr){2-2}
-& $\d R / \d\lna$: time derivatives of the above \\
-\midrule
-& particle spacing \\
-\cmidrule(lr){2-2}
-other scales & $l$: PM force mesh cell size \\
-\cmidrule(lr){2-2}
-& softening length \\
-\midrule
-& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
-\cmidrule(lr){2-2}
-& $\d\ln G_m / \d\lna$ \\
-\cmidrule(lr){2-2}
-dimensionless factors & $\Omegam(a)$ \\
-\cmidrule(lr){2-2}
-& $\d\ln\!H / \d\lna$ \\
-\cmidrule(lr){2-2}
-& $\Delta\lna$: time step size in $\lna$ \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-The dimensionless power spectrum
-%
-\begin{equation}
-\Delta^2(k) \triangleq \frac{k^3 P(k)}{2 \pi^2}.
-\end{equation}
-
-Variance of linear matter overdensity in a top-hat window
-%
-\begin{equation}
-\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
-  \Delta_\linear^2(k) W_\tophat^2(kR),
-\end{equation}
-where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
-Fourier transform of the top-hat window.
-Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
-$W_\gauss(kR) = e^{-(kR)^2/2}$.
-
-The RMS linear displacement
-%
-\begin{equation}
-R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
-\end{equation}
-%
-
-Relevant scale-independent dimensionless factors are
-\begin{itemize}
-\item $G_m(a) \triangleq D_m(a) / a^m$: linear growth given in
-  suppression factors for $m \in \{1, 2\}$;
-\item $\d\ln G_m / \d\lna$: with $m=1$ case related to the growth rate
-  $f \triangleq \d\ln D_1 / \d\lna$;
-\item $\Omegam(a)$: matter density parameter at $a$;
-\item $\d\ln\!H / \d\lna$: time derivative of Hubble expansion;
-\item $\Delta\lna$: time step size in $\lna$.
-\end{itemize}
-
-We combine the above list of parameters in $\vvartheta$.
-
-In Fourier space, the gravitational force field reads
-\begin{equation}
-- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
-\end{equation}
-Force sharpening
-%
-\begin{equation}
-\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
-\end{equation}
-%
-where $f$, $g$ are neural networks, each approximating a nonlinear
-function of the wavevector components and all the features considered
-above in dimensionless combinations, e.g.,
-%
-\begin{equation}
-f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
-\end{equation}
-%
-$f$ and $g$ are optimized during training to correct and sharpen the
-gravitational forces.
-Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
-broken by the PM solver to the O\textsubscript{h} discrete symmetry with
-48 elements involving permutations and reversals of the 3 principal axes
-of the simulation box.
-The PM evaluation of each force component can further break the symmetry
-along that axis, leaving the remaining 2 permutationally symmetric.
-Therefore, we implement $f$ and $g$ to be even functions of the
-wavevector components, and $g$ to be permutation equivariant in all 3
-components, so that $f g$ is the most general modification to the force
-kernel.
-An even and real kernel in configuration space is also even in Fourier
-space.
-We take the absolute value for the evenness, and sort the 3 wavevector
-compontents to achieve the permutation equivariance.
-
-\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
-etc to see which one gives simpler expression, given the possible
-different rankings by complexity.
-For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
-function.}
-
 \begin{table*}
@@
 \vspace{1em}
-\subsection{Simulations}
+\subsection{Target Data for Training and Testing}
 
@@
 \vspace{1em}
-\subsection{Training}
+\subsection{Training of SO neural networks}
 
@@
 Hyperparameters:
+
+\vspace{1em}
+\subsection{Symbolic Regression}
+
+We use the \texttt{PySR} \citep{Cranmer2023} symbolic regression code to fit for
+analytic function expressions of the trained SO neural networks.
+
+\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
+etc to see which one gives simpler expression, given the possible
+different rankings by complexity.
+For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
+function.}
 

### 2023-06-25 docs/papers/sto/so.tex
FIXUP draft: add some text on training

@@
 
+The differentiable cosmological simulation with adjoint method \citep{Li2022a}
+and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
+foundation of the optimization discussed in this work.
 We briefly recap the discussion in \citet{Li2022a} and introduce the notations.
@@
 The forward simulation is evolved using a symplectic integrator.
-From $a_{i-1}$ to $a_i$, the particles' state is updated alternately for $\vx$
-and $\vp$ through a loop over the symplectic integration indexed with $j$
+From time step $a_{i-1}$ to $a_i$, the particles' state is updated alternately
+for $\vx$ and $\vp$ through a loop over the symplectic integration indexed with
+$j$
 \begin{alignat}{2}
@@
 % where $\vtheta$ could include cosmological parameters, initial conditions, as
-% well as STO parameters in this work.
+% well as neural network parameters for optimization in this work.
 
@@
 We refer to the quantities that are directly related to the objective function
-as the observables. Our observable $\vo$ is a snapshot $\hat\vz = (\hat\vx,
+as the observables.
+Our observable $\vo$ is a snapshot $\hat\vz = (\hat\vx,
 \hat\vp)^\intercal$ at a given time $a_o$, which in general can be given by the
 interpolation of two snapshots $\vz_{k-1}$ and $\vz_k$ where $a_{k-1} \leq a_o <
-a_k$. Then ${\p\vo}/{\p\vz_i}$ would be zero except for the relevant $\vz_{k-1}$
+a_k$.
+Then ${\p\vo}/{\p\vz_i}$ would be zero except for the relevant $\vz_{k-1}$
 and $\vz_k$. We have in the initial conditions of the adjoint equations
@@
 \vspace{1em}
-\subsection{Spatial Optimization of Dynamics}
+\subsection{Optimization of Dynamics}
 \label{sec:so}
@@
 compontents to achieve the permutation equivariance.
+
+\YZ{
+The parameters of the neural networks, which enter every force evaluation in
+every time integration step, can be treated just the same as other cosmological
+parameters.
+The objective gradients are accumulated, based on which the parameters are then
+optimized.
+}
 
@@
 
-We perform parallel training using a number of nodes that each hosts
-multiple GPU devices.
+We perform parallel training using distributed GPU devices.
 The training on each snapshot is performed on a single GPU device.
@@
 Hyperparameters:
+
+\YZ{
+We implement the neural networks with \texttt{Flax} \citep{flax2020github}, and
+use the optimizers in \texttt{Optax}, part of the DeepMind JAX Ecosystem
+\citep{deepmind2020jax}.
+}
 

### 2023-06-25 docs/papers/sto/so.tex
FIXUP draft: clean up the recap on adjoint

@@
 foundation of the optimization discussed in this work.
-We briefly recap the discussion in \citet{Li2022a} and introduce the notations.
 We denote the particles' state with $\vz = (\vx, \vp)^\intercal$ and their
 adjoint variables with $\vlambda = (\vxi, \vpi)$.
-The forward simulation is evolved using a symplectic integrator.
-From time step $a_{i-1}$ to $a_i$, the particles' state is updated alternately
-for $\vx$ and $\vp$ through a loop over the symplectic integration indexed with
-$j$
-\begin{alignat}{2}
-&D_{j-1}^j: &\qquad \vx_j &= \vx_{j-1} +
-  \vp_{j-1} D(a^p_{j-1}, a^x_{j-1}, a^x_j), \nonumber\\
-%
-&F_j: & \va_j &\triangleq -\vnabla\varphi(\vx_j), \nonumber\\
-%
-&K_{j-1}^j: & \vp_j &= \vp_{j-1} + \va_j K(a^x_j, a^p_{j-1}, a^p_j).
-\label{DFK}
-\end{alignat}
-The intermediate times are given by
-\begin{align}
-a^x_j &= (1 - C^x_j) a_{i-1} + C^x_j a_i, \nonumber\\
-a^p_j &= (1 - C^p_j) a_{i-1} + C^p_j a_i,
-\end{align}
-where $C^x$ and $C^p$ are the cumulative coefficients of the symplectic
-integrator, e.g. $C^x = (0, 0, 1)$ and $C^p = (0, 0.5, 1)$ for leapfrog.
-The reverse time integration from $a_i$ to $a_{i-1}$ and adjoint equations
-\begin{alignat}{2}
-&K_j^{j-1}: &\qquad \vp_{j-1} &= \vp_j +
-  \va_j K(a^x_j, a^p_j, a^p_{j-1}), \nonumber\\
-%
-& & \vxi_{j-1} &= \vxi_j - \valpha_j K(a^x_j, a^p_j, a^p_{j-1}) \nonumber\\
-%
-&D_j^{j-1}: & \vx_{j-1} &= \vx_j +
-  \vp_{j-1} D(a^p_{j-1}, a^x_j, a^x_{j-1}), \nonumber\\
-%
-& & \vpi_{j-1} &= \vpi_j - \vxi_{j-1} D(a^p_{j-1}, a^x_j, a^x_{j-1}), \nonumber\\
-%
-&F_{j-1}: & \va_{j-1} &\triangleq -\vnabla\vvarphi(\vx_{j-1}), \nonumber\\
-%
-& & \valpha_{j-1} &\triangleq -\vpi_{j-1} \cdot
-  \frac{\p\vnabla\vvarphi_{j-1}}{\p\vx_{j-1}}, \nonumber\\
-& & \vzeta_{j-1} &\triangleq - \vpi_{j-1} \cdot
-  \frac{\p\vnabla\vvarphi_{j-1}}{\p\vtheta}.
-\label{KDF_adj}
-\end{alignat}
-
-We consider an objective function $\cJ(\vo, \theta)$ defined on the observable
-$\vo(\vz_0,...,\vz_n,\vtheta)$.
-The observe operator from $a_i$ to $a_{i-1}$ reads
-\begin{align}
-  O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
-  + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vx_{i-1}}, \nonumber\\
-                    \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
-  + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vp_{i-1}}.
-\end{align}
-
-% % TODO update the final gradient for loop over j
-% The final gradient is accumulated through
-% \begin{multline}
-%   \frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
-%   %
-%   + \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vtheta}
-%   %
-%   + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
-%   + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
-%   %
-%   - \sum_{i=1}^n \Bigl[
-%     \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
-%       \bigr) K(t_i, t_i, t_{i-\half}) \\
-%     + \vxi_i \cdot \vp_{i-\half}
-%       \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta}
-%     \Bigr],
-% \label{KDFO_grad}
-% \end{multline}
-% where $\vtheta$ could include cosmological parameters, initial conditions, as
-% well as neural network parameters for optimization in this work.
 

### 2023-06-25 docs/papers/sto/so.tex
FIXUP draft: minor reorg of SO subsection

@@
 
+In Fourier space, the gravitational force field reads
+\begin{equation}
+- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\end{equation}
+Force sharpening
+%
+\begin{equation}
+\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
+  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
+\end{equation}
+%
+where $f$, $g$ are neural networks, each approximating a nonlinear
+function of the wavevector components and all possible relevant features
+$\vvartheta$ summarized in \autoref{tab:feat} and discussed below.
+
 \begin{table}
@@
 
-In Fourier space, the gravitational force field reads
-\begin{equation}
-- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
-\end{equation}
-Force sharpening
-%
-\begin{equation}
-\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
-\end{equation}
-%
-where $f$, $g$ are neural networks, each approximating a nonlinear
-function of the wavevector components and all the features considered
-above in dimensionless combinations, e.g.,
+Dimensionless combinations, e.g.,
 %

### 2023-06-26 docs/papers/sto/so.tex
FIXUP draft a few sentences on so training

@@
 \begin{equation}
-f(k_i; \vvartheta) = f(k_i/k_P, kR_\tophat, \cdots; D, f, \cdots).
+f(k_i; \vvartheta) = f(k_iR_d, k_iR_\tophat, \cdots; G_m, \Omegam, \cdots)
+\end{equation}
+%
+\begin{equation}
+g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_\tophat, \cdots\}_{j=1,2,3}; G_m, \Omegam, \cdots)
 \end{equation}
@@
 optimized.
+For both $f$ and $g$ neural networks, we adopt the multilayer perceptron (MLP)
+structure, each with one input layer, two hidden layers and one output layer.
 }
@@
 The training on each snapshot is performed on a single GPU device.
-The gradients are collected and averaged to update the SO parameters.
+During each training step, the gradients are collected and averaged across
+devices to update the SO parameters.
 

### 2023-06-26 pmwd/sto/sr.py
Add a func for feature importance with rand forest

@@ -1,4 +1,7 @@
 """Some util functions related to the Symbolic Regression."""
+import numpy as np
+from pysr import PySRRegressor
+from sklearn.ensemble import RandomForestRegressor
 
 
 def gen_data_sr():
@@ -7,3 +10,13 @@ def gen_data_sr():
     X, y = None, None
     X_names = None
     return X, y, X_names
+
+
+def get_feature_importance(X, y, n_estimators=100, max_depth=3, random_state=None):
+    """Get the feature importance and indices, from high to low."""
+    clf = RandomForestRegressor(n_estimators=n_estimators, max_depth=max_depth,
+                                random_state=random_state)
+    clf.fit(X, y)
+    idx = np.argsort(clf.feature_importances_)[::-1]  # high to low
+
+    return clf.feature_importances_[idx], idx

### 2023-06-27 pmwd/gravity.py
FIXUP refactor so functions

@@ -60,14 +60,14 @@ def gravity(a, ptcl, cosmo, conf):
 
     if conf.so_type is not None:  # spatial optimization
         theta = sotheta(cosmo, conf, a)
-        pot = pot_sharp(kvec, theta, pot, cosmo, conf, a)
+        pot = pot_sharp(pot, kvec, theta, cosmo, conf, a)
 
     acc = []
     for k in kvec:
         grad = neg_grad(k, pot, conf.cell_size)
 
         if conf.so_type is not None:  # spatial optimization
-            grad = grad_sharp(k, theta, grad, cosmo, conf, a)
+            grad = grad_sharp(grad, k, theta, cosmo, conf, a)
 
         grad = fftinv(grad, shape=conf.mesh_shape)
         grad = grad.astype(conf.float_dtype)  # no jnp.complex32

### 2023-06-27 pmwd/sto/so.py

@@ -86,29 +86,7 @@ def soft(k, theta):
     """SO features for neural nets input, with k being a scalar."""
     theta_l, theta_o = theta
     k_theta_l = k * theta_l
-    # k_theta_l = jnp.log(jnp.where(k_theta_l != 0., jnp.abs(k_theta_l), 1.))
     return jnp.concatenate((k_theta_l, theta_o))
-
-
-def soft_bc(k, theta):
-    """SO features for neural nets input, broadcast with k being an array."""
-    # multiply each element of k with theta_l, and append theta_o
-    theta_l, theta_o = theta
-    k_shape = k.shape
-    k = k.reshape(k_shape + (1,))
-    theta_l = theta_l.reshape((1,) * len(k_shape) + theta_l.shape)
-    ft = k * theta_l
-    # ft = jnp.log(jnp.where(ft != 0., jnp.abs(ft), 1.))
-    ft = jnp.concatenate((ft, jnp.broadcast_to(theta_o, k_shape+theta_o.shape)),
-                         axis=-1)
-    return ft
-
-
-def sonn_bc(k, theta, cosmo, conf, nid):
-    """Evaluate the neural net, broadcast with k being processed in one pass."""
-    ft = soft_bc(k, theta)
-    net = MLP(features=conf.so_nodes[nid])
-    return net.apply(cosmo.so_params[nid], ft)[..., 0]  # rm the trailing axis of dim one
 
 
 def sonn_vmap(k, theta, cosmo, conf, nid):
@@ -120,32 +98,45 @@ def sonn_vmap(k, theta, cosmo, conf, nid):
     return vmap(_sonn)(k.ravel()).reshape(k.shape)
 
 
-def sonn_g(kv, theta, cosmo, conf, nid):
-    # kv shape: e.g. (128, 128, 65, 3)
-    kv_shape = kv.shape
-    kv = kv.reshape(kv_shape + (1,))
-
-    theta_l, theta_o = theta
-    theta_l = theta_l.reshape((1,) * len(kv_shape) + theta_l.shape)
-    ft = (kv * theta_l).reshape(kv_shape[:-1] + (-1,))
-    ft = jnp.concatenate((ft, jnp.broadcast_to(theta_o, kv_shape[:-1]+theta_o.shape)),
-                         axis=-1)
+def sonn_k(k, theta, cosmo, conf, nid):
+    """SO net with input: (k * l, o)."""
[…]

### 2023-06-27 docs/papers/sto/so.tex
FIXUP draft: todo & sentences on feature selection

@@
   compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
+\item (test) Generate a sample of $(X, y)$ data with the trained neural network
+  for feature selection and symbolic regression.
 \item Fix cross references and other journal quirks as a patch
@@
 
-We use the \texttt{PySR} \citep{Cranmer2023} symbolic regression code to fit for
-analytic function expressions of the trained SO neural networks.
+We generate a sample of (input, output) data with the trained neural network to
+perform feature selection and SR.
+
+We use the random forest regressor for feature selection using impurity-based
+importance.
+\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
+importance?}
+We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
+% more details on the hyperparams in random forest
+
+The selected important features are then used for SR.
+
+We use the \texttt{PySR} \citep{Cranmer2023} SR code to fit for analytic
+function expressions of the trained SO neural networks.
 

### 2023-06-28 pmwd/sto/vis.py
FIXUP log for cc & misusage of norm

@@ -25,12 +25,13 @@ def plt_tf(k, tf, ylim=(-1.1, 1.1)):
     return fig
 
 
-def plt_cc(k, cc, ylim=(-0.011, 1.1)):
+def plt_cc(k, cc, ylim=(1e-4, 1.1)):
     """Plot the correlation coefficient."""
     fig, ax = plt.subplots(1, 1, figsize=(4.8, 3.6), tight_layout=True)
     ax.plot(k, 1 - cc**2)
     ax.set_ylabel(r'$1-r^2$')
-    ax.set_yscale('symlog', linthresh=0.01)
+    # ax.set_yscale('symlog', linthresh=0.01)
+    ax.set_yscale('log')
     ax.set_ylim(*ylim)
     ax.set_xlabel(r'$k$')
     ax.set_xscale('log')
@@ -97,8 +98,8 @@ def vis_inspect(tgt, so_params, pmwd_params, vis_mesh_shape=1,
 
     # plot the density slab
     if p_slab:
-        norm = CosmicWebNorm(dens_t)
         slab = int(dens.shape[0] * 0.2)
+        norm = CosmicWebNorm(dens_t[:slab].mean(axis=0))
         figs['dens_target'] = simshow(dens_t[:slab].mean(axis=0), norm=norm)[0]
         figs['dens_target'].tight_layout()
         figs['dens'] = simshow(dens[:slab].mean(axis=0), norm=norm)[0]

### 2023-07-13 docs/examples/sto/train/run_train.py
Add support for dropout

@@ -29,7 +29,7 @@ from pmwd.sto.data import G4sobolDataset
 from pmwd.sto.hypars import (
     n_epochs, sobol_ids_global, snap_ids, shuffle_epoch,
     learning_rate, get_optimizer, lr_scheduler,
-    so_type, so_nodes, so_params)
+    so_type, so_nodes, so_params, dropout_rate)
 
 
 def _printinfo(s, flush=False):
@@ -48,6 +48,7 @@ def _checkpoint(epoch, so_params):
 # RNGs with fixed seeds, for same randomness across processes
 np_rng = np.random.default_rng(0)  # for pmwd MC sampling
 tc_rng = torch.Generator().manual_seed(0)  # for dataloader shuffle
+jax_key = jax.random.PRNGKey(0)  # for dropout
 
 # the corresponding sobol ids of training data for current proc
 # each proc must have the same number of sobol ids
@@ -88,9 +89,12 @@ for epoch in range(1, n_epochs+1):
         # n_steps = np.rint(10**np_rng.uniform(1, 3)).astype(int)
         n_steps = 100
 
+        jax_key, dropout_key = jax.random.split(jax_key)
+
         tgts = g4sobol['snapshots']
         pmwd_params = (g4sobol['a_snaps'], g4sobol['sidx'], g4sobol['sobol'],
-                       mesh_shape, n_steps, so_type, so_nodes)
+                       mesh_shape, n_steps, so_type, so_nodes, dropout_rate,
+                       dropout_key)
         opt_params = (optimizer, opt_state)
         so_params, loss, opt_state = train_step(tgts, so_params, pmwd_params,
                                                 opt_params)

### 2023-07-13 pmwd/configuration.py

@@ -158,6 +158,8 @@ class Configuration:
     # list of the number of nodes (no input layer) of so nn
     so_nodes: Optional[list] = None
     softening_length: Optional[float] = None
+    dropout_rate: Optional[float] = None
+    dropout_key: Optional[list] = None
 
     def __post_init__(self):
         if self._is_transforming():

### 2023-07-13 pmwd/sto/mlp.py

@@ -10,18 +10,18 @@ class MLP(nn.Module):
     activator: Callable[[jnp.ndarray], jnp.ndarray] = nn.softplus
     outivator: Callable[[jnp.ndarray], jnp.ndarray] = None
 
-    def setup(self):
-        self.layers = [nn.Dense(f, param_dtype=jnp.float64) for f in self.features]
+    @nn.compact
+    def __call__(self, x, dropout=False, dropout_rate=0.5):
+        # hidden layers
+        for i, fts in enumerate(self.features[:-1]):
+            x = nn.Dense(fts, param_dtype=jnp.float64)(x)
+            x = self.activator(x)
+            x = nn.Dropout(rate=dropout_rate, deterministic=not dropout)(x)
 
-    def __call__(self, inputs):
-        x = inputs
-        for i, lyr in enumerate(self.layers):
-            x = lyr(x)
-            if i != len(self.layers)-1:
-                x = self.activator(x)
-            else:
-                if self.outivator is not None:
-                    x = self.outivator(x)
+        # output layer
+        x = nn.Dense(self.features[-1], param_dtype=jnp.float64)(x)
+        if self.outivator is not None:
+            x = self.outivator(x)
 
         return x
 
@@ -39,9 +39,9 @@ def init_mlp_params(n_input, nodes, scheme=None):
     if scheme == 'last_w0_b1':
         for i, p in enumerate(params):
             p = unfreeze(p)
-            p['params'][f'layers_{len(nodes[i])-1}']['kernel'] = (
+            p['params'][f'Dense_{len(nodes[i])-1}']['kernel'] = (
                 jnp.zeros((nodes[i][-2], nodes[i][-1])))
-            p['params'][f'layers_{len(nodes[i])-1}']['bias'] = (
+            p['params'][f'Dense_{len(nodes[i])-1}']['bias'] = (
                 jnp.ones(nodes[i][-1]))
             params[i] = freeze(p)
 
@@ -50,9 +50,9 @@ def init_mlp_params(n_input, nodes, scheme=None):
         keys = random.split(random.PRNGKey(1), len(params))
         for i, p in enumerate(params):
             p = unfreeze(p)
-            p['params'][f'layers_{len(nodes[i])-1}']['kernel'] = (
+            p['params'][f'Dense_{len(nodes[i])-1}']['kernel'] = (
[…]

### 2023-07-13 pmwd/sto/so.py

@@ -89,12 +89,23 @@ def soft(k, theta):
     return jnp.concatenate((k_theta_l, theta_o))
 
 
+def apply_net(nid, conf, cosmo, x):
+    net = MLP(features=conf.so_nodes[nid])
+    if conf.dropout_rate is not None:
+        dropout = True
+        rngs = {'dropout': jnp.asarray(conf.dropout_key, dtype=jnp.uint32)}
+    else:
+        dropout = False
+        rngs = None
+    return net.apply(cosmo.so_params[nid], x, dropout=dropout,
+                     dropout_rate=conf.dropout_rate, rngs=rngs)
+
+
 def sonn_vmap(k, theta, cosmo, conf, nid):
     """Evaluate the neural net, using vmap over k."""
-    net = MLP(features=conf.so_nodes[nid])
     def _sonn(_k):
         _ft = soft(_k, theta)
-        return net.apply(cosmo.so_params[nid], _ft)[0]
+        return apply_net(nid, conf, cosmo, _ft)[0]
     return vmap(_sonn)(k.ravel()).reshape(k.shape)
 
 
@@ -107,8 +118,7 @@ def sonn_k(k, theta, cosmo, conf, nid):
     ft = k * theta_l  # (128, 1, 1, 8)
     theta_o = jnp.broadcast_to(theta_o, k_shape+theta_o.shape)  # (128, 1, 1, 6)
     ft = jnp.concatenate((ft, theta_o), axis=-1)  # (128, 1, 1, 8+6)
-    net = MLP(features=conf.so_nodes[nid])
-    return net.apply(cosmo.so_params[nid], ft)[..., 0]  # rm the trailing axis of dim one
+    return apply_net(nid, conf, cosmo, ft)[..., 0]  # rm the trailing axis of dim one
 
 
 def sonn_kvec(kv, theta, cosmo, conf, nid):
@@ -122,8 +132,7 @@ def sonn_kvec(kv, theta, cosmo, conf, nid):
     ft = ft.reshape(kv_shape[:-1] + (-1,))  # (128, 128, 65, 3*8)
     theta_o = jnp.broadcast_to(theta_o, kv_shape[:-1]+theta_o.shape)  # (128, 128, 65, 6)
     ft = jnp.concatenate((ft, theta_o), axis=-1)  # (128, 128, 65, 3*8+6)
-    net = MLP(features=conf.so_nodes[nid])
[…]

### 2023-07-13 pmwd/sto/train.py

@@ -33,11 +33,13 @@ def global_mean(tree):
 
 
 def init_pmwd(pmwd_params):
-    a_snaps, sidx, sobol, mesh_shape, n_steps, so_type, so_nodes = pmwd_params
+    (a_snaps, sidx, sobol, mesh_shape, n_steps, so_type, so_nodes,
+     dropout_rate, dropout_key) = pmwd_params
 
     # generate ic, cosmo, conf
     conf, cosmo = gen_cc(sobol, mesh_shape=(mesh_shape,)*3, a_snapshots=a_snaps,
-                         a_nbody_num=n_steps, so_type=so_type, so_nodes=so_nodes)
+                         a_nbody_num=n_steps, so_type=so_type, so_nodes=so_nodes,
+                         dropout_rate=dropout_rate, dropout_key=dropout_key)
     ptcl_ic = gen_ic(sidx, conf, cosmo)
 
     return ptcl_ic, cosmo, conf
@@ -46,7 +48,7 @@ def init_pmwd(pmwd_params):
 def train_step(tgts, so_params, pmwd_params, opt_params):
     ptcl_ic, cosmo, conf = init_pmwd(pmwd_params)
 
-    # loss and grad
+    # get loss and grad
     obj_valgrad = jax.value_and_grad(obj, argnums=2)
     loss, grad = obj_valgrad(tgts, ptcl_ic, so_params, cosmo, conf)
 
@@ -57,5 +59,8 @@ def train_step(tgts, so_params, pmwd_params, opt_params):
     optimizer, opt_state = opt_params
     updates, opt_state = optimizer.update(grad, opt_state, so_params)
     so_params = optax.apply_updates(so_params, updates)
+
+    # not necessary, but no harm
+    so_params = global_mean(so_params)
 
     return so_params, loss, opt_state

### 2023-07-13 pmwd/sto/data.py
Add support for dropout

@@ -59,7 +59,8 @@ def scale_Sobol(fn='sobol.txt', ind=slice(None)):
 
 
 def gen_cc(sobol, mesh_shape=1, a_snapshots=(1,), a_nbody_num=63,
-           so_type=None, so_nodes=None, a_start=1/16, a_stop=1+1/128):
+           so_type=None, so_nodes=None, a_start=1/16, a_stop=1+1/128,
+           dropout_rate=None, dropout_key=None):
     """Setup conf and cosmo given a sobol."""
     conf = Configuration(
         ptcl_spacing = sobol[0] / 128,
@@ -73,6 +74,8 @@ def gen_cc(sobol, mesh_shape=1, a_snapshots=(1,), a_nbody_num=63,
         so_type = so_type,
         so_nodes = so_nodes,
         softening_length = sobol[8],
+        dropout_rate = dropout_rate,
+        dropout_key = dropout_key.tolist(),  # array could cause ValueError in conf
     )
 
     cosmo = Cosmology(

### 2023-07-13 docs/papers/sto/so.tex
FIXUP draft minor
FIXUP draft some updates

@@
     linear regime), say 1 Gpc/$h$.\newline
-    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
-    consistent at the field level. To further check power spectrum. For odd mesh
-    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
-    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
-    slightly different. To figure out this weird issue.}
+    \YZ{\textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are
+    visually consistent at the field level. For power spectrum: For
+    odd mesh shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent
+    on large scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers
+    look slightly different. To figure out this weird issue.}}
 \item (model; YL) Probably necessary to first incorporate List \& Hahn's
@@
   losses at convergence. Note that 512 realization are not too many, and
-  we have features that can be very correlated.
+  we have features that can be very correlated.\newline
+  \YZ{Dropout with Flax has been coded. We can use Optax AdamW for weight decay.}
 \item (train) optuna hyperparam search on a subset of training data
   before production training
-  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna example}).
+  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna
+  example}).\newline
+  \YZ{Do we also use another subset of the training data as validation data?}
 \item (train) Reduce learning rate on (mean epoch) loss plateau in
-  production
+  production\newline
+  \YZ{A simple scheduler has been coded, to be fine-tuned for production training.}
 \item (test) Interpolation test: compare SO DM and halo fields and
@@
   for feature selection and symbolic regression.
+\item (paper) Think about what figures and tables to show in the paper. Update
+to 6 items could be included. Some tables could be included in the context as
+list if we have too many figures to show?\newline
+  \YZ{figures:
+  1. $f$ and $g$ functions, both NN and Symbolic.
+  2. Density field (and halo?) with zoom in for small scale sharpening.
+  3. Transfer function and correlation coefficient.
+  4. Halo related stats.}
 \item Fix cross references and other journal quirks as a patch
@@
 \section{Results}
+% Results should be divided by topical subheadings
+
+\subsection{Trained SO neural networks and symbolic expressions}
+
+\subsection{Dark matter fields}
+
+\subsection{Halo catalogs}
 
@@
 \section{Discussion}
+% Discussion does not contain subheadings
 
@@
 \centering
-\caption{Possible features that can affect force sharpening. TODO}
+\caption{TODO Possible features that can affect force sharpening.
+All scales are then multiplied by the wavenumber $k$ to form dimensionless
+factors.}
 \label{tab:feat}

### 2023-07-20 docs/papers/sto/so.tex
FIXUP draft minor
FIXUP draft: add some notes
FIXUP draft: minor edits

@@
 \begin{abstract}
+% up to 150 words
 
@@
   while\_loop can help to save compilation when trained on different
-  snapshots. Performance difference between scan and while?
+  snapshots. Performance difference between scan and while?\newline
+  \YZ{It seems that lax.while\_loop only supports fwd grad}
+  \href{https://jax.readthedocs.io/en/latest/notebooks/Common_Gotchas_in_JAX.html#summary}{link}
 \item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
@@
   for feature selection and symbolic regression.
-\item (paper) Think about what figures and tables to show in the paper. Update
+\item (paper) Think about what figures and tables to show in the paper. Up
 to 6 items could be included. Some tables could be included in the context as
@@
 
+% Main text: up to 3500 words: introduction + results + discussion
+% so we should keep introduction and discussion brief
 \vspace{1em}
 \section{Introduction}
+% without heading
 
@@
 
+\YZ{We further sum the loss of all snapshots across a single forward simulation
+at different times, i.e.
+\begin{equation}
+\cJ = \sum_{a_i} \cJ |_{a_i}
+\end{equation}
+}
+
 \begin{table*}
@@
 In practice, we sample one conditioned on the other and the box size.
-*Number of time steps from $a=1/16$ to $a=1$.
+*Number of time steps from $a=1/16$ to $a=1+1/128$ (in order to cover the
+snapshot offset).
 }
@@
 \vspace{1em}
-\subsection{Parameters and Configurations}
+\subsection{Target Data and Configurations}
 
+We generate our target data for training and testing with the \GADGET\
+simulation code \citep{GADGET-4}.
 We run 512 \GADGET\ simulations that cover a wide range of
@@
 
-\vspace{1em}
-\subsection{Target Data for Training and Testing}
-
-\citet{GADGET-4}
-
 We configure \GADGET\ with high force and time integration accuracy
@@
 
-Given 512 \GADGET\ simulations and 121 output snapshots each, in total
-we have 61952 snapshots for training.
-Following \autoref{tab:param}, in each iteration, we randomly sample the
-snapshots, the number of time steps, and the mesh shape (or its cell
-size), and configure \pmwd\ with the same initial coniditions,
-cosmological parameters and other common configurations.
+Following \autoref{tab:param}, in each iteration, we randomly sample the number
+of time steps and the mesh shape (or its cell size), and configure \pmwd\ with
+the same initial coniditions, cosmological parameters and other common
+configurations.
 To balance the work of GPUs during training, in each iteration we use
-the same snapshot $a$, number of time steps, and mesh shape on all GPU.
+the same number of time steps and mesh shape on all GPUs.
 
 We perform parallel training using distributed GPU devices.
-The training on each snapshot is performed on a single GPU device.
-During each training step, the gradients are collected and averaged across
-devices to update the SO parameters.
+The training on all snapshots of a single Sobol configuration is performed on a
+single GPU device.
+In each training step, the loss and gradients are collected and averaged across
+devices.
+The SO parameters are initialized to be the same for all devices, and being
+updated with the same averaged gradients.
 
 Hyperparameters:
+\YZ{
+\begin{itemize}
+  \item Model: number of hidden layers and nodes of f and g nets
+  \item Optimization: learning rate, weight decay and dropout rate etc
+\end{itemize}
+}
 

### 2023-08-01 docs/papers/sto/so.tex
FIXUP draft minor
FIXUP draft: a few sentences on training

@@
 \vspace{1em}
-\subsection{Target Data and Configurations}
+\subsection{Configurations and Data for Training and Testing}
 
@@
 
+\YZ{Do we give this dataset a name and release it? This could be useful for
+future training projects.}
+
 \vspace{1em}
@@
 
-We perform parallel training using distributed GPU devices.
-The training on all snapshots of a single Sobol configuration is performed on a
-single GPU device.
-In each training step, the loss and gradients are collected and averaged across
-devices.
+We perform distributed data-parallel training using GPU devices hosted on
+multiple CPU nodes.
+% TODO hardware details
+The 512 training data samples are uniformly distributed to the processes, with
+each process being binded to a single GPU device.
+In each training step, a data sample is fed into the GPU from the CPU memory
+where the preloading takes place to save the reading time from the disk.
+
+The loss and gradients are collected and averaged across the GPU devices.
 The SO parameters are initialized to be the same for all devices, and being
@@
   \item Model: number of hidden layers and nodes of f and g nets
-  \item Optimization: learning rate, weight decay and dropout rate etc
+  \item Optimization: learning rate scheduler, weight decay and dropout rate etc
+  \item Loss: fields to include and loss function forms
 \end{itemize}

### 2023-08-09 docs/papers/sto/so.tex
FIXUP draft: a few sentences

@@
 The particle velocity is then given by the derivative $\d\vx / \d a$.
+\YZ{Notice that the interpolation is a linear combination of the particles at
+two consecutive time steps, which therefore can be done in either two steps or
+just one step by carrying the particles from the previous time step.
+We find the two-steps scheme more convenient for the backward evolution of the
+adjoint variables.}
 
@@
 \begin{itemize}
-  \item Model: number of hidden layers and nodes of f and g nets
-  \item Optimization: learning rate scheduler, weight decay and dropout rate etc
+  \item Model: number of hidden layers and nodes of f and g nets, activation function
+  \item Optimization: learning rate scheduler, momentum, weight decay and dropout rate etc
   \item Loss: fields to include and loss function forms

### 2023-08-17 pmwd/sto/mlp.py
FIXUP explicitly set he_normal init of kernel

@@ -1,38 +1,43 @@
 import jax.numpy as jnp
 from jax import random
-import flax.linen as nn
 from typing import Sequence, Callable
+import flax.linen as nn
 from flax.core.frozen_dict import unfreeze, freeze
+from flax.linen.initializers import he_normal, zeros_init
 
 
 class MLP(nn.Module):
     features: Sequence[int]
-    activator: Callable[[jnp.ndarray], jnp.ndarray] = nn.leaky_relu
+    activator: Callable[[jnp.ndarray], jnp.ndarray] = nn.relu
     outivator: Callable[[jnp.ndarray], jnp.ndarray] = None
+    kernel_init: Callable = he_normal()
+    bias_init: Callable = zeros_init()
 
     @nn.compact
     def __call__(self, x, dropout=False, dropout_rate=0.5):
         # hidden layers
         for i, fts in enumerate(self.features[:-1]):
-            x = nn.Dense(fts, param_dtype=jnp.float64)(x)
+            x = nn.Dense(fts, param_dtype=jnp.float64, kernel_init=self.kernel_init,
+                         bias_init=self.bias_init)(x)
             x = self.activator(x)
             x = nn.Dropout(rate=dropout_rate, deterministic=not dropout)(x)
 
         # output layer
-        x = nn.Dense(self.features[-1], param_dtype=jnp.float64)(x)
+        x = nn.Dense(self.features[-1], param_dtype=jnp.float64,
+                     kernel_init=self.kernel_init, bias_init=self.bias_init)(x)
         if self.outivator is not None:
             x = self.outivator(x)
 
         return x
 
 
-def init_mlp_params(n_input, nodes, scheme=None):
+def init_mlp_params(n_input, nodes, kernel_init=he_normal(), bias_init=zeros_init(),
+                    scheme=None):
     """Initialize MLP parameters."""
-    nets = [MLP(features=n) for n in nodes]
+    nets = [MLP(features=n, kernel_init=kernel_init, bias_init=bias_init) for n in nodes]
     xs = [jnp.ones(n) for n in n_input]  # dummy inputs
     keys = random.split(random.PRNGKey(0), len(n_input))
 
-    # by default in flax.linen.Dense, kernel: lecun_norm, bias: 0
     params = [nn.init(key, x) for nn, key, x in zip(nets, keys, xs)]
 
     # for the last layer: set bias to one & weights to zero

### 2023-08-17 docs/papers/sto/so.tex
FIXUP draft minor

@@
 
-\title{\large Neural, Tree, and Symbolic Optimization of Cosmological
+\title{\large Neural and Symbolic Optimization of Cosmological
 Particle-Mesh Simulation
@@
 
+The SO MLP parameters are initialized in a way such that the MLP output is very
+close to one, i.e. no SO.
+We achieve this by setting the kernel of the output layer to be very small
+random values and the bias to be one.
+
 Hyperparameters:

### 2023-08-23 pmwd/sto/soft_a.py
FIXUP add another simpler set of SO input features

@@ -1,0 +1,58 @@
+import jax.numpy as jnp
+
+from pmwd.configuration import Configuration
+from pmwd.cosmology import Cosmology, H_deriv, Omega_m_a, SimpleLCDM
+from pmwd.boltzmann import boltzmann, growth, linear_power
+
+
+def _sotheta(cosmo, conf, a):
+    """Physical quantities to be used in SO input features along with k."""
+    theta = jnp.asarray([
+        a,
+        cosmo.A_s_1e9,
+        cosmo.n_s,
+        cosmo.Omega_m,
+        cosmo.Omega_b / cosmo.Omega_m,
+        cosmo.Omega_k,
+        cosmo.h,
+        conf.box_size[0],
+        conf.ptcl_spacing,
+        conf.cell_size,
+        conf.softening_length,
+    ])
+    return theta
+
+
+def _sotheta_names():
+    # str names of the variables returned by the sotheta function above
+    # currently hardcoded, should be updated along with sotheta function
+    pass
+
+
+def _soft_len(k_fac=1):
+    # get the length of SO input features with dummy conf and cosmo
+    conf = Configuration(1., (128,)*3)
+    cosmo = SimpleLCDM(conf)
+    cosmo = boltzmann(cosmo, conf)
+    theta = _sotheta(cosmo, conf, conf.a_start)
+    return k_fac + len(theta)
+
+
+def soft(k, theta):
+    """SO features for neural nets input, with k being a scalar."""
+    return jnp.concatenate((jnp.atleast_1d(k), theta))
+
+
+def soft_k(k, theta, log=False):
+    """Get SO input features (k, theta)."""
+    theta = jnp.broadcast_to(theta, k.shape+theta.shape)
+    k = k.reshape(k.shape + (1,))
+    ft = jnp.concatenate((k, theta), axis=-1)
+    return ft
+
+
+def soft_kvec(kv, theta, log=False):
+    """Get SO input features (k1, k2, k3, theta)."""
+    theta = jnp.broadcast_to(theta, kv.shape[:-1]+theta.shape)
+    ft = jnp.concatenate((kv, theta), axis=-1)
+    return ft

### 2023-08-23 docs/papers/sto/so.tex
FIXUP draft minor

@@
   with 0.01 linthresh, and $1-r^2$ in log at 3 different $a$'s; Show $f$
-  and $g$ (along axis and face diagonal) at 3 different $a$'s; Track
-  mean epoch loss
-\item (train) Randomize snapshots of each simulation instead of
-  sequential; alternatively accumulate their gradients; Normalize loss
-  by number of time steps\newline
-  \YZ{\textbf{Update}: Normalizing loss by number of time steps seems to halt
-  the training of late time snapshots, i.e. they stop improving.}
+  and $g$ (along axis and face diagonal) at 3 different $a$'s
 \item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add
@@
 % TODO hardware details
-The 512 training data samples are uniformly distributed to the processes, with
+The training data samples are uniformly distributed to the processes, with
 each process being binded to a single GPU device.

### 2023-08-29 docs/papers/sto/so.tex
FIXUP draft minor
FIXUP update draft

@@
 
-\correspondingauthor{XYZ, Yin Li, and XYZ}
-\email{XYZ, eelregit@gmail.com, and XYZ}
+\correspondingauthor{Yucheng Zhang \& Yin Li}
+\email{yucheng.zhang@nyu.edu \& eelregit@gmail.com}
 
@@
 % without heading
+Keywords: AI for science, differentiable simulations, deep neural networks,
+symbolic interpretability
+
+Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD),
 
@@
 
-\subsection{Trained SO neural networks and symbolic expressions}
+\subsection{SO neural networks and symbolic expressions}
 
@@
 and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
-foundation of the optimization discussed in this work.
-We denote the particles' state with $\vz = (\vx, \vp)^\intercal$ and their
-adjoint variables with $\vlambda = (\vxi, \vpi)$.
+foundation of the neural optimization discussed in this work.
+Within this fully differentiable framework, we are able to embed neural networks
+in any physical process and train them for many purposes, including the
+optimization of the gravitational force field in this work.
+
+Following these two works, we denote the particles' state with $\vz = (\vx,
+\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
+We refer to the quantities that are directly related to the final objective
+function $\cJ$ as the observables.
+In this work, our observable $\vo$ includes multiple snapshots at given times.
+In the observation operator $O_{i}^{i-1}$, the adjoint variables are updated
+through
+\begin{align}
+  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
+  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
+\end{align}
+where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
+be non-zero if they were used in the interpolation of any snapshots in $\vo$.
+The initial conditions are
+\begin{align}
+  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
+  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
+\end{align}
+which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
+The derivative of the objective function with respect to the observable and the
+VJP of the observation funciton (mainly the interpolation function discussed
+below) can be conveniently computed by AD.
 
@@
 
-Given two snapshots at $a_i$ and $a_{i+1}$, we get the particle displacements
-in between at $a$ using interpolation with the cubic Hermite spline,
+The N-body time integration is carried out following discrete time steps.
+Given the particle displacements and velocities at two consecutive time steps
+$a_i$ and $a_{i+1}$, we get the status of particles in between at $a$ using the
+cubic Hermite interpolation.
+The interpolated particle displacements are given by
 \begin{align}
@@
 \end{align}
-The particle velocity is then given by the derivative $\d\vx / \d a$.
-\YZ{Notice that the interpolation is a linear combination of the particles at
-two consecutive time steps, which therefore can be done in either two steps or
-just one step by carrying the particles from the previous time step.
+The particle velocities are then given by the derivative $\d\vx / \d a$.
+Note that the interpolation is a linear combination of the particles at two
+consecutive time steps, which therefore can be done in either two steps or just
+one step by carrying the particles from the previous time step.
 We find the two-steps scheme more convenient for the backward evolution of the
-adjoint variables.}
-
-We refer to the quantities that are directly related to the objective function
-as the observables.
-Our observable $\vo$ is a snapshot $\hat\vz = (\hat\vx,
-\hat\vp)^\intercal$ at a given time $a_o$, which in general can be given by the
-interpolation of two snapshots $\vz_{k-1}$ and $\vz_k$ where $a_{k-1} \leq a_o <
-a_k$.
-Then ${\p\vo}/{\p\vz_i}$ would be zero except for the relevant $\vz_{k-1}$
-and $\vz_k$. We have in the initial conditions of the adjoint equations
-\begin{align}
-  \vxi_n &= \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vx_n}, \nonumber\\
-  \vpi_n &= \frac{\p\cJ}{\p\vo} \cdot \frac{\p\vo}{\p\vp_n}.
-\end{align}
-In the reverse time evolution of the adjoint equations, these adjoint varibales
-remain zero until the time stepping from $a_{k+1}$ to $a_k$.
+adjoint variables.
 
@@
 
-In Fourier space, the gravitational force field reads
+In cosmological $N$-body simulations, the dynamics of the particles is governed
+by the Poisson equation
+%
+\begin{equation}
+  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
+\end{equation}
+%
+In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
+%
 \begin{equation}
@@
 \end{equation}
-Force sharpening
+%
+We introduce force sharpening through
 %
@@
 %
-where $f$, $g$ are neural networks, each approximating a nonlinear
+where $f$ and $g$ are neural networks, each approximating a nonlinear
 function of the wavevector components and all possible relevant features
@@
 \cmidrule(lr){2-2}
-& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
+& $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
 \cmidrule(lr){2-2}
@@
 
-The dimensionless power spectrum
+Relevant typical scales are defined through
+\begin{itemize}
+\item The RMS linear displacement is given by
+%
+\begin{equation}
+R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
+\end{equation}
+%
+\item The dimensionless power spectrum is defined as
 %
@@
 \end{equation}
-
-Variance of linear matter overdensity in a top-hat window
+%
+\item The variance of linear matter overdensity in a top-hat window is given by
 %
@@
 \end{equation}
+%
 where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
 Fourier transform of the top-hat window.
-Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
+\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
 $W_\gauss(kR) = e^{-(kR)^2/2}$.
-
-The RMS linear displacement
-%
-\begin{equation}
-R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
-\end{equation}
-%
+\end{itemize}
 
@@
 
-Dimensionless combinations, e.g.,
+Dimensionless combinations by multiplying the wavevector components with the
+scales, e.g.,
 %
@@
 \YZ{
-The parameters of the neural networks, which enter every force evaluation in
+The parameters of the neural networks, which enter the force evaluation in
 every time integration step, can be treated just the same as other cosmological
@@
 optimized.
-For both $f$ and $g$ neural networks, we adopt the multilayer perceptron (MLP)
-structure, each with one input layer, two hidden layers and one output layer.
+For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
+fully connected) structures.
 }
@@
 
+The loss function is constructed by comparing the simulated field $\hat f$ with
+the target field $f$.
 \YL{
@@
 %
-where $N_a$ is the time steps it take to generate the snapshot for each
-iteration.
+where $N_a$ is the number of time steps it takes to generate the snapshot for
+each iteration
 (or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
-their Fourier wavevector $\vk$ in both cases)
+their Fourier wavevector $\vk$ in both cases).
 %
@@
 intermediate scales.
+Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
+Lagrangian space, while the density field is in the Eulerian space:
+%
+\begin{equation}
+1 + \delta(\vx) = \int\!\d^3\vq\, \deltaD[ \vx - \vq - \vs(\vq)],
+\end{equation}
+%
+which we compute by scattering particles to the choice of grid for loss.
 
@@
 %
-
-\YL{TODO move this above:}
-Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
-Lagrangian space, while the density field is in the Eulerian space:
-%
-\begin{equation}
-1 + \delta(\vx) = \int\!\d^3\vq\, \deltaD[ \vx - \vq - \vs(\vq)],
-\end{equation}
-%
-which we compute by scattering particles to the choice of grid for loss.
 
@@
 \vspace{1em}
-\subsection{Configurations and Data for Training and Testing}
+\subsection{Configurations and Training Data}
 
@@
 \vspace{1em}
-\subsection{Training of SO neural networks}
+\subsection{Training of SO Neural Networks}
 
@@
 }
+
+We empoly a learning rate scheduler that reduces the learning rate on plateau
+where the loss function stops decreasing.
 

### 2023-09-01 docs/papers/sto/so.tex
FIXUP draft notes

@@
   for feature selection and symbolic regression.
+\item \YZ{Do we also consider pruning and distilling of the trained networks?}
 \item (paper) Think about what figures and tables to show in the paper. Up
@@
 Keywords: AI for science, differentiable simulations, deep neural networks,
-symbolic interpretability
+symbolic interpretability, fast cosmological simulations
 
-Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD),
+Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD), neural
+network (NN)
+
+% introduce PM and its limitation on small scales
+Using a hybrid embedding of neural networks in differentiable physical
+simulations, we improve the performance of PM method on small scales in
+cosmological $N$-body simulations by sharpening the gravitational force.
 
@@
 
-\subsection{SO neural networks and symbolic expressions}
+\subsection{SO Neural Networks and Symbolic Expressions}
 
-\subsection{Dark matter fields}
+% visualization & comparison of neural nets & symbolic functions
+For the $f(k_i; \vvartheta)$ NN, we show it as a function of $k_i$ at multiple
+different $\vvartheta$ values.
 
-\subsection{Halo catalogs}
+For the $g(k_1,k_2,k_3;\vvartheta)$ NN, which is permutationally symmetric about
+$(k_1, k_2, k_3)$, we show it as a function of $k_i$ along axis and face
+diagonal directions.
+
+\subsection{Dark Matter Density Fields}
+
+% transfer function & correlation coefficient for different confs
+
+% density field with zoom in (with halos painted as circles?)
+
+\subsection{Dark Matter Halo Catalogs}
 
@@
 In this work, our observable $\vo$ includes multiple snapshots at given times.
-In the observation operator $O_{i}^{i-1}$, the adjoint variables are updated
-through
+For the observation operator $O_{i}^{i-1}$ in the adjoint equations from time
+step $a_i$ back to $a_{i-1}$, the adjoint variables are updated through
 \begin{align}

### 2023-09-04 docs/papers/sto/so.tex
FIXUP update paper draft

@@
 % without heading
-Keywords: AI for science, differentiable simulations, deep neural networks,
-symbolic interpretability, fast cosmological simulations
 
 Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD), neural
-network (NN)
+network (NN), particle-mesh (PM), fast-Fourier transform (FFT)
+
+% AI for science and cosmology
+
+% a brief review of (differentiable) cosmological simulations
+BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
+\texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
 
 % introduce PM and its limitation on small scales
-Using a hybrid embedding of neural networks in differentiable physical
-simulations, we improve the performance of PM method on small scales in
-cosmological $N$-body simulations by sharpening the gravitational force.
+The PM method solves the Poisson equation by scattering the particles onto a
+mesh, on which FFT is performed and the Laplacian operator reduces to simple
+multiplication of wavenumbers in Fourier space.
+Although being computationally efficient, the PM method is mainly limited by its
+resolution on small scales, i.e. scales around the mesh cell size.
+This is due to the loss of information during the scattering process of
+continuously distributed particles onto the discrete mesh.
+We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to compute the
+mass fractions of a particle scattered to nearby grid points.
+Other schemes include the nearest grid point (NGP) and triangle shaped cloud
+(TSC) etc.
+
+Using a hybrid embedding of NNs in differentiable physical simulations, we
+optimize the performance of PM method on small scales in cosmological $N$-body
+simulations by sharpening the gravitational force.
 
@@
 % Results should be divided by topical subheadings
+% Put main ideas in Results first, then more technical details in Methods.
 
-\subsection{SO Neural Networks and Symbolic Expressions}
+\subsection{Optimization of Dynamics}
+In cosmological $N$-body simulations, the dynamics of the particles is governed
+by the Poisson equation
+%
+\begin{equation}
+  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
+\end{equation}
+%
+In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
+%
+\begin{equation}
+- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\end{equation}
+%
+We introduce force sharpening through
+%
+\begin{equation}
+\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
+  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
+\end{equation}
+%
+where $f$ and $g$ are NNs, each approximating a nonlinear function of the
+wavevector components and all possible relevant features $\vvartheta$ discussed
+in \autoref{sec:features}.
+Being embedded in the differentiable simulation, the NNs can be trained using
+gradient-based optimization methods as in usual deep learning tasks.
+The parameters of the neural networks, which enter the force evaluation in
+every time integration step, can be treated just the same as other cosmological
+parameters.
+The objective gradients are accumulated, based on which the parameters are then
+optimized.
+For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
+fully connected) structures.
+
+We generate our target data for training and testing with the \GADGET\
+simulation code \citep{GADGET-4}.
+
+\subsection{Trained Neural Networks}
 
@@
 
+\subsection{Symbolic Expressions}
+
+We use the random forest regressor for feature selection using impurity-based
+importance.
+\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
+importance?}
+We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
+% more details on the hyperparams in random forest
+
+The selected important features are then used for SR.
+We use the \texttt{PySR} \citep{PySR} SR code to fit for analytic
+function expressions of the trained SO neural networks.
+
 \subsection{Dark Matter Density Fields}
@@
 % transfer function & correlation coefficient for different confs
+To compare two fields $\hat f$ and $f$, we introduce for each mode the square
+root ratio of power
+%
+\begin{equation}
+T(\vk) \triangleq
+\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
+\label{T}
+\end{equation}
+%
+which is also referred to as the transfer function, and the cross correlation
+coefficient
+%
+\begin{equation}
+r(\vk) \triangleq
+\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
+     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
+\label{r}
+\end{equation}
+%
+The two fields are identical when both $T$ and $r$ are unity.
 
 % density field with zoom in (with halos painted as circles?)
+The effect of force sharpening can also be visually shown by looking directly at
+the density field.
 
@@
 \vspace{1em}
-\subsection{Optimization of Dynamics}
-\label{sec:so}
-
-In cosmological $N$-body simulations, the dynamics of the particles is governed
-by the Poisson equation
-%
-\begin{equation}
-  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
-\end{equation}
-%
-In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
-%
-\begin{equation}
-- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
-\end{equation}
-%
-We introduce force sharpening through
-%
-\begin{equation}
-\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
-\end{equation}
-%
-where $f$ and $g$ are neural networks, each approximating a nonlinear
-function of the wavevector components and all possible relevant features
-$\vvartheta$ summarized in \autoref{tab:feat} and discussed below.
+\subsection{Input Features}
+\label{sec:features}
 
@@
 
-\YZ{
-The parameters of the neural networks, which enter the force evaluation in
-every time integration step, can be treated just the same as other cosmological
-parameters.
-The objective gradients are accumulated, based on which the parameters are then
-optimized.
-For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
-fully connected) structures.
-}
-
 \vspace{1em}
@@
 
-To understand the form of the loss function, let's introduce for each
-mode the square root ratio of power
-%
-\begin{equation}
-T(\vk) \triangleq
-\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
-\end{equation}
-%
-and the cross correlation coefficient
-%
-\begin{equation}
-r(\vk) \triangleq
-\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
-     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
-\end{equation}
-%
-Now, under the mild assumptions that $T$ and $r$ are smooth and
-isotropic, we can show
+Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
+and isotropic, we can show
 %
@@
 
-We generate our target data for training and testing with the \GADGET\
-simulation code \citep{GADGET-4}.
 We run 512 \GADGET\ simulations that cover a wide range of
@@
 
-We use the random forest regressor for feature selection using impurity-based
-importance.
-\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
-importance?}
-We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
-% more details on the hyperparams in random forest
-
-The selected important features are then used for SR.
-
-We use the \texttt{PySR} \citep{Cranmer2023} SR code to fit for analytic
-function expressions of the trained SO neural networks.
-
 \YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$

### 2023-09-06 pmwd/cosmology.py
FIXUP minor

@@ -75,7 +75,7 @@ class Cosmology:
 
     varlin_g: Optional[jnp.ndarray] = field(default=None, compare=False)
 
-    # list of parameters of SO neural nets, order: f, g, h
+    # list of parameters of SO neural nets
     so_params: Optional[list[FrozenDict]] = None
 
     def __post_init__(self):

### 2023-09-06 docs/papers/sto/so.tex
FIXUP updating draft

@@
 \subsection{Optimization of Dynamics}
+
+\begin{table}
+  \centering
+  \caption{Possible features that can affect force sharpening.
+  All scales are then multiplied by the wavenumber $k$ to form dimensionless
+  factors.}
+  \label{tab:feat}
+  \begin{tabular}{rl}
+  \toprule
+  type & feature \\
+  \midrule
+  & $R_\mathrm{d}$: RMS linear displacement \\
+  \cmidrule(lr){2-2}
+  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  & $\d R / \d\lna$: time derivatives of the above \\
+  \midrule
+  & particle spacing \\
+  \cmidrule(lr){2-2}
+  other scales & $l$: PM force mesh cell size \\
+  \cmidrule(lr){2-2}
+  & softening length \\
+  \midrule
+  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
+  \cmidrule(lr){2-2}
+  & $\d\ln G_m / \d\lna$ \\
+  \cmidrule(lr){2-2}
+  dimensionless factors & $\Omegam(a)$ \\
+  \cmidrule(lr){2-2}
+  & $\d\ln\!H / \d\lna$ \\
+  \cmidrule(lr){2-2}
+  & $\Delta\lna$: time step size in $\lna$ \\
+  \bottomrule
+  \end{tabular}
+  \end{table}
+
 In cosmological $N$-body simulations, the dynamics of the particles is governed
@@
 where $f$ and $g$ are NNs, each approximating a nonlinear function of the
-wavevector components and all possible relevant features $\vvartheta$ discussed
-in \autoref{sec:features}.
+wavevector components and all possible relevant features $\vvartheta$ summarized
+in \autoref{tab:feat} and more details discussed in \autoref{sec:features}.
+Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
+broken by the PM solver to the O\textsubscript{h} discrete symmetry with
+48 elements involving permutations and reversals of the 3 principal axes
+of the simulation box.
+The PM evaluation of each force component can further break the symmetry
+along that axis, leaving the remaining 2 permutationally symmetric.
+Therefore, we implement $f$ and $g$ to be even functions of the
+wavevector components, and $g$ to be permutation equivariant in all 3
+components, so that $f g$ is the most general modification to the force
+kernel.
+An even and real kernel in configuration space is also even in Fourier
+space.
+We take the absolute value for the evenness, and sort the 3 wavevector
+compontents to achieve the permutation equivariance.
 Being embedded in the differentiable simulation, the NNs can be trained using
@@
 
+\begin{table*}
+  \centering
+  \caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
+  parameters.
+  Note that the grid ratio need next fast len to determine the mesh shape
+  for fast FFT.
+  Given the box size, the mesh shape determines the cell size.
+  The softening parameter gives the ratio of the comoving softening length
+  to the mean particle spacing.
+  The curvature $\Omega_k$ is related to the separate universe simulation.
+  We sample parameters applicable to \GADGET\ during data generation, and
+  those applicable to \pmwd\ during training.
+  \textsuperscript\dag{}The box size is determined jointly by the \pmwd\
+  mesh shape and mesh cell size below, which are assumed to be sampled
+  independently. \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
+  In practice, we sample one conditioned on the other and the box size.
+  *Number of time steps from $a=1/16$ to $a=1+1/128$ (in order to cover the
+  snapshot offset).
+  }
+  \label{tab:param}
+  \begin{tabular}{lcccr}
+  \toprule
+  applicability & parameter & distribution & range & sampling \\
+  \midrule
+  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
+  \cmidrule(lr){2-5}
+  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
+  \cmidrule(lr){2-5}
+  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
+  \cmidrule(lr){2-4}
+  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
+  \cmidrule(lr){2-4}
+  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
+  \cmidrule(lr){2-4}
+  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
+  \cmidrule(lr){2-4}
+  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
+  \cmidrule(lr){2-4}
+  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
+  \cmidrule(lr){2-4}
+  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
+  \cmidrule(lr){2-4}
+  & $h$ & log-uniform & [0.5, 1) \\
+  \cmidrule(lr){1-4}
+  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
+  \cmidrule(lr){1-5}
+  & mesh shape & log-uniform & [128, 512] \\
+  \cmidrule(lr){2-4}
+  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
+  \cmidrule(lr){2-4}
+  & number of time steps* & log-uniform & [10, 1000] \\
+  \cmidrule(lr){1-4}
+  training & loss grid offset in its own spacing & uniform & $[0, 1)^3$ \\
+  \bottomrule
+  \end{tabular}
+  \end{table*}
+
 We generate our target data for training and testing with the \GADGET\
 simulation code \citep{GADGET-4}.
+We run 512 \GADGET\ simulations that cover a wide range of
+configurations and cosmological parameters, as listed in
+\autoref{tab:param}.
+Each simulation has $128^3$ particles and writes 121 snapshots linearly
+spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
+spacing $1/128$.
+$\Delta a$ is an offset, common for all snapshots in each simulation,
+that varies as in \autoref{tab:param} and helps to sample different
+alignment between snapshot and start/stop times.
+In our \pmwd\ model, we generate the snapshot at an arbitrary given time using
+the cubic Hermite interpolation, with more details in \autoref{sec:snapobs}.
+
+The curvature $\Omega_k$ is related to the separate universe simulation
+\citep{LiEtAl2014, WagnerEtAl2015}.
 
@@
 \subsection{Differentiable Simulation}
+\label{sec:pmwd}
 
@@
 in any physical process and train them for many purposes, including the
-optimization of the gravitational force field in this work.
+optimization of the gravitational force evaluation in this work.
 
@@
 
-\begin{table}
-\centering
-\caption{TODO Possible features that can affect force sharpening.
-All scales are then multiplied by the wavenumber $k$ to form dimensionless
-factors.}
-\label{tab:feat}
-\begin{tabular}{rl}
-\toprule
-type & feature \\
-\midrule
-& $R_\mathrm{d}$: RMS linear displacement \\
-\cmidrule(lr){2-2}
-& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
-\cmidrule(lr){2-2}
-nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
-\cmidrule(lr){2-2}
-& $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
-\cmidrule(lr){2-2}
-& $\d R / \d\lna$: time derivatives of the above \\
-\midrule
-& particle spacing \\
-\cmidrule(lr){2-2}
-other scales & $l$: PM force mesh cell size \\
-\cmidrule(lr){2-2}
-& softening length \\
-\midrule
-& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
-\cmidrule(lr){2-2}
-& $\d\ln G_m / \d\lna$ \\
-\cmidrule(lr){2-2}
-dimensionless factors & $\Omegam(a)$ \\
-\cmidrule(lr){2-2}
-& $\d\ln\!H / \d\lna$ \\
-\cmidrule(lr){2-2}
-& $\Delta\lna$: time step size in $\lna$ \\
-\bottomrule
-\end{tabular}
-\end{table}
+Multiple scales and dimensionless factors could possibly affect force
+sharpening, which are denoted as $\vvartheta$ and enter the NNs of $f(k_i;
+\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ as input features.
+A list of these features is summarized in \autoref{tab:feat}, and here we
+present relevant definations and more details.
 
-Relevant typical scales are defined through
+Below are some definations related to the typical scales.
 \begin{itemize}
-\item The RMS linear displacement is given by
+\item The root-mean-square (RMS) linear displacement is given by
 %
 \begin{equation}
-R_\mathrm{d} = \int \frac{k P_\linear}{6\pi^2} \d\lnk.
+R_\mathrm{d} = \int \frac{k P_\linear(k)}{6\pi^2} \d\lnk,
 \end{equation}
 %
-\item The dimensionless power spectrum is defined as
+where $P_\linear(k)$ is the linear power spectrum.
+\item The dimensionless linear power spectrum is defined as
 %
 \begin{equation}
-\Delta^2(k) \triangleq \frac{k^3 P(k)}{2 \pi^2}.
+\Delta^2_\linear(k) \triangleq \frac{k^3 P_\linear(k)}{2 \pi^2},
 \end{equation}
 %
+based on which we define a nonlinear scale $R_P$ through
+$\Delta_\linear^2(R_P^{-1}) \triangleq 1$.
 \item The variance of linear matter overdensity in a top-hat window is given by
@@
 Fourier transform of the top-hat window.
+Then we have $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$.
 \item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
-$W_\gauss(kR) = e^{-(kR)^2/2}$.
+$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$: $\sigma_\gauss^2(R_\gauss)
+\triangleq 1$.
 \end{itemize}
+These nonlinear scales are functions of time since the linear power spectrum
+evolves with time through the growth factor $D(a)$, we also take the time
+derivatives of these scales as input features.
+Other relevant scales include the particle spacing in Lagrangian space, the cell
+size of the chosen mesh grid for force evaluation, and the softening length in
+the target \GADGET\ simulations.
 
@@
 
-We combine the above list of parameters in $\vvartheta$.
-
-Dimensionless combinations by multiplying the wavevector components with the
-scales, e.g.,
+Dimensionless combinations are formed by multiplying the wavevector components
+with the scales, i.e.,
 %
 \begin{equation}
-f(k_i; \vvartheta) = f(k_iR_d, k_iR_\tophat, \cdots; G_m, \Omegam, \cdots)
+f(k_i; \vvartheta) = f(k_iR_d, k_iR_P, \cdots; G_m, \Omegam, \cdots),
 \end{equation}
@@
 \begin{equation}
-g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_\tophat, \cdots\}_{j=1,2,3}; G_m, \Omegam, \cdots)
+g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_P, \cdots\}_{j=1,2,3}; G_m, \Omegam, \cdots).
 \end{equation}
 %
-$f$ and $g$ are optimized during training to correct and sharpen the
-gravitational forces.
-Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
-broken by the PM solver to the O\textsubscript{h} discrete symmetry with
-48 elements involving permutations and reversals of the 3 principal axes
-of the simulation box.
-The PM evaluation of each force component can further break the symmetry
-along that axis, leaving the remaining 2 permutationally symmetric.
-Therefore, we implement $f$ and $g$ to be even functions of the
-wavevector components, and $g$ to be permutation equivariant in all 3
-components, so that $f g$ is the most general modification to the force
-kernel.
-An even and real kernel in configuration space is also even in Fourier
-space.
-We take the absolute value for the evenness, and sort the 3 wavevector
-compontents to achieve the permutation equivariance.
 
@@
 
-\begin{table*}
-\centering
-\caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
-parameters.
-Note that the grid ratio need next fast len to determine the mesh shape
-for fast FFT.
-Given the box size, the mesh shape determines the cell size.
-The softening parameter gives the ratio of the comoving softening length
-to the mean particle spacing.
-The curvature $\Omega_k$ is related to the separate universe simulation.
-We sample parameters applicable to \GADGET\ during data generation, and
-those applicable to \pmwd\ during training.
-\textsuperscript\dag{}The box size is determined jointly by the \pmwd\
-mesh shape and mesh cell size below, which are assumed to be sampled
-independently. \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
-In practice, we sample one conditioned on the other and the box size.
-*Number of time steps from $a=1/16$ to $a=1+1/128$ (in order to cover the
-snapshot offset).
-}
-\label{tab:param}
-\begin{tabular}{lcccr}
-\toprule
-applicability & parameter & distribution & range & sampling \\
-\midrule
-& $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
-\cmidrule(lr){2-5}
-& 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
-\cmidrule(lr){2-5}
-& box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
-\cmidrule(lr){2-4}
-& snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
-\cmidrule(lr){2-4}
-\GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
-\cmidrule(lr){2-4}
-\& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
-\cmidrule(lr){2-4}
-& $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
-\cmidrule(lr){2-4}
-& $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
-\cmidrule(lr){2-4}
-& $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
-\cmidrule(lr){2-4}
-& $h$ & log-uniform & [0.5, 1) \\
-\cmidrule(lr){1-4}
-\GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
-\cmidrule(lr){1-5}
-& mesh shape & log-uniform & [128, 512] \\
-\cmidrule(lr){2-4}
-\pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
-\cmidrule(lr){2-4}
-& number of time steps* & log-uniform & [10, 1000] \\
-\cmidrule(lr){1-4}
-training & loss grid offset in its own spacing & uniform & $[0, 1)^3$ \\
-\bottomrule
-\end{tabular}
-\end{table*}
-
 \vspace{1em}
 \subsection{Configurations and Training Data}
-
-We run 512 \GADGET\ simulations that cover a wide range of
-configurations and cosmological parameters, as listed in
-\autoref{tab:param}.
-Each simulation has $128^3$ particles and writes 121 snapshots linearly
-spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
-spacing $1/128$.
-$\Delta a$ is an offset, common for all snapshots in each simulation,
-that varies as in \autoref{tab:param} and helps to sample different
-alignment between snapshot and start/stop times.
-
-The curvature $\Omega_k$ is related to the separate universe simulation
-\citep{LiEtAl2014, WagnerEtAl2015}.
 

### 2023-09-07 docs/papers/sto/so.tex
FIXUP updating paper draft

@@
 wavevector components and all possible relevant features $\vvartheta$ summarized
-in \autoref{tab:feat} and more details discussed in \autoref{sec:features}.
+in \autoref{tab:feat} with more details discussed in \autoref{sec:features}.
 Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
@@
 
-\subsection{Symbolic Expressions}
-
+\subsection{Feature Importance}
 We use the random forest regressor for feature selection using impurity-based
@@
 importance?}
+
+\subsection{Symbolic Expressions}
 We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
@@
 
-We combine Lagrangian and Eulerian fields to form the total loss:
+We combine Lagrangian and Eulerian fields to form the total loss for a single
+snapshot:
 %
 \begin{equation}
-\cJ = \cL_\vs + [\cL_\vv] + \cL_\delta.
+\cJ = \cL_\vs + \cL_\delta.
 \end{equation}
@@
 \subsection{Configurations and Training Data}
-
-We have verified IC generation on GPUs is deterministic though different
-from those on CPUs, so we generate ICs for both training data and
-training on GPUs.
 
@@
 
+Small numerical differences in initial conditions (ICs) could become more
+significant with the $N$-body evolution.
+We have verified that the IC generation on GPUs is deterministic though
+different from those on CPUs, therefore we generate ICs for both the training
+data and the model on GPUs.
+
 We configure \GADGET\ with high force and time integration accuracy
@@
 %FIXME list other from Config.sh and param.txt
-
-\YZ{Do we give this dataset a name and release it? This could be useful for
-future training projects.}
 
@@
 \vspace{1em}
-\subsection{Symbolic Regression}
+\subsection{Feature Importance Analysis}
 
 We generate a sample of (input, output) data with the trained neural network to
-perform feature selection and SR.
+perform feature importance analysis.
+
+\vspace{1em}
+\subsection{Symbolic Regression}
 
@@
 \vspace{1em}
+\textit{\large Data Availability:}
+\YZ{Do we give \GADGET\ Sobol 512 dataset a name and make it public? This could
+be useful for future training projects.}
+
+\vspace{1em}
 \textit{\large Code Availability:}
 \pmwd\ \citep{Li2022b} is open-source on GitHub
-\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
-files and scripts of this paper
+\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source files
+and scripts of this paper
 \href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
+\GADGET\ \citep{GADGET-4}, which can be obtained through a public git-repository
+\href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab}, hosted by a gitlab
+server at the Max-Planck Computing and Data Facility (MPCDF).
 
@@
 
+\appendix
+% supplementary information of Nat. Mach. Intell.
+
 %\listofchanges

### 2023-09-07 docs/papers/sto/so.tex
FIXUP updating draft test on so param gradients

@@
 
+\vspace{1em}
 \appendix
 % supplementary information of Nat. Mach. Intell.
+
+\vspace{1em}
+\section{Tests on the Gradients of NN Parameters}
+As discussed in the main text, our objective loss function is defined on
+multiple snapshots at given times, which are constructed through interpolation
+during the observation operation.
+Here we perform tests on the gradients of NN parameters to check our
+implementation of the observation inside the adjoint method of the $N$-body time
+integration.
+The gradients given by the adjoint method are compared against those given by
+the AD where we disable the custom VJP of the $N$-body function, see
+\autoref{fig:nn_grads_cmp}.
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.5\linewidth]{nn_grads_cmp.pdf}
+  \caption{\YZ{TODO make this figure better}.
+  A comparison of adjoint gradients of all parameters in the $f$ and $g$ NNs to
+  those by AD.
+  This test runs a \pmwd\ simulation of $128^3$ particles, with a $128^3$ mesh
+  and 15 time steps.
+  The size and time steps of this test case is mainly limited by the memory cost
+  of the AD method.
+  The two NNs $f$ and $g$ both have two hidden layers of size $[n, n]$, where
+  $n$ is the size of the input layer.
+  The loss function involves 3 snapshots at $a=(0.6, 0.8, 1)$.
+  }
+  \label{fig:nn_grads_cmp}
+\end{figure*}
 

### 2023-09-08 docs/papers/sto/so.tex
FIXUP draft minor

@@
 \vspace{1em}
-\section{Tests on the Gradients of NN Parameters}
+\section{Tests of Snapshots Observation in Adjoint Method}
 As discussed in the main text, our objective loss function is defined on
@@
 Here we perform tests on the gradients of NN parameters to check our
-implementation of the observation inside the adjoint method of the $N$-body time
-integration.
+implementation of the snapshots observation inside the adjoint method of the
+$N$-body time integration.
 The gradients given by the adjoint method are compared against those given by
@@
   \includegraphics[width=0.5\linewidth]{nn_grads_cmp.pdf}
-  \caption{\YZ{TODO make this figure better}.
-  A comparison of adjoint gradients of all parameters in the $f$ and $g$ NNs to
-  those by AD.
+  \caption{A comparison of adjoint gradients of all parameters in the $f$ and
+  $g$ NNs to those by AD.
   This test runs a \pmwd\ simulation of $128^3$ particles, with a $128^3$ mesh

### 2023-09-12 docs/papers/sto/so.tex
FIXUP draft minor

@@
   mesh shape and mesh cell size below, which are assumed to be sampled
-  independently. \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
+  independently.
+  \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
+  \YZ{We are using ``Mpc/h'' in the code.}
   In practice, we sample one conditioned on the other and the box size.
@@
   \cmidrule(lr){2-5}
-  & box size in Mpc\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
+  & box size in Mpc\YZ{/h}\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
   \cmidrule(lr){2-4}
@@
   \cmidrule(lr){2-4}
-  \pmwd\ & mesh cell size in Mpc & log-uniform & [0.2, 5] & MC \\
+  \pmwd\ & mesh cell size in Mpc\YZ{/h} & log-uniform & [0.2, 5] & MC \\
   \cmidrule(lr){2-4}

### 2023-09-12 docs/papers/sto/so.tex
FIXUP draft minor

@@
 
+% hardware setups
+We perform distributed data-parallel training using GPU devices hosted on
+multiple CPU nodes.
+With each process being binded to a single GPU device, the training data samples
+are uniformly distributed to the processes.
+In total, we use [...] \GPU\ GPUs, each accompanied with [...] CPU cores and
+[...] GB CPU memory mainly for preloading the training data.
+In each training step, a data sample is fed into the GPU from the CPU memory
+where the preloading takes place to save the reading time from the disk.
+
 Following \autoref{tab:param}, in each iteration, we randomly sample the number
@@
 the same number of time steps and mesh shape on all GPUs.
-
-We perform distributed data-parallel training using GPU devices hosted on
-multiple CPU nodes.
-% TODO hardware details
-The training data samples are uniformly distributed to the processes, with
-each process being binded to a single GPU device.
-In each training step, a data sample is fed into the GPU from the CPU memory
-where the preloading takes place to save the reading time from the disk.
 

### 2023-09-13 docs/papers/sto/so.tex
FIXUP draft updating

@@
 \newcommand{\GADGET}{{{\fontsize{10pt}{12pt}\selectfont GADGET}-4}}
+\newcommand{\GSDATA}{{GS512}}
 
@@
 the cubic Hermite interpolation, with more details in \autoref{sec:snapobs}.
-
 The curvature $\Omega_k$ is related to the separate universe simulation
 \citep{LiEtAl2014, WagnerEtAl2015}.
+\YZ{We refer to this training dataset of 512 \GADGET\ simulations sampled with
+Sobol sequence as \GSDATA.}
 
@@
 \subsection{Loss Function}
+\label{sec:loss}
 
@@
 Following \autoref{tab:param}, in each iteration, we randomly sample the number
-of time steps and the mesh shape (or its cell size), and configure \pmwd\ with
-the same initial coniditions, cosmological parameters and other common
-configurations.
+of time steps and the mesh shape, and configure \pmwd\ with the same initial
+coniditions, cosmological parameters and other common configurations.
 To balance the work of GPUs during training, in each iteration we use
@@
 
-The loss and gradients are collected and averaged across the GPU devices.
+As discussed in \autoref{sec:loss}, the loss in a single simulation sums over
+all snapshots.
+Our batch size is [...], i.e. the number of GPU devices since one simulation is
+run on a given GPU and compared to the corresponding training data sample for
+loss.
+For each training step, the loss and gradients are collected and averaged across
+all GPU devices.
 The SO parameters are initialized to be the same for all devices, and being
@@
 Hyperparameters:
-\YZ{
 \begin{itemize}
@@
 \end{itemize}
-}
 
@@
 
-\YZ{
 We implement the neural networks with \texttt{Flax} \citep{flax2020github}, and
@@
 \citep{deepmind2020jax}.
-}
 
@@
 \textit{\large Data Availability:}
-\YZ{Do we give \GADGET\ Sobol 512 dataset a name and make it public? This could
-be useful for future training projects.}
+\YZ{Do we make the \GSDATA\ data public?}
 

### 2023-09-15 docs/papers/sto/so.tex
FIXUP draft add two prelim figures

@@
 
-We sharpen the PM force with first neural networks and then symbolic
-regression, taking symmetry and dimensional analysis in consideration.
+We optimize the spatial resolution of cosmological particle-mesh (PM)
+simulations by sharpening the PM force with first neural networks and then
+symbolic regression, taking symmetry and dimensional analysis in consideration.
 
@@
 %
-We introduce force sharpening through
+where we introduce force sharpening for spatial optimization through
 %
@@
 
-% transfer function & correlation coefficient for different confs
+\begin{figure*}
+  \centering
+  \includegraphics[width=.8\columnwidth]{sobol0snap120epoch300.pdf}
+  \caption{The transfer function $T$ and correlation coefficient $r$ of the
+  \GADGET\ and \pmwd\ simulated density fields, with and without force
+  sharpening.
+  Sobol 0 snap 120}
+  \label{fig:tfcc}
+\end{figure*}
+
+% transfer function & correlation coefficient
 To compare two fields $\hat f$ and $f$, we introduce for each mode the square
@@
 The two fields are identical when both $T$ and $r$ are unity.
+With force sharpening,
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.8\columnwidth]{denslab_0_120_e300.pdf}
+  \caption{Projected dark matter density fields of \GADGET, \pmwd\ without and
+  with force sharpening.
+  Sobol 0 snap 120}
+  \label{fig:denslab}
+\end{figure*}
 
@@
 where the loss function stops decreasing.
+The initial learning rate is set at [...] with a reduction factor of [...] after
+[...] epochs of no decreasing on loss.
 

### 2023-09-15 docs/papers/sto/so.tex
FIXUP draft

@@
 and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
-foundation of the neural optimization discussed in this work.
+foundation of the gradient-based optimization discussed in this work.
 Within this fully differentiable framework, we are able to embed neural networks
-in any physical process and train them for many purposes, including the
-optimization of the gravitational force evaluation in this work.
+in physical processes and train them for many purposes, including the
+optimization of spatial resolution by sharpening the gravitational force in this
+work.
 
-Following these two works, we denote the particles' state with $\vz = (\vx,
+Here we briefly recap the equations that are closely related to our
+extensions to the original framework, and setup the notations of the variables.
+Following \cite{Li2022a}, we denote the particles' state with $\vz = (\vx,
 \vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
 We refer to the quantities that are directly related to the final objective
-function $\cJ$ as the observables.
-In this work, our observable $\vo$ includes multiple snapshots at given times.
-For the observation operator $O_{i}^{i-1}$ in the adjoint equations from time
-step $a_i$ back to $a_{i-1}$, the adjoint variables are updated through
+function $\cJ$ as the observable $\vo$.
+In this work, our observable includes multiple snapshots at given times.
+For the observation operator $O_{i}^{i-1}$ in the adjoint equations (check
+Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to $a_{i-1}$, the adjoint
+variables are updated through
 \begin{align}
@@
 where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
-be non-zero if they were used in the interpolation of any snapshots in $\vo$.
-The initial conditions are
+be non-zero only if they were used in the generation of any snapshots in $\vo$.
+The initial conditions of these adjoint variables are given by
 \begin{align}
@@
 which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
-The derivative of the objective function with respect to the observable and the
-VJP of the observation funciton (mainly the interpolation function discussed
-below) can be conveniently computed by AD.
+The derivative of the objective function with respect to the observable
+($\p\cJ/\p\vo$) and the VJPs of the observation operation ($\p\vo/\p\vx_n$ and
+$\p\vo/\p\vp_n$, mainly the interpolation function discussed in
+\autoref{sec:snapobs} below) can be conveniently computed by AD.
 
@@
 
-The N-body time integration is carried out following discrete time steps.
-Given the particle displacements and velocities at two consecutive time steps
-$a_i$ and $a_{i+1}$, we get the status of particles in between at $a$ using the
-cubic Hermite interpolation.
+The N-body time integration is carried out following discrete time steps. Given
+the particle displacements $\vx$ and velocities $\vv = \vp/(a^3H)$ at two
+consecutive time steps $a_i$ and $a_{i+1}$, we get the snapshot of particles in
+between at $a$ using
+the cubic Hermite interpolation.
 The interpolated particle displacements are given by
@@
 The particle velocities are then given by the derivative $\d\vx / \d a$.
-Note that the interpolation is a linear combination of the particles at two
-consecutive time steps, which therefore can be done in either two steps or just
-one step by carrying the particles from the previous time step.
+Note that the interpolation is a linear combination of the particle snapshots at
+two consecutive time steps, which therefore can be done in either two steps or
+just one step by carrying the particles from the previous time step.
 We find the two-steps scheme more convenient for the backward evolution of the
 adjoint variables.
+The interpolated snapshots at different times make the observable in this work,
+and the VJPs of the interpolation operation are computed by AD and put into the
+adjoint equations as mentioned in \autoref{sec:pmwd}.
 

### 2023-09-25 docs/papers/sto/so.tex
FIXUP draft minor

@@
 \subsection{Feature Importance}
+
+We generate a sample of (input, output) data of the trained neural network to
+perform feature importance analysis.
+We use Sobol sequence to generate the sample of (parameters in
+\autoref{tab:param}, $k$, $a$).
+
 We use the random forest regressor for feature selection using impurity-based
 importance.
+We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
+% more details on the hyperparams in random forest
 \YZ{Maybe we can add another column in \autoref{tab:feat} for feature
@@
 \subsection{Symbolic Expressions}
-We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
-% more details on the hyperparams in random forest
 
@@
 
-We generate a sample of (input, output) data with the trained neural network to
-perform feature importance analysis.
-
 \vspace{1em}

### 2023-09-28 docs/papers/sto/so.tex
FIXUP draft gadget cpu hour

@@
 Sobol sequence as \GSDATA.}
+On average each \GADGET\ simulation takes roughly 2K CPU hours to run, and the
+generation of the whole \GSDATA\ dataset consumes 1M CPU hours.
+% to be compare with the speed of pmwd with SO
 

### 2023-11-02 docs/papers/adjoint/response.txt
Add ApJS revision

We thank the referee for the careful review, which really helped improve
the manuscript. And YL apologize for the delay in this response, due to
mostly job related reasons. We address each point below.

> The authors present a novel framework for differentiable cosmological
> simulations based on the adjoint method, implemented in the code
> package “pmwd”. A great advantage of their method is that the memory
> usage is independent of the number of timesteps performed; hence, this
> paper is pioneering in the development of differentiable simulations
> with a large number of timesteps. The authors achieve this by
> computing the gradients of some objective function w.r.t. the
> cosmological parameters and the initial white noise field via the
> adjoint equations, unlike previous differentiable simulations that
> backpropagate the gradients through all time steps of the simulation
> with automatic differentiation. Using the adjoint method for
> cosmological N-body simulations is an excellent idea, and while the
> authors only consider a simple toy example as the only application in
> this work, which does not reveal the full potential of the method, I
> am certain that their framework will find use in many applications in
> the context of forward modeling.
>
> The paper is generally well written and mostly well structured;
> however, the presentation of the adjoint method could still be
> improved, so I would suggest the authors to restructure that section.
> Altogether, I am confident that the manuscript will eventually merit
> publication in ApJS, but I would like to kindly ask the authors to
> address my detailed comments below.
>
> General comment:
> • For readers unfamiliar with the adjoint method, the intuitive
> meaning of the adjoint variables is somewhat opaque, and the adjoint
> equations are not well motivated in the main text. When the adjoint
> equation is introduced above l.239, it is unclear how the adjoint
> variables xi and pi relate to x, p, and to the objective function J,
> and why they are needed in the first place. I would suggest to state
> the initial conditions for the adjoint variables xi and pi (i.e.
> ∂J/∂x_n and ∂J/∂p_n) already before the detailed equations for their
> evolution. In this way, it immediately becomes clear that xi and pi
> are needed for tracking the gradient of the objective function J
> w.r.t. to x and p backwards from the final time to the initial
> conditions. Also, it might be more pedagogical to start Sec. 2.5 with
> Eq. (20) to emphasize that your aim is to compute the total derivative
> dJ/dθ, which can be expressed in terms of auxiliary variables (namely
> the adjoint variables) and only then provide the detailed equations
> (18) that the adjoint variables satisfy. It should be highlighted
> that, crucially, the definition of the adjoint variables depends on
> the objective J (and changing the objective requires re-computing the
> backward pass of the simulation with a different set of adjoint
> variables).

We have expanded the introduction section extensively with a high-level
introduction of the adjoint method, including some intuition on the
adjoint variables, the adjoint equations, their initial conditions, and
the gradients on parameters. We also think following through the
pedagogical development in the previous App. B is probably the best way
to get a better understanding of the method, since the formalism there
is both simpler and more generic than its specific application in Sec.
2.5. So we have moved that appendix before the previous Sec. 2.5 and
merge them into a new Sec. 3.

> Comments:
> • Introduction: the authors emphasize the use of automatic
> differentiation in their framework (as compared to analytic
> derivatives used in earlier works). For completeness, could the
> authors please add a brief explanation (1-2 sentences) of how AD works
> and how it compares to analytic and numerical differentiation?

Explanations of the analytic and automatic differentiations are added.

> • l.67: Could the authors please define the “discretize-then-optimize”
> approach (in comparison to the “optimize-then-discretize” approach)?

Explanations of both approaches added.

> • l.192: When the authors mention that FastPM chooses G_D and G_K
> “according to the ZA growth history”, they should make the connection
> to the 2LPT equation (2) to emphasize that the ZA is actually 1LPT.

Connection added.

> • Sec. 2.4: The observations “on our past light cone” seem somewhat of
> an overkill because the authors only consider a toy example in the
> results section, where the observable is given by the density field in
> the simulation box at final time. But I guess the authors are planning
> on using their framework in more complex settings in future work, so
> this is probably fine.

Indeed, we have derived the formalism for the light cone in Sec. 2.4 and
Sec. 3.4-3.6. And we have implemented snapshots as pmwd "observables"
and hopefully will have on-the-fly light cones and even post-Born ray
tracing available very soon.

> • Sec. 2.5: The authors should clearly state the meaning of the
> adjoint variables when they appear for the first time (see my comment
> above). In fact, this should already be mentioned in the introduction.
> I think it would already provide a lot of intuition for the somewhat
> abstract notion of adjoint variables and equations if the authors
> explained earlier in the paper that the adjoint variables are
> initialized as the gradient of the objective function J w.r.t. the
> state variables (x, p), i.e. the first two expressions in Eq. (19).
> Note that the expression that the adjoint method aims to compute,
> namely dJ/dθ, only appears at the very end of Sec. 2.5, rather than at
> the beginning.

Clarified in the introduction.

> • l.239: “...and left the ∂J/∂z_i term to the O operator”: based on
> the main text alone, it is unclear which ∂J/∂z_i term the authors are
> referring to here.

Some intuition added.

> • l.262: “Also note that θ can stand for both itself and ω, which we
> have omitted for brevity”: Given that the authors emphasize the
> distinction between θ and ω in Fig. 1, I would suggest changing the
> notation to e.g. θ* := (θ, ω) for clarity and consistency.

Equation for ω added explicitly.

> • l.280: “Because GPUs are inherently parallel devices, they can
> output different results for identical inputs.” This is not clear to
> me – where does the non- determinism come from specifically? I was
> under the impression that at least in the newer versions of Jax (or
> when setting TF_CUDNN_DETERMINISTIC) Jax is deterministic even on a
> GPU, see the discussion here:
> https://github.com/google/jax/issues/565. Could the authors please
> comment on this?

The non-determinism result for example from the different orders in
reduction operations among multiple processors. Enforcing determinism on
GPUs hurt the performance of those opertators. Some later issue
(https://github.com/google/jax/issues/13672) mentioned the flag to set
is now XLA_FLAGS='--xla_gpu_deterministic_ops=true'. A test run on a
problem 32768 times smaller than that in Fig. 2 (with 16^3 particles)
takes 10s to finish, so the deterministic run seems be more than 10^5
times slower (also accounting for the logN factor).

> • l.296: “... in practice the reconstructed trajectory can differ from
> the forward one due to loss of small-scale information”. What exactly
> do the authors mean by “small-scale information” here? As the authors
> state, the N-body equations themselves are time- reversible, so any
> non-reversibility should be caused by numerical errors, and the
> authors should be more specific as to what exactly causes the
> non-reversibility.

Clarified following the suggestion.

> • Tables 1 and 2: since the number of particles seems to be fixed at
> 384^3 in the reproducibility and forward-vs-backward experiments, the
> authors vary the particle mass by effectively varying the box size if
> I am not mistaken. For completeness, it would be good to state the box
> size that belongs to each particle mass.

Clarified following the suggestion. The pmwd API actually ask user to
provide the Lagrangian particle spacing first, which is directly related
to the particle mass. This allows us to specify a non-cubic box more
easily.

> • Table 2: The relative difference in the backwards-in-time
> displacement and velocity field towards the 2LPT ICs can be quite
> large, e.g. 5.2% / 7.1% for the first case. It would be interesting to
> mention how this error compares to the magnitude of the 2LPT
> displacement component s^(2) in Eq. (3). For example, if one starts
> with 2LPT at z = z_ini, evolves the system forward in time to z = 0,
> then goes backwards in time again and finally ends up with an error on
> the same order of magnitude as the 2LPT correction to the Zel’dovich
> term or even larger at z = z_ini, it might be questionable if the
> accuracy of the backward evolution is sufficient.

An interesting point. Since high-order LPTs affect the forward model
accuracy of the late time observables, and reversibility affects the
gradient accuracy on the simulation inputs, it seems to me that they are
orthogonal effects and a fair comparison is more complicated than one at
the 2LPT level. Ideally, we want both accuracy. And we are implementing
the 3LPT which can directly help the former, and indirectly the latter
too. We plan to test their impacts on the inference in future works.

> • Caption of Table 2: “Their smallness contributes partly to the fact
> that the quoted relative differences here being much greater than
> those in Table 1.” If this is indeed the problem, the authors could
> modify / expand their time reversibility experiment and evolve their
> backwards-in-time solution at the initial time once again forward in
> time and compare the difference at the final time (i.e. forward vs.
> forward-backward- forward) to make Tables 1 and 2 more comparable.

A fairer and more direct comparison is actually already in Fig. 3, where
the adjoint gradients suffer from both the reproducibility and the
reversibility, while the AD results only from the former. We have edited
the caption and the text to highlight this.

> • l.307: in the AD vs. adjoint gradient comparison, it is unclear to
> me which gradients are compared. From the text, I understand the
> gradients are w.r.t. the white noise field and the cosmological
> parameters, but what is the objective function J for this comparison?
> Please add this information.

Added in the caption of Fig. 3.

> • l.314: “we perform both adjoint and AD runs for 8 times”: I assume
> these 8 runs are done with the same cosmological parameters, but with
> different white noise fields – could the authors please comment on
> this and add this information to their manuscript?

Actually we repeated 8 times with the same cosmology and the same white
noise modes. This allows us to compare not only just the adjoint and AD
gradients, but also the adjoint or AD with itself, i.e., the
reproducibility of either gradients. We have clarified this in the text.

> • l.323: “possible breaking of order- or time-invariance in AD”: what
> exactly do the authors mean by this (in particular by “order
> invariance in AD”)?

This is related to the last point and should hopefully be clearer now.
The 8 AD runs give slightly different results that correlate with their
order, so that their asymmetric differences has asymmetric tails in the
histogram. However, with a new combination of JAX, CUDA and GPU, this
problem has disappeared, which we have further verified by increasing
the repetition to 64 times. So we have removed this comment.

> • Fig. 3: Could the authors please provide a description of the
> red-blue graphic and also add a color bar?

Description of the upper panel and color bar are added.

> • Fig. 4: The toy example is very illustrative; however, it would be
> interesting to also see the optimized ICs (i.e., the ICs for which the
> forward simulation leads to the “pmwd” pattern) – I encourage the
> authors to add them to the figure.

We found that the variance of the modes became bigger than 1 (that of
the standard normal white noise) after the optimization. We plot the
projected mean and standard deviation of these modes confirming that
trend. In addition, the mean figures show some level of spatial
correlation not present in the white noise modes. We have added these
figures to the repository
(https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint
optim_modes_{mean,std}_{0,10,100,1000}.pdf) and summarize the finding in
the text.

> Minor comments:
> • l.40: “rapid advances ... opens” → “open”
> • l.53: “..., thus requires saving the states...” → “..., which
>   requires saving the states...”?
> • l.56: “much smaller than” → “much less than”?
> • l.62: “these variables evolves” → “evolve”
> • l.192: “thereby improves” → “thereby improving”
> • l.245: “but enter” → “but enters”
> • l.256: “we can accumulate objective gradient following” → “the
>   objective gradient”?
> • l.271: it seems as if an equal sign “=” is missing here: “θ(As = 2 ×
>   10−9” → “θ = (As = ...)
> • l.293: “To save memory, trajectory of the model state” → “the
>   trajectory”?
> • Caption of Table 2: “... that the quoted relative differences here
>   being much greater than those in Table 1.” → “...are much greater”?
> • l.313: “To compensate for good statistics” → this sounds somewhat
>   odd to me (shouldn’t it rather be “to compensate for bad
>   statistics”?), maybe reformulate
> • l.318: “agree very well, and so are their gradients” → “so do their
>   gradients”?
> • l.441: “...using the facts that d(M^{−1}) = −M^{−1} dM M^{−1}”: it
>   took me a second to realize that M stands for an arbitrary matrix
>   here – maybe mention this briefly
> • l.444: “when one optimize” → “optimizes”

Corrected following the referee's suggestions.

We have also updated the results, especially Fig. 5, using an H100 GPU
instead of the previous A100.

> Data/Software Review:

We thank the data editor for suggestions. And YL apologize for the delay
in this response, due to mostly job related reasons.

> I. It is not clear the state of the pmwd JOSS submission. Is there a
> submission/review link started at JOSS? Do you wish this to be a joint
> AAS+JOSS review article [1]? This information would be useful both for
> the editing of the manuscript and for the eventual reader to
> understand how pmwd relates to the ApJS article. Please reply to this
> in the reply to science editor indicating if you want these journal
> articles related.
> [1] https://blog.joss.theoj.org/2018/12/a-new-collaboration-with-aas-publishing

I have not submitted to JOSS yet as there are still some unit testing,
documentation, and cleaning to be done. This is also my first AAS+JOSS
attempt. Should the two reviewing processes be coordinated? I will do my
best to submit to JOSS as soon as possible.

> II. I would suggest depositing the specific scripts related to this
> ApJS submission into a DOI-issuing repository and linking the text to
> that DOI -- in addition to, but not replacing the current GitHub
> script link. More information on this can be found in our tutorial [2]
> [2] https://github.com/AASJournals/Tutorials/tree/master/Repositories

To accommodate some new features, we plan to change the API slightly
in the following weeks. I will then modify the scripts accordingly for
the permanent repository and provide the DOI.

### 2023-11-02 docs/papers/adjoint/grads.py

@@ -27,11 +27,7 @@ def model(modes, cosmo, conf):
 
 
 def obj(tgt_dens, modes, cosmo, conf):
-    cosmo = boltzmann(cosmo, conf)
-    modes = linear_modes(modes, cosmo, conf)
-    ptcl, obsvbl = lpt(modes, cosmo, conf)
-    ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
-    dens = scatter(ptcl, conf)
+    dens = model(modes, cosmo, conf)
     return (dens - tgt_dens).var()
 
 obj_grad = jax.grad(obj, argnums=(1, 2))
@@ -65,7 +61,7 @@ if not os.path.exists(fname):
 modes = jnp.load(fname)
 
 
-n = 8
+n = 64
 fname_am = 'grads_am{}.npy'  # adjoint mode gradients
 fname_ad = 'grads_ad{}.npy'  # AD mode gradients
 
@@ -89,8 +85,9 @@ else:  # making plots
     from matplotlib.colors import SymLogNorm, LogNorm
     plt.style.use('adjoint.mplstyle')
 
-    fig, _ = simshow(gam[0, 32], cmap='RdBu_r',
-                     norm=SymLogNorm(0.01, vmin=-0.1, vmax=0.1), colorbar=False)
+    fig, _ = simshow(gam[0, 32], figsize=(3.5, 2.7), cmap='RdBu_r',
+                     norm=SymLogNorm(0.01, vmin=-0.1, vmax=0.1), colorbar=True,
+                     interpolation='none')
     fig.savefig('grads.pdf')
     plt.close(fig)
 
@@ -128,7 +125,7 @@ else:  # making plots
     print('AD-AD:  ', gd.std())
     ax.hist(gd, color='gray', ls=':', lw=1.5, label='AD-AD', **kwargs)
     ax.set_xlim(bins[0], bins[-1])
-    ax.set_ylim(5e2, 5e7)
+    ax.set_ylim(2e4, 5e9)
     ax.set_yscale('log')
     ax.ticklabel_format(axis='x', scilimits=(0, 0))
     ax.tick_params(axis='y', which='minor', left=False, right=False)

### 2023-11-02 docs/papers/adjoint/optim.py

@@ -23,7 +23,8 @@ def model(modes, cosmo, conf):
     modes = linear_modes(modes, cosmo, conf)
     ptcl, obsvbl = lpt(modes, cosmo, conf)
     ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
-    return ptcl
+    dens = scatter_2d(ptcl, conf)
+    return dens
 
 
 def scatter_2d(ptcl, conf):
@@ -32,12 +33,13 @@ def scatter_2d(ptcl, conf):
     return dens.sum(axis=2)
 
 
-pmwdshow = partial(simshow, figsize=(27/2, 8), cmap='cividis', colorbar=False)
+pmwdshow = partial(simshow, figsize=(2.7, 1.6), cmap='cividis', colorbar=False,
+                   interpolation='none')
+modeshow = partial(simshow, figsize=(3.2, 1.6), colorbar=True, interpolation='none')
 
 
 def obj(tgt, modes, cosmo, conf):
-    ptcl = model(modes, cosmo, conf)
-    dens = scatter_2d(ptcl, conf)
+    dens = model(modes, cosmo, conf)
     return (dens - tgt).var() / tgt.var()
 
 obj_valgrad = jax.value_and_grad(obj, argnums=1)
@@ -62,7 +64,7 @@ def optim(tgt, modes, cosmo, conf, iters=100, lr=0.1):
 
 
 text = 'pmwd'
-font = ImageFont.truetype('../nova/NovaRoundSlim-BookOblique.ttf', 32)
+font = ImageFont.truetype('../../nova/NovaRoundSlim-BookOblique.ttf', 32)
 
 ptcl_spacing = 10.
 ptcl_grid_shape = (16, 27, 16)
@@ -87,6 +89,7 @@ cosmo = boltzmann(cosmo, conf)
 
 
 plt.style.use('adjoint.mplstyle')
+plt.rcParams['savefig.pad_inches'] = 0
 
 ptcl, obsvbl = lpt(linear_modes(modes, cosmo, conf), cosmo, conf)
 fig, _ = pmwdshow(scatter_2d(ptcl, conf))
@@ -98,22 +101,32 @@ fig, _ = pmwdshow(scatter_2d(ptcl, conf))
 fig.savefig('optim_0.pdf')
 plt.close(fig)
 
[…]

### 2023-11-02 docs/papers/adjoint/perf.py

@@ -1,6 +1,7 @@
 import os
 
 import numpy as np
+import jax
 import matplotlib.pyplot as plt
 
 
@@ -9,7 +10,7 @@ ptcl_grid_sizes = [2**n for n in range(5, 10)]
 
 
 number = 1
-repeat = 8
+repeat = 16
 exec_times = np.empty((2 + len(mesh_shapes), len(ptcl_grid_sizes)))
 fname = 'perf.txt'
 
@@ -39,28 +40,26 @@ if not os.path.exists(fname):
             modes = white_noise(seed, conf)
 
 
-            boltzmann(cosmo, conf)  # do not overwrite cosmo
-            t = timeit.repeat('boltzmann(cosmo, conf).h.block_until_ready()',
+            t = timeit.repeat('jax.block_until_ready(boltzmann(cosmo, conf, varlin=False))',
                               repeat=repeat, number=number, globals=globals())
             t = min(t) / number
+            print(f'growth {t}')
             exec_times[0, j] = t
-            print(f'boltz {t}')
-            cosmo = boltzmann(cosmo, conf)  # overwrite cosmo
+            cosmo = boltzmann(cosmo, conf)
 
 
-            ptcl, obsvbl = lpt(linear_modes(modes, cosmo, conf), cosmo, conf)  # do not overwrite modes
-            t = timeit.repeat('lpt(linear_modes(modes, cosmo, conf), cosmo, conf)[0].vel.block_until_ready()',
+            t = timeit.repeat('jax.block_until_ready(lpt(linear_modes(modes, cosmo, conf), cosmo, conf))',
                               repeat=repeat, number=number, globals=globals())
             t = min(t) / number
-            print(f'  lpt {t}')
+            print(f'   lpt {t}')
             exec_times[1, j] = t
+            ptcl, obsvbl = lpt(linear_modes(modes, cosmo, conf), cosmo, conf)
 
 
-            nbody(ptcl, obsvbl, cosmo, conf)  # do not overwrite ptcl and obsvbl
-            t = timeit.repeat('nbody(ptcl, obsvbl, cosmo, conf)[0].vel.block_until_ready()',
+            t = timeit.repeat('jax.block_until_ready(nbody(ptcl, obsvbl, cosmo, conf))',
                               repeat=repeat, number=number, globals=globals())
             t = min(t) / number / (1 + conf.a_nbody_num)
-            print(f'nbody {t}')
+            print(f' nbody {t}')
             exec_times[2+i, j] = t
 
     np.savetxt(fname, exec_times)

### 2023-11-02 docs/papers/adjoint/slab.py

@@ -18,7 +18,8 @@ def model(modes, cosmo, conf):
     modes = linear_modes(modes, cosmo, conf)
     ptcl, obsvbl = lpt(modes, cosmo, conf)
     ptcl, obsvbl = nbody(ptcl, obsvbl, cosmo, conf)
-    return ptcl, obsvbl
+    dens = scatter(ptcl, conf)
+    return dens
 
 
 ptcl_spacing = 1.
@@ -30,9 +31,8 @@ cosmo = SimpleLCDM(conf)
 seed = 0
 modes = white_noise(seed, conf)
 
-ptcl, obsvbl = model(modes, cosmo, conf)
+dens = model(modes, cosmo, conf)
 
-dens = scatter(ptcl, conf)
 fig, _ = simshow(dens[:16].mean(axis=0), norm='CosmicWebNorm')
 fig.savefig('slab.pdf')
 plt.close(fig)

### 2023-11-09 pmwd/cuda_ops.py
workspace memory managed in jax breaks, moving to cuda memory for now. Need to figure out the new way to use workspace memory managed in jax side

@@ -2,17 +2,15 @@ __all__ = ["scatter_cuda", "gather_cuda"]
 
 from functools import partial
 import numpy as np
-import jaxlib.mlir.ir as ir
+from jax.interpreters.mlir import ir
 from jax import core, dtypes, lax
 from jax.core import Primitive
 from jax import numpy as jnp
 from jax.core import ShapedArray
-from jax._src.interpreters import ad, batching, mlir, xla
+from jax.interpreters import mlir,xla
 from jax import jit
 from jax.lib import xla_client
 from jaxlib.hlo_helpers import custom_call
-from jax._src.lib.mlir.dialects import hlo
-from jax._src import dispatch
 from . import _jaxpmwd
 
 # Registering ops for XLA
@@ -58,7 +56,7 @@ def _scatter_lowering(ctx, pmid, disp, val, mesh, *, offset, ptcl_grid, ptcl_spa
     else:
         raise NotImplementedError(f"Unsupported dtype {np_dtype}")
 
-    workspace = mlir.full_like_aval(ctx, 0, core.ShapedArray(shape=[workspace_size], dtype=np.byte))
+    #workspace = mlir.full_like_aval(ctx, 0, ShapedArray(shape=[workspace_size,], dtype=np.byte))
 
     # And then the following is what changes between the GPU and CPU
     if platform == "cpu":
@@ -70,27 +68,30 @@ def _scatter_lowering(ctx, pmid, disp, val, mesh, *, offset, ptcl_grid, ptcl_spa
             )
 
         # TODO: if we use shared mem with bin sort, bin sort work mem allocate by XLA here and pass to cuda
-        result = custom_call(
+        #result = custom_call(
+        return custom_call(
             op_name,
             # Output types
+            #result_types=[out_type, ir.RankedTensorType.get([workspace_size],ir.IntegerType.get_signless(8))],
             result_types=[out_type],
             # The inputs:
-            operands=[pmid,disp,val,mesh,workspace],
+            operands=[pmid,disp,val,mesh],
             # Layout specification:
-            operand_layouts=[in_layout1, in_layout1, in_layout2, out_layout, (0,)],
+            operand_layouts=[in_layout1, in_layout1, in_layout2, out_layout],
             result_layouts=[out_layout],
             operand_output_aliases={3:0},
[…]

### 2023-11-09 pmwdlib/kernels.cc.cu

@@ -674,15 +674,19 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
     uint32_t *pmid = reinterpret_cast<uint32_t *>(buffers[0]);
     T *disp = reinterpret_cast<T *>(buffers[1]);
     T *particle_values = reinterpret_cast<T *>(buffers[2]);
-    void *work_d = buffers[4];
-    T *grid_values = reinterpret_cast<T *>(buffers[5]);
-    char *work_i_d = static_cast<char *>(work_d);
+    T *grid_values = reinterpret_cast<T *>(buffers[4]);
+    //void *work_d = buffers[5];
+    //char *work_i_d = static_cast<char *>(work_d);
+    char *work_i_d;
+
 
     // parameters for shared mem using bins to group cells
     uint32_t bin_size = BINSIZE;
     uint32_t nbinx = static_cast<uint32_t>(std::ceil(1.0*stride[0]/bin_size));
     uint32_t nbiny = static_cast<uint32_t>(std::ceil(1.0*stride[1]/bin_size));
     uint32_t nbinz = static_cast<uint32_t>(std::ceil(1.0*stride[2]/bin_size));
+
+    CUDA_SAFE_CALL(cudaMalloc((void**)&work_i_d, sizeof(uint32_t)*(n_particle*4+2*nbinx*nbiny*nbinz+1)+temp_storage_bytes));
 
     uint32_t npts_mem_size = sizeof(uint32_t) * n_particle;
     uint32_t nbins_mem_size = sizeof(uint32_t) * nbinx*nbiny*nbinz;
@@ -797,6 +801,7 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
     cudaEventElapsedTime(&milliseconds, start, stop);
     printf("cuda kernel scatter: %f milliseconds\n", milliseconds);
 #endif
+    cudaFree(work_i_d);
 
 }
 
@@ -814,9 +819,9 @@ void gather_sm(cudaStream_t stream, void** buffers, const char* opaque, std::siz
 
     // parameters for shared mem using bins to group cells
     uint32_t bin_size = BINSIZE;
-    uint32_t nbinx = stride[0]/bin_size+1;
-    uint32_t nbiny = stride[1]/bin_size+1;
-    uint32_t nbinz = stride[2]/bin_size+1;
[…]

### 2023-11-09 docs/papers/adjoint/adjoint.tex

@@
-\documentclass[modern, trackchanges, dvipsnames]{aastex631}
+\documentclass[modern, dvipsnames]{aastex631}
   \urlstyle{sf}
@@
 
-\newcommand{\GPU}{NVIDIA A100 80GB SXM4}
-
-\newcommand{\HL}[1]{\textcolor{Bittersweet}{#1}}
+\newcommand{\GPU}{NVIDIA H100 PCIe}
+
+\newcommand{\HL}[1]{\textcolor{Bittersweet}{\textbf{#1}}}
 
@@
 with differentiable simulations.
-Current differentiable cosmological simulations are limited by memory,
-thus are subject to a trade-off between time and space/mass resolution.
-They typically integrate for only tens of time steps, unlike the
-standard non-differentiable simulations.
+Based on analytic or automatic backpropagation, current differentiable
+cosmological simulations are limited by memory, and thus are subject to
+a trade-off between time and space/mass resolution, usually sacrificing
+both.
 We present a new approach free of such constraints, using the adjoint
 method and reverse time integration.
-It enables larger and more accurate forward modeling, and will improve
-gradient based optimization and inference.
-We implement it in a particle-mesh (PM) $N$-body library \pmwd\
-(particle-mesh with derivatives).
+It enables larger and more accurate forward modeling at the field level,
+and will improve gradient based optimization and inference.
+We implement it in an open-source particle-mesh (PM) $N$-body library
+\pmwd\ (particle-mesh with derivatives).
 Based on the powerful AD system \texttt{JAX}, \pmwd\ is fully
@@
 on conventional CPU clusters, but rapid advances in accelerator
-technology like GPUs opens the possibility of simulation-based modeling
+technology like GPUs open the possibility of simulation-based modeling
 and inference \citep{CranmerEtAl2020}.
@@
 and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
-advent of modern AD systems, and were based on implementations of
-analytic derivatives \citep[see e.g.,][App.~D]{SeljakEtAl2017}).
+advent of modern AD systems, and were based on \HL{the analytic derivatives,
+which involve first convoluted derivation by hand using the chain rule
+\citep[see e.g.,][App.~D]{SeljakEtAl2017} before implementing them in
+code.}
 Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
-vmad, SeljakEtAl2017, FlowPM} compute gradients using AD engines, namely
-\texttt{vmad} (written by the same authors) and \texttt{TensorFlow},
-respectively.
+vmad, SeljakEtAl2017, FlowPM} compute gradients using the AD engines,
+namely \texttt{vmad} (written by the same authors) and
+\texttt{TensorFlow}, respectively.
+\HL{The AD frameworks automatically apply the chain rule to the primitive
+operations that comprise the whole simulation, relieving the burden of
+derivation and implementation of the derivatives.}
 Both analytic differentiation and AD backpropagate the gradients through
-the whole history, thus requires saving the states at all time steps in
+the whole history, which requires saving the states at all time steps in
 memory.
 Therefore, they are subject to a trade-off between time and space/mass
-resolution, and typically can integrate for only tens of time steps,
-much smaller than the standard non-differentiable simulations.
+resolution, usually sacrificing both.
 As a result, they lose accuracy on small scales and in dense regions
@@
 
-Alternatively, the adjoint method provides systematic ways of deriving
-model gradients under constraints \citep{Pontryagin1962}, such as those
-imposed by the $N$-body equations of motion in a simulated Universe.
-It identifies a set of adjoint variables, dual to the state variables in
-the simulation.
-For time-dependent problems, these variables evolves backward in time by
-a dual set of equations, known as the adjoint equations.
-Their evolution depends on the states in the forward run, which can be
-re-simulated with reverse time integration if the dynamics is
-reversible, thereby dramatically reducing the memory cost
+\HL{Alternatively, the adjoint method provides systematic ways of deriving
+the gradients of an objective function $\cJ$ under constraints
+\citep{Pontryagin1962}, such as those imposed by the $N$-body equations
+of motion in a simulated Universe.
+It identifies a set of \emph{adjoint variables} $\vlambda$, dual to the
+state variables $\vz$ of the model, and carrying the gradient
+information of the objective function with respect to the model state
+$\p\cJ / \p\vz$.
+For time-dependent problems, the adjoint variables evolve
+\emph{backward} in time by a set of equations dual to that of the
+forward evolution, known as the \emph{adjoint equations}.
+For continuous time, the adjoint equations are a set of differential
+equations, while in the discrete case they become difference equations
+which are practically a systematic way to structure the chain rule or
+backpropagation.
+Their initial conditions are set by the explicit dependence of the
+objective on the simulation state, e.g., $\vlambda_n = \p\cJ / \p\vz_n$
+if $\cJ$ is a function of the final state $\vz_n$.
+Solving the adjoint equations can help us to propagate the gradient
+information via the adjoint variables to the input parameters $\vtheta$,
+to compute the objective gradients $\d\cJ / \d\vtheta$.
+And we will see later that the propagated and accumulated gradients on
+parameters come naturally from multiple origins, each reflecting a
+$\vtheta$-dependence at one stage of the modeling in
+\autoref{fig:model}.}
+
+The backward adjoint evolutions depend on the states in the forward run,
+which we can re-simulate with reverse time integration if the dynamics
+is reversible, thereby dramatically reducing the memory cost
 \citep{NeuralODE}.
-Furthermore, we take the discretize-then-optimize approach
-\citep[e.g.,][]{ANODE} to ensure gradients propagate backward along the
-\emph{same} discrete trajectory as taken by the forward time
+Furthermore, we \HL{derive the discrete adjoint equations dual to the
+discrete forward time integration, known as the discretize-then-optimize
+approach \citep[e.g.,][]{ANODE}, to ensure gradients propagate backward
+along the \emph{same} trajectory as taken by the forward time
 integration.
+This is in contrast with the optimize-then-discretize approach, that
+numerically integrates the continuous adjoint equations, and is prone to
+larger deviations between the forward and the backward trajectories due
+to different discretizations \citep{LanzieriLanusseEtAl2022}.}
 In brief, to compute the gradient we only need to evolve a simulation
 forward, and then backward jointly with its dual adjoint equations.
-We introduce the adjoint method for generic time-dependent problems in
-both continuous and discrete cases in \autoref{app:adj}.
+We introduce the adjoint method first for generic time-dependent
+problems in both continuous and discrete cases in \autoref{sec:bwd}, and
+then present its application on cosmological simulation in
+\autoref{sec:adj}.
 
@@
 Single arrows from the cosmological parameters $\vtheta$ and white noise
-field $\vomega$ indicate dependence on $\vtheta$ only, while double
+modes $\vomega$ indicate dependence on $\vtheta$ only, while double
 arrows imply dependence on both.
@@
 expands the dashed box in \subref{fig:loop}.
-We describe different operators in \autoref{sec:method}: Boltzmann
-solver (``Boltz'') and initial condition generator (``IC'') in
+We describe different operators in \autoref{sec:fwd}: Boltzmann solver
+(``Boltz'') and initial condition generator (``IC'') in
 \autoref{sec:ic}; force solver ($F$) in \autoref{sec:force}; time
@@
 \vspace{1em}
-\section{Methods}
-\label{sec:method}
-
-We first review all components of the $N$-body simulation-based forward
-model of the cosmological structure formation, and then introduce the
-adjoint method for each part, and how they can be combined to
-differentiate a whole simulation at low memory cost.
+\section{Forward Simulation}
+\label{sec:fwd}
+
+We first review and formulate all components of the $N$-body
+simulation-based forward model of the cosmological structure formation.
 
@@
 $|\vnabla \cdot \vs|$, the linear approximation of the density
-fluctuation, is much smaller than 1.
+fluctuation, is much less than 1.
 We compute the initial displacements and momenta using the second order
@@
   = a^2 H \bigl( D_1' \vs^{(1)} + D_2' \vs^{(2)} \bigr),
-\label{ic}
+\label{lpt}
 \end{align}
@@
 %
-with the scalar potentials sourced by\footnote{Note the difference in
-the sign of the source of $\phi_\vs^{(2)}$ from the usual LPT
-convention.}
+with the scalar potentials sourced by
 %
@@
 as\footnote{Here the time dependence of the linear power, $\Plin(k, a) =
-\Plin(k) D_1^2(a)$, has been left to \eqref{ic}}
+\Plin(k) D_1^2(a)$, has been left to \eqref{lpt}}
 %
@@
 W(\vx, \vx') = \prod_{i=1}^3
-  \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_\mathrm{cell}}, 0 \Bigr),
+  \max \Bigl( 1 - \frac{|x_i-x'_i|}{l}, 0 \Bigr),
 \label{cic}
@@
 %
-where $l_\mathrm{cell}$ is the mesh cell size.
+where $l$ is the mesh cell size.
 
@@
 We use the \texttt{FastPM} time stepping \citep{FastPM}, designed to
-reproduce the first order Lagrangian perturbation theory, also known as
-the Zel'dovich approximation (hereafter ZA), in the linear regime.
+reproduce in the linear regime the linear Lagrangian perturbation
+theory, i.e., the 1LPT as the first order in \eqref{lpt}, also known as
+the the Zel'dovich approximation (hereafter ZA).
 We present a simplified derivation below.
@@
 \texttt{FastPM} chooses $G_D$ and $G_K$ according to the ZA growth
-history, thereby improves the accuracy on large scales and at early
+history, thereby improving the accuracy on large scales and at early
 times.
@@
 \end{align}
-%
 
@@
 Note that we only ``observe'' the dark matter phase space here, and
-leave more realistic observables to a future work, including the forward
+leave more realistic observables to future works, including the forward
 modeling of real observational effects.
@@
 \vspace{1em}
-\subsection{Adjoint Method}
+\section{Backward Differentiation -- the Adjoint Method}
+\label{sec:bwd}
+
+We first introduce the adjoint method for generic time-dependent ODEs,
+and derive the adjoint equations following a pedagogical tutorial by
+\citet{Bradley2019}.
+We then adopt the discretize-then-optimize approach, and derive the
+discrete adjoint equations, that is more suitable for the $N$-body
+symplectic time integration.
+Finally we apply them to derive the adjoint equations and the gradients
+for cosmological simulations described in \autoref{sec:fwd}, and couple
+them with the reverse time integration to reduce the space complexity.
+
+\vspace{1em}
+\subsection{Variational (Tangent) and Adjoint Equations}
+
+Consider a vector state $\vz(t)$ subject to the following ODEs and
+initial conditions
+%
+\begin{equation}
+\dot\vz = \vf(\vz, t, \vtheta),
+\qquad
+\vz(t_0) = \vz_0(\vtheta),
+\label{ode}
+\end{equation}
+%
+for $t \in [t_0, t_1]$.
+Here the initial conditions can depend on the parameters $\vtheta$.
+
+A perturbation in the initial conditions propagates forward in time.
+For $\vz(t, t_0, \vz_0)$, the Jacobian of state variables describing
+this,
+%
+\begin{equation}
+\vDelta = \frac{\p\vz}{\p\vz_0},
+\end{equation}
+%
+evolves from identity $\vDelta_0 = \vI$ by
+%
+\begin{equation}
+\dot\vDelta = \frac{\p\vf}{\p\vz} \cdot \vDelta,
+\label{variational}
+\end{equation}
+%
+following from \eqref{ode}.
+\eqref{variational} is known as the \emph{variational or tangent
+equation}.
+
+The backward version of $\vDelta$,
+%
+\begin{equation}
+\vLambda = \frac{\p\vz_1}{\p\vz},
+\end{equation}
+%
+evolves backward in time from identity $\vLambda_1 = \vI$ by
+%
+\begin{equation}
+\dot\vLambda = - \vLambda \cdot \frac{\p\vf}{\p\vz}.
+\label{adjoint}
+\end{equation}
+%
+\eqref{adjoint} is called the \emph{adjoint equation}, for the
+right-hand side is $(\p\vf / \p\vz)^\intercal \vLambda$.
+It can be derived from the time-invariance of $\vLambda \cdot \vDelta$:
+%
+\begin{equation}
+\vLambda \cdot \dot\vDelta + \dot\vLambda \cdot \vDelta
+= \frac\d{\d t} (\vLambda \cdot \vDelta)
+= \frac\d{\d t} \Bigl( \frac{\p\vz_1}{\p\vz}
+  \cdot \frac{\p\vz}{\p\vz_0} \Bigr)
+= \frac\d{\d t} \frac{\p\vz_1}{\p\vz_0}
+= \vzero.
+\end{equation}
+
+Alternatively, the adjoint equation can be derived from the variational
+equation, using the facts that $\d \bm{M}^{-1} / \d t = - \bm{M}^{-1} \,
+\dot{\bm{M}} \, \bm{M}^{-1}$ \HL{for any invertible matrix $\bm{M}$}, and
+$(\p\vz / \p\tilde\vz)^{-1} = \p\tilde\vz / \p\vz$.
+Like \eqref{variational} and \eqref{adjoint}
+%
+\begin{equation}
+\frac\d{\d t} \frac{\p\vz}{\p\tilde\vz}
+= \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\tilde\vz},
+\qquad
+\frac\d{\d t} \frac{\p\tilde\vz}{\p\vz}
+= - \frac{\p\tilde\vz}{\p\vz} \cdot \frac{\p\vf}{\p\vz}.
+\end{equation}
+%
+As we can see next, the adjoint equation takes a similar form when one
+optimizes an objective function of the state.
+
+\vspace{1em}
+\subsection{Objective on the Final State}
+
+In the simplest case, the objective function depends only on the final
+state, e.g., the last snapshot of a simulation, and possibly the
+parameters too in the form of regularization or prior information, i.e.,
+$\cJ(\vz_1, \vtheta)$.
+To optimize the objective under the constraint given by the ODEs, we can
+introduce a time-dependent function $\vlambda(t)$ as the Lagrange
+multiplier:
+%
+\begin{equation}
+\cL = \cJ(\vz_1, \vtheta)
+- \int_{t_0}^{t_1} \vlambda(t)
+  \cdot \bigl[ \dot\vz - \vf(\vz, t, \vtheta) \bigr] \, \d t.
+\end{equation}
+%
+Note the minus sign we have introduced in front of $\vlambda$ for later
+convenience.
+
+Its total derivative with respect to $\vtheta$ is
+%
+\begin{equation}
+\frac{\d\cL}{\d\vtheta}
+= \frac{\p\cJ}{\p\vtheta}
++ \frac{\p\cJ}{\p\vz_1} \cdot \frac{\p\vz_1}{\p\vtheta}
+- \int_{t_0}^{t_1} \vlambda
+  \cdot \Bigl( \frac{\p\dot\vz}{\p\vtheta}
+    - \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
+    - \frac{\p\vf}{\p\vtheta} \Bigr) \, \d t.
+\end{equation}
+%
+Integrating the first term of the integrand by parts:
+%
+\begin{equation}
+\int_{t_0}^{t_1} \vlambda \cdot \frac{\p\dot\vz}{\p\vtheta} \, \d t
+= \int_{t_0}^{t_1} \vlambda \cdot \frac{\p^2\vz}{\p\vtheta\p t} \, \d t
+= \Bigl[ \vlambda \cdot \frac{\p\vz}{\p\vtheta} \Bigr]_{t_0}^{t_1}
+- \int_{t_0}^{t_1} \dot \vlambda \cdot \frac{\p\vz}{\p\vtheta} \, \d t,
+\end{equation}
+%
+and plugging it back:
+%
+\begin{equation}
+\frac{\d\cL}{\d\vtheta}
+= \frac{\p\cJ}{\p\vtheta}
++ \Bigl( \frac{\p\cJ}{\p\vz_1} - \vlambda_1 \Bigr)
+  \cdot \frac{\p\vz_1}{\p\vtheta}
++ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \int_{t_0}^{t_1} \Bigl[
+  \Bigl( \vlambda \cdot \frac{\p\vf}{\p\vz} + \dot\vlambda \Bigr)
+  \cdot \frac{\p\vz}{\p\vtheta}
+  + \vlambda \cdot \frac{\p\vf}{\p\vtheta} \Bigr] \, \d t.
+\end{equation}
+
+Now we are free to choose
+%
+\begin{equation}
+\dot\vlambda = - \vlambda \cdot \frac{\p\vf}{\p\vz},
+\qquad
+\vlambda_1 = \frac{\p\cJ}{\p\vz_1},
+\label{adjoint1}
+\end{equation}
+%
+that allow us to avoid all $\p\vz/\p\vtheta$ terms in the final
+objective gradient:
+%
+\begin{equation}
+\frac{\d\cJ}{\d\vtheta}
+= \frac{\d\cL}{\d\vtheta}
+= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \int_{t_0}^{t_1} \vlambda \cdot \frac{\p\vf}{\p\vtheta} \, \d t,
+\end{equation}
+%
+in which the first two terms come from the regularization and initial
+conditions, respectively.
+\eqref{adjoint1} is the adjoint equation for the objective $\cJ(\vz_1,
+\vtheta)$.
+With the initial conditions set at the final time, we can integrate it
+backward in time to obtain $\vlambda(t)$, which enters the above
+equation and yields $\d\cJ / \d\vtheta$.
+
+Note that \eqref{adjoint1} has the same form as \eqref{adjoint}, and
+their solutions are related by
+%
+\begin{equation}
+\vlambda = \vlambda_1 \cdot \vLambda
+= \frac{\p\cJ}{\p\vz_1} \cdot \frac{\p\vz_1}{\p\vz}
+= \frac{\p\cJ}{\p\vz}.
+\end{equation}
+%
+And like $\vLambda \cdot \vDelta$, $\vlambda \cdot \vDelta$ is
+time-invariant:
+%
+\begin{equation}
+\vlambda_0 \cdot \vDelta_0
+= \vlambda_1 \cdot \vDelta_1
+= \vlambda \cdot \vDelta
+= \frac{\p\cJ}{\p\vz} \cdot \frac{\p\vz}{\p\vz_0}
+= \frac{\p\cJ}{\p\vz_0}
+= \vlambda_0.
+\end{equation}
+%
+Computing $\vlambda_1 \cdot \vDelta_1$ directly is expensive, but
+solving \eqref{adjoint1} backward for $\vlambda_0$ is cheap.
+This is related to the fact that the reverse-mode AD or backpropagation
+is cheaper than the forward mode for optimization.
+
+\vspace{1em}
+\subsection{Objective on the State History}
+
+The adjoint method applies to more complex cases too.
+Let's consider an objective as a functional of the evolution history
+with some regularization $\cR$ on $\vtheta$
+%
+\begin{equation}
+\cJ = \cR(\vtheta) + \int_{t_0}^{t_1} g(\vz, t, \vtheta) \, \d t.
+\end{equation}
+%
+The derivation is similar:
+%
+\begin{equation}
+\cL = \cR(\vtheta)
++ \int_{t_0}^{t_1} \bigl\{ g(\vz, t, \vtheta) -
+  \vlambda(t) \cdot \bigl[ \dot\vz - \vf(\vz, t, \vtheta) \bigr]
+  \bigr\} \, \d t
+\end{equation}
+%
+has gradient
+%
+\begin{align}
+\frac{\d\cL}{\d\vtheta}
+&= \frac{\d\cR}{\d\vtheta}
++ \int_{t_0}^{t_1} \Bigl[
+    \frac{\p g}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
+    + \frac{\p g}{\p\vtheta}
+    - \vlambda \cdot \Bigl( \frac{\p\dot\vz}{\p\vtheta}
+    - \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
+    - \frac{\p\vf}{\p\vtheta} \Bigr) \Bigr] \, \d t \\
+%
+&= \frac{\d\cR}{\d\vtheta}
+- \vlambda_1 \cdot \frac{\p\vz_1}{\p\vtheta}
++ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \int_{t_0}^{t_1} \Bigl[
+  \Bigl( \frac{\p g}{\p\vz} + \vlambda \cdot \frac{\p\vf}{\p\vz}
+    + \dot\vlambda \Bigr) \cdot \frac{\p\vz}{\p\vtheta}
+  + \frac{\p g}{\p\vtheta} + \vlambda \cdot \frac{\p\vf}{\p\vtheta}
+\Bigr] \, \d t. \nonumber
+\end{align}
+
+The adjoint equation becomes
+%
+\begin{equation}
+\dot\vlambda = - \frac{\p g}{\p\vz}
+  - \vlambda \cdot \frac{\p\vf}{\p\vz},
+\qquad
+\vlambda_1 = \vzero,
+\end{equation}
+%
+with the objective gradient
+%
+\begin{equation}
+\frac{\d\cJ}{\d\vtheta}
+= \frac{\d\cL}{\d\vtheta}
+= \frac{\d\cR}{\d\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \int_{t_0}^{t_1} \Bigl( \frac{\p g}{\p\vtheta}
+  + \vlambda \cdot \frac{\p\vf}{\p\vtheta} \Bigr) \, \d t.
+\end{equation}
+
+The time-invariant is now
+%
+\begin{equation}
+\vlambda_1 \cdot \vDelta_1
+= \vlambda_0 \cdot \vDelta_0
+  + \int_{t_0}^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t
+= \vlambda \cdot \vDelta
+  + \int_t^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t'
+= \vzero,
+\end{equation}
+%
+so
+%
+\begin{equation}
+\vlambda \cdot \vDelta
+= - \int_t^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t'
+= - \int_t^{t_1} \frac{\p g}{\p\vz_0} \, \d t'.
+\end{equation}
+
+\vspace{1em}
+\subsection{Objective on the Observables}
+\label{sec:lightcone}
+
+Now let's consider an objective that depends on different state
+components at different times, e.g., in the Universe where further
+objects intersected our past light cone earlier.
+It falls between the previous two scenarios, and we can derive its
+adjoint equation similarly.
+
+We denote the observables by $\hat\vz$, with different components $\hat
+z_j$ affecting the objective $\cJ(\hat\vz, \vtheta)$ at different times
+$\hat t_j$, i.e., $\hat z_j \triangleq z_j(\hat t_j)$.
+The Lagrangian becomes
+%
+\begin{equation}
+\cL = \cJ(\hat\vz, \vtheta)
+- \sum_j \int_{t_0}^{\hat t_j} \lambda_j(t)
+  \bigl[ \dot z_j - f_j(\vz, t, \vtheta) \bigr] \, \d t,
+\end{equation}
+%
+constraining only the parts of trajectories inside the light cone.
+Its gradient is
+%
+\begin{align}
+\frac{\d\cL}{\d\vtheta}
+&= \frac{\p\cJ}{\p\vtheta}
++ \frac{\p\cJ}{\p\hat\vz} \cdot \frac{\p\hat\vz}{\p\vtheta}
+- \sum_j \int_{t_0}^{\hat t_j}
+  \Bigl( \lambda_j \frac{\p\dot z_j}{\p\vtheta}
+    - \vlambda \cdot \frac{\p\vf}{\p z_j} \frac{\p z_j}{\p\vtheta}
+    - \lambda_j \frac{\p f_j}{\p\vtheta} \Bigr) \, \d t \\
+%
+&= \frac{\p\cJ}{\p\vtheta}
++ \Bigl( \frac{\p\cJ}{\p\hat\vz} - \hat\vlambda \Bigr)
+  \cdot \frac{\p\hat\vz}{\p\vtheta}
++ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \sum_j \int_{t_0}^{\hat t_j} \Bigl[
+  \Bigl( \vlambda \cdot \frac{\p\vf}{\p z_j} + \dot\lambda_j \Bigr)
+  \frac{\p z_j}{\p\vtheta}
+  + \lambda_j \frac{\p f_j}{\p\vtheta} \Bigr] \, \d t, \nonumber
+\end{align}
+%
+where we have defined $\hat\vlambda$ similarly with components
+$\hat\lambda_j \triangleq \lambda_j(\hat t_j)$.
+In the first equality, we have also dropped a vanishing term $\sum_j
+\hat\lambda_j \bigl[ \dot z_j - f_j(\hat t_j) \bigr] \p\hat t_j /
+\p\vtheta$, i.e., $\p\hat t_j / \p\vtheta$ does not directly enter the
+gradient.
+
+Now we find the adjoint equation
+%
+\begin{equation}
+\dot\vlambda = - \vlambda \cdot \frac{\p\vf}{\p\vz},
+\qquad
+\hat\vlambda = \frac{\p\cJ}{\p\hat\vz},
+\end{equation}
+%
+that has the same form as \eqref{adjoint1}, with a slightly different
+initial condition given at the respective observation time of each
+component.
+The objective gradient is also similar to the previous cases:
+%
+\begin{equation}
+\frac{\d\cJ}{\d\vtheta}
+= \frac{\d\cL}{\d\vtheta}
+= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \sum_j \int_{t_0}^{\hat t_j} \lambda_j \frac{\p f_j}{\p\vtheta} \, \d t.
+\end{equation}
+
+Note that even though $\lambda_j(t)$ for $t > \hat t_j$ does not affect
+the final gradient, they can enter the right-hand side of the adjoint
+equation, and affect those $\lambda_k$ with $t < \hat t_k$, i.e., inside
+the light cone.
+Physically however, $\p f_j / \p z_k$ should vanish for spacelike
+separated pairs of $z_j$ and $z_k$, even though the Newtonian
+approximation we adopt introduces some small deviation.
+Therefore, we can set $\lambda_j(t)$ to 0 for $t > \hat t_j$, and bump
+it to $\p\cJ / \p\hat z_j$ at $\hat t_j$.
+
+\vspace{1em}
+\subsection{Discretize Then Optimize}
+
+In practice, the time integration of \eqref{ode} is discrete.
+Consider the explicit methods,
+%
+\begin{equation}
+\vz_{i+1} = \vz_i + \vF_i(\vz_i, \vtheta), \qquad i = 0, \cdots, n-1,
+\label{explicit}
+\end{equation}
+%
+which include the leapfrog integrator commonly used for Hamiltonian
+dynamics.
+We want to propagate the gradients backward along the \emph{same}
+discrete trajectory as taken by the forward integration.
+Therefore, instead of the continuous adjoint equations derived above, we
+need the adjoint method for the discrete integrator.
+
+Without loss of generality, we derive the adjoint equation for an
+objective depending on the state at all time steps, which can be easily
+specialized to the 3 cases discussed above with only slight
+modifications.
+The discretized Lagrangian is now
+%
+\begin{equation}
+\cL = \cJ(\vz_0, \cdots, \vz_n, \vtheta)
+- \sum_{i=0}^{n-1} \vlambda_{i+1}
+  \cdot \bigl[ \vz_{i+1} - \vz_i - \vF_i(\vz_i, \vtheta) \bigr],
+\end{equation}
+%
+whose gradient is
+%
+\begin{align}
+\frac{\d\cL}{\d\vtheta}
+&= \frac{\p\cJ}{\p\vtheta}
++ \sum_{i=0}^n \frac{\p\cJ}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
+- \sum_{i=0}^{n-1} \vlambda_{i+1}
+  \cdot \Bigl( \frac{\p\vz_{i+1}}{\p\vtheta} - \frac{\p\vz_i}{\p\vtheta}
+    - \frac{\p\vF_i}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
+    - \frac{\p\vF_i}{\p\vtheta} \Bigr) \nonumber\\
+%
+&= \frac{\p\cJ}{\p\vtheta}
++ \sum_{i=0}^n \frac{\p\cJ}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
+- \sum_{i=1}^n \vlambda_i \cdot \frac{\p\vz_i}{\p\vtheta}
++ \sum_{i=0}^{n-1} \vlambda_{i+1}
+  \cdot \Bigl( \frac{\p\vz_i}{\p\vtheta}
+    + \frac{\p\vF_i}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
+    + \frac{\p\vF_i}{\p\vtheta} \Bigr) \nonumber\\
+&= \frac{\p\cJ}{\p\vtheta}
++ \Bigl( \frac{\p\cJ}{\p\vz_n} - \vlambda_n \Bigr)
+  \cdot \frac{\p\vz_n}{\p\vtheta}
++ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta} \nonumber\\
+&\hspace{4em} + \sum_{i=0}^{n-1} \Bigl[
+  \Bigl( \frac{\p\cJ}{\p\vz_i} - \vlambda_i + \vlambda_{i+1}
+    + \vlambda_{i+1} \cdot \frac{\p\vF_i}{\p\vz_i} \Bigr)
+  \cdot \frac{\p\vz_i}{\p\vtheta}
+  + \vlambda_{i+1} \cdot \frac{\p\vF_i}{\p\vtheta} \Bigr].
+\end{align}
+
+So the discrete adjoint equation is
+%
+\begin{equation}
+\vlambda_{i-1} = \vlambda_i
+  + \vlambda_i \cdot \frac{\p\vF_{i-1}}{\p\vz_{i-1}}
+  + \frac{\p\cJ}{\p\vz_{i-1}},
+\quad
+i = n, \cdots, 1;
+\qquad
+\vlambda_n = \frac{\p\cJ}{\p\vz_n}.
+\label{explicit_adj}
+\end{equation}
+%
+We can iterate it backward in time to compute the final objective
+gradient:
+%
+\begin{equation}
+\frac{\d\cJ}{\d\vtheta}
+= \frac{\d\cL}{\d\vtheta}
+= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
++ \sum_{i=1}^n \vlambda_i \cdot \frac{\p\vF_{i-1}}{\p\vtheta}.
+\label{explicit_grad}
+\end{equation}
+%
+These equations are readily adaptable to simulated observables.
+For snapshots at $t_n$ or interpolated between $t_{n-1}$ and $t_n$,
+all $\p\cJ / \p\vz_i$ vanish except for the last one or two,
+respectively.
+For light cones, as discussed in \autoref{sec:lightcone}, each component
+of $\hat\vz$ is interpolated at different times, thus all $\p\cJ /
+\p\vz_i$ vanish except for those times relevant for its interpolation,
+and the corresponding $\lambda_i$ can be set to zero for $i$ greater
+than the intersection time.
+
+At the $i$-th iteration the adjoint variable requires the
+vector-Jacobian product (VJP) $\vlambda_i \cdot \p\vF_{i-1} /
+\p\vz_{i-1}$ and the partial objective derivative $\p\cJ / \p\vz_{i-1}$
+at the next time step, which can be easily computed by AD if the whole
+forward history of \eqref{explicit} has been saved.
+However, this can be extremely costly in memory, which can be alleviated
+by checkpointing algorithms such as Revolve and its successors
+\citep{Revolve}.
+Alternatively, if a solution to \eqref{ode} is unique, we can integrate
+it backward and recover the history, which is easy for reversible
+Hamiltonian dynamics and with reversible integrators such as leapfrog.
+When the $N$-body dynamics become too chaotic, one can use more precise
+floating-point numbers and/or save multiple checkpoints\footnote{This is
+different from the checkpointing in the Revolve algorithm, which needs
+to rerun the forward iterations.} during the forward evolution, from
+which the backward evolution can be resumed piecewise.
+
+\vspace{1em}
+\subsection{Application to Simulation}
 \label{sec:adj}
@@
 $N$-body equations of motion.
-In \autoref{app:adj} we introduce the adjoint method for generic
-time-dependent problems in both continuous and discrete cases.
+We have introduced above the adjoint method for generic time-dependent
+problems in both continuous and discrete cases.
 The continuous case is easier to understand and has pedagogical values,
 while the discrete case is the useful one in our application, for we
-want to propagate the gradients backward along the \emph{same} path as
-that of the forward time integration.
+want to propagate numerically the gradients backward along the
+\emph{same} path as that of the forward time integration.
 
@@
 where we have used the fact that $D(t_c, t_a, t_b) = - D(t_c, t_b,
-t_a)$, and left the $\p\cJ / \p\vz_i$ term to the $O$ operator.
+t_a)$, and left the $\p\cJ / \p\vz_i$ term, \HL{the explicit dependence of
+the objective on the intermediate states (from, e.g., observables on the
+light cone), to the observation operator $O$ below.}
 This also naturally determines the subscripts of $\xi$ and $\pi$.
@@
 Like $\va$, we have introduced $\valpha$ and $\vzeta$ to cache the
-vector-Jacobian products (VJPs) on their right-hand sides, for the next
-time step in the kick operator and the objective gradient (see below),
+vector-Jacobian products on their right-hand sides, for the next time
+step in the kick operator and the objective gradient (see below),
 respectively.
@@
 Likewise, the gradient of $O_n^{n-1}$ at $t_n$ is absent in
-\eqref{KDFKO_adj} but enter via the initial conditions following
+\eqref{KDFKO_adj} but enters via the initial conditions following
 \eqref{explicit_adj}.
@@
 and $\p\cJ / \p\vz$'s.
-In practice, the numerical reversibility can be challenged by the
-chaotic $N$-body dynamics, which we find is generally not a concern for
-our applications in the result section.
-
-Finally, during the reverse time integration, we can accumulate
+In practice, the numerical reversibility suffers from the finite
+precision and the chaotic $N$-body dynamics, which we find is generally
+not a concern for our applications in the result section.
+
+Finally, during the reverse time integration, we can accumulate the
 objective gradient following \eqref{explicit_grad}:
 %
-\begin{multline}
-\frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
+\begin{align}
+\frac{\d\cJ}{\d\vtheta} =& \frac{\p\cJ}{\p\vtheta}
 %
@@
   + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
-+ \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
-%
-- \sum_{i=1}^n \Bigl[
+  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \nonumber\\
+%
+  &- \sum_{i=1}^n \Bigl[
     \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
-    \bigr) K(t_i, t_i, t_{i-\half}) \\
+      \bigr) K(t_i, t_i, t_{i-\half})
     + \vxi_i \cdot \vp_{i-\half}
-    \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \\
-  + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta} + \vzeta_{i-1}
-    \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
-  \Bigr],
+      \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \nonumber\\
+    &\qquad + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta}
+      + \vzeta_{i-1} \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
+    \Bigr], \nonumber\\
+%
+\frac{\d\cJ}{\d\vomega} =& \frac{\p\cJ}{\p\vomega}
+%
+  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vomega}
+  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vomega},
 \label{KDFKO_grad}
-\end{multline}
-%
-in which the derivatives and VJPs of the regularization/prior, the
-observation, the initial conditions, the kick and drift factors, and the
-growth and transfer functions can all be conveniently computed by AD.
-Also note that $\vtheta$ can stand for both itself and $\vomega$, which
-we have omitted for brevity.
+\end{align}
+%
+\HL{where the latter backpropagates from fewer sources than the former as
+shown in \autoref{fig:model}.}
 To implement \eqref{KDFKO_adj}-\eqref{KDFKO_grad} in \pmwd\ with
-\texttt{JAX}, we only need to write custom VJP operators for the
-high-level $N$-body integration-observation loop.
+\texttt{JAX}, we only need to write custom VJP rules for the high-level
+$N$-body integration-observation loop, while the derivatives and VJPs of
+the remaining parts including regularization/prior, the observation, the
+initial conditions, the kick and drift factors, the growth and transfer
+functions, etc., can all be conveniently computed by AD.
 
@@
 $(512\,\Mpc/h)^3$ box for 63 time steps.
-The simulation takes only 20 seconds to finish on a \GPU\ GPU.
+The simulation takes only 13 seconds to finish on an \GPU\ GPU.
 }
@@
 PM code \pmwd.
-All results assume a simple $\Lambda$CDM cosmology: $\vtheta (\As =
+All results assume a simple $\Lambda$CDM cosmology: $\vtheta = (\As =
 2\times10^{-9}, \ns = 0.96, \Omegam = 0.3, \Omegab = 0.05, h = 0.7)$,
-and use a \GPU\ GPU.
+and use an \GPU\ GPU with 80GB memory.
 
@@
 We simulate $384^3$ particles from $a=1/64$ to $a=1$, with two
-floating-point precisions, two mesh sizes, two particle masses, and two
-time step sizes.
+floating-point precisions, two mesh sizes, two particle masses (with box
+sizes of $192 \, \Mpc / h$ and $384 \, \Mpc / h$), and two time step
+sizes.
 We take the root-mean-square deviations (RMSDs) of particle
@@
 Model differentiation evolves the adjoint equations backward in time.
-To save memory, trajectory of the model state in the forward run is not
-saved, but re-simulated together with the adjoint equations.
+To save memory, the trajectory of the model state in the forward run is
+not saved, but re-simulated together with the adjoint equations.
 Even though in principle the $N$-body systems are reversible, in
 practice the reconstructed trajectory can differ from the forward one
-due to loss of small-scale information.
-To test the reversibility, in \autoref{tab:rev} we compare the RMSDs of
-the forward-then-reversed particle displacements and velocities from the
-LPT initial conditions, relative to their respective standard
-deviations, that are very small at the \emph{initial} time.
+due to \HL{the finite numerical precision and exacerbated by the chaotic
+dynamics.
+Better reversibility means the gradients propagate backward along a
+trajectory closer to the forward path, and thus would be more accurate.}
+To test this, in \autoref{tab:rev} we compare the RMSDs of the
+forward-then-reverse particle displacements and velocities from the LPT
+initial conditions, relative to their respective standard deviations,
+that are very small at the \emph{initial} time.
 As before, we vary the floating-point precision, the mesh size, the
@@
 One way to improve reversibility is use higher order LPT to initialize
-the $N$-body simulations at later times, when the displacements and
-velocities are not as small.
+the $N$-body simulations at later times \citep{MichauxEtAl2021}, when
+the displacements and velocities are not as small.
 We leave this for future development.
@@
 \, \Mpc / h$ and $0.7 \, \mathrm{km} / \mathrm{s}$.
-Their smallness contributes partly to the fact that the quoted relative
-differences here being much greater than those in \autoref{tab:rep}.
+Their smallness is the main reason that the quoted relative differences
+here are \HL{orders of magnitude greater than those in \autoref{tab:rep}
+(see \autoref{fig:grads} where the reversibility is few times more
+important than the reproducibility in their impacts on the gradients).}
 With the same setup as that in \autoref{tab:rep}, we see the same
@@
 number of time steps or increase the space/mass resolution by
-$2^3\times$.
-To compensate for good statistics, we perform both adjoint and AD runs
-for 8 times, and compare their results by an asymmetric difference of
-$X_i - Y_j$ only for $j<i$, where $i$ and $j$ index the 8 runs.
-We can set $X$ and $Y$ to adjoint and AD respectively for comparison, or
-set both to either adjoint or AD to check their respective
+$2^3\times$ from the listed specifications in the caption.
+\HL{For better statistics, we repeat both adjoint and AD runs for 64 times,
+with the same cosmology and white noise modes, and compare their results
+by an asymmetric difference of $X_i - Y_j$, where $1 \leq j < i \leq
+64$.
+First we set $X$ and $Y$ to the adjoint and AD gradients respectively,
+and find they agree very well on the real white noise (so do their
+gradients on cosmological parameters not shown here; see
+\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint/grads.txt}{\faFile}).
+In addition, we can set both $X$ and $Y$ to either adjoint or AD to
+check their respective \emph{reproducibility}.
+We find both gradients are consistent among different runs of
+themselves, with AD being a lot more reproducible without uncertainty
+from the reverse time integration but only that from the GPU
 reproducibility.
-The adjoint and AD gradients on the real white noise agree very well,
-and so are their gradients on cosmological parameters.
-We find both gradients are consistent among different runs, with AD
-being more reproducible without the uncertainty from reverse time
-integration.
-Interestingly, the asymmetric difference of AD gradients is not
-symmetric about 0, implying some difference between the early and late
-AD runs, i.e., possible breaking of order- or time-invariance in AD.
+This implies that we can ignore the reproducibility errors
+(\autoref{tab:rep}) when the reversibility ones dominate
+(\autoref{tab:rev}).
+Though this statement should be verified again in the future when we
+reduce the reversibility errors using, e.g., 3LPT.}
+%NOTE: this behavior has disappeared with new JAX/CUDA/GPU
+%Interestingly, the asymmetric difference of AD gradients is not
+%symmetric about 0, implying some difference between the early and late
+%AD results, i.e., possible breaking of order- or time-invariance by
+%\texttt{JAX} AD.
 
@@
 \centering
-\includegraphics[width=0.4\linewidth]{grads.pdf}
+\hspace{7em}
+\includegraphics[width=0.5\linewidth]{grads.pdf}
 \\
@@
 \caption{Adjoint gradients of a $128\times128$ slice of the real white
-noise fields, in a \pmwd\ simulation of $128^3$ particles in a
-$(128\,\Mpc/h)^3$ box, with a $256^3$ mesh, 15 time steps, and single
-precision.
+noise field $\d\cJ / \d\vomega$ (top panel), in a \pmwd\ simulation of
+$128^3$ particles in a $(128\,\Mpc/h)^3$ box, with a $256^3$ mesh, 15
+time steps, and single precision.
+\HL{We choose a mean squared error (MSE) objective between two realizations
+with the same cosmology but different initial modes, on their density
+fields on the $256^3$ mesh at $a=1$, and then compute the gradients with
+respect to one realization while holding the other fixed.}
 We compare the adjoint gradients to those by AD, for which we have
@@
 problem in \autoref{fig:optim}.
-We use a simple mean squared error loss between the projected density at
-$a=1$ and a target image, and the Adam optimizer \citep{Adam} with a
-learning rate of $0.1$ and other default hyperparameters.
+We use the Adam optimizer \citep{Adam} with a learning rate of $0.1$ and
+the default values for other hyperparameters.
 Holding the cosmological parameters fixed, we can optimize the real
-white noise field to source initial particles to evolve into our target
+white noise modes to source initial particles to evolve into our target
 pattern, in hundreds to thousands of iterations.
+\HL{Interestingly, we find that the variance of the modes becomes bigger
+than 1 (that of the standard normal white noise) after the optimization,
+and the optimized modes show some level of spatial correlation not
+present in the white noise fields, suggesting that the optimized initial
+conditions are probably no longer Gaussian.}
 
@@
 \, \Mpc^3/h^3$ box.
+We compute their projected density in $64\times108$ pixels and compare
+that to the target image at the same resolution with an MSE objective.
 We use the adjoint method and reverse time integration, assuming the
@@
 nature of $N$-body Hamiltonian dynamics, and implement it with
-\texttt{JAX} in a new PM library, \pmwd{}.
+\texttt{JAX} in a new PM library, \pmwd.
 We have validated the numerical reversibility and the accuracy of the
@@
 
-Being both computation and memory efficient, \pmwd\ enables larger and
+\pmwd\ is both computation and memory efficient, enabling larger and
 more accurate cosmological dark matter simulations.
@@
 conditions of the Universe.
-The efficiency of \pmwd{} also makes it a promising route to generate
+The efficiency of \pmwd\ also makes it a promising route to generate
 the large amount of training data needed by the likelihood-free
@@
 Currently, these applications require more development, including
-distributed parallel capability on multiple GPUs \citep{FlowPM},
+distributed parallel capability on multiple GPUs \citep{FlowPM}, more
+accurate time integration beyond \texttt{FastPM} \citep{ListHahn2023},
 optimization of spatiotemporal resolution of the PM solvers \citep{LDL,
@@
 The 2LPT growth functions follow the following ODEs:
+%
 \begin{align}
@@
   &= 0, \nonumber\\
+%
 D_2'' + \Bigl( 2 + \frac{H'}H \Bigr) D_2' - \frac32 \Omegam(a) D_2
@@
 \end{align}
+%
 where
+%
 \begin{equation}
@@
 \end{equation}
+%
 If the universe has been sufficiently matter dominated at $a_\ic$, with
-$\Omegam(a_\ic) \to 1$ and $H'/H \to -3/2$, the initial conditions of
-the ODEs can be set as the growing mode in this limiting case
+$\Omegam(a_\ic) \simeq 1$ and $H'/H \simeq -3/2$, the initial conditions
+of the ODEs can be set as the growing mode in this limiting case
+%
 \begin{align}
@@
 \end{align}
-$D_1 \approx a$ in the matter dominated era, before being suppressed by
+%
+$D_1 \simeq a$ in the matter dominated era, before being suppressed by
 dark energy.
@@
 The above growth equations can be written in such suppression factors,
+%
 \begin{equation}
@@
 \end{equation}
+%
 for $m \in \{1, 2\}$, as
+%
 \begin{align}
@@
   &= 0, \nonumber\\
+%
 G_2'' + \Bigl( 6 + \frac{H'}H \Bigr) G_2'
@@
 \end{align}
+%
 with initial conditions
+%
 \begin{align}
@@
 \end{align}
+%
 We solve the growth equations in $G_m$ instead of $D_m$, with the
-\texttt{JAX} adaptive ODE integrator implementing the adjoint
-method\footnote{The growth equation adjoint method aims to solve the
-ODEs in their continuous limit, i.e., with the optimize-then-discretize
-approach, this is different from our discretize-then-optimize $N$-body
-adjoint method.}.
+\texttt{JAX} adaptive ODE integrator implementing the adjoint method in
+the optimize-then-discretize approach.
 This is because the former can be integrated backward in time more
@@
 \eqref{G_ZA} by
+%
 \begin{align}
 G_D &= a^2 \dot D_1 = a^2 H D_1', \nonumber\\
+%
 G_K &= a \dot G_D = a^3 H^2 \Bigl[
-    D_1'' + \Bigl(2 + \frac{H'}H \Bigr) D_1'
-  \Bigr],
+  D_1'' + \Bigl(2 + \frac{H'}H \Bigr) D_1' \Bigr],
 \label{G_eval}
 \end{align}
+%
 and
+%
 \begin{align}
@@
 \end{align}
-
-\vspace{1em}
-\section{Adjoint Method}
-\label{app:adj}
-
-We introduce the adjoint method for generic time-dependent ODEs, and
-derive the adjoint equations following a pedagogical tutorial by
-\citet{Bradley2019}.
-We then adopt the discretize-then-optimize approach, and derive the
-discrete adjoint equations, that is more suitable for the $N$-body
-symplectic time integration.
-
-\vspace{1em}
-\subsection{Variational (Tangent) and Adjoint Equations}
-
-Consider a vector state $\vz(t)$ subject to the following ODEs and
-initial conditions
-%
-\begin{equation}
-\dot\vz = \vf(\vz, t, \vtheta),
-\qquad
-\vz(t_0) = \vz_0(\vtheta),
-\label{ode}
-\end{equation}
-%
-for $t \in [t_0, t_1]$.
-Here the initial conditions can depend on the parameters $\vtheta$.
-
-A perturbation in the initial conditions propagates forward in time.
-For $\vz(t, t_0, \vz_0)$, the Jacobian of state variables describing
-this,
-%
-\begin{equation}
-\vDelta = \frac{\p\vz}{\p\vz_0},
-\end{equation}
-%
-evolves from identity $\vDelta_0 = \vI$ by
-%
-\begin{equation}
-\dot\vDelta = \frac{\p\vf}{\p\vz} \cdot \vDelta,
-\label{variational}
-\end{equation}
-%
-following from \eqref{ode}.
-\eqref{variational} is known as the \emph{variational or tangent
-equation}.
-
-The backward version of $\vDelta$,
-%
-\begin{equation}
-\vLambda = \frac{\p\vz_1}{\p\vz},
-\end{equation}
-%
-evolves backward in time from identity $\vLambda_1 = \vI$ by
-%
-\begin{equation}
-\dot\vLambda = - \vLambda \cdot \frac{\p\vf}{\p\vz}.
-\label{adjoint}
-\end{equation}
-%
-\eqref{adjoint} is called the \emph{adjoint equation}, for the
-right-hand side is $(\p\vf / \p\vz)^\intercal \vLambda$.
-It can be derived from the time-invariance of $\vLambda \cdot \vDelta$:
-%
-\begin{equation}
-\vLambda \cdot \dot\vDelta + \dot\vLambda \cdot \vDelta
-= \frac\d{\d t} (\vLambda \cdot \vDelta)
-= \frac\d{\d t} \Bigl( \frac{\p\vz_1}{\p\vz}
-  \cdot \frac{\p\vz}{\p\vz_0} \Bigr)
-= \frac\d{\d t} \frac{\p\vz_1}{\p\vz_0}
-= \vzero.
-\end{equation}
-
-Alternatively, the adjoint equation can be derived from the variational
-equation, using the facts that $\d (\bm{M}^{-1}) = - \bm{M}^{-1} \, \d
-\bm{M} \, \bm{M}^{-1}$ and $(\p\vz / \p\tilde\vz)^{-1} = \p\tilde\vz /
-\p\vz$.
-Like \eqref{variational} and \eqref{adjoint}
-%
-\begin{equation}
-\frac\d{\d t} \frac{\p\vz}{\p\tilde\vz}
-= \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\tilde\vz},
-\qquad
-\frac\d{\d t} \frac{\p\tilde\vz}{\p\vz}
-= - \frac{\p\tilde\vz}{\p\vz} \cdot \frac{\p\vf}{\p\vz}.
-\end{equation}
-As we can see next, the adjoint equation takes a similar form when one
-optimize an objective function of the state.
-
-\vspace{1em}
-\subsection{Objective on the Final State}
-
-In the simplest case, the objective function depends only on the final
-state, e.g., the last snapshot of a simulation, and possibly the
-parameters too in the form of regularization or prior information, i.e.,
-$\cJ(\vz_1, \vtheta)$.
-To optimize the objective under the constraint given by the ODEs, we can
-introduce a time-dependent function $\vlambda(t)$ as the Lagrange
-multiplier:
-%
-\begin{equation}
-\cL = \cJ(\vz_1, \vtheta)
-- \int_{t_0}^{t_1} \vlambda(t)
-  \cdot \bigl[ \dot\vz - \vf(\vz, t, \vtheta) \bigr] \, \d t.
-\end{equation}
-%
-Note the minus sign we have introduced in front of $\vlambda$ for later
-convenience.
-
-Its total derivative with respect to $\vtheta$ is
-%
-\begin{equation}
-\frac{\d\cL}{\d\vtheta}
-= \frac{\p\cJ}{\p\vtheta}
-+ \frac{\p\cJ}{\p\vz_1} \cdot \frac{\p\vz_1}{\p\vtheta}
-- \int_{t_0}^{t_1} \vlambda
-  \cdot \Bigl( \frac{\p\dot\vz}{\p\vtheta}
-    - \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
-    - \frac{\p\vf}{\p\vtheta} \Bigr) \, \d t.
-\end{equation}
-%
-Integrating the first term of the integrand by parts:
-%
-\begin{equation}
-\int_{t_0}^{t_1} \vlambda \cdot \frac{\p\dot\vz}{\p\vtheta} \, \d t
-= \int_{t_0}^{t_1} \vlambda \cdot \frac{\p^2\vz}{\p\vtheta\p t} \, \d t
-= \Bigl[ \vlambda \cdot \frac{\p\vz}{\p\vtheta} \Bigr]_{t_0}^{t_1}
-- \int_{t_0}^{t_1} \dot \vlambda \cdot \frac{\p\vz}{\p\vtheta} \, \d t,
-\end{equation}
-%
-and plugging it back:
-%
-\begin{equation}
-\frac{\d\cL}{\d\vtheta}
-= \frac{\p\cJ}{\p\vtheta}
-+ \Bigl( \frac{\p\cJ}{\p\vz_1} - \vlambda_1 \Bigr)
-  \cdot \frac{\p\vz_1}{\p\vtheta}
-+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \int_{t_0}^{t_1} \Bigl[
-  \Bigl( \vlambda \cdot \frac{\p\vf}{\p\vz} + \dot\vlambda \Bigr)
-  \cdot \frac{\p\vz}{\p\vtheta}
-  + \vlambda \cdot \frac{\p\vf}{\p\vtheta} \Bigr] \, \d t.
-\end{equation}
-
-Now we are free to choose
-%
-\begin{equation}
-\dot\vlambda = - \vlambda \cdot \frac{\p\vf}{\p\vz},
-\qquad
-\vlambda_1 = \frac{\p\cJ}{\p\vz_1},
-\label{adjoint1}
-\end{equation}
-%
-that allow us to avoid all $\p\vz/\p\vtheta$ terms in the final
-objective gradient:
-%
-\begin{equation}
-\frac{\d\cJ}{\d\vtheta}
-= \frac{\d\cL}{\d\vtheta}
-= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \int_{t_0}^{t_1} \vlambda \cdot \frac{\p\vf}{\p\vtheta} \, \d t,
-\end{equation}
-%
-in which the first two terms come from the regularization and initial
-conditions, respectively.
-\eqref{adjoint1} is the adjoint equation for the objective $\cJ(\vz_1,
-\vtheta)$.
-With the initial conditions set at the final time, we can integrate it
-backward in time to obtain $\vlambda(t)$, which enters the above
-equation and yields $\d\cJ / \d\vtheta$.
-
-Note that \eqref{adjoint1} has the same form as \eqref{adjoint}, and
-their solutions are related by
-%
-\begin{equation}
-\vlambda = \vlambda_1 \cdot \vLambda
-= \frac{\p\cJ}{\p\vz_1} \cdot \frac{\p\vz_1}{\p\vz}
-= \frac{\p\cJ}{\p\vz}.
-\end{equation}
-And like $\vLambda \cdot \vDelta$, $\vlambda \cdot \vDelta$ is
-time-invariant:
-%
-\begin{equation}
-\vlambda_0 \cdot \vDelta_0
-= \vlambda_1 \cdot \vDelta_1
-= \vlambda \cdot \vDelta
-= \frac{\p\cJ}{\p\vz} \cdot \frac{\p\vz}{\p\vz_0}
-= \frac{\p\cJ}{\p\vz_0}
-= \vlambda_0.
-\end{equation}
-%
-Computing $\vlambda_1 \cdot \vDelta_1$ directly is expensive, but
-solving \eqref{adjoint1} backward for $\vlambda_0$ is cheap.
-This is related to the fact that the reverse-mode AD or backpropagation
-is cheaper than the forward mode for optimization.
-
-\vspace{1em}
-\subsection{Objective on the State History}
-
-The adjoint method applies to more complex cases too.
-Let's consider an objective as a functional of the evolution history
-with some regularization $\cR$ on $\vtheta$
-%
-\begin{equation}
-\cJ = \cR(\vtheta) + \int_{t_0}^{t_1} g(\vz, t, \vtheta) \, \d t.
-\end{equation}
-%
-The derivation is similar:
-%
-\begin{equation}
-\cL = \cR(\vtheta)
-+ \int_{t_0}^{t_1} \bigl\{ g(\vz, t, \vtheta) -
-  \vlambda(t) \cdot \bigl[ \dot\vz - \vf(\vz, t, \vtheta) \bigr]
-  \bigr\} \, \d t
-\end{equation}
-%
-has gradient
-%
-\begin{align}
-\frac{\d\cL}{\d\vtheta}
-&= \frac{\d\cR}{\d\vtheta}
-+ \int_{t_0}^{t_1} \Bigl[
-    \frac{\p g}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
-    + \frac{\p g}{\p\vtheta}
-    - \vlambda \cdot \Bigl( \frac{\p\dot\vz}{\p\vtheta}
-    - \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\vtheta}
-    - \frac{\p\vf}{\p\vtheta} \Bigr) \Bigr] \, \d t \\
-%
-&= \frac{\d\cR}{\d\vtheta}
-- \vlambda_1 \cdot \frac{\p\vz_1}{\p\vtheta}
-+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \int_{t_0}^{t_1} \Bigl[
-  \Bigl( \frac{\p g}{\p\vz} + \vlambda \cdot \frac{\p\vf}{\p\vz}
-    + \dot\vlambda \Bigr) \cdot \frac{\p\vz}{\p\vtheta}
-  + \frac{\p g}{\p\vtheta} + \vlambda \cdot \frac{\p\vf}{\p\vtheta}
-\Bigr] \, \d t. \nonumber
-\end{align}
-
-The adjoint equation becomes
-%
-\begin{equation}
-\dot\vlambda = - \frac{\p g}{\p\vz}
-  - \vlambda \cdot \frac{\p\vf}{\p\vz},
-\qquad
-\vlambda_1 = \vzero,
-\end{equation}
-%
-with the objective gradient
-%
-\begin{equation}
-\frac{\d\cJ}{\d\vtheta}
-= \frac{\d\cL}{\d\vtheta}
-= \frac{\d\cR}{\d\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \int_{t_0}^{t_1} \Bigl( \frac{\p g}{\p\vtheta}
-  + \vlambda \cdot \frac{\p\vf}{\p\vtheta} \Bigr) \, \d t.
-\end{equation}
-
-The time-invariant is now
-%
-\begin{equation}
-\vlambda_1 \cdot \vDelta_1
-= \vlambda_0 \cdot \vDelta_0
-  + \int_{t_0}^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t
-= \vlambda \cdot \vDelta
-  + \int_t^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t'
-= \vzero,
-\end{equation}
-%
-so
-%
-\begin{equation}
-\vlambda \cdot \vDelta
-= - \int_t^{t_1} \frac{\p g}{\p\vz} \cdot \vDelta \, \d t'
-= - \int_t^{t_1} \frac{\p g}{\p\vz_0} \, \d t'.
-\end{equation}
-
-\vspace{1em}
-\subsection{Objective on the Observables}
-\label{app:lightcone}
-
-Now let's consider an objective that depends on different state
-components at different times, e.g., in the Universe where further
-objects intersected our past light cone earlier.
-It falls between the previous scenarios, and we can derive its adjoint
-equation similarly.
-
-We denote the observables by $\hat\vz$, with different components $\hat
-z_j$ affecting the objective $\cJ(\hat\vz, \vtheta)$ at different times
-$\hat t_j$, i.e., $\hat z_j \triangleq z_j(\hat t_j)$.
-The Lagrangian becomes
-%
-\begin{equation}
-\cL = \cJ(\hat\vz, \vtheta)
-- \sum_j \int_{t_0}^{\hat t_j} \lambda_j(t)
-  \bigl[ \dot z_j - f_j(\vz, t, \vtheta) \bigr] \, \d t,
-\end{equation}
-%
-constraining only the parts of trajectories inside the light cone.
-Its gradient is
-%
-\begin{align}
-\frac{\d\cL}{\d\vtheta}
-&= \frac{\p\cJ}{\p\vtheta}
-+ \frac{\p\cJ}{\p\hat\vz} \cdot \frac{\p\hat\vz}{\p\vtheta}
-- \sum_j \int_{t_0}^{\hat t_j}
-  \Bigl( \lambda_j \frac{\p\dot z_j}{\p\vtheta}
-    - \vlambda \cdot \frac{\p\vf}{\p z_j} \frac{\p z_j}{\p\vtheta}
-    - \lambda_j \frac{\p f_j}{\p\vtheta} \Bigr) \, \d t \\
-%
-&= \frac{\p\cJ}{\p\vtheta}
-+ \Bigl( \frac{\p\cJ}{\p\hat\vz} - \hat\vlambda \Bigr)
-  \cdot \frac{\p\hat\vz}{\p\vtheta}
-+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \sum_j \int_{t_0}^{\hat t_j} \Bigl[
-  \Bigl( \vlambda \cdot \frac{\p\vf}{\p z_j} + \dot\lambda_j \Bigr)
-  \frac{\p z_j}{\p\vtheta}
-  + \lambda_j \frac{\p f_j}{\p\vtheta} \Bigr] \, \d t, \nonumber
-\end{align}
-%
-where we have defined $\hat\vlambda$ similarly with components
-$\hat\lambda_j \triangleq \lambda_j(\hat t_j)$.
-In the first equality, we have also dropped a vanishing term $\sum_j
-\hat\lambda_j \bigl[ \dot z_j - f_j(\hat t_j) \bigr] \p\hat t_j /
-\p\vtheta$, i.e., $\p\hat t_j / \p\vtheta$ does not directly enter the
-gradient.
-
-Now we find the adjoint equation
-%
-\begin{equation}
-\dot\vlambda = - \vlambda \cdot \frac{\p\vf}{\p\vz},
-\qquad
-\hat\vlambda = \frac{\p\cJ}{\p\hat\vz},
-\end{equation}
-%
-that has the same form as \eqref{adjoint1}, with a slightly different
-initial condition given at the respective observation time of each
-component.
-The objective gradient is also similar to the previous cases:
-%
-\begin{equation}
-\frac{\d\cJ}{\d\vtheta}
-= \frac{\d\cL}{\d\vtheta}
-= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \sum_j \int_{t_0}^{\hat t_j} \lambda_j \frac{\p f_j}{\p\vtheta} \, \d t.
-\end{equation}
-
-Note that even though $\lambda_j(t)$ for $t > \hat t_j$ does not affect
-the final gradient, they can enter the right-hand side of the adjoint
-equation, and affect those $\lambda_k$ with $t < \hat t_k$, i.e., inside
-the light cone.
-Physically however, $\p f_j / \p z_k$ should vanish for spacelike
-separated pairs of $z_j$ and $z_k$, even though the Newtonian
-approximation we adopt introduces some small deviation.
-Therefore, we can set $\lambda_j(t)$ to 0 for $t > \hat t_j$, and bump
-it to $\p\cJ / \p\hat z_j$ at $\hat t_j$.
-
-\vspace{1em}
-\subsection{Discretize Then Optimize}
-
-In practice, the time integration of \eqref{ode} is discrete.
-Consider the explicit methods,
-%
-\begin{equation}
-\vz_{i+1} = \vz_i + \vF_i(\vz_i, \vtheta), \qquad i = 0, \cdots, n-1,
-\label{explicit}
-\end{equation}
-%
-which include the leapfrog integrator commonly used for Hamiltonian
-dynamics.
-We want to propagate the gradients backward along the \emph{same}
-discrete trajectory as taken by the forward integration.
-Therefore, instead of the continuous adjoint equations derived above, we
-need the adjoint method for the discrete integrator.
-
-Without loss of generality, we derive the adjoint equation for an
-objective depending on the state at all time steps, which can be easily
-specialized to the 3 cases discussed above with only slight
-modifications.
-The discretized Lagrangian is now
-%
-\begin{equation}
-\cL = \cJ(\vz_0, \cdots, \vz_n, \vtheta)
-- \sum_{i=0}^{n-1} \vlambda_{i+1}
-  \cdot \bigl[ \vz_{i+1} - \vz_i - \vF_i(\vz_i, \vtheta) \bigr],
-\end{equation}
-%
-whose gradient is
-%
-\begin{align}
-\frac{\d\cL}{\d\vtheta}
-&= \frac{\p\cJ}{\p\vtheta}
-+ \sum_{i=0}^n \frac{\p\cJ}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
-- \sum_{i=0}^{n-1} \vlambda_{i+1}
-  \cdot \Bigl( \frac{\p\vz_{i+1}}{\p\vtheta} - \frac{\p\vz_i}{\p\vtheta}
-    - \frac{\p\vF_i}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
-    - \frac{\p\vF_i}{\p\vtheta} \Bigr) \nonumber\\
-%
-&= \frac{\p\cJ}{\p\vtheta}
-+ \sum_{i=0}^n \frac{\p\cJ}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
-- \sum_{i=1}^n \vlambda_i \cdot \frac{\p\vz_i}{\p\vtheta}
-+ \sum_{i=0}^{n-1} \vlambda_{i+1}
-  \cdot \Bigl( \frac{\p\vz_i}{\p\vtheta}
-    + \frac{\p\vF_i}{\p\vz_i} \cdot \frac{\p\vz_i}{\p\vtheta}
-    + \frac{\p\vF_i}{\p\vtheta} \Bigr) \nonumber\\
-&= \frac{\p\cJ}{\p\vtheta}
-+ \Bigl( \frac{\p\cJ}{\p\vz_n} - \vlambda_n \Bigr)
-  \cdot \frac{\p\vz_n}{\p\vtheta}
-+ \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta} \nonumber\\
-&\hspace{4em} + \sum_{i=0}^{n-1} \Bigl[
-  \Bigl( \frac{\p\cJ}{\p\vz_i} - \vlambda_i + \vlambda_{i+1}
-    + \vlambda_{i+1} \cdot \frac{\p\vF_i}{\p\vz_i} \Bigr)
-  \cdot \frac{\p\vz_i}{\p\vtheta}
-  + \vlambda_{i+1} \cdot \frac{\p\vF_i}{\p\vtheta} \Bigr].
-\end{align}
-
-So the discrete adjoint equation is
-%
-\begin{equation}
-\vlambda_{i-1} = \vlambda_i
-  + \vlambda_i \cdot \frac{\p\vF_{i-1}}{\p\vz_{i-1}}
-  + \frac{\p\cJ}{\p\vz_{i-1}},
-\quad
-i = n, \cdots, 1;
-\qquad
-\vlambda_n = \frac{\p\cJ}{\p\vz_n}.
-\label{explicit_adj}
-\end{equation}
-%
-We can iterate it backward in time to compute the final objective
-gradient:
-%
-\begin{equation}
-\frac{\d\cJ}{\d\vtheta}
-= \frac{\d\cL}{\d\vtheta}
-= \frac{\p\cJ}{\p\vtheta} + \vlambda_0 \cdot \frac{\p\vz_0}{\p\vtheta}
-+ \sum_{i=1}^n \vlambda_i \cdot \frac{\p\vF_{i-1}}{\p\vtheta}.
-\label{explicit_grad}
-\end{equation}
-These equations are readily adaptable to simulated observables.
-For snapshots at $t_n$ or interpolated between $t_{n-1}$ and $t_n$,
-all $\p\cJ / \p\vz_i$ vanish except for the last one or two,
-respectively.
-For light cones, as discussed in \autoref{app:lightcone}, each component
-of $\hat\vz$ is interpolated at different times, thus all $\p\cJ /
-\p\vz_i$ vanish except for those times relevant for its interpolation,
-and the corresponding $\lambda_i$ can be set to zero for $i$ greater
-than the intersection time.
-
-At the $i$-th iteration the adjoint variable requires the
-vector-Jacobian product (VJP) $\vlambda_i \cdot \p\vF_{i-1} /
-\p\vz_{i-1}$ and the partial objective derivative $\p\cJ / \p\vz_{i-1}$
-at the next time step, which can be easily computed by AD if the whole
-forward history of \eqref{explicit} has been saved.
-However, this can be extremely costly in memory, which can be alleviated
-by checkpointing algorithms such as Revolve and its successors
-\citep{Revolve}.
-Alternatively, if a solution to \eqref{ode} is unique, we can integrate
-it backward and recover the history, which is easy for reversible
-Hamiltonian dynamics and with reversible integrators such as leapfrog.
-When the $N$-body dynamics become too chaotic, one can use more precise
-floating-point numbers and/or save multiple checkpoints\footnote{This is
-different from the checkpointing in the Revolve algorithm, which needs
-to rerun the forward iterations.} during the forward evolution, from
-which the backward evolution can be resumed piecewise.
 

### 2023-11-10 pmwdlib/kernels.cc.cu
fix integer out of range

@@ -688,8 +688,8 @@ void scatter_sm(cudaStream_t stream, void** buffers, const char* opaque, std::si
 
     CUDA_SAFE_CALL(cudaMalloc((void**)&work_i_d, sizeof(uint32_t)*(n_particle*4+2*nbinx*nbiny*nbinz+1)+temp_storage_bytes));
 
-    uint32_t npts_mem_size = sizeof(uint32_t) * n_particle;
-    uint32_t nbins_mem_size = sizeof(uint32_t) * nbinx*nbiny*nbinz;
+    int64_t npts_mem_size = sizeof(uint32_t) * (int64_t)n_particle;
+    int64_t nbins_mem_size = sizeof(uint32_t) * (int64_t)nbinx*nbiny*nbinz;
     uint32_t* d_sortidx = (uint32_t*)work_i_d;
     uint32_t* d_sortidx_buff = (uint32_t*)&work_i_d[npts_mem_size];
     uint32_t* d_index = (uint32_t*)&work_i_d[2*npts_mem_size];

### 2023-11-24 docs/examples/sto/train/job.sh
FIXUP minor

@@ -14,8 +14,8 @@
 module purge
 
 source /mnt/home/yzhang/mambaforge/bin/activate
-# srun python run_train.py
-srun python optune.py
+srun python run_train.py
+# srun python optune.py
 
 
 # Notes:

### 2023-11-24 docs/examples/sto/train/optune.py

@@ -21,14 +21,15 @@ def objective(trial, sobol_ids, gsdata, snap_ids):
     learning_rate = trial.suggest_float('learning_rate', 1e-6, 1e-3, log=True)
     if opt_name == 'Adam':
         b1 = trial.suggest_float('b1', 0.5, 0.9)
-        optimizer = optax.adam(learning_rate, b1=b1)
+        b2 = trial.suggest_float('b2', 0.5, 0.999)
+        optimizer = optax.adam(learning_rate, b1=b1, b2=b2)
     if opt_name == 'AdamW':
         b1 = trial.suggest_float('b1', 0.5, 0.9)
         weight_decay = trial.suggest_float('weight_decay', 1e-5, 1e-3, log=True)
         optimizer = optax.adamw(learning_rate, b1=b1, weight_decay=weight_decay)
     if opt_name == 'SGD':
         optimizer = optax.sgd(learning_rate)
-    n_layers = trial.suggest_int('n_layers', 2, 5)
+    n_layers = trial.suggest_int('n_layers', 2, 8)
 
     so_nodes = [[n] * n_layers + [1] for n in n_input]
     so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')

### 2023-11-24 pmwd/sto/data.py

@@ -170,8 +170,6 @@ class G4snapDataset(Dataset):
         sobol_id = self.sobol_ids[idx // self.snaps_per_sim]
         snap_id = self.snap_ids[idx % self.snaps_per_sim]
 
-        # TODO generate pmwd parameters and IC here?
-
         return self.g4data[sobol_id][snap_id]
 
     def getsnap(self, sidx, snap):

### 2023-11-24 pmwd/sto/hypars.py

@@ -6,7 +6,7 @@ import pickle
 from pmwd.sto.so import soft_len
 from pmwd.sto.mlp import init_mlp_params
 
-n_epochs = 300
+n_epochs = 600
 
 # data
 sobol_ids_global = np.arange(0, 64)
@@ -15,7 +15,13 @@ shuffle_epoch = True  # shuffle the order of sobols across epochs
 
 # optimizer
 learning_rate = 3e-5
-optimizer = optax.adam(learning_rate)
+optimizer = optax.adam(learning_rate, b1=0.9, b2=0.9)
+# optimizer = optax.chain(
+#     optax.clip_by_global_norm(3.),
+#     optax.adam(learning_rate),
+# )
+# optimizer = optax.amsgrad(learning_rate, eps=1e-4)
+# optimizer = optax.adamw(learning_rate, weight_decay=1e-4)
 
 def get_optimizer(lr):
     return optax.adam(lr)
@@ -47,10 +53,10 @@ def lr_scheduler(lr, skd_state, loss, patience=15, factor=0.5, threshold=1e-4, l
 so_type = 'NN'
 n_input = [soft_len(k_fac=3), soft_len()]
 
-so_nodes = [[n, n, n, 1] for n in n_input]
-# so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')
-with open('params/2967953/e288.pickle', 'rb') as f:
-    so_params = pickle.load(f)['so_params']
+so_nodes = [[3*n, 3*n, 3*n, 1] for n in n_input]
+so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')
+# with open('params/2980166/e107.pickle', 'rb') as f:
+#     so_params = pickle.load(f)['so_params']
 
 dropout_rate = None  # set to None for no dropout applied
 

### 2023-11-24 docs/papers/sto/so.tex
FIXUP draft add optuna ref
FIXUP draft change loss

@@
 The two fields are identical when both $T$ and $r$ are unity.
-With force sharpening,
+The PM method is accurate on scales much larger than the cell size, therefore we
+expect $T$ and $r$ to be very close to being unity both before and after force
+sharpening optimization.
 
@@
 \begin{equation}
-\cL_f = \frac1{N_a N_K} \sum_i \ln
-\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} | f(\vk')|^2},
+\cL_f = \frac1{N_K N_a} \sum_{ij} \ln \biggl[
+\frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk, a_j) - f(\vk, a_j) \bigr|^2}
+     {\sum_{\vk' \in K_i} | f(\vk', a_j)|^2} + \epsilon \biggr],
 \end{equation}
 %
-and
-%
-\begin{equation}
-\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
-\frac{\sum_{k \in K_i} w(\vk)
-      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
-\end{equation}
-%
+where $\epsilon$ sets the relative threshold tolerance below which we
+consider $\hat f$ in each $K$ accurate enough, and the gradients from
+that bin unimportant.
+%and
+%%
+%\begin{equation}
+%\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
+%\frac{\sum_{k \in K_i} w(\vk)
+%      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+%     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
+%\end{equation}
+%%
 Then the question is which loss we should use for which $f$, to get
@@
 
+We use Optuna \citep{optuna_2019} to search for the optimal Hyperparameters.
+
 We empoly a learning rate scheduler that reduces the learning rate on plateau

### 2023-11-24 docs/papers/sto/so.tex
FIXUP draft tentative changes

@@
 \newcommand{\Omegab}{\Omega_\mathrm{b}}
+\newcommand{\OmegaK}{\Omega_K}
 \newcommand{\Mpc}{\mathrm{Mpc}}
 \newcommand{\ic}{\mathrm{i}}
+\newcommand{\f}{\mathrm{f}}
 \newcommand{\linear}{\mathrm{lin}}
@@
   \cmidrule(lr){2-2}
-  dimensionless factors & $\Omegam(a)$ \\
+  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
   \cmidrule(lr){2-2}
@@
   \cmidrule(lr){2-2}
-  & $\Delta\lna$: time step size in $\lna$ \\
+  & $\Delta\lna$: time step size in $\lna$\YL{. Now this can be an important feature with varying $a_\ic$}\\
   \bottomrule
@@
   to the mean particle spacing.
-  The curvature $\Omega_k$ is related to the separate universe simulation.
+  The curvature $\OmegaK$ is related to the separate universe simulation.
   We sample parameters applicable to \GADGET\ during data generation, and
   those applicable to \pmwd\ during training.
-  \textsuperscript\dag{}The box size is determined jointly by the \pmwd\
-  mesh shape and mesh cell size below, which are assumed to be sampled
-  independently.
-  \YL{I forgot why I wrote ``Mpc'' without the $h$ here.}
-  \YZ{We are using ``Mpc/h'' in the code.}
-  In practice, we sample one conditioned on the other and the box size.
-  *Number of time steps from $a=1/16$ to $a=1+1/128$ (in order to cover the
-  snapshot offset).
+  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
+  Now it seems more natural to me to have the ptcl spacing sampled
+  uniformly instead...)}
   }
@@
   \cmidrule(lr){2-5}
-  & 121 snapshots, $a$ & uniform & [1/16, 1] & deterministic \\
-  \cmidrule(lr){2-5}
-  & box size in Mpc\YZ{/h}\textsuperscript\dag & log-trapezoidal & [25.6, 2560) \\
+  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
   \cmidrule(lr){2-4}
-  & snapshot offset, $\Delta\!a$ & uniform & [0, 1/128) \\
+  & snapshot offset & uniform & (0, 1] \\
   \cmidrule(lr){2-4}
@@
   \cmidrule(lr){2-4}
-  & $\Omega_k / (1 - \Omega_k)$ & uniform & [-1/3, 1/3) \\
+  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
   \cmidrule(lr){2-4}
@@
   \cmidrule(lr){1-5}
-  & mesh shape & log-uniform & [128, 512] \\
+  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
   \cmidrule(lr){2-4}
-  \pmwd\ & mesh cell size in Mpc\YZ{/h} & log-uniform & [0.2, 5] & MC \\
-  \cmidrule(lr){2-4}
-  & number of time steps* & log-uniform & [10, 1000] \\
   \cmidrule(lr){1-4}
-  training & loss grid offset in its own spacing & uniform & $[0, 1)^3$ \\
+  training & loss grid offset & uniform & $[0, 1)^3$ \\
   \bottomrule
@@
 \autoref{tab:param}.
-Each simulation has $128^3$ particles and writes 121 snapshots linearly
-spaced in scale factor from $1/16 + \Delta a$ to $1 + \Delta a$, with
-spacing $1/128$.
-$\Delta a$ is an offset, common for all snapshots in each simulation,
-that varies as in \autoref{tab:param} and helps to sample different
-alignment between snapshot and start/stop times.
+\YL{
+Each simulation evolves $128^3$ particles from the initial time $a_\ic$
+to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
+factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
+We determine $a_\ic$ for each configuration depending on its cosmology
+and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why and how here.)} which then sets the
+snapshot spacing $\Delta a = (1 - a_\ic) / 64$.
+To sample different alignment between the snapshots and the start/stop
+times and avoid possible related numerical errors in \pmwd, we introduce
+a fractional random offset $\gamma$ that varies between 0 and 1 for
+different simulations.
+}
 In our \pmwd\ model, we generate the snapshot at an arbitrary given time using
 the cubic Hermite interpolation, with more details in \autoref{sec:snapobs}.
-The curvature $\Omega_k$ is related to the separate universe simulation
+The curvature $\OmegaK$ is related to the separate universe simulation
 \citep{LiEtAl2014, WagnerEtAl2015}.
@@
 % to be compare with the speed of pmwd with SO
+
+\YL{
+In this work, we train \pmwd\ using $XXX$ time steps from $a_\ic$ to
+$a_\f$, generate snapshots at the same times as in \GADGET, and compare
+them in the loss function.
+Although the number of time steps are fixed here, we emphasize that our
+spatial optimization of course also applies for different number of time
+steps, which we will explore, together with the optimal choice of time
+variable, in a follow-up work on the temporal optimization.
+}
 
@@
 \YL{
-For a field $f$ in either Lagrangian or Eulerian space, we have thought
-of 3 forms of MSE-inspired loss:
-%
-\begin{equation}
-\cL_f = \frac1{N_a} \ln
-  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
-       {\sum_{\vq'} f(\vq')^2},
-\end{equation}
-%
-where $N_a$ is the number of time steps it takes to generate the snapshot for
-each iteration
-(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
-their Fourier wavevector $\vk$ in both cases).
+For a field $f$ in either Lagrangian or Eulerian space%, we have thought
+%of 3 forms of MSE-inspired loss:
+%%
+%\begin{equation}
+%\cL_f = \frac1{N_a} \ln
+%  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
+%       {\sum_{\vq'} f(\vq')^2},
+%\end{equation}
+%%
+%where $N_a$ is the number of time steps it takes to generate the snapshot for
+%each iteration
+%(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
+%their Fourier wavevector $\vk$ in both cases).
 %
@@
 %
-where $\epsilon$ sets the relative threshold tolerance below which we
-consider $\hat f$ in each $K$ accurate enough, and the gradients from
-that bin unimportant.
+where $N_a$ is the number of snapshots and $N_K$ is the number of
+$k$-bins.
+$\epsilon$ sets the relative threshold tolerance, below which we
+consider $\hat f$ in each $K$ accurate enough and the gradients from
+that bin less important.
 %and
@@
 %%
-Then the question is which loss we should use for which $f$, to get
-stable and robust results, e.g., no bump for large or small boxes.
-To further limit our scope in this paper, we should probably only
-consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
-velocity $\vv(\vq)$.
+%Then the question is which loss we should use for which $f$, to get
+%stable and robust results, e.g., no bump for large or small boxes.
+%To further limit our scope in this paper, we should probably only
+%consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
+%velocity $\vv(\vq)$.
+We can use $f \in \{\vs(\vq), \delta(\vx), \vv(\vq), (1+\delta)\vv(\vx),
+(1+\delta)\vv\vv(\vx)\}$.
 }
@@
 
-\YL{Even though the log should be enough for adaptive optimizer and
-fixed sim config, this may create problem when training on all config at
-the same time. So normalizations inside the log is added above and
-below.}
+%\YL{Even though the log should be enough for adaptive optimizer and
+%fixed sim config, this may create problem when training on all config at
+%the same time. So normalizations inside the log is added above and
+%below.}
 
@@
 Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
-Lagrangian space, while the density field is in the Eulerian space:
+Lagrangian space, while the density and momentum fields are in the
+Eulerian space:
 %
-\begin{equation}
-1 + \delta(\vx) = \int\!\d^3\vq\, \deltaD[ \vx - \vq - \vs(\vq)],
-\end{equation}
+\begin{align}
+1 + \delta(\vx) &= \int\!\d^3\vq\,
+  \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+(1 + \delta)\vv(\vx) &= \int\!\d^3\vq\,
+  \vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+(1 + \delta)\vv\vv(\vx) &= \int\!\d^3\vq\,
+  \vv\vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+\end{align}
 %
@@
 \textit{\large Data Availability:}
-\YZ{Do we make the \GSDATA\ data public?}
+We make the \GSDATA\ simulation data set public at ...
 

### 2023-11-27 pmwd/sto/loss.py
FIXUP add epsilon to ln power loss

@@ -39,10 +39,10 @@ def loss_power_w(f, g, spacing=1, log=True, w=None, cut_nyq=False):
     return loss
 
 
-def loss_power_ln(f, g, spacing=1, cut_nyq=False):
+def loss_power_ln(f, g, spacing=1, cut_nyq=False, eps=1e-3):
     k, P_d, N, bins = powspec(f - g, spacing, cut_nyq=cut_nyq)
     k, P_g, N, bins = powspec(g, spacing, cut_nyq=cut_nyq)
-    loss = jnp.log(P_d / P_g).sum() / len(k)
+    loss = jnp.log(P_d / P_g + eps).sum() / len(k)
     return loss
 
 

### 2023-11-27 docs/examples/sto/train/run_train.py
FIXUP add loading existing optimizer state

@@ -110,7 +110,7 @@ def prep_train(sobol_ids_global, snap_ids):
 
 
 def run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch,
-              learning_rate, optimizer, so_type, so_nodes, so_params,
+              learning_rate, optimizer, opt_state, so_type, so_nodes, so_params,
               ret=False, log_id=None, verbose=True):
 
     # RNGs with fixed seeds
@@ -121,7 +121,6 @@ def run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch,
     # jax: dropout layer
     jax_key = jax.random.PRNGKey(0)
 
-    opt_state = optimizer.init(so_params)
     skd_state = None
 
     jax_device_sync()
@@ -193,9 +192,9 @@ if __name__ == "__main__":
 
     from pmwd.sto.hypars import (
         n_epochs, sobol_ids_global, snap_ids, shuffle_epoch,
-        learning_rate, optimizer, so_type, so_nodes, so_params)
+        learning_rate, optimizer, opt_state, so_type, so_nodes, so_params)
 
     sobol_ids, gsdata = prep_train(sobol_ids_global, snap_ids)
 
     run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch,
-              learning_rate, optimizer, so_type, so_nodes, so_params)
+              learning_rate, optimizer, opt_state, so_type, so_nodes, so_params)

### 2023-11-27 pmwd/sto/hypars.py
FIXUP add loading existing optimizer state

@@ -6,7 +6,7 @@ import pickle
 from pmwd.sto.so import soft_len
 from pmwd.sto.mlp import init_mlp_params
 
-n_epochs = 600
+n_epochs = 1000
 
 # data
 sobol_ids_global = np.arange(0, 64)
@@ -15,13 +15,31 @@ shuffle_epoch = True  # shuffle the order of sobols across epochs
 
 # optimizer
 learning_rate = 3e-5
-optimizer = optax.adam(learning_rate, b1=0.9, b2=0.9)
+optimizer = optax.adam(learning_rate)
 # optimizer = optax.chain(
 #     optax.clip_by_global_norm(3.),
 #     optax.adam(learning_rate),
 # )
 # optimizer = optax.amsgrad(learning_rate, eps=1e-4)
 # optimizer = optax.adamw(learning_rate, weight_decay=1e-4)
+
+# so neural nets
+so_type = 'NN'
+n_input = [soft_len(k_fac=3), soft_len()]
+
+so_nodes = [[3*n, 3*n, 3*n, 1] for n in n_input]
+
+# start a new training
+# so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')
+# opt_state = optimizer.init(so_params)
+
+# load and continue a training
+with open('params/3016932/e1000.pickle', 'rb') as f:
+    dic = pickle.load(f)
+    so_params = dic['so_params']
+    opt_state = dic['opt_state']
+
+
 
 def get_optimizer(lr):
     return optax.adam(lr)
@@ -47,16 +65,6 @@ def lr_scheduler(lr, skd_state, loss, patience=15, factor=0.5, threshold=1e-4, l
     skd_state = (counter, loss_low)
 
     return lr, skd_state
-
-
-# so neural nets
-so_type = 'NN'
-n_input = [soft_len(k_fac=3), soft_len()]
-
-so_nodes = [[3*n, 3*n, 3*n, 1] for n in n_input]
-so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')
-# with open('params/2980166/e107.pickle', 'rb') as f:
-#     so_params = pickle.load(f)['so_params']
 
 dropout_rate = None  # set to None for no dropout applied
 

### 2023-11-27 docs/examples/sto/train/job.sh
FIXUP add log power loss eps in hypars for tuning

@@ -2,12 +2,12 @@
 
 #SBATCH --job-name=trainso
 #SBATCH --partition=gpu
-#SBATCH --nodes=4
-#SBATCH --ntasks-per-node=4
-#SBATCH --gpus-per-node=4
-#SBATCH --cpus-per-task=16
-#SBATCH --constraint=a100-80gb
-#SBATCH --mem-per-gpu=250G
+#SBATCH --nodes=2
+#SBATCH --ntasks-per-node=8
+#SBATCH --gpus-per-node=8
+#SBATCH --cpus-per-task=8
+#SBATCH --constraint=h100-80gb
+#SBATCH --mem-per-gpu=125G
 #SBATCH --time=7-0
 #SBATCH --output=outs/%j.out
 

### 2023-11-27 pmwd/sto/hypars.py

@@ -13,6 +13,9 @@ sobol_ids_global = np.arange(0, 64)
 snap_ids = np.arange(0, 121, 1)
 shuffle_epoch = True  # shuffle the order of sobols across epochs
 
+# loss
+log_loss_eps = 1e-2
+
 # optimizer
 learning_rate = 3e-5
 optimizer = optax.adam(learning_rate)
@@ -30,14 +33,14 @@ n_input = [soft_len(k_fac=3), soft_len()]
 so_nodes = [[3*n, 3*n, 3*n, 1] for n in n_input]
 
 # start a new training
-# so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')
-# opt_state = optimizer.init(so_params)
+so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws_b1')
+opt_state = optimizer.init(so_params)
 
 # load and continue a training
-with open('params/3016932/e1000.pickle', 'rb') as f:
-    dic = pickle.load(f)
-    so_params = dic['so_params']
-    opt_state = dic['opt_state']
+# with open('params/3016932/e1000.pickle', 'rb') as f:
+#     dic = pickle.load(f)
+#     so_params = dic['so_params']
+#     opt_state = dic['opt_state']
 
 
 

### 2023-11-27 pmwd/sto/loss.py

@@ -39,14 +39,14 @@ def loss_power_w(f, g, spacing=1, log=True, w=None, cut_nyq=False):
     return loss
 
 
-def loss_power_ln(f, g, spacing=1, cut_nyq=False, eps=1e-3):
+def loss_power_ln(f, g, eps, spacing=1, cut_nyq=False):
     k, P_d, N, bins = powspec(f - g, spacing, cut_nyq=cut_nyq)
     k, P_g, N, bins = powspec(g, spacing, cut_nyq=cut_nyq)
     loss = jnp.log(P_d / P_g + eps).sum() / len(k)
     return loss
 
 
-def loss_ptcl_disp(ptcl, ptcl_t, conf):
+def loss_ptcl_disp(ptcl, ptcl_t, conf, log_loss_eps):
     # get the disp from particles' grid Lagrangian positions
     # may be necessary since we have it divided in the mse
     disp, disp_t = (ptcl_rpos(p, Particles.gen_grid(p.conf), p.conf)
@@ -57,39 +57,39 @@ def loss_ptcl_disp(ptcl, ptcl_t, conf):
     disp_t = disp_t.T.reshape(shape_)
 
     # loss = loss_mse(disp, disp_t)
-    loss = loss_power_ln(disp, disp_t)
+    loss = loss_power_ln(disp, disp_t, log_loss_eps)
     return loss
 
 
-def loss_ptcl_dens(ptcl, ptcl_t, conf, loss_mesh_shape):
+def loss_ptcl_dens(ptcl, ptcl_t, conf, log_loss_eps, loss_mesh_shape):
     # get the density fields
     (dens, dens_t), cell_size = scatter_dens((ptcl, ptcl_t), conf, loss_mesh_shape)
 
     # loss = loss_power_w(dens, dens_t)
-    loss = loss_power_ln(dens, dens_t)
+    loss = loss_power_ln(dens, dens_t, log_loss_eps)
     return loss
 
 
-def loss_snap(snap, snap_t, a_snap, conf, loss_mesh_shape=1):
+def loss_snap(snap, snap_t, a_snap, conf, log_loss_eps, loss_mesh_shape=1):
     loss = 0.
     # displacement
-    loss += loss_ptcl_disp(snap, snap_t, conf)
+    loss += loss_ptcl_disp(snap, snap_t, conf, log_loss_eps)
     # density field
-    loss += loss_ptcl_dens(snap, snap_t, conf, loss_mesh_shape)
[…]

### 2023-11-27 pmwd/sto/train.py

@@ -17,9 +17,9 @@ def pmodel(ptcl_ic, so_params, cosmo, conf):
     return obsvbl, cosmo
 
 
-def obj(tgts, ptcl_ic, so_params, cosmo, conf):
+def obj(tgts, ptcl_ic, so_params, cosmo, conf, log_loss_eps):
     obsvbl, cosmo = pmodel(ptcl_ic, so_params, cosmo, conf)
-    loss = loss_func(obsvbl, tgts, conf)
+    loss = loss_func(obsvbl, tgts, conf, log_loss_eps)
     return loss
 
 
@@ -36,12 +36,12 @@ def init_pmwd(pmwd_params):
     return ptcl_ic, cosmo, conf
 
 
-def train_step(tgts, so_params, pmwd_params, opt_params):
+def train_step(tgts, so_params, pmwd_params, opt_params, log_loss_eps):
     ptcl_ic, cosmo, conf = init_pmwd(pmwd_params)
 
     # get loss and grad
     obj_valgrad = jax.value_and_grad(obj, argnums=2)
-    loss, grad = obj_valgrad(tgts, ptcl_ic, so_params, cosmo, conf)
+    loss, grad = obj_valgrad(tgts, ptcl_ic, so_params, cosmo, conf, log_loss_eps)
 
     # average over global devices
     loss, grad = global_mean((loss, grad))
@@ -58,7 +58,7 @@ def train_step(tgts, so_params, pmwd_params, opt_params):
 
 
 def train_epoch(procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, so_params,
-                opt_state, optimizer, jax_key, verbose):
+                opt_state, optimizer, jax_key, log_loss_eps, verbose):
     loss_epoch = 0.  # the sum of loss of the whole epoch
 
     tic = time.perf_counter()
@@ -79,7 +79,7 @@ def train_epoch(procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, so_pa
                        dropout_rate, dropout_key)
         opt_params = (optimizer, opt_state)
         so_params, loss, opt_state = train_step(tgts, so_params, pmwd_params,
-                                                opt_params)
+                                                opt_params, log_loss_eps)
 
         loss = float(loss)
         loss_epoch += loss
@@ -97,7 +97,7 @@ def train_epoch(procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, so_pa
 
 
[…]

### 2023-11-27 docs/examples/sto/train/run_train.py
FIXUP add log power loss eps in hypars for tuning

@@ -111,7 +111,7 @@ def prep_train(sobol_ids_global, snap_ids):
 
 def run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch,
               learning_rate, optimizer, opt_state, so_type, so_nodes, so_params,
-              ret=False, log_id=None, verbose=True):
+              log_loss_eps, ret=False, log_id=None, verbose=True):
 
     # RNGs with fixed seeds
     # pmwd MC sampling
@@ -146,11 +146,11 @@ def run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch,
         if epoch == 0:  # evaluate the loss before training, with init so_params
             loss_epoch_mean = loss_epoch(
                 procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, so_params,
-                jax_key, verbose)
+                jax_key, log_loss_eps, verbose)
         else:
             loss_epoch_mean, so_params, opt_state, optimizer = train_epoch(
                 procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, so_params,
-                opt_state, optimizer, jax_key, verbose)
+                opt_state, optimizer, jax_key, log_loss_eps, verbose)
 
             # learning rate scheduler
             # learning_rate, skd_state = lr_scheduler(learning_rate, skd_state, loss_epoch_mean)
@@ -192,9 +192,11 @@ if __name__ == "__main__":
 
     from pmwd.sto.hypars import (
         n_epochs, sobol_ids_global, snap_ids, shuffle_epoch,
-        learning_rate, optimizer, opt_state, so_type, so_nodes, so_params)
+        learning_rate, optimizer, opt_state, so_type, so_nodes, so_params,
+        log_loss_eps)
 
     sobol_ids, gsdata = prep_train(sobol_ids_global, snap_ids)
 
     run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch,
-              learning_rate, optimizer, opt_state, so_type, so_nodes, so_params)
+              learning_rate, optimizer, opt_state, so_type, so_nodes, so_params,
+              log_loss_eps)

### 2023-12-08 pmwd/nbody.py
WIP speed up observe

@@ -187,18 +187,20 @@ def coevolve_init(a, ptcl, cosmo, conf):
 
 
 def observe(a_prev, a_next, ptcl, obsvbl, cosmo, conf):
-    def itp(a_snap, snap):
+    def itp(a_snap, i, obsvbl):
         snap_itp = interptcl(obsvbl['ptcl_prev'], ptcl, a_prev, a_next, a_snap, cosmo)
-        return snap_itp
+        obsvbl['snaps'] = obsvbl['snaps'].replace(
+            disp=obsvbl['snaps'].disp.at[i].set(snap_itp.disp),
+            vel=obsvbl['snaps'].vel.at[i].set(snap_itp.vel))
+        return obsvbl
 
-    def itp_cond(carry, x):
-        a_snap, snap = x
-        y = cond(jnp.logical_and(a_prev < a_snap, a_snap <= a_next),
-                 itp, lambda *args: snap, a_snap, snap)
-        return None, y
+    def itp_cond(obsvbl, i):
+        a_snap = obsvbl['a_snaps'][i]
+        obsvbl = cond(jnp.logical_and(a_prev < a_snap, a_snap <= a_next),
+                      itp, lambda *args: obsvbl, a_snap, i, obsvbl)
+        return obsvbl, None
 
-    # TODO maybe two copies of obsvbl['snaps'] exist here? use donate to save mem?
-    obsvbl['snaps'] = scan(itp_cond, None, (obsvbl['a_snaps'], obsvbl['snaps']))[1]
+    obsvbl, _ = scan(itp_cond, obsvbl, jnp.arange(0, len(obsvbl['a_snaps'])))
 
     obsvbl['ptcl_prev'] = ptcl
 

### 2023-12-08 pmwd/nbody.py
WIP speed up observe - using while_loop

@@ -4,7 +4,7 @@ import jax
 from jax import value_and_grad, jit, vjp, custom_vjp
 import jax.numpy as jnp
 from jax.tree_util import tree_map
-from jax.lax import cond, fori_loop, scan
+from jax.lax import cond, scan, while_loop
 
 from pmwd.boltzmann import growth
 from pmwd.cosmology import E2, H_deriv
@@ -187,20 +187,24 @@ def coevolve_init(a, ptcl, cosmo, conf):
 
 
 def observe(a_prev, a_next, ptcl, obsvbl, cosmo, conf):
-    def itp(a_snap, i, obsvbl):
+    i = jnp.searchsorted(obsvbl['a_snaps'], a_prev, side='left')
+    j = jnp.searchsorted(obsvbl['a_snaps'], a_next, side='left')
+    init_state = (i, j, obsvbl)
+
+    def cond_fun(state):
+        i, j, _ = state
+        return i < j
+
+    def body_fun(state):
+        i, j, obsvbl = state
+        a_snap = obsvbl['a_snaps'][i]
         snap_itp = interptcl(obsvbl['ptcl_prev'], ptcl, a_prev, a_next, a_snap, cosmo)
         obsvbl['snaps'] = obsvbl['snaps'].replace(
             disp=obsvbl['snaps'].disp.at[i].set(snap_itp.disp),
             vel=obsvbl['snaps'].vel.at[i].set(snap_itp.vel))
-        return obsvbl
+        return (i + 1, j, obsvbl)
 
-    def itp_cond(obsvbl, i):
-        a_snap = obsvbl['a_snaps'][i]
-        obsvbl = cond(jnp.logical_and(a_prev < a_snap, a_snap <= a_next),
-                      itp, lambda *args: obsvbl, a_snap, i, obsvbl)
-        return obsvbl, None
-
-    obsvbl, _ = scan(itp_cond, obsvbl, jnp.arange(0, len(obsvbl['a_snaps'])))
+    _, _, obsvbl = while_loop(cond_fun, body_fun, init_state)
 
     obsvbl['ptcl_prev'] = ptcl
 

### 2023-12-08 docs/papers/sto/so.tex
FIXUP draft acknowledge
FIXUP draft pcl affiliation

@@
 \author[0000-0002-9300-2632]{\normalsize Yucheng Zhang}
-\affiliation{Department of Mathematics and Theory, Peng Cheng
-Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
 \affiliation{Center for Cosmology and Particle Physics, New York
@@
 New York, New York 10010, USA}
-\affiliation{Department of Mathematics and Theory, Peng Cheng
-Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
 \affiliation{Center for Computational Astrophysics, Flatiron Institute,
@@
 \textit{\large Acknowledgements:}
+We thank Qiufan Lin, Paulo Montero-Camacho, and Yang Wang for helpful
+discussions.
 YZ and YL were supported by The Major Key Project of PCL.

### 2023-12-19 docs/papers/sto/so.tex
FIXUP draft update figures

@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sobol0snap120epoch300.pdf}
+  \includegraphics[width=.8\columnwidth]{tf_snap120_3063821_e888.pdf} \\
+  \includegraphics[width=.8\columnwidth]{cc_snap120_3063821_e888.pdf}
   \caption{The transfer function $T$ and correlation coefficient $r$ of the
   \GADGET\ and \pmwd\ simulated density fields, with and without force
-  sharpening.
-  Sobol 0 snap 120}
+  sharpening. The solid line and shaded region indicate the mean and standard
+  deviation of the 64 Sobol simulation setups. mesh shape=1}
   \label{fig:tfcc}
@@
 To compare two fields $\hat f$ and $f$, we introduce for each mode the square
-root ratio of power
+root ratio of the power spectrum
 %
@@
 sharpening optimization.
+To show the $T$ and $r$ for different box sizes, we normalize the wavenumber by
+the Nyquist wavenumber $k_{\rm Nyquist}$ of the particle grid.
+\YZ{We show T and r separately for different mesh shapes?}
 
@@
 
-\subsection{Dark Matter Halo Catalogs}
+\subsection{Dark Matter Halos}
 
@@
 
-We use Optuna \citep{optuna_2019} to search for the optimal Hyperparameters.
+We use Optuna \citep{optuna_2019} to search for the optimal hyperparameters.
 

### 2023-12-19 pmwd/sto/mlp.py
FIXUP add the squareplus func

@@ -1,9 +1,16 @@
 import jax.numpy as jnp
-from jax import random
+from jax import random, jit
 from typing import Sequence, Callable
 import flax.linen as nn
 from flax.core.frozen_dict import unfreeze, freeze
 from flax.linen.initializers import he_normal, zeros_init
+
+
+@jit
+def squareplus(x, b=4):
+  """Squareplus activation function in https://arxiv.org/abs/2112.11687."""
+  y = x + jnp.sqrt(jnp.square(x) + b)
+  return y / 2
 
 
 class MLP(nn.Module):

### 2024-02-01 docs/papers/sto/so.tex
FIXUP draft updates

--- 2023-12-08 docs/papers/sto/so.tex
@@
 We optimize the spatial resolution of cosmological particle-mesh (PM)
-simulations by sharpening the PM force with first neural networks and then
-symbolic regression, taking symmetry and dimensional analysis in consideration.
+simulations by sharpening the PM force with first neural networks and
+then symbolic regression, taking symmetry and dimensional analysis in
+consideration.
 
 We combine differentiable simulation, deep learning, [decision tree
-(feature selection)], and symbolic regression to extract the final kernel
-applicable to all other cosmological simulation code.
+(feature selection)], and symbolic regression to extract the final
+kernel applicable to all other cosmological simulation code.
 
@@
 
-Abbrs: vector-Jacobian product (VJP), automatic differentiation (AD), neural
-network (NN), particle-mesh (PM), fast-Fourier transform (FFT)
-
 % AI for science and cosmology
+Rapid advances in artificial intelligence (AI) and more specifically
+deep learning (DL) have provided powerful tools that are revolutionizing
+scientific research across disciplines.
+AI for science consists of not only the state-of-art deep neural
+networks, but also more fundamental computational tools and hardwares
+like automatic differentiation (AD) and GPUs.
+One particular application is the numerical simulation of the Universe.
 
@@
 % introduce PM and its limitation on small scales
-The PM method solves the Poisson equation by scattering the particles onto a
-mesh, on which FFT is performed and the Laplacian operator reduces to simple
-multiplication of wavenumbers in Fourier space.
-Although being computationally efficient, the PM method is mainly limited by its
-resolution on small scales, i.e. scales around the mesh cell size.
+The particle-mesh (PM) method solves the Poisson equation by scattering
+the particles onto a mesh, on which FFT is performed and the Laplacian
+operator reduces to simple multiplication of wavenumbers in Fourier
+space.
+Although being computationally efficient, the PM method is mainly
+limited by its resolution on small scales, i.e. scales around the mesh
+cell size.
 This is due to the loss of information during the scattering process of
 continuously distributed particles onto the discrete mesh.
-We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to compute the
-mass fractions of a particle scattered to nearby grid points.
-Other schemes include the nearest grid point (NGP) and triangle shaped cloud
-(TSC) etc.
+We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to
+compute the mass fractions of a particle scattered to nearby grid
+points.
+Other schemes include the nearest grid point (NGP) and triangle shaped
+cloud (TSC) etc.
 
-Using a hybrid embedding of NNs in differentiable physical simulations, we
-optimize the performance of PM method on small scales in cosmological $N$-body
-simulations by sharpening the gravitational force.
+% motivation of this work
+Optimizing the accuracy of fast and differentiable simulations is
+crucial for their applications in, e.g. generation of mocks for
+cosmological surveys, simulation-based inference or field-level
+inference.
+
+% introduction of this work
+The differentiable cosmological simulation with adjoint method
+\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
+\citep{Li2022b} lay the foundation of the gradient-based optimization
+discussed in this work.
+Within this fully differentiable framework, we sharpen the gravitational
+force by embedding neural networks (NNs) in the Poisson equation and
+train them for the spatial optimization (SO) of the simulation
+resolution on small scales.
+For better interpretability, we further fit the trained NNs using
+symbolic regression (SR).
 
@@
 
-In cosmological $N$-body simulations, the dynamics of the particles is governed
-by the Poisson equation
+In cosmological $N$-body simulations, the dynamics of the particles is
+governed by the Poisson equation
 %
@@
 %
-In Fourier space, the gravitational force field $- \vnabla \varphi$ reads
+where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
+constant, and the two fields $\delta(\vx)$ and $\varphi(\vx)$ are the
+gravitational potential and matter density fluctuation respectively.
+In Fourier space, the gravitational force field $- \vnabla \varphi$
+follows
 %
@@
 %
-where we introduce force sharpening for spatial optimization through
+where for each dimension $i\in\{1,2,3\}$ of the 3D field, we then
+introduce SO force sharpening through
 %
@@
 %
-where $f$ and $g$ are NNs, each approximating a nonlinear function of the
-wavevector components and all possible relevant features $\vvartheta$ summarized
-in \autoref{tab:feat} with more details discussed in \autoref{sec:features}.
+where $f$ and $g$ are NNs, each approximating a nonlinear function of
+the wavevector components and all possible relevant features
+$\vvartheta$ summarized in \autoref{tab:feat} with more details
+discussed in \autoref{sec:features}.
 Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
@@
 compontents to achieve the permutation equivariance.
-Being embedded in the differentiable simulation, the NNs can be trained using
-gradient-based optimization methods as in usual deep learning tasks.
-The parameters of the neural networks, which enter the force evaluation in
-every time integration step, can be treated just the same as other cosmological
-parameters.
-The objective gradients are accumulated, based on which the parameters are then
-optimized.
-For both $f$ and $g$ neural networks, we adopt multilayer perceptron (MLP, i.e.
-fully connected) structures.
+
+Being embedded in the differentiable simulation, the NNs can be trained
+using gradient-based optimization methods as in usual deep learning tasks.
+The parameters of the neural networks, which enter the force evaluation
+in every time integration step, can be treated just the same as other
+cosmological parameters.
+The objective gradients are accumulated, based on which the parameters
+are then optimized.
+For both $f$ and $g$ neural networks, we adopt multilayer perceptron
+(MLP, i.e. fully connected) structures.
 
@@
 }
-In our \pmwd\ model, we generate the snapshot at an arbitrary given time using
-the cubic Hermite interpolation, with more details in \autoref{sec:snapobs}.
+In our \pmwd\ model, we generate the snapshot at an arbitrary given time
+using the cubic Hermite interpolation, with more details in
+\autoref{sec:snapobs}.
 The curvature $\OmegaK$ is related to the separate universe simulation
@@
 
-\subsection{Trained Neural Networks}
+\subsection{Trained Neural Networks \& Symbolic Expressions}
 
-% visualization & comparison of neural nets & symbolic functions
-For the $f(k_i; \vvartheta)$ NN, we show it as a function of $k_i$ at multiple
-different $\vvartheta$ values.
+\begin{figure*}
+  \centering
+  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
+  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
+  functions }
+  \label{fig:func}
+\end{figure*}
 
-For the $g(k_1,k_2,k_3;\vvartheta)$ NN, which is permutationally symmetric about
-$(k_1, k_2, k_3)$, we show it as a function of $k_i$ along axis and face
-diagonal directions.
+With discrete Fourier transform (DFT) on a regular mesh of length $N_c$
+with cell size $l_c$, the range of sampled 1D discrete wavenumbers is
+given by the fundamental $k_{\rm fund} = 2\pi/(l_c N_c)$ and the
+maximum $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is simply the box
+size.
+To visualize the functions for different box sizes, we normzalize $k$ by
+$k_{\rm Nyquist}$, which gives the same $[2/N_c, 1]$ regardless of the
+box size.
+For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum value
+is then $\sqrt{3}$.
 
-\subsection{Feature Importance}
-
-We generate a sample of (input, output) data of the trained neural network to
-perform feature importance analysis.
-We use Sobol sequence to generate the sample of (parameters in
-\autoref{tab:param}, $k$, $a$).
-
-We use the random forest regressor for feature selection using impurity-based
-importance.
-We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
-% more details on the hyperparams in random forest
-\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
-importance?}
-
-\subsection{Symbolic Expressions}
-
-The selected important features are then used for SR.
-We use the \texttt{PySR} \citep{PySR} SR code to fit for analytic
-function expressions of the trained SO neural networks.
+For the $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally
+symmetric about $(k_1, k_2, k_3)$, we show it as a function of $k_i$
+along axis, face and body diagonal directions.
 
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{sobol0snap120epoch300.pdf}
-  \caption{The transfer function $T$ and correlation coefficient $r$ of the
-  \GADGET\ and \pmwd\ simulated density fields, with and without force
-  sharpening.
-  Sobol 0 snap 120}
+  \includegraphics[width=.48\columnwidth]{tf_snap120_3091776_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{cc_snap120_3091776_e3000.pdf}
+  \caption{The TF and CC of PM and PM-SONN simulated density fields with
+  respect to that of the \GADGET. The solid line and shaded region
+  indicate the median and the [16, 84] percentiles of the 64 Sobol
+  simulation setups. last snapshot $a\simeq 1$, mesh shape=1}
   \label{fig:tfcc}
@@
 
-% transfer function & correlation coefficient
-To compare two fields $\hat f$ and $f$, we introduce for each mode the square
-root ratio of power
+To compare two fields $\hat f$ and $f$, we introduce for each mode the
+square root ratio of the auto power spectrum
 %
@@
 %
-which is also referred to as the transfer function, and the cross correlation
-coefficient
+which is referred to as the transfer function (TF), and the correlation
+coefficient (CC) defined with the cross power spectrum
 %
@@
 The two fields are identical when both $T$ and $r$ are unity.
-The PM method is accurate on scales much larger than the cell size, therefore we
-expect $T$ and $r$ to be very close to being unity both before and after force
-sharpening optimization.
+The PM method is accurate on scales much larger than the mesh cell size,
+therefore we expect $T$ and $r$ to be very close to being unity both
+before and after the force sharpening optimization.
+To show the $T$ and $r$ for different box sizes, we normalize the
+wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
+the particle grid, where $l_p$ is the particle spacing.
+\YZ{We show T and r separately for different mesh shapes?}
+
+comparison of PM, PM with SO NN (PM-SONN), PM with SO SR (PM-SOSR)
 
@@
 % density field with zoom in (with halos painted as circles?)
-The effect of force sharpening can also be visually shown by looking directly at
-the density field.
+The effect of force sharpening can also be visually shown by looking
+directly at the density field.
 
-\subsection{Dark Matter Halo Catalogs}
+\subsection{Dark Matter Halos}
+
+\subsection{Benchmarks}
+We compare the computational cost of \GADGET, \pmwd\ with NNs, and
+\pmwd\ with SR.
 
@@
 
-The differentiable cosmological simulation with adjoint method \citep{Li2022a}
-and the corresponding PM $N$-body library \pmwd\ \citep{Li2022b} are the
-foundation of the gradient-based optimization discussed in this work.
-Within this fully differentiable framework, we are able to embed neural networks
-in physical processes and train them for many purposes, including the
-optimization of spatial resolution by sharpening the gravitational force in this
-work.
+The spatial optimization using gradient descent discussed in this work
+is built in the framework of the PM $N$-body library \pmwd\
+\citep{Li2022b}, which is an implementation of the differentiable
+cosmological simulation with adjoint method \citep{Li2022a}.
 
 Here we briefly recap the equations that are closely related to our
-extensions to the original framework, and setup the notations of the variables.
-Following \cite{Li2022a}, we denote the particles' state with $\vz = (\vx,
-\vp)^\intercal$ and their adjoint variables with $\vlambda = (\vxi, \vpi)$.
-We refer to the quantities that are directly related to the final objective
-function $\cJ$ as the observable $\vo$.
-In this work, our observable includes multiple snapshots at given times.
-For the observation operator $O_{i}^{i-1}$ in the adjoint equations (check
-Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to $a_{i-1}$, the adjoint
-variables are updated through
+extensions of the original framework, and introduce the notations of the
+variables.
+Following \cite{Li2022a}, we denote the particles' state with $\vz =
+(\vx, \vp)^\intercal$, where $\vx$ and $\vp$ are the positions and
+momenta of the particles, and their adjoint variables with $\vlambda =
+(\vxi, \vpi)$.
+We refer to the quantities that are directly related to the final
+objective function $\cJ$ as the observable $\vo$.
+In this work, our observables include multiple snapshots at given times.
+For the observation operator $O_{i}^{i-1}$ in the adjoint equations
+(check Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to
+$a_{i-1}$, the adjoint variables are updated through
 \begin{align}
@@
 \end{align}
-which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$.
+which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$, the
+state of particles at the final time step.
 The derivative of the objective function with respect to the observable
-($\p\cJ/\p\vo$) and the VJPs of the observation operation ($\p\vo/\p\vx_n$ and
-$\p\vo/\p\vp_n$, mainly the interpolation function discussed in
-\autoref{sec:snapobs} below) can be conveniently computed by AD.
+($\p\cJ/\p\vo$) and the VJPs of the observation operation
+($\p\vo/\p\vx_n$ and $\p\vo/\p\vp_n$, mainly the interpolation function
+discussed in \autoref{sec:snapobs} below) can be conveniently computed
+by AD.
 
@@
 
-The N-body time integration is carried out following discrete time steps. Given
-the particle displacements $\vx$ and velocities $\vv = \vp/(a^3H)$ at two
-consecutive time steps $a_i$ and $a_{i+1}$, we get the snapshot of particles in
-between at $a$ using
-the cubic Hermite interpolation.
+The N-body time integration is carried out along discrete time steps in
+terms of the scale factor $a$, and the state of particles at any given
+time can be achieved using interpolation.
+Given the particle displacements $\vs$ and velocities\footnote{This
+velocity is related to the canonical momentum $\vp$ through $\vv =
+\vp/(a^3H)$.} $\vv := \d\vs/\d a$ at two consecutive time steps $a_i$ and
+$a_{i+1}$, we get the snapshot of particles in between at $a$ using the
+cubic Hermite interpolation.
 The interpolated particle displacements are given by
 \begin{align}
-  \vx(a) =\ &h_{00}(\alpha)\vx_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
-           &h_{01}(\alpha)\vx_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
+  \vs(a) =\ &h_{00}(\alpha)\vs_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
+           &h_{01}(\alpha)\vs_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
 \end{align}
-where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis functions read
+where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis
+functions read
 \begin{align}
@@
 \end{align}
-The particle velocities are then given by the derivative $\d\vx / \d a$.
-Note that the interpolation is a linear combination of the particle snapshots at
-two consecutive time steps, which therefore can be done in either two steps or
-just one step by carrying the particles from the previous time step.
-We find the two-steps scheme more convenient for the backward evolution of the
-adjoint variables.
-The interpolated snapshots at different times make the observable in this work,
-and the VJPs of the interpolation operation are computed by AD and put into the
-adjoint equations as mentioned in \autoref{sec:pmwd}.
+The particle velocities are then given by the derivative $\vv(a) =
+\d\vs(a) / \d a$.
+The interpolation is a linear combination of the particle snapshots at
+two consecutive time steps, which therefore can be done in either two
+steps or just one step by carrying the particles from the previous step.
+We find the two-steps scheme more convenient for the backward evolution
+of the adjoint variables.
+The interpolated snapshots at different times make the observable in
+this work, and the VJPs of the interpolation functionx are computed by
+AD and put into the adjoint equations as mentioned in
+\autoref{sec:pmwd}.
 
@@
 \vspace{1em}
-\subsection{Training of SO Neural Networks}
+\subsection{Training of Neural Networks}
 
@@
 
+Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
+input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
+input layer, which is [...] and [...] for $f$ and $g$ respectively.
+A rectified linear unit (ReLU) activation function is used for hidden
+layers.
+We further apply an exponential function to the output as a regulator
+for positive output.
+
 The SO MLP parameters are initialized in a way such that the MLP output is very
@@
 \vspace{1em}
-\subsection{Feature Importance Analysis}
+\subsection{Symbolic Regression}
 
-\vspace{1em}
-\subsection{Symbolic Regression}
+We generate a sample of (input, output) data of the trained neural network to
+perform feature importance analysis.
+We use Sobol sequence to generate the sample of (parameters in
+\autoref{tab:param}, $k$, $a$).
+
+We use the random forest regressor for feature selection using impurity-based
+importance.
+We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
+% more details on the hyperparams in random forest
+\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
+importance?}
+
+The selected important features are then used for SR.
+We use the \texttt{PySR} \citep{PySR} SR code to fit for analytic
+function expressions of the trained SO neural networks.
 
@@
 YZ and YL were supported by The Major Key Project of PCL.
+YZ was also supported by the China Postdoctoral Science Foundation under award
+number 2023M731831.
 The Flatiron Institute is supported by the Simons Foundation.
@@
 \pmwd\ \citep{Li2022b} is open-source on GitHub
-\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source files
-and scripts of this paper
+\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
+files and scripts of this paper
 \href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
-\GADGET\ \citep{GADGET-4}, which can be obtained through a public git-repository
-\href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab}, hosted by a gitlab
-server at the Max-Planck Computing and Data Facility (MPCDF).
+\GADGET\ \citep{GADGET-4} can be obtained through a public
+git-repository \href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab},
+hosted by a gitlab server at the Max-Planck Computing and Data Facility
+(MPCDF).
 

### 2024-02-02 docs/papers/sto/so.tex
FIXUP draft updates

@@
 where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
-constant, and the two fields $\delta(\vx)$ and $\varphi(\vx)$ are the
+constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
 gravitational potential and matter density fluctuation respectively.
@@
 space.
-We take the absolute value for the evenness, and sort the 3 wavevector
-compontents to achieve the permutation equivariance.
-
-Being embedded in the differentiable simulation, the NNs can be trained
-using gradient-based optimization methods as in usual deep learning tasks.
-The parameters of the neural networks, which enter the force evaluation
-in every time integration step, can be treated just the same as other
-cosmological parameters.
-The objective gradients are accumulated, based on which the parameters
-are then optimized.
-For both $f$ and $g$ neural networks, we adopt multilayer perceptron
-(MLP, i.e. fully connected) structures.
+We take the absolute value of the wavevectors for the evenness, and sort
+the 3 wavevector compontents to achieve the permutation equivariance.
 
@@
 
-We generate our target data for training and testing with the \GADGET\
-simulation code \citep{GADGET-4}.
+Our target data samples for training and testing are generated with the
+\GADGET\ simulation code \citep{GADGET-4}.
 We run 512 \GADGET\ simulations that cover a wide range of
@@
 
-\subsection{Trained Neural Networks \& Symbolic Expressions}
+\subsection{Spatial Optimization Neural Networks}
 
@@
   \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
-  functions }
+  functions. The solid line and shaded region indicate the median and
+  the [16, 84] percentiles of the 64 Sobol simulation setups. $a=1$,
+  mesh shape=1}
   \label{fig:func}
 \end{figure*}
+
+For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
+fully connected) structures.
+Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
+input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
+input layer, which is [...] and [...] for $f$ and $g$ respectively.
+A rectified linear unit (ReLU) activation function is used for hidden
+layers.
+We further apply an exponential function to the output as a regulator
+for positive output.
+Being embedded in the differentiable simulation library \pmwd, the NNs
+can be trained using gradient-based optimization methods as in usual
+deep learning tasks.
+The parameters of the NNs, which enter the force evaluation in every
+time integration step, can be treated just the same as other
+cosmological parameters.
+The objective gradients are accumulated, based on which the parameters
+are then optimized.
 
@@
 along axis, face and body diagonal directions.
+
+\subsection{Analytic Expressions with Symbolic Regression}
+
+We fit the trained NNs with symbolic regression.
 
@@
 
-The loss function is constructed by comparing the simulated field $\hat f$ with
-the target field $f$.
+The loss function is constructed by comparing the simulated field $\hat
+f$ with the target field $f$.
 \YL{
@@
 
-We combine Lagrangian and Eulerian fields to form the total loss for a single
-snapshot:
+We combine Lagrangian and Eulerian fields to form the total loss for a
+single snapshot:
 %
@@
 
-\YZ{We further sum the loss of all snapshots across a single forward simulation
-at different times, i.e.
+\YZ{We further sum the loss of all snapshots across a single forward
+simulation at different times, i.e.
 \begin{equation}
@@
 
-Small numerical differences in initial conditions (ICs) could become more
-significant with the $N$-body evolution.
+Small numerical differences in initial conditions (ICs) could become
+more significant with the $N$-body evolution.
 We have verified that the IC generation on GPUs is deterministic though
-different from those on CPUs, therefore we generate ICs for both the training
-data and the model on GPUs.
+different from those on CPUs, therefore we generate ICs for both the
+training data and the model on GPUs.
 
@@
 
-Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
-input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
-input layer, which is [...] and [...] for $f$ and $g$ respectively.
-A rectified linear unit (ReLU) activation function is used for hidden
-layers.
-We further apply an exponential function to the output as a regulator
-for positive output.
-
 The SO MLP parameters are initialized in a way such that the MLP output is very
@@
 \vspace{1em}
-\subsection{Symbolic Regression}
-
-We generate a sample of (input, output) data of the trained neural network to
-perform feature importance analysis.
+\subsection{Symbolic Regression Fitting of the Neural Networks}
+
+We generate samples of (input, output) data of the trained neural
+networks to perform symbolic regression.
 We use Sobol sequence to generate the sample of (parameters in

### 2024-02-06 docs/papers/sto/so.tex
FIXUP draft updates

@@
 
-% hardware setups
-We perform distributed data-parallel training using GPU devices hosted on
-multiple CPU nodes.
-With each process being binded to a single GPU device, the training data samples
-are uniformly distributed to the processes.
-In total, we use [...] \GPU\ GPUs, each accompanied with [...] CPU cores and
-[...] GB CPU memory mainly for preloading the training data.
-In each training step, a data sample is fed into the GPU from the CPU memory
-where the preloading takes place to save the reading time from the disk.
-
-Following \autoref{tab:param}, in each iteration, we randomly sample the number
-of time steps and the mesh shape, and configure \pmwd\ with the same initial
-coniditions, cosmological parameters and other common configurations.
-To balance the work of GPUs during training, in each iteration we use
-the same number of time steps and mesh shape on all GPUs.
-
-As discussed in \autoref{sec:loss}, the loss in a single simulation sums over
-all snapshots.
-Our batch size is [...], i.e. the number of GPU devices since one simulation is
-run on a given GPU and compared to the corresponding training data sample for
-loss.
-For each training step, the loss and gradients are collected and averaged across
-all GPU devices.
-The SO parameters are initialized to be the same for all devices, and being
-updated with the same averaged gradients.
-
-The SO MLP parameters are initialized in a way such that the MLP output is very
-close to one, i.e. no SO.
-We achieve this by setting the kernel of the output layer to be very small
-random values and the bias to be one.
+We perform distributed data-parallel training using GPU devices hosted
+on multiple CPU nodes.
+With each process being binded to a single GPU device, the training data
+samples are uniformly distributed to the processes.
+In total, we use [...] \GPU\ GPUs, each accompanied with [...] CPU cores
+and [...] GB CPU memory mainly for preloading the training data.
+In each training step, a data sample is fed into the GPU from the CPU
+memory where the preloading takes place to save the reading time from
+the disk.
+
+For each training step, we configure and run \pmwd\ with the same
+initial coniditions, cosmological parameters and other common
+configurations as the training data sample.
+As discussed in \autoref{sec:loss}, the loss in a single simulation sums
+over all snapshots.
+Our batch size is [...], i.e. the number of GPU devices, since one
+simulation is run on a single GPU and compared to the corresponding
+training data sample for loss.
+Then the loss and gradients are collected and averaged across all GPU
+devices, with which the optimizer is also updated.
+The SO parameters are initialized to be the same for all devices, and
+hence remain the same for all processes given the same averaged
+gradients.
+We use the Adam \citep{Adam} optimizer.
+
+The SO NN parameters are initialized in a way such that the output
+is very close to one, i.e. no SO.
+We achieve this by setting the kernel of the output layer to be very
+small random values and the bias to be zero, which is around one after
+the exponential function.
 
@@
 
-We use Optuna \citep{optuna_2019} to search for the optimal hyperparameters.
-
-We empoly a learning rate scheduler that reduces the learning rate on plateau
-where the loss function stops decreasing.
-The initial learning rate is set at [...] with a reduction factor of [...] after
-[...] epochs of no decreasing on loss.
-
-We implement the neural networks with \texttt{Flax} \citep{flax2020github}, and
-use the optimizers in \texttt{Optax}, part of the DeepMind JAX Ecosystem
+We use Optuna \citep{optuna_2019} to search for the optimal
+hyperparameters.
+
+We empoly a learning rate scheduler that reduces the learning rate on
+plateau where the loss function stops decreasing.
+The initial learning rate is set at [...] with a reduction factor of
+[...] after [...] epochs of no decreasing on loss.
+
+We implement the neural networks with \texttt{Flax}
+\citep{flax2020github}, and use the optimizers in \texttt{Optax}, part
+of the DeepMind JAX Ecosystem
 \citep{deepmind2020jax}.
@@
 \vspace{1em}
-\subsection{Symbolic Regression Fitting of the Neural Networks}
+\subsection{Symbolic Regression Fitting}
 

### 2024-02-06 docs/papers/adjoint/adjoint.tex
Finalize ApJS adjoint paper

@@
 
+\clearpage
 \end{abstract}
 
-\vspace{1em}
 \section{Introduction}
@@
 and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
-advent of modern AD systems, and were based on \HL{the analytic derivatives,
+advent of modern AD systems, and were based on the analytic derivatives,
 which involve first convoluted derivation by hand using the chain rule
 \citep[see e.g.,][App.~D]{SeljakEtAl2017} before implementing them in
-code.}
+code.
 Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
@@
 \texttt{TensorFlow}, respectively.
-\HL{The AD frameworks automatically apply the chain rule to the primitive
+The AD frameworks automatically apply the chain rule to the primitive
 operations that comprise the whole simulation, relieving the burden of
-derivation and implementation of the derivatives.}
+derivation and implementation of the derivatives.
 Both analytic differentiation and AD backpropagate the gradients through
@@
 
-\HL{Alternatively, the adjoint method provides systematic ways of deriving
+Alternatively, the adjoint method provides systematic ways of deriving
 the gradients of an objective function $\cJ$ under constraints
@@
 $\vtheta$-dependence at one stage of the modeling in
-\autoref{fig:model}.}
+\autoref{fig:model}.
 
@@
 \citep{NeuralODE}.
-Furthermore, we \HL{derive the discrete adjoint equations dual to the
+Furthermore, we derive the discrete adjoint equations dual to the
 discrete forward time integration, known as the discretize-then-optimize
@@
 larger deviations between the forward and the backward trajectories due
-to different discretizations \citep{LanzieriLanusseEtAl2022}.}
+to different discretizations \citep{LanzieriLanusseEtAl2022}.
 In brief, to compute the gradient we only need to evolve a simulation
@@
   \node (boltz) {Boltz};
-  \node (ic) [right=of boltz] {IC} edge [from] (boltz);
+  \node (ic) [right=of boltz] {LPT} edge [from] (boltz);
   \node (integ) [right=of ic] {Integ} edge [from] (ic);
@@
 \begin{tikzpicture}[node style]
-  \node (ic) {IC};
+  \node (ic) {LPT};
   \node (integ_1) [right=of ic] {$I_1$} edge [from] (ic);
@@
 We describe different operators in \autoref{sec:fwd}: Boltzmann solver
-(``Boltz'') and initial condition generator (``IC'') in
-\autoref{sec:ic}; force solver ($F$) in \autoref{sec:force}; time
-integration (``Integ''), kick ($K$), and drift ($D$) in
-\autoref{sec:integ}; observation (``Obs'' and $O$) and objective
-(``Obj'') in \autoref{sec:obsobj}.
+(``Boltz'') and initial condition generator by the Lagrangian
+perturbation theory (``LPT'') in \autoref{sec:ic}; force solver ($F$) in
+\autoref{sec:force}; time integration (``Integ''), kick ($K$), and drift
+($D$) in \autoref{sec:integ}; observation (``Obs'' and $O$) and
+objective (``Obj'') in \autoref{sec:obsobj}.
 Gradients flow backward with all arrows reversed (\autoref{sec:adj}).
@@
 equation, using the facts that $\d \bm{M}^{-1} / \d t = - \bm{M}^{-1} \,
-\dot{\bm{M}} \, \bm{M}^{-1}$ \HL{for any invertible matrix $\bm{M}$}, and
+\dot{\bm{M}} \, \bm{M}^{-1}$ for any invertible matrix $\bm{M}$, and
 $(\p\vz / \p\tilde\vz)^{-1} = \p\tilde\vz / \p\vz$.
@@
 where we have used the fact that $D(t_c, t_a, t_b) = - D(t_c, t_b,
-t_a)$, and left the $\p\cJ / \p\vz_i$ term, \HL{the explicit dependence of
+t_a)$, and left the $\p\cJ / \p\vz_i$ term, the explicit dependence of
 the objective on the intermediate states (from, e.g., observables on the
-light cone), to the observation operator $O$ below.}
+light cone), to the observation operator $O$ below.
 This also naturally determines the subscripts of $\xi$ and $\pi$.
@@
       \bigr) K(t_i, t_i, t_{i-\half})
-    + \vxi_i \cdot \vp_{i-\half}
+    + \vxi_{i-\half} \cdot \vp_{i-\half}
       \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \nonumber\\
@@
 %
-\HL{where the latter backpropagates from fewer sources than the former as
-shown in \autoref{fig:model}.}
+where the latter backpropagates from fewer sources than the former as
+shown in \autoref{fig:model}.
 To implement \eqref{KDFKO_adj}-\eqref{KDFKO_grad} in \pmwd\ with
@@
 practice the reconstructed trajectory can differ from the forward one
-due to \HL{the finite numerical precision and exacerbated by the chaotic
+due to the finite numerical precision and exacerbated by the chaotic
 dynamics.
 Better reversibility means the gradients propagate backward along a
-trajectory closer to the forward path, and thus would be more accurate.}
+trajectory closer to the forward path, and thus would be more accurate.
 To test this, in \autoref{tab:rev} we compare the RMSDs of the
@@
 Their smallness is the main reason that the quoted relative differences
-here are \HL{orders of magnitude greater than those in \autoref{tab:rep}
+here are orders of magnitude greater than those in \autoref{tab:rep}
 (see \autoref{fig:grads} where the reversibility is few times more
-important than the reproducibility in their impacts on the gradients).}
+important than the reproducibility in their impacts on the gradients).
 With the same setup as that in \autoref{tab:rep}, we see the same
@@
 $2^3\times$ from the listed specifications in the caption.
-\HL{For better statistics, we repeat both adjoint and AD runs for 64 times,
+For better statistics, we repeat both adjoint and AD runs for 64 times,
 with the same cosmology and white noise modes, and compare their results
@@
 Though this statement should be verified again in the future when we
-reduce the reversibility errors using, e.g., 3LPT.}
+reduce the reversibility errors using, e.g., 3LPT.
 %NOTE: this behavior has disappeared with new JAX/CUDA/GPU
@@
 time steps, and single precision.
-\HL{We choose a mean squared error (MSE) objective between two realizations
+We choose a mean squared error (MSE) objective between two realizations
 with the same cosmology but different initial modes, on their density
 fields on the $256^3$ mesh at $a=1$, and then compute the gradients with
-respect to one realization while holding the other fixed.}
+respect to one realization while holding the other fixed.
 We compare the adjoint gradients to those by AD, for which we have
@@
 pattern, in hundreds to thousands of iterations.
-\HL{Interestingly, we find that the variance of the modes becomes bigger
+Interestingly, we find that the variance of the modes becomes bigger
 than 1 (that of the standard normal white noise) after the optimization,
@@
 present in the white noise fields, suggesting that the optimized initial
-conditions are probably no longer Gaussian.}
+conditions are probably no longer Gaussian.
 
@@
 YL and YZ were supported by The Major Key Project of PCL.
+YZ was supported by the China Postdoctoral Science Foundation under
+award number 2023M731831.
 The Flatiron Institute is supported by the Simons Foundation.
@@
 
-\vspace{1em}
+\clearpage
 \appendix

### 2024-02-07 docs/papers/adjoint/adjoint.tex
Show hidden appendices

@@
 \usepackage{nicefrac}
+\usepackage{booktabs}
 \usepackage[caption=false]{subfig}
-\usepackage[figure,figure*]{hypcap}
+\usepackage[figure, figure*]{hypcap}
 \usepackage{tikz}
   \usetikzlibrary{positioning, fit, calc, arrows.meta}
-\usepackage{booktabs}
+\usepackage{framed}
 
@@
 
-%\vspace{1em}
-%\section{Hamiltonian Dynamics}
-%
-%Consider a Hamiltonian system, with $\vz = (\vx, \vp)$ where $\vx$ and
-%$\vp$ are the canonical coordinates and momenta.
-%Other variables split likewise: $\vDelta = (\vDelta_\vx, \vDelta_\vp)$,
-%$\vLambda = (\vLambda_\vx, \vLambda_\vp)$, and $\vlambda =
-%(\vlambda_\vx, \vlambda_\vp)$.
-%Hamilton's equation of motion is
-%%
-%\begin{equation}
-%\vf = \begin{pmatrix}
-%  \hphantom{-} \p\cH/\p \vp \\ - \p\cH/\p \vx
-%\end{pmatrix}.
-%\label{Hamilton}
-%\end{equation}
-%%
-%The dynamics is time-reversible: flipping the arrow of time $\d/\d t \to
-%- \d/\d t$ and the direction of momenta $\vp \to -\vp$ preserve the form
-%of \eqref{ode}.
-%It can be reversible even after discretization if appropriate integrator
-%is used.
-%
-%The variational equation \eqref{variational} and the adjoint equation
-%\eqref{adjoint} become
-%%
-%\begin{align}
-%\begin{pmatrix}
-%  \dot\vDelta_{\vx\vx} & \dot\vDelta_{\vx\vp} \\
-%  \dot\vDelta_{\vp\vx} & \dot\vDelta_{\vp\vp}
-%\end{pmatrix}
-%&= \begin{pmatrix}
-%  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vx}
-%  & \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp} \\
-%  - \dfrac{\p^2\cH}{\p\vx\p\vx}
-%  & - \dfrac{\p^2\cH}{\p\vx\p\vp} \\
-%\end{pmatrix}
-%\cdot
-%\begin{pmatrix}
-%  \vDelta_{\vx\vx} & \vDelta_{\vx\vp} \\
-%  \vDelta_{\vp\vx} & \vDelta_{\vp\vp}
-%\end{pmatrix}, \nonumber \\
-%%
-%\qquad
-%- \begin{pmatrix}
-%  \dot\vLambda_{\vx\vx} & \dot\vLambda_{\vx\vp} \\
-%  \dot\vLambda_{\vp\vx} & \dot\vLambda_{\vp\vp}
-%\end{pmatrix}
-%&= \begin{pmatrix}
-%  \hphantom{-} \dfrac{\p^2\cH}{\p\vx\p\vp}
-%  & - \dfrac{\p^2\cH}{\p\vx\p\vx} \\
-%  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp}
-%  & - \dfrac{\p^2\cH}{\p\vp\p\vx} \\
-%\end{pmatrix}
-%\cdot
-%\begin{pmatrix}
-%  \vLambda_{\vx\vx} & \vLambda_{\vx\vp} \\
-%  \vLambda_{\vp\vx} & \vLambda_{\vp\vp}
-%\end{pmatrix}.
-%\end{align}
-%%
-%Interestingly, their forms are identical once we swap rows of
-%$\vLambda$, and flip the directions of time and momenta.
-%
-%As discussed before, $\vlambda$ follows adjoint equations similar to
-%that of $\vLambda$ above.
-%One can integrate it and \eqref{Hamilton} backward together to compute
-%$\d\cJ / \d\vtheta$.
-%
-%
-%\vspace{1em}
-%\section{\texorpdfstring{$N$}{N}-body Dynamics}
-%
-%We model self-gravitating collisionless dark matter with $N$ particles
-%each of mass $m$ in a periodic box of comoving volume $V$, whose
-%physical length expands proportional to a scale factor $a(t)$.
-%The canonical coordinates are the comoving positions $\vx = (\vx_1,
-%\cdots, \vx_N)$, and their canonical momenta are $\vp = (\vp_1, \cdots,
-%\vp_N)$ with $\vp_i = a^2 m \dot\vx_i$ for each particle.
-%
-%The $N$-body Hamiltonian is
-%%
-%\begin{equation}
-%\cH = \sum_i \frac{\vp_i^2}{2 m a^2}
-%+ \sum_{i<j} \frac{m^2 \varphi(\vx_i - \vx_j)}{a},
-%\end{equation}
-%%
-%where $\varphi$ is the potential of a unit-mass particle at $a=1$ under
-%periodic boundary condition
-%%
-%\begin{equation}
-%\nabla^2 \varphi
-%= 4\pi G \bigl[ \deltaD(\vx - \vx') - V^{-1} \bigr].
-%\end{equation}
-%%
-%So the only nonzero second $\cH$ derivatives are
-%%
-%\begin{equation}
-%\frac{\p^2\cH}{\p\vp_i\p\vp_j} = \frac{\vI_3}{2 m a^2},
-%\qquad
-%\frac{\p^2\cH}{\p\vx_i\p\vx_j}
-%= \deltaK_{ij} \sum_{k \neq i} \frac{m^2}a
-%  \frac{\p\varphi(\vx_i-\vx_k)}{\p\vx_i\p\vx_i}
-%+ \frac{m^2}a \frac{\p\varphi(\vx_i-\vx_j)}{\p\vx_i\p\vx_j},
-%\end{equation}
-%%
-%where $\vI_3$ is the $3\times3$ identity matrix, the sum is the tidal
-%tensor at $\vx_i$, and the second term vanishes when $i=j$.
+\begin{framed}
+\vspace{1em}
+\section*{Hamiltonian Dynamics}
+
+Consider a Hamiltonian system, with $\vz = (\vx, \vp)$ where $\vx$ and
+$\vp$ are the canonical coordinates and momenta.
+Other variables split likewise: $\vDelta = (\vDelta_\vx, \vDelta_\vp)$,
+$\vLambda = (\vLambda_\vx, \vLambda_\vp)$, and $\vlambda =
+(\vlambda_\vx, \vlambda_\vp)$.
+Hamilton's equation of motion is
+%
+\begin{equation}
+\vf = \begin{pmatrix}
+  \hphantom{-} \p\cH/\p \vp \\ - \p\cH/\p \vx
+\end{pmatrix}.
+\label{Hamilton}
+\end{equation}
+%
+The dynamics is time-reversible: flipping the arrow of time $\d/\d t \to
+- \d/\d t$ and the direction of momenta $\vp \to -\vp$ preserve the form
+of \eqref{ode}.
+It can be reversible after discretization if appropriate integrator is
+used.
+
+The variational equation \eqref{variational} and the adjoint equation
+\eqref{adjoint} become
+%
+\begin{align}
+\begin{pmatrix}
+  \dot\vDelta_{\vx\vx} & \dot\vDelta_{\vx\vp} \\
+  \dot\vDelta_{\vp\vx} & \dot\vDelta_{\vp\vp}
+\end{pmatrix}
+&= \begin{pmatrix}
+  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vx}
+  & \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp} \\
+  - \dfrac{\p^2\cH}{\p\vx\p\vx}
+  & - \dfrac{\p^2\cH}{\p\vx\p\vp} \\
+\end{pmatrix}
+\cdot
+\begin{pmatrix}
+  \vDelta_{\vx\vx} & \vDelta_{\vx\vp} \\
+  \vDelta_{\vp\vx} & \vDelta_{\vp\vp}
+\end{pmatrix}, \nonumber \\
+%
+\qquad
+- \begin{pmatrix}
+  \dot\vLambda_{\vx\vx} & \dot\vLambda_{\vx\vp} \\
+  \dot\vLambda_{\vp\vx} & \dot\vLambda_{\vp\vp}
+\end{pmatrix}
+&= \begin{pmatrix}
+  \hphantom{-} \dfrac{\p^2\cH}{\p\vx\p\vp}
+  & - \dfrac{\p^2\cH}{\p\vx\p\vx} \\
+  \hphantom{-} \dfrac{\p^2\cH}{\p\vp\p\vp}
+  & - \dfrac{\p^2\cH}{\p\vp\p\vx} \\
+\end{pmatrix}
+\cdot
+\begin{pmatrix}
+  \vLambda_{\vx\vx} & \vLambda_{\vx\vp} \\
+  \vLambda_{\vp\vx} & \vLambda_{\vp\vp}
+\end{pmatrix}.
+\end{align}
+%
+Interestingly, their forms are identical once we swap rows of
+$\vLambda$, and flip the directions of time and momenta.
+
+As discussed before, $\vlambda$ follows adjoint equations similar to
+that of $\vLambda$ above.
+One can integrate it and \eqref{Hamilton} backward together to compute
+$\d\cJ / \d\vtheta$.
+
+\vspace{1em}
+\section*{\texorpdfstring{$N$}{N}-body Dynamics}
+
+We model self-gravitating collisionless dark matter with $N$ particles
+each of mass $m$ in a periodic box of comoving volume $V$, whose
+physical length expands proportional to a scale factor $a(t)$.
+The canonical coordinates are the comoving positions $\vx = (\vx_1,
+\cdots, \vx_N)$, and their canonical momenta are $\vp = (\vp_1, \cdots,
+\vp_N)$ with $\vp_i = a^2 m \dot\vx_i$ for each particle.
+
+The $N$-body Hamiltonian is
+%
+\begin{equation}
+\cH = \sum_i \frac{\vp_i^2}{2 m a^2}
++ \sum_{i<j} \frac{m^2 \varphi(\vx_i - \vx_j)}{a},
+\end{equation}
+%
+where $\varphi$ is the potential of a unit-mass particle at $a=1$ under
+periodic boundary condition
+%
+\begin{equation}
+\nabla^2 \varphi
+= 4\pi G \bigl[ \deltaD(\vx - \vx') - V^{-1} \bigr].
+\end{equation}
+%
+So the only nonzero second $\cH$ derivatives are
+%
+\begin{equation}
+\frac{\p^2\cH}{\p\vp_i\p\vp_j} = \frac{\vI_3}{2 m a^2},
+\qquad
+\frac{\p^2\cH}{\p\vx_i\p\vx_j}
+= \deltaK_{ij} \sum_{k \neq i} \frac{m^2}a
+  \frac{\p\varphi(\vx_i-\vx_k)}{\p\vx_i\p\vx_i}
++ \frac{m^2}a \frac{\p\varphi(\vx_i-\vx_j)}{\p\vx_i\p\vx_j},
+\end{equation}
+%
+where $\vI_3$ is the $3\times3$ identity matrix, the sum is the tidal
+tensor at $\vx_i$, and the second term vanishes when $i=j$.
+\end{framed}
 
@@
 
-%\listofchanges
-
 \end{document}

### 2024-02-26 docs/papers/sto/so.tex
FIXUP draft update adjoint paper bib
FIXUP draft updates

@@
 
+% Main text: up to 3500 words: introduction + results + discussion
+% so we should keep introduction and discussion brief
+\vspace{1em}
+\section{Introduction}
+% without heading
+
+% AI for science and cosmology
+Rapid advances in artificial intelligence (AI) and more specifically
+deep learning (DL) have provided powerful tools that are revolutionizing
+scientific research across disciplines.
+AI for science consists of not only the state-of-art deep neural
+networks, but also more fundamental computational tools and hardwares
+like automatic differentiation (AD) and GPUs.
+One particular application is the numerical simulation of the Universe.
+
+% a brief review of (differentiable) cosmological simulations
+BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
+\texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
+
+% introduce PM and its limitation on small scales
+The particle-mesh (PM) method solves the Poisson equation by scattering
+the particles onto a mesh, on which FFT is performed and the Laplacian
+operator reduces to simple multiplication of wavenumbers in Fourier
+space.
+Although being computationally efficient, the PM method is mainly
+limited by its resolution on small scales, i.e. scales around the mesh
+cell size.
+This is due to the loss of information during the scattering process of
+continuously distributed particles onto the discrete mesh.
+We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to
+compute the mass fractions of a particle scattered to nearby grid
+points.
+Other schemes include the nearest grid point (NGP) and triangle shaped
+cloud (TSC) etc.
+
+% motivation of this work
+Optimizing the accuracy of fast and differentiable simulations is
+crucial for their applications in, e.g. generation of mocks for
+cosmological surveys, simulation-based inference or field-level
+inference.
+
+% introduction of this work
+The differentiable cosmological simulation with adjoint method
+\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
+\citep{Li2022b} lay the foundation of the gradient-based optimization
+discussed in this work.
+Within this fully differentiable framework, we sharpen the gravitational
+force by embedding neural networks (NNs) in the Poisson equation and
+train them for the spatial optimization (SO) of the simulation
+resolution on small scales.
+For better interpretability, we further fit the trained NNs using
+symbolic regression (SR).
+
+\vspace{1em}
+\section{Results}
+% Results should be divided by topical subheadings
+% Put main ideas in Results first, then more technical details in Methods.
+
+\subsection{Optimization of Dynamics}
+
+\begin{table*}
+  \centering
+  \caption{Ranges of \GADGET\ and \pmwd\ configurations and cosmological
+  parameters.
+  Note that the grid ratio need next fast len to determine the mesh shape
+  for fast FFT.
+  Given the box size, the mesh shape determines the cell size.
+  The softening parameter gives the ratio of the comoving softening length
+  to the mean particle spacing.
+  The curvature $\OmegaK$ is related to the separate universe simulation.
+  We sample parameters applicable to \GADGET\ during data generation, and
+  those applicable to \pmwd\ during training.
+  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
+  Now it seems more natural to me to have the ptcl spacing sampled
+  uniformly instead...)}
+  }
+  \label{tab:param}
+  \begin{tabular}{lcccr}
+  \toprule
+  applicability & parameter & distribution & range & sampling \\
+  \midrule
+  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
+  \cmidrule(lr){2-5}
+  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
+  \cmidrule(lr){2-4}
+  & snapshot offset & uniform & (0, 1] \\
+  \cmidrule(lr){2-4}
+  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
+  \cmidrule(lr){2-4}
+  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
+  \cmidrule(lr){2-4}
+  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
+  \cmidrule(lr){2-4}
+  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
+  \cmidrule(lr){2-4}
+  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
+  \cmidrule(lr){2-4}
+  & $h$ & log-uniform & [0.5, 1) \\
+  \cmidrule(lr){1-4}
+  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
+  \cmidrule(lr){1-5}
+  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
+  \cmidrule(lr){2-4}
+  \cmidrule(lr){1-4}
+  training & loss grid offset & uniform & $[0, 1)^3$ \\
+  \bottomrule
+  \end{tabular}
+  \end{table*}
+
+\begin{table}
+  \centering
+  \caption{Possible features that can affect force sharpening.
+  All scales are then multiplied by the wavenumber $k$ to form dimensionless
+  factors.}
+  \label{tab:feat}
+  \begin{tabular}{rl}
+  \toprule
+  type & feature \\
+  \midrule
+  & $R_\mathrm{d}$: RMS linear displacement \\
+  \cmidrule(lr){2-2}
+  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  & $\d R / \d\lna$: time derivatives of the above \\
+  \midrule
+  & particle spacing \\
+  \cmidrule(lr){2-2}
+  other scales & $l$: PM force mesh cell size \\
+  \cmidrule(lr){2-2}
+  & softening length \\
+  \midrule
+  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
+  \cmidrule(lr){2-2}
+  & $\d\ln G_m / \d\lna$ \\
+  \cmidrule(lr){2-2}
+  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
+  \cmidrule(lr){2-2}
+  & $\d\ln\!H / \d\lna$ \\
+  \cmidrule(lr){2-2}
+  & $\Delta\lna$: time step size in $\lna$ \\
+  & \YL{Now this can be an important feature with varying $a_\ic$} \\
+  \bottomrule
+  \end{tabular}
+  \end{table}
+
+Cosmological $N$-body simulations evolve dark matter particles under the
+gravitational force.
+The dynamics of the particles is governed by the Poisson equation
+%
+\begin{equation}
+  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
+\end{equation}
+%
+where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
+constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
+gravitational potential and matter density fluctuation respectively.
+In Fourier space, the gravitational force field $- \vnabla \varphi$
+follows
+%
+\begin{equation}
+- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\end{equation}
+%
+we then introduce SO force sharpening through
+%
+\begin{equation}
+\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
+  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
+\end{equation}
+%
+where $f$ and $g$ are NNs, each approximating a nonlinear function of
+the wavevector components and all possible relevant features
+$\vvartheta$.
+The subscript $i\in\{1,2,3\}$ denotes the 3 principal axes of the
+simulation box.
+
+Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
+broken by the PM solver to the O\textsubscript{h} discrete symmetry with
+48 elements involving permutations and reversals of the 3 principal axes
+of the simulation box.
+The PM evaluation of each force component can further break the symmetry
+along that axis, leaving the remaining 2 permutationally symmetric.
+Therefore, we implement $f$ and $g$ to be even functions of the
+wavevector components, and $g$ to be permutation equivariant in all 3
+components, so that $f g$ is the most general modification to the force
+kernel.
+An even and real kernel in configuration space is also even in Fourier
+space.
+We take the absolute value of the wavevectors for the evenness, and sort
+the 3 wavevector compontents to achieve the permutation equivariance.
+
+Our target data samples are generated with the \GADGET\ simulation code
+\citep{GADGET-4}.
+We run 512 \GADGET\ simulations that cover a wide range of
+configurations and cosmological parameters, as listed in
+\autoref{tab:param}.
+\YL{
+Each simulation evolves $128^3$ particles from the initial time $a_\ic$
+to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
+factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
+We determine $a_\ic$ for each configuration depending on its cosmology
+and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why
+and how here.)} which then sets the snapshot spacing $\Delta a = (1 -
+a_\ic) / 64$.
+To sample different alignment between the snapshots and the start/stop
+times and avoid possible related numerical errors in \pmwd, we introduce
+a fractional random offset $\gamma$ that varies between 0 and 1 for
+different simulations.
+}
+The curvature $\OmegaK$ is related to the separate universe simulation
+\citep{LiEtAl2014, WagnerEtAl2015}.
+\YZ{We refer to this training dataset of 512 \GADGET\ simulations
+sampled with Sobol sequence as \GSDATA.}
+On average each \GADGET\ simulation takes roughly 2K CPU hours to run,
+and the generation of the whole \GSDATA\ dataset consumes 1M CPU hours.
+
+\YL{
+In this work, we train \pmwd\ using $XXX$ time steps from $a_\ic$ to
+$a_\f$, generate snapshots at the same times as in \GADGET, and compare
+them in the loss function.}
+In our \pmwd\ model, the snapshot at an arbitrary given time is
+generated using the cubic Hermite interpolation, with more details in
+\autoref{sec:snapobs}.
+\YL{Although the number of time steps are fixed here, we emphasize that
+our spatial optimization of course also applies for different number of
+time steps, which we will explore, together with the optimal choice of
+time variable, in a follow-up work on the temporal optimization.
+}
+
+\subsection{Spatial Optimization Neural Networks}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
+  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
+  functions. The solid line and shaded region indicate the median and
+  the [16, 84] percentiles of the 64 Sobol simulation setups. $a=1$,
+  mesh shape=1}
+  \label{fig:func}
+\end{figure*}
+
+For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
+fully connected) structures to learn the nonlinear functions.
+Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
+input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
+input layer, which is [...] and [...] for $f$ and $g$ respectively.
+The rectified linear unit (ReLU) activation function is used for hidden
+layers.
+We further apply the exponential function to the output as a regulator
+for positive output.
+Being embedded in the differentiable simulation library \pmwd, the NNs
+can be trained using gradient-based optimization methods as in usual
+deep learning tasks.
+The parameters of the NNs, which enter the force evaluation in every
+time integration step, can be treated just the same as other
+cosmological parameters.
+The objective gradients are accumulated, based on which the parameters
+are then optimized.
+
+With discrete Fourier transform (DFT) on a regular grid of length $N_c$
+and cell size $l_c$, the range of sampled 1D discrete wavenumbers is
+limited by the fundamental and Nyquist values, i.e. $k_{\rm fund} =
+2\pi/(l_c N_c)$ and $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is
+simply the box size.
+Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
+$[2/N_c, 1]$ for different box sizes.
+For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum
+normalized value is then $\sqrt{3}$.
+
+For $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally symmetric
+about $(k_1, k_2, k_3)$, we show it as a function of the wavevector
+along axis, face, and body diagonal directions.
+
+\subsection{Symbolic Expressions}
+
+We fit the trained NNs with symbolic regression for analytic
+expressions of the $f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3;
+\vvartheta)$ functions.
+
+\subsection{Dark Matter Density Fields}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.48\columnwidth]{tf_snap120_3091776_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{cc_snap120_3091776_e3000.pdf}
+  \caption{The TF and CC of PM and PM-SONN simulated density fields with
+  respect to that of the \GADGET. The solid line and shaded region
+  indicate the median and the [16, 84] percentiles of the 64 Sobol
+  simulation setups. last snapshot $a\simeq 1$, mesh shape=1}
+  \label{fig:tfcc}
+\end{figure*}
+
+Improvements of SO on the dark matter density fields are the first
+things to check.
+We construct these fields by scattering the particles onto a mesh.
+To compare two fields $\hat f$ and $f$, we introduce for each mode the
+square root ratio of the auto power spectrum
+%
+\begin{equation}
+T(\vk) \triangleq
+\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
+\label{T}
+\end{equation}
+%
+which is referred to as the transfer function (TF), and the correlation
+coefficient (CC) defined with the cross power spectrum
+%
+\begin{equation}
+r(\vk) \triangleq
+\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
+     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
+\label{r}
+\end{equation}
+%
+The two fields are identical when both $T$ and $r$ are unity.
+The PM method is accurate on scales much larger than the mesh cell size,
+therefore we expect $T$ and $r$ to be very close to being unity both
+before and after the force sharpening optimization.
+To show the $T$ and $r$ for different box sizes, we normalize the
+wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
+the particle grid, where $l_p$ is the particle spacing.
+\YZ{We show T and r separately for different mesh shapes?}
+
+comparison of PM, PM with SO NN (PM-SONN), PM with SO SR (PM-SOSR)
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=.8\columnwidth]{denslab_0_120_e300.pdf}
+  \caption{Projected dark matter density fields of \GADGET, \pmwd\ without and
+  with force sharpening.
+  Sobol 0 snap 120}
+  \label{fig:denslab}
+\end{figure*}
+
+% density field with zoom in (with halos painted as circles?)
+The effect of force sharpening can also be visually shown by looking
+directly at the density field.
+
+\subsection{Dark Matter Halos}
+
+\subsection{Benchmarks}
+We compare the computational cost of \GADGET, \pmwd\ with NNs, and
+\pmwd\ with SR.
+
+We time the computation of force sharpening for one step using NNs and
+SR respectively.
+
+\vspace{1em}
+\section{Discussion}
+% Discussion does not contain subheadings
+
+\vspace{1em}
+\section{Methods}
+
+\vspace{1em}
+\subsection{Differentiable Simulation}
+\label{sec:pmwd}
+
+The SO using gradient descent training discussed in this work is built
+in the framework of the PM $N$-body library \pmwd\ \citep{Li2022b},
+which is an implementation of the differentiable cosmological simulation
+with the adjoint method \citep{Li2022a}.
+
+Here we briefly recap the equations that are closely related to our
+extensions of the original framework, and introduce the notations of the
+variables.
+Following \cite{Li2022a}, we denote the particles' state with $\vz =
+(\vx, \vp)^\intercal$, where $\vx$ and $\vp$ are the positions and
+momenta of the particles, and their adjoint variables with $\vlambda =
+(\vxi, \vpi)$.
+We refer to the quantities that are directly related to the final
+objective function $\cJ$ as the observable $\vo$.
+In this work, our observables include multiple snapshots at different
+times.
+For the observation operator $O_{i}^{i-1}$ in the adjoint equations
+(check Eq.(49) in \cite{Li2022a}) from time step $a_i$ back to
+$a_{i-1}$, the adjoint variables are updated through
+\begin{align}
+  O_i^{i-1}: \qquad \vxi_{i-1} &:= \vxi_{i-1} +
+              \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
+  \vpi_{i-1} &:= \vpi_{i-1} +
+                          \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
+\end{align}
+where the derivatives of $\vo$ with respect to $\vx_{i-1}$ or
+$\vp_{i-1}$ would be non-zero only if $\vx_{i-1}$ or $\vp_{i-1}$ was
+used in the generation of any snapshots in $\vo$.
+The initial conditions of these adjoint variables are given by
+\begin{align}
+  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
+  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
+\end{align}
+which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$, the
+state of particles at the final time step.
+The derivative of the objective function with respect to the observable,
+i.e. $\p\cJ/\p\vo$, depends on our construction of the loss function on
+the snapshots (\autoref{sec:loss}).
+The VJPs of the observation operation, i.e. $\p\vo/\p\vx_n$ and
+$\p\vo/\p\vp_n$, mainly involves the interpolation function
+(\autoref{sec:snapobs}).
+Both can be conveniently computed by AD.
+
+\vspace{1em}
+\subsection{Snapshot Observable}
+\label{sec:snapobs}
+
+The $N$-body time integration is carried out along discrete time steps
+in terms of the scale factor $a$, and the state of particles at any
+given time can be achieved using interpolation.
+Given the particle displacements $\vs$ and velocities\footnote{This
+velocity is related to the canonical momentum $\vp$ through $\vv =
+\vp/(a^3H)$.} $\vv := \d\vs/\d a$ at two consecutive time steps $a_i$ and
+$a_{i+1}$, we get the snapshot of particles in between at $a$ using the
+cubic Hermite interpolation.
+The interpolated particle displacements are given by
+\begin{align}
+  \vs(a) =\ &h_{00}(\alpha)\vs_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
+           &h_{01}(\alpha)\vs_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
+\end{align}
+where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis
+functions read
+\begin{align}
+  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
+  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
+  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
+  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
+\end{align}
+The particle velocities are then given by the derivative $\vv(a) =
+\d\vs(a) / \d a$.
+The interpolation is a linear combination of the particle snapshots at
+two consecutive time steps, which therefore can be done in either two
+steps or just one step by carrying the particles from the previous step.
+We find the two-steps scheme more convenient for the backward evolution
+of the adjoint variables.
+The interpolated snapshots at different times make the observable in
+this work, and the VJPs of the interpolation functions are computed by
+AD and put into the adjoint equations as mentioned in
+\autoref{sec:pmwd}.
+
+\vspace{1em}
+\subsection{Configurations and Training Data}
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.9\linewidth]{sobol.pdf}
+  \caption{Randomized Quasi-Monte Carlo (RQMC) configuration with
+    scrambled Sobol sequence of 512 points in 9D.
+    Lower triangular panels show the 2D projections and the diagonal
+    panels are the 1D cumulative histograms.
+    From left to right (top to bottom), we use each dimension of the
+    sample to scale the parameters as ordered in \autoref{tab:param}.
+    We use the \texttt{scipy.stats.qmc} package \citep{SciPy} to
+    generate the Sobol sequence \citep{Sobol1967}, which uses the
+    direction number from \citet{JoeKuo2008} and the Owen scrambling
+    \citep{Owen1998}.
+    We search among 65536 scrambling seeds to minimize the mixture
+    discrepancy (a uniformity measure) proposed in \citet{Zhou2013MD}.
+  }
+  \label{fig:sobol}
+\end{figure*}
+
+We configure \GADGET\ with high force and time integration accuracy
+settings.
+For gravitational force, we choose pure FMM with $p=5$ and opening angle
+0.4.
+%FIXME list other from Config.sh and param.txt
+
+We sample 512 data points in the 9D space of configurations and
+cosmological parameters using randomized quasi-Monte Carlo (RQMC) method
+with scrambled Sobol sequence (\autoref{fig:sobol}).
+We use the \texttt{scipy.stats.qmc} package \citep{SciPy} to generate
+the Sobol sequence \citep{Sobol1967}, which uses the direction number
+from \citet{JoeKuo2008} and the Owen scrambling \citep{Owen1998}.
+We search among 65536 scrambling seeds to minimize the mixture
+discrepancy (a uniformity measure) proposed in \citet{Zhou2013MD}.
+
+Small numerical differences in initial conditions (ICs) could become
+more significant with the $N$-body evolution.
+We verified that the IC generation on GPUs is deterministic though
+different from those on CPUs, therefore we generate ICs for both the
+training data and the \pmwd\ model on GPUs.
+
+\vspace{1em}
+\subsection{Input Features}
+\label{sec:features}
+
+We design the input features of the NNs in a way such that they could
+properly fit and interpolate given our wide coverage of the
+configurations and cosmological parameters.
+
+Multiple scales and dimensionless factors could possibly affect force
+sharpening, which are denoted as $\vvartheta$ and enter the NNs of $f(k_i;
+\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ as input features.
+A list of these features is summarized in \autoref{tab:feat}, and here we
+present relevant definations and more details.
+
+Below are some definations related to the typical scales.
+\begin{itemize}
+\item The root-mean-square (RMS) linear displacement is given by
+%
+\begin{equation}
+R_\mathrm{d} = \int \frac{k P_\linear(k)}{6\pi^2} \d\lnk,
+\end{equation}
+%
+where $P_\linear(k)$ is the linear power spectrum.
+\item The dimensionless linear power spectrum is defined as
+%
+\begin{equation}
+\Delta^2_\linear(k) \triangleq \frac{k^3 P_\linear(k)}{2 \pi^2},
+\end{equation}
+%
+based on which we define a nonlinear scale $R_P$ through
+$\Delta_\linear^2(R_P^{-1}) \triangleq 1$.
+\item The variance of linear matter overdensity in a top-hat window is given by
+%
+\begin{equation}
+\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
+  \Delta_\linear^2(k) W_\tophat^2(kR),
+\end{equation}
+%
+where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
+Fourier transform of the top-hat window.
+Then we have $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$.
+\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
+$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$: $\sigma_\gauss^2(R_\gauss)
+\triangleq 1$.
+\end{itemize}
+These nonlinear scales are functions of time since the linear power
+spectrum evolves with time through the growth factor $D(a)$, we also
+take the time derivatives of these scales as input features.
+Other relevant scales include the particle spacing in Lagrangian space,
+the cell size of the chosen mesh grid for force evaluation, and the
+softening length in the target \GADGET\ simulations.
+
+Relevant scale-independent dimensionless factors are
+\begin{itemize}
+\item $G_m(a) \triangleq D_m(a) / a^m$: linear growth given in
+  suppression factors for $m \in \{1, 2\}$;
+\item $\d\ln G_m / \d\lna$: with $m=1$ case related to the growth rate
+  $f \triangleq \d\ln D_1 / \d\lna$;
+\item $\Omegam(a)$: matter density parameter at $a$;
+\item $\d\ln\!H / \d\lna$: time derivative of Hubble expansion;
+\item $\Delta\lna$: time step size in $\lna$.
+\end{itemize}
+
+Dimensionless combinations are formed by multiplying the wavevector
+components with the scales, i.e.,
+%
+\begin{equation}
+f(k_i; \vvartheta) = f(k_iR_d, k_iR_P, \cdots; G_m, \Omegam, \cdots),
+\end{equation}
+%
+\begin{equation}
+g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_P, \cdots\}_{j=1,2,3};
+                                 G_m, \Omegam, \cdots).
+\end{equation}
+%
+
+\vspace{1em}
+\subsection{Loss Function}
+\label{sec:loss}
+
+The loss function is constructed by comparing the simulated field $\hat
+f$ with the target field $f$.
+\YL{
+For a field $f$ in either Lagrangian or Eulerian space%, we have thought
+%of 3 forms of MSE-inspired loss:
+%%
+%\begin{equation}
+%\cL_f = \frac1{N_a} \ln
+%  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
+%       {\sum_{\vq'} f(\vq')^2},
+%\end{equation}
+%%
+%where $N_a$ is the number of time steps it takes to generate the snapshot for
+%each iteration
+%(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
+%their Fourier wavevector $\vk$ in both cases).
+%
+\begin{equation}
+\cL_f = \frac1{N_K N_a} \sum_{ij} \ln \biggl[
+\frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk, a_j) - f(\vk, a_j) \bigr|^2}
+     {\sum_{\vk' \in K_i} | f(\vk', a_j)|^2} + \epsilon \biggr],
+\end{equation}
+%
+where $N_a$ is the number of snapshots and $N_K$ is the number of
+$k$-bins.
+$\epsilon$ sets the relative threshold tolerance, below which we
+consider $\hat f$ in each $K$ accurate enough and the gradients from
+that bin less important.
+%and
+%%
+%\begin{equation}
+%\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
+%\frac{\sum_{k \in K_i} w(\vk)
+%      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+%     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
+%\end{equation}
+%%
+%Then the question is which loss we should use for which $f$, to get
+%stable and robust results, e.g., no bump for large or small boxes.
+%To further limit our scope in this paper, we should probably only
+%consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
+%velocity $\vv(\vq)$.
+We can use $f \in \{\vs(\vq), \delta(\vx), \vv(\vq), (1+\delta)\vv(\vx),
+(1+\delta)\vv\vv(\vx)\}$.
+}
+
+For the displacement field, we adopt the mean squared error (MSE) loss
+that proved effective in many previous works
+\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
+%
+\begin{equation}
+\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
+                   {\sum_{\vq'} s(\vq')^2},
+\end{equation}
+%
+in which a particle originating from its Lagrangian position $\vq$ is
+displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
+prediction of \pmwd\ model in training.
+Note that we have further taken the logarithm to combine it with the
+other loss(es) below.
+% Say that only s loss is not enough even though in principle it is,
+% and cite DrewEtAl x2 too
+
+%\YL{Even though the log should be enough for adaptive optimizer and
+%fixed sim config, this may create problem when training on all config at
+%the same time. So normalizations inside the log is added above and
+%below.}
+
+However, in the Eulerian space, an MSE loss is dominated by the small
+scale modes.
+In this work we choose uniform weights in the scale space on the
+relative MSE.
+We bin the modes logarithmically by \nicefrac13 octave in scale, with
+the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
+2^{1+i/3} \pi / L$.
+Our loss on the overdensity field $\delta$ sums up its MSE relative to
+the target field power in logarithmic bins:
+%
+\begin{equation}
+\cL_\delta = \frac1{N_K} \sum_i \ln
+\frac{\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |\delta(\vk')|^2},
+\end{equation}
+%
+\begin{equation}
+\cL_\delta = \frac1{N_K} \sum_i
+\frac{\sum_{k \in K_i} w(\vk)
+      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
+     {\sum_{k' \in K_i} |\delta(\vk')|^2},
+\end{equation}
+%
+where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
+factor that suppresses the high $k$ contribution in Eulerian fields such
+as the density field (with $w=1$ for the Lagrangian displacement field).
+This is because a PM solver gradually loses small scale information step
+by step, and without $w$ setting the priority, the loss function could
+encourage the network to fit the small scales at the cost of the
+intermediate scales.
+Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
+Lagrangian space, while the density and momentum fields are in the
+Eulerian space:
+%
+\begin{align}
+1 + \delta(\vx) &= \int\!\d^3\vq\,
+  \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+(1 + \delta)\vv(\vx) &= \int\!\d^3\vq\,
+  \vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+(1 + \delta)\vv\vv(\vx) &= \int\!\d^3\vq\,
+  \vv\vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
+\end{align}
+%
+which we compute by scattering particles to the choice of grid for loss.
+
+Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
+and isotropic, we can show
+%
+\begin{equation}
+\cL_\delta \propto \int \!\d\ln\!k\, \ln
+\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
+\end{equation}
+%
+replacing
+%
+\begin{equation}
+\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
+\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
+\end{equation}
+%
+that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
+when $\hat\delta$ is identical to $\delta$.
+Above we choose the binned version instead for the robustness against
+the noisy modewise $|\delta(\vk)|^2$ in the denominator.
+The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
+loss, while our modification is important to balance among all scales so
+that the neural network does not focus too much on either high or low
+$k$'s.
+
+We combine Lagrangian and Eulerian fields to form the total loss for a
+single snapshot:
+%
+\begin{equation}
+\cJ = \cL_\vs + \cL_\delta.
+\end{equation}
+%
+
+\YZ{We further sum the loss of all snapshots across a single forward
+simulation at different times, i.e.
+\begin{equation}
+\cJ = \sum_{a_i} \cJ |_{a_i}
+\end{equation}
+}
+
+\vspace{1em}
+\subsection{Training of Neural Networks}
+
+We perform distributed data-parallel training using GPU devices hosted
+on multiple CPU nodes.
+With each process being binded to a single GPU device, the training data
+samples are uniformly distributed to the processes.
+In total, we use [...] \GPU\ GPUs, each accompanied with [...] CPU cores
+and [...] GB CPU memory mainly for preloading the training data.
+In each training step, a data sample is fed into the GPU from the CPU
+memory where the preloading takes place to save the reading time from
+the disk.
+
+For each training step, we configure and run \pmwd\ with the same
+initial coniditions, cosmological parameters and other common
+configurations as the training data sample.
+As discussed in \autoref{sec:loss}, the loss in a single simulation sums
+over all snapshots.
+Our batch size is [...], i.e. the number of GPU devices, since one
+simulation is run on a single GPU and compared to the corresponding
+training data sample for loss.
+Then the loss and gradients are collected and averaged across all GPU
+devices, with which the optimizer is also updated.
+The SO parameters are initialized to be the same for all devices, and
+hence remain the same for all processes given the same averaged
+gradients.
+We use the Adam \citep{Adam} optimizer.
+
+The SO NN parameters are initialized in a way such that the output
+is very close to one, i.e. no SO.
+We achieve this by setting the kernel of the output layer to be very
+small random values and the bias to be zero, which is around one after
+the exponential function.
+
+Besides the mesh for the force evaluation, we also need meshes when
+computing the power spectra of the density field for the loss function
+as well as for visualization.
+
+Hyperparameters:
+\begin{itemize}
+  \item Model: number of hidden layers and nodes of f and g nets,
+  activation function
+  \item Optimization: learning rate scheduler, momentum, weight decay
+  and dropout rate etc
+  \item Loss: fields to include and loss function forms
+\end{itemize}
+
+We use Optuna \citep{optuna_2019} to search for the optimal
+hyperparameters.
+
+We empoly a learning rate scheduler that reduces the learning rate on
+plateau where the loss function stops decreasing.
+The initial learning rate is set at [...] with a reduction factor of
+[...] after [...] epochs of no decreasing on loss.
+
+We implement the neural networks with \texttt{Flax}
+\citep{flax2020github}, and use the optimizers in \texttt{Optax}, part
+of the DeepMind JAX Ecosystem
+\citep{deepmind2020jax}.
+
+\vspace{1em}
+\subsection{Symbolic Regression Fitting}
+
+To perform further analyses of the trained NNs, including feature
+importance analysis and symbolic regression fitting, we need to generate
+samples of (input, output) data.
+The input features of the NNs are constructed based on the RQMC
+parameters in \autoref{tab:param}, the wavenumber $k$, and the scale
+factor $a$.
+Again, we use Sobol sequence to sample data points in this high
+dimensional space.
+
+We use the random forest regressor for feature selection using
+impurity-based importance.
+We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
+% more details on the hyperparams in random forest
+\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
+importance?}
+
+The selected important features are then used for SR.
+We use the \texttt{PySR} \citep{PySR} SR code to fit for analytic
+function expressions of the trained SO neural networks.
+
+\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
+etc to see which one gives simpler expression, given the possible
+different rankings by complexity.
+For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
+function.}
+
+\vspace{1em}
+\textit{\large Acknowledgements:}
+We thank Qiufan Lin, Paulo Montero-Camacho, and Yang Wang for helpful
+discussions.
+YZ and YL were supported by The Major Key Project of PCL.
+YZ was also supported by the China Postdoctoral Science Foundation under
+award number 2023M731831.
+The Flatiron Institute is supported by the Simons Foundation.
+
+\vspace{1em}
+\textit{\large Data Availability:}
+We make the \GSDATA\ simulation data set public at ...
+
+\vspace{1em}
+\textit{\large Code Availability:}
+\pmwd\ \citep{Li2022b} is open-source on GitHub
+\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
+files and scripts of this paper
+\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
+\GADGET\ \citep{GADGET-4} can be obtained through a public
+git-repository \href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab},
+hosted by a gitlab server at the Max-Planck Computing and Data Facility
+(MPCDF).
+
+\bibliographystyle{aasjournal}
+\bibliography{sto}
+
+\vspace{1em}
+\appendix
+% supplementary information of Nat. Mach. Intell.
+
+\vspace{1em}
+\section{Tests of Snapshots Observation in Adjoint Method}
+As discussed in the main text, our objective loss function is defined on
+multiple snapshots at given times, which are constructed through
+interpolation during the observation operation.
+Here we perform tests on the gradients of NN parameters to check our
+implementation of the snapshots observation inside the adjoint method of
+the $N$-body time integration.
+The gradients given by the adjoint method are compared against those
+given by the AD where we disable the custom VJP of the $N$-body
+function, see
+\autoref{fig:nn_grads_cmp}.
+
+\begin{figure*}
+  \centering
+  \includegraphics[width=0.5\linewidth]{nn_grads_cmp.pdf}
+  \caption{A comparison of adjoint gradients of all parameters in the
+  $f$ and $g$ NNs to those by AD.
+  This test runs a \pmwd\ simulation of $128^3$ particles, with a
+  $128^3$ mesh and 15 time steps.
+  The size and time steps of this test case is mainly limited by the
+  memory cost of the AD method.
+  The two NNs $f$ and $g$ both have two hidden layers of size $[n, n]$,
+  where $n$ is the size of the input layer.
+  The loss function involves 3 snapshots at $a=(0.6, 0.8, 1)$.
+  }
+  \label{fig:nn_grads_cmp}
+\end{figure*}
+
+%\listofchanges
+
 \section*{TODO}
 \begin{itemize}
+\item
+
 \item (paper) Moving some of the following ongoing and completed tasks
@@
   (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna
-  example}).\newline
-  \YZ{Do we also use another subset of the training data as validation data?}
+  example}).
 \item (train) Reduce learning rate on (mean epoch) loss plateau in
@@
   compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
-\item (test) Generate a sample of $(X, y)$ data with the trained neural network
-  for feature selection and symbolic regression.
-\item \YZ{Do we also consider pruning and distilling of the trained networks?}
 \item (paper) Think about what figures and tables to show in the paper. Up
@@
 
-% Main text: up to 3500 words: introduction + results + discussion
-% so we should keep introduction and discussion brief
-\vspace{1em}
-\section{Introduction}
-% without heading
-
-% AI for science and cosmology
-Rapid advances in artificial intelligence (AI) and more specifically
-deep learning (DL) have provided powerful tools that are revolutionizing
-scientific research across disciplines.
-AI for science consists of not only the state-of-art deep neural
-networks, but also more fundamental computational tools and hardwares
-like automatic differentiation (AD) and GPUs.
-One particular application is the numerical simulation of the Universe.
-
-% a brief review of (differentiable) cosmological simulations
-BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
-\texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
-
-% introduce PM and its limitation on small scales
-The particle-mesh (PM) method solves the Poisson equation by scattering
-the particles onto a mesh, on which FFT is performed and the Laplacian
-operator reduces to simple multiplication of wavenumbers in Fourier
-space.
-Although being computationally efficient, the PM method is mainly
-limited by its resolution on small scales, i.e. scales around the mesh
-cell size.
-This is due to the loss of information during the scattering process of
-continuously distributed particles onto the discrete mesh.
-We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to
-compute the mass fractions of a particle scattered to nearby grid
-points.
-Other schemes include the nearest grid point (NGP) and triangle shaped
-cloud (TSC) etc.
-
-% motivation of this work
-Optimizing the accuracy of fast and differentiable simulations is
-crucial for their applications in, e.g. generation of mocks for
-cosmological surveys, simulation-based inference or field-level
-inference.
-
-% introduction of this work
-The differentiable cosmological simulation with adjoint method
-\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
-\citep{Li2022b} lay the foundation of the gradient-based optimization
-discussed in this work.
-Within this fully differentiable framework, we sharpen the gravitational
-force by embedding neural networks (NNs) in the Poisson equation and
-train them for the spatial optimization (SO) of the simulation
-resolution on small scales.
-For better interpretability, we further fit the trained NNs using
-symbolic regression (SR).
-
-\vspace{1em}
-\section{Results}
-% Results should be divided by topical subheadings
-% Put main ideas in Results first, then more technical details in Methods.
-
-\subsection{Optimization of Dynamics}
-
-\begin{table}
-  \centering
-  \caption{Possible features that can affect force sharpening.
-  All scales are then multiplied by the wavenumber $k$ to form dimensionless
-  factors.}
-  \label{tab:feat}
-  \begin{tabular}{rl}
-  \toprule
-  type & feature \\
-  \midrule
-  & $R_\mathrm{d}$: RMS linear displacement \\
-  \cmidrule(lr){2-2}
-  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
-  \cmidrule(lr){2-2}
-  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
-  \cmidrule(lr){2-2}
-  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
-  \cmidrule(lr){2-2}
-  & $\d R / \d\lna$: time derivatives of the above \\
-  \midrule
-  & particle spacing \\
-  \cmidrule(lr){2-2}
-  other scales & $l$: PM force mesh cell size \\
-  \cmidrule(lr){2-2}
-  & softening length \\
-  \midrule
-  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
-  \cmidrule(lr){2-2}
-  & $\d\ln G_m / \d\lna$ \\
-  \cmidrule(lr){2-2}
-  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
-  \cmidrule(lr){2-2}
-  & $\d\ln\!H / \d\lna$ \\
-  \cmidrule(lr){2-2}
-  & $\Delta\lna$: time step size in $\lna$\YL{. Now this can be an important feature with varying $a_\ic$}\\
-  \bottomrule
-  \end{tabular}
-  \end{table}
-
-In cosmological $N$-body simulations, the dynamics of the particles is
-governed by the Poisson equation
-%
-\begin{equation}
-  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
-\end{equation}
-%
-where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
-constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
-gravitational potential and matter density fluctuation respectively.
-In Fourier space, the gravitational force field $- \vnabla \varphi$
-follows
-%
-\begin{equation}
-- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
-\end{equation}
-%
-where for each dimension $i\in\{1,2,3\}$ of the 3D field, we then
-introduce SO force sharpening through
-%
-\begin{equation}
-\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
-\end{equation}
-%
-where $f$ and $g$ are NNs, each approximating a nonlinear function of
-the wavevector components and all possible relevant features
-$\vvartheta$ summarized in \autoref{tab:feat} with more details
-discussed in \autoref{sec:features}.
-Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
-broken by the PM solver to the O\textsubscript{h} discrete symmetry with
-48 elements involving permutations and reversals of the 3 principal axes
-of the simulation box.
-The PM evaluation of each force component can further break the symmetry
-along that axis, leaving the remaining 2 permutationally symmetric.
-Therefore, we implement $f$ and $g$ to be even functions of the
-wavevector components, and $g$ to be permutation equivariant in all 3
-components, so that $f g$ is the most general modification to the force
-kernel.
-An even and real kernel in configuration space is also even in Fourier
-space.
-We take the absolute value of the wavevectors for the evenness, and sort
-the 3 wavevector compontents to achieve the permutation equivariance.
-
-\begin{table*}
-  \centering
-  \caption{Ranges of \GADGET\ and \pmwd\ configuration and cosmological
-  parameters.
-  Note that the grid ratio need next fast len to determine the mesh shape
-  for fast FFT.
-  Given the box size, the mesh shape determines the cell size.
-  The softening parameter gives the ratio of the comoving softening length
-  to the mean particle spacing.
-  The curvature $\OmegaK$ is related to the separate universe simulation.
-  We sample parameters applicable to \GADGET\ during data generation, and
-  those applicable to \pmwd\ during training.
-  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
-  Now it seems more natural to me to have the ptcl spacing sampled
-  uniformly instead...)}
-  }
-  \label{tab:param}
-  \begin{tabular}{lcccr}
-  \toprule
-  applicability & parameter & distribution & range & sampling \\
-  \midrule
-  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
-  \cmidrule(lr){2-5}
-  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
-  \cmidrule(lr){2-4}
-  & snapshot offset & uniform & (0, 1] \\
-  \cmidrule(lr){2-4}
-  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
-  \cmidrule(lr){2-4}
-  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
-  \cmidrule(lr){2-4}
-  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
-  \cmidrule(lr){2-4}
-  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
-  \cmidrule(lr){2-4}
-  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
-  \cmidrule(lr){2-4}
-  & $h$ & log-uniform & [0.5, 1) \\
-  \cmidrule(lr){1-4}
-  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
-  \cmidrule(lr){1-5}
-  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
-  \cmidrule(lr){2-4}
-  \cmidrule(lr){1-4}
-  training & loss grid offset & uniform & $[0, 1)^3$ \\
-  \bottomrule
-  \end{tabular}
-  \end{table*}
-
-Our target data samples for training and testing are generated with the
-\GADGET\ simulation code \citep{GADGET-4}.
-We run 512 \GADGET\ simulations that cover a wide range of
-configurations and cosmological parameters, as listed in
-\autoref{tab:param}.
-\YL{
-Each simulation evolves $128^3$ particles from the initial time $a_\ic$
-to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
-factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
-We determine $a_\ic$ for each configuration depending on its cosmology
-and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why and how here.)} which then sets the
-snapshot spacing $\Delta a = (1 - a_\ic) / 64$.
-To sample different alignment between the snapshots and the start/stop
-times and avoid possible related numerical errors in \pmwd, we introduce
-a fractional random offset $\gamma$ that varies between 0 and 1 for
-different simulations.
-}
-In our \pmwd\ model, we generate the snapshot at an arbitrary given time
-using the cubic Hermite interpolation, with more details in
-\autoref{sec:snapobs}.
-The curvature $\OmegaK$ is related to the separate universe simulation
-\citep{LiEtAl2014, WagnerEtAl2015}.
-\YZ{We refer to this training dataset of 512 \GADGET\ simulations sampled with
-Sobol sequence as \GSDATA.}
-On average each \GADGET\ simulation takes roughly 2K CPU hours to run, and the
-generation of the whole \GSDATA\ dataset consumes 1M CPU hours.
-% to be compare with the speed of pmwd with SO
-
-\YL{
-In this work, we train \pmwd\ using $XXX$ time steps from $a_\ic$ to
-$a_\f$, generate snapshots at the same times as in \GADGET, and compare
-them in the loss function.
-Although the number of time steps are fixed here, we emphasize that our
-spatial optimization of course also applies for different number of time
-steps, which we will explore, together with the optimal choice of time
-variable, in a follow-up work on the temporal optimization.
-}
-
-\subsection{Spatial Optimization Neural Networks}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
-  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
-  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
-  functions. The solid line and shaded region indicate the median and
-  the [16, 84] percentiles of the 64 Sobol simulation setups. $a=1$,
-  mesh shape=1}
-  \label{fig:func}
-\end{figure*}
-
-For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
-fully connected) structures.
-Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
-input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
-input layer, which is [...] and [...] for $f$ and $g$ respectively.
-A rectified linear unit (ReLU) activation function is used for hidden
-layers.
-We further apply an exponential function to the output as a regulator
-for positive output.
-Being embedded in the differentiable simulation library \pmwd, the NNs
-can be trained using gradient-based optimization methods as in usual
-deep learning tasks.
-The parameters of the NNs, which enter the force evaluation in every
-time integration step, can be treated just the same as other
-cosmological parameters.
-The objective gradients are accumulated, based on which the parameters
-are then optimized.
-
-With discrete Fourier transform (DFT) on a regular mesh of length $N_c$
-with cell size $l_c$, the range of sampled 1D discrete wavenumbers is
-given by the fundamental $k_{\rm fund} = 2\pi/(l_c N_c)$ and the
-maximum $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is simply the box
-size.
-To visualize the functions for different box sizes, we normzalize $k$ by
-$k_{\rm Nyquist}$, which gives the same $[2/N_c, 1]$ regardless of the
-box size.
-For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum value
-is then $\sqrt{3}$.
-
-For the $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally
-symmetric about $(k_1, k_2, k_3)$, we show it as a function of $k_i$
-along axis, face and body diagonal directions.
-
-\subsection{Analytic Expressions with Symbolic Regression}
-
-We fit the trained NNs with symbolic regression.
-
-\subsection{Dark Matter Density Fields}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=.48\columnwidth]{tf_snap120_3091776_e3000.pdf}
-  \includegraphics[width=.48\columnwidth]{cc_snap120_3091776_e3000.pdf}
-  \caption{The TF and CC of PM and PM-SONN simulated density fields with
-  respect to that of the \GADGET. The solid line and shaded region
-  indicate the median and the [16, 84] percentiles of the 64 Sobol
-  simulation setups. last snapshot $a\simeq 1$, mesh shape=1}
-  \label{fig:tfcc}
-\end{figure*}
-
-To compare two fields $\hat f$ and $f$, we introduce for each mode the
-square root ratio of the auto power spectrum
-%
-\begin{equation}
-T(\vk) \triangleq
-\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk)|^2}},
-\label{T}
-\end{equation}
-%
-which is referred to as the transfer function (TF), and the correlation
-coefficient (CC) defined with the cross power spectrum
-%
-\begin{equation}
-r(\vk) \triangleq
-\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
-     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
-\label{r}
-\end{equation}
-%
-The two fields are identical when both $T$ and $r$ are unity.
-The PM method is accurate on scales much larger than the mesh cell size,
-therefore we expect $T$ and $r$ to be very close to being unity both
-before and after the force sharpening optimization.
-To show the $T$ and $r$ for different box sizes, we normalize the
-wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
-the particle grid, where $l_p$ is the particle spacing.
-\YZ{We show T and r separately for different mesh shapes?}
-
-comparison of PM, PM with SO NN (PM-SONN), PM with SO SR (PM-SOSR)
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=.8\columnwidth]{denslab_0_120_e300.pdf}
-  \caption{Projected dark matter density fields of \GADGET, \pmwd\ without and
-  with force sharpening.
-  Sobol 0 snap 120}
-  \label{fig:denslab}
-\end{figure*}
-
-% density field with zoom in (with halos painted as circles?)
-The effect of force sharpening can also be visually shown by looking
-directly at the density field.
-
-\subsection{Dark Matter Halos}
-
-\subsection{Benchmarks}
-We compare the computational cost of \GADGET, \pmwd\ with NNs, and
-\pmwd\ with SR.
-
-\vspace{1em}
-\section{Discussion}
-% Discussion does not contain subheadings
-
-\vspace{1em}
-\section{Methods}
-
-\vspace{1em}
-\subsection{Differentiable Simulation}
-\label{sec:pmwd}
-
-The spatial optimization using gradient descent discussed in this work
-is built in the framework of the PM $N$-body library \pmwd\
-\citep{Li2022b}, which is an implementation of the differentiable
-cosmological simulation with adjoint method \citep{Li2022a}.
-
-Here we briefly recap the equations that are closely related to our
-extensions of the original framework, and introduce the notations of the
-variables.
-Following \cite{Li2022a}, we denote the particles' state with $\vz =
-(\vx, \vp)^\intercal$, where $\vx$ and $\vp$ are the positions and
-momenta of the particles, and their adjoint variables with $\vlambda =
-(\vxi, \vpi)$.
-We refer to the quantities that are directly related to the final
-objective function $\cJ$ as the observable $\vo$.
-In this work, our observables include multiple snapshots at given times.
-For the observation operator $O_{i}^{i-1}$ in the adjoint equations
-(check Eq.(18) in \cite{Li2022a}) from time step $a_i$ back to
-$a_{i-1}$, the adjoint variables are updated through
-\begin{align}
-  \vxi_{i-1} &:= \vxi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vx_{i-1}}, \nonumber\\
-  \vpi_{i-1} &:= \vpi_{i-1} + \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p \vp_{i-1}},
-\end{align}
-where the derivatives of $\vo$ with respect to $\vx_{i-1}$ and $\vp_{i-1}$ would
-be non-zero only if they were used in the generation of any snapshots in $\vo$.
-The initial conditions of these adjoint variables are given by
-\begin{align}
-  \vxi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vx_n}, \nonumber\\
-  \vpi_n &= \frac{\p\cJ}{\p\vo}\frac{\p\vo}{\p\vp_n},
-\end{align}
-which are zero if no snapshot in $\vo$ depends on $(\vx_n, \vp_n)$, the
-state of particles at the final time step.
-The derivative of the objective function with respect to the observable
-($\p\cJ/\p\vo$) and the VJPs of the observation operation
-($\p\vo/\p\vx_n$ and $\p\vo/\p\vp_n$, mainly the interpolation function
-discussed in \autoref{sec:snapobs} below) can be conveniently computed
-by AD.
-
-\vspace{1em}
-\subsection{Snapshot Observable}
-\label{sec:snapobs}
-
-The N-body time integration is carried out along discrete time steps in
-terms of the scale factor $a$, and the state of particles at any given
-time can be achieved using interpolation.
-Given the particle displacements $\vs$ and velocities\footnote{This
-velocity is related to the canonical momentum $\vp$ through $\vv =
-\vp/(a^3H)$.} $\vv := \d\vs/\d a$ at two consecutive time steps $a_i$ and
-$a_{i+1}$, we get the snapshot of particles in between at $a$ using the
-cubic Hermite interpolation.
-The interpolated particle displacements are given by
-\begin{align}
-  \vs(a) =\ &h_{00}(\alpha)\vs_i + h_{10}(\alpha)(a_{i+1} - a_i)\vv_i + \nonumber\\
-           &h_{01}(\alpha)\vs_{i+1} + h_{11}(\alpha)(a_{i+1} - a_i)\vv_{i+1},
-\end{align}
-where $\alpha := (a - a_i)/(a_{i+1} - a_i)$ and the Hermite basis
-functions read
-\begin{align}
-  h_{00}(\alpha) &= 2\alpha^3 - 3\alpha^2 + 1, \nonumber\\
-  h_{10}(\alpha) &= \alpha^3 - 2\alpha^2 + \alpha, \nonumber\\
-  h_{01}(\alpha) &= -2\alpha^3 + 3\alpha^2, \nonumber\\
-  h_{11}(\alpha) &= \alpha^3 - \alpha^2.
-\end{align}
-The particle velocities are then given by the derivative $\vv(a) =
-\d\vs(a) / \d a$.
-The interpolation is a linear combination of the particle snapshots at
-two consecutive time steps, which therefore can be done in either two
-steps or just one step by carrying the particles from the previous step.
-We find the two-steps scheme more convenient for the backward evolution
-of the adjoint variables.
-The interpolated snapshots at different times make the observable in
-this work, and the VJPs of the interpolation functionx are computed by
-AD and put into the adjoint equations as mentioned in
-\autoref{sec:pmwd}.
-
-\vspace{1em}
-\subsection{Input Features}
-\label{sec:features}
-
-Multiple scales and dimensionless factors could possibly affect force
-sharpening, which are denoted as $\vvartheta$ and enter the NNs of $f(k_i;
-\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ as input features.
-A list of these features is summarized in \autoref{tab:feat}, and here we
-present relevant definations and more details.
-
-Below are some definations related to the typical scales.
-\begin{itemize}
-\item The root-mean-square (RMS) linear displacement is given by
-%
-\begin{equation}
-R_\mathrm{d} = \int \frac{k P_\linear(k)}{6\pi^2} \d\lnk,
-\end{equation}
-%
-where $P_\linear(k)$ is the linear power spectrum.
-\item The dimensionless linear power spectrum is defined as
-%
-\begin{equation}
-\Delta^2_\linear(k) \triangleq \frac{k^3 P_\linear(k)}{2 \pi^2},
-\end{equation}
-%
-based on which we define a nonlinear scale $R_P$ through
-$\Delta_\linear^2(R_P^{-1}) \triangleq 1$.
-\item The variance of linear matter overdensity in a top-hat window is given by
-%
-\begin{equation}
-\sigma_\tophat^2(R) = \int_0^\infty \frac{\d k}k
-  \Delta_\linear^2(k) W_\tophat^2(kR),
-\end{equation}
-%
-where $W_\tophat(kR) = 3[\sin(kR) - kR\cos(kR)] / (kR)^3$ is the
-Fourier transform of the top-hat window.
-Then we have $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$.
-\item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
-$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$: $\sigma_\gauss^2(R_\gauss)
-\triangleq 1$.
-\end{itemize}
-These nonlinear scales are functions of time since the linear power spectrum
-evolves with time through the growth factor $D(a)$, we also take the time
-derivatives of these scales as input features.
-Other relevant scales include the particle spacing in Lagrangian space, the cell
-size of the chosen mesh grid for force evaluation, and the softening length in
-the target \GADGET\ simulations.
-
-Relevant scale-independent dimensionless factors are
-\begin{itemize}
-\item $G_m(a) \triangleq D_m(a) / a^m$: linear growth given in
-  suppression factors for $m \in \{1, 2\}$;
-\item $\d\ln G_m / \d\lna$: with $m=1$ case related to the growth rate
-  $f \triangleq \d\ln D_1 / \d\lna$;
-\item $\Omegam(a)$: matter density parameter at $a$;
-\item $\d\ln\!H / \d\lna$: time derivative of Hubble expansion;
-\item $\Delta\lna$: time step size in $\lna$.
-\end{itemize}
-
-Dimensionless combinations are formed by multiplying the wavevector components
-with the scales, i.e.,
-%
-\begin{equation}
-f(k_i; \vvartheta) = f(k_iR_d, k_iR_P, \cdots; G_m, \Omegam, \cdots),
-\end{equation}
-%
-\begin{equation}
-g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_P, \cdots\}_{j=1,2,3}; G_m, \Omegam, \cdots).
-\end{equation}
-%
-
-\vspace{1em}
-\subsection{Loss Function}
-\label{sec:loss}
-
-The loss function is constructed by comparing the simulated field $\hat
-f$ with the target field $f$.
-\YL{
-For a field $f$ in either Lagrangian or Eulerian space%, we have thought
-%of 3 forms of MSE-inspired loss:
-%%
-%\begin{equation}
-%\cL_f = \frac1{N_a} \ln
-%  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
-%       {\sum_{\vq'} f(\vq')^2},
-%\end{equation}
-%%
-%where $N_a$ is the number of time steps it takes to generate the snapshot for
-%each iteration
-%(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
-%their Fourier wavevector $\vk$ in both cases).
-%
-\begin{equation}
-\cL_f = \frac1{N_K N_a} \sum_{ij} \ln \biggl[
-\frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk, a_j) - f(\vk, a_j) \bigr|^2}
-     {\sum_{\vk' \in K_i} | f(\vk', a_j)|^2} + \epsilon \biggr],
-\end{equation}
-%
-where $N_a$ is the number of snapshots and $N_K$ is the number of
-$k$-bins.
-$\epsilon$ sets the relative threshold tolerance, below which we
-consider $\hat f$ in each $K$ accurate enough and the gradients from
-that bin less important.
-%and
-%%
-%\begin{equation}
-%\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
-%\frac{\sum_{k \in K_i} w(\vk)
-%      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-%     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
-%\end{equation}
-%%
-%Then the question is which loss we should use for which $f$, to get
-%stable and robust results, e.g., no bump for large or small boxes.
-%To further limit our scope in this paper, we should probably only
-%consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
-%velocity $\vv(\vq)$.
-We can use $f \in \{\vs(\vq), \delta(\vx), \vv(\vq), (1+\delta)\vv(\vx),
-(1+\delta)\vv\vv(\vx)\}$.
-}
-
-For the displacement field, we adopt the mean squared error (MSE) loss
-that proved effective in many previous works
-\citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
-%
-\begin{equation}
-\cL_\vs = \ln \frac{\sum_\vq \bigl[ \hat\vs(\vq) - \vs(\vq) \bigr]^2}
-                   {\sum_{\vq'} s(\vq')^2},
-\end{equation}
-%
-in which a particle originating from its Lagrangian position $\vq$ is
-displaced by $\vs$ in the \GADGET\ snapshot and by $\hat\vs$ in the
-prediction of \pmwd\ model in training.
-Note that we have further taken the logarithm to combine it with the
-other loss(es) below.
-% Say that only s loss is not enough even though in principle it is, and cite DrewEtAl x2 too
-
-%\YL{Even though the log should be enough for adaptive optimizer and
-%fixed sim config, this may create problem when training on all config at
-%the same time. So normalizations inside the log is added above and
-%below.}
-
-However, in the Eulerian space, an MSE loss is dominated by the small
-scale modes.
-In this work we choose uniform weights in the scale space on the
-relative MSE.
-We bin the modes logarithmically by \nicefrac13 octave in scale, with
-the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
-2^{1+i/3} \pi / L$.
-Our loss on the overdensity field $\delta$ sums up its MSE relative to
-the target field power in logarithmic bins:
-%
-\begin{equation}
-\cL_\delta = \frac1{N_K} \sum_i \ln
-\frac{\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |\delta(\vk')|^2},
-\end{equation}
-%
-\begin{equation}
-\cL_\delta = \frac1{N_K} \sum_i
-\frac{\sum_{k \in K_i} w(\vk)
-      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |\delta(\vk')|^2},
-\end{equation}
-%
-where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
-factor that suppresses the high $k$ contribution in Eulerian fields such
-as the density field (with $w=1$ for the Lagrangian displacement field).
-This is because a PM solver gradually loses small scale information step
-by step, and without $w$ setting the priority, the loss function could
-encourage the network to fit the small scales at the cost of the
-intermediate scales.
-Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
-Lagrangian space, while the density and momentum fields are in the
-Eulerian space:
-%
-\begin{align}
-1 + \delta(\vx) &= \int\!\d^3\vq\,
-  \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
-(1 + \delta)\vv(\vx) &= \int\!\d^3\vq\,
-  \vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
-(1 + \delta)\vv\vv(\vx) &= \int\!\d^3\vq\,
-  \vv\vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
-\end{align}
-%
-which we compute by scattering particles to the choice of grid for loss.
-
-Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
-and isotropic, we can show
-%
-\begin{equation}
-\cL_\delta \propto \int \!\d\ln\!k\, \ln
-\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
-  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
-\end{equation}
-%
-replacing
-%
-\begin{equation}
-\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
-\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
-  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
-\end{equation}
-%
-that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
-when $\hat\delta$ is identical to $\delta$.
-Above we choose the binned version instead for the robustness against
-the noisy modewise $|\delta(\vk)|^2$ in the denominator.
-The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
-loss, while our modification is important to balance among all scales so
-that the neural network does not focus too much on either high or low
-$k$'s.
-
-We combine Lagrangian and Eulerian fields to form the total loss for a
-single snapshot:
-%
-\begin{equation}
-\cJ = \cL_\vs + \cL_\delta.
-\end{equation}
-%
-
-\YZ{We further sum the loss of all snapshots across a single forward
-simulation at different times, i.e.
-\begin{equation}
-\cJ = \sum_{a_i} \cJ |_{a_i}
-\end{equation}
-}
-
-\vspace{1em}
-\subsection{Configurations and Training Data}
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=0.9\linewidth]{sobol.pdf}
-  \caption{Randomized Quasi-Monte Carlo (RQMC) configuration with
-    scrambled Sobol sequence of 512 points in 9D.
-    Lower triangular panels show the 2D projections and the diagonal
-    panels are the 1D cumulative histograms.
-    From left to right (top to bottom), we use each dimension of the
-    sample to scale the parameters as ordered in \autoref{tab:param}.
-    We use the \texttt{scipy.stats.qmc} package \citep{SciPy} to
-    generate the Sobol sequence \citep{Sobol1967}, which uses the
-    direction number from \citet{JoeKuo2008} and the Owen scrambling
-    \citep{Owen1998}.
-    We search among 65536 scrambling seeds to minimize the mixture
-    discrepancy (a uniformity measure) proposed in \citet{Zhou2013MD}.
-  }
-  \label{fig:sobol}
-\end{figure*}
-
-Small numerical differences in initial conditions (ICs) could become
-more significant with the $N$-body evolution.
-We have verified that the IC generation on GPUs is deterministic though
-different from those on CPUs, therefore we generate ICs for both the
-training data and the model on GPUs.
-
-We configure \GADGET\ with high force and time integration accuracy
-settings.
-For gravitational force, we choose pure FMM with $p=5$ and opening angle
-0.4.
-%FIXME list other from Config.sh and param.txt
-
-\vspace{1em}
-\subsection{Training of Neural Networks}
-
-We perform distributed data-parallel training using GPU devices hosted
-on multiple CPU nodes.
-With each process being binded to a single GPU device, the training data
-samples are uniformly distributed to the processes.
-In total, we use [...] \GPU\ GPUs, each accompanied with [...] CPU cores
-and [...] GB CPU memory mainly for preloading the training data.
-In each training step, a data sample is fed into the GPU from the CPU
-memory where the preloading takes place to save the reading time from
-the disk.
-
-For each training step, we configure and run \pmwd\ with the same
-initial coniditions, cosmological parameters and other common
-configurations as the training data sample.
-As discussed in \autoref{sec:loss}, the loss in a single simulation sums
-over all snapshots.
-Our batch size is [...], i.e. the number of GPU devices, since one
-simulation is run on a single GPU and compared to the corresponding
-training data sample for loss.
-Then the loss and gradients are collected and averaged across all GPU
-devices, with which the optimizer is also updated.
-The SO parameters are initialized to be the same for all devices, and
-hence remain the same for all processes given the same averaged
-gradients.
-We use the Adam \citep{Adam} optimizer.
-
-The SO NN parameters are initialized in a way such that the output
-is very close to one, i.e. no SO.
-We achieve this by setting the kernel of the output layer to be very
-small random values and the bias to be zero, which is around one after
-the exponential function.
-
-Hyperparameters:
-\begin{itemize}
-  \item Model: number of hidden layers and nodes of f and g nets, activation function
-  \item Optimization: learning rate scheduler, momentum, weight decay and dropout rate etc
-  \item Loss: fields to include and loss function forms
-\end{itemize}
-
-We use Optuna \citep{optuna_2019} to search for the optimal
-hyperparameters.
-
-We empoly a learning rate scheduler that reduces the learning rate on
-plateau where the loss function stops decreasing.
-The initial learning rate is set at [...] with a reduction factor of
-[...] after [...] epochs of no decreasing on loss.
-
-We implement the neural networks with \texttt{Flax}
-\citep{flax2020github}, and use the optimizers in \texttt{Optax}, part
-of the DeepMind JAX Ecosystem
-\citep{deepmind2020jax}.
-
-\vspace{1em}
-\subsection{Symbolic Regression Fitting}
-
-We generate samples of (input, output) data of the trained neural
-networks to perform symbolic regression.
-We use Sobol sequence to generate the sample of (parameters in
-\autoref{tab:param}, $k$, $a$).
-
-We use the random forest regressor for feature selection using impurity-based
-importance.
-We use the \texttt{scikit-learn} \citep{scikit-learn} code for this.
-% more details on the hyperparams in random forest
-\YZ{Maybe we can add another column in \autoref{tab:feat} for feature
-importance?}
-
-The selected important features are then used for SR.
-We use the \texttt{PySR} \citep{PySR} SR code to fit for analytic
-function expressions of the trained SO neural networks.
-
-\YL{It is interesting to symbolically regress $f$, $k_i f$, and $f g$
-etc to see which one gives simpler expression, given the possible
-different rankings by complexity.
-For example, $k_i f(k_i)$ may have a simpler form as a nonlinear odd
-function.}
-
-\vspace{1em}
-\textit{\large Acknowledgements:}
-We thank Qiufan Lin, Paulo Montero-Camacho, and Yang Wang for helpful
-discussions.
-YZ and YL were supported by The Major Key Project of PCL.
-YZ was also supported by the China Postdoctoral Science Foundation under award
-number 2023M731831.
-The Flatiron Institute is supported by the Simons Foundation.
-
-\vspace{1em}
-\textit{\large Data Availability:}
-We make the \GSDATA\ simulation data set public at ...
-
-\vspace{1em}
-\textit{\large Code Availability:}
-\pmwd\ \citep{Li2022b} is open-source on GitHub
-\href{https://github.com/eelregit/pmwd}{\faGithub}, including the source
-files and scripts of this paper
-\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{\faFile}.
-\GADGET\ \citep{GADGET-4} can be obtained through a public
-git-repository \href{http://gitlab.mpcdf.mpg.de/vrs/gadget4}{\faGitlab},
-hosted by a gitlab server at the Max-Planck Computing and Data Facility
-(MPCDF).
-
-\bibliographystyle{aasjournal}
-\bibliography{sto}
-
-\vspace{1em}
-\appendix
-% supplementary information of Nat. Mach. Intell.
-
-\vspace{1em}
-\section{Tests of Snapshots Observation in Adjoint Method}
-As discussed in the main text, our objective loss function is defined on
-multiple snapshots at given times, which are constructed through interpolation
-during the observation operation.
-Here we perform tests on the gradients of NN parameters to check our
-implementation of the snapshots observation inside the adjoint method of the
-$N$-body time integration.
-The gradients given by the adjoint method are compared against those given by
-the AD where we disable the custom VJP of the $N$-body function, see
-\autoref{fig:nn_grads_cmp}.
-
-\begin{figure*}
-  \centering
-  \includegraphics[width=0.5\linewidth]{nn_grads_cmp.pdf}
-  \caption{A comparison of adjoint gradients of all parameters in the $f$ and
-  $g$ NNs to those by AD.
-  This test runs a \pmwd\ simulation of $128^3$ particles, with a $128^3$ mesh
-  and 15 time steps.
-  The size and time steps of this test case is mainly limited by the memory cost
-  of the AD method.
-  The two NNs $f$ and $g$ both have two hidden layers of size $[n, n]$, where
-  $n$ is the size of the input layer.
-  The loss function involves 3 snapshots at $a=(0.6, 0.8, 1)$.
-  }
-  \label{fig:nn_grads_cmp}
-\end{figure*}
-
-%\listofchanges
-
 \end{document}

### 2024-02-27 docs/papers/sto/so.tex
WIP draft loss function

@@
 f$ with the target field $f$.
-\YL{
-For a field $f$ in either Lagrangian or Eulerian space%, we have thought
-%of 3 forms of MSE-inspired loss:
-%%
-%\begin{equation}
-%\cL_f = \frac1{N_a} \ln
-%  \frac{\sum_\vq \bigl[ \hat f(\vq) - f(\vq) \bigr]^2}
-%       {\sum_{\vq'} f(\vq')^2},
-%\end{equation}
-%%
-%where $N_a$ is the number of time steps it takes to generate the snapshot for
-%each iteration
-%(or with $\vq$ replaced by $\vx$ for Eulerian fields, or equivalently by
-%their Fourier wavevector $\vk$ in both cases).
-%
-\begin{equation}
-\cL_f = \frac1{N_K N_a} \sum_{ij} \ln \biggl[
-\frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk, a_j) - f(\vk, a_j) \bigr|^2}
-     {\sum_{\vk' \in K_i} | f(\vk', a_j)|^2} + \epsilon \biggr],
-\end{equation}
-%
-where $N_a$ is the number of snapshots and $N_K$ is the number of
-$k$-bins.
-$\epsilon$ sets the relative threshold tolerance, below which we
-consider $\hat f$ in each $K$ accurate enough and the gradients from
-that bin less important.
-%and
-%%
-%\begin{equation}
-%\cL_f = \frac1{N_a} \ln \biggl[ \frac1{N_K} \sum_i
-%\frac{\sum_{k \in K_i} w(\vk)
-%      \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
-%     {\sum_{k' \in K_i} |f(\vk')|^2} \biggl].
-%\end{equation}
-%%
-%Then the question is which loss we should use for which $f$, to get
-%stable and robust results, e.g., no bump for large or small boxes.
-%To further limit our scope in this paper, we should probably only
-%consider $f \in \{\vs, \delta\}$, and optionally the Lagrangian particle
-%velocity $\vv(\vq)$.
-We can use $f \in \{\vs(\vq), \delta(\vx), \vv(\vq), (1+\delta)\vv(\vx),
-(1+\delta)\vv\vv(\vx)\}$.
-}
-
-For the displacement field, we adopt the mean squared error (MSE) loss
-that proved effective in many previous works
+In this work, we consider two fields in our loss function, the
+Lagrangian particle displacement field $\vs(\vq)$ and the Eulerian
+density field $\delta(\vx)$.
+
+For the displacement field, the mean squared error (MSE) loss has been
+proved effective in many previous works
 \citep[e.g.,][]{HeEtAl2019, LiEtAl2021}:
@@
 Note that we have further taken the logarithm to combine it with the
-other loss(es) below.
+other loss(es).
+Even though the logarithm should be enough for adaptive optimizer and
+fixed simulation configuration, this may create problem in our training
+on all configurations at the same time.
+So normalizations inside the logarithm is added above and below.
 % Say that only s loss is not enough even though in principle it is,
 % and cite DrewEtAl x2 too
-
-%\YL{Even though the log should be enough for adaptive optimizer and
-%fixed sim config, this may create problem when training on all config at
-%the same time. So normalizations inside the log is added above and
-%below.}
 
@@
 2^{1+i/3} \pi / L$.
-Our loss on the overdensity field $\delta$ sums up its MSE relative to
-the target field power in logarithmic bins:
-%
-\begin{equation}
-\cL_\delta = \frac1{N_K} \sum_i \ln
-\frac{\sum_{k \in K_i} \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |\delta(\vk')|^2},
-\end{equation}
-%
-\begin{equation}
-\cL_\delta = \frac1{N_K} \sum_i
-\frac{\sum_{k \in K_i} w(\vk)
-      \bigl| \hat\delta(\vk) - \delta(\vk) \bigr|^2}
-     {\sum_{k' \in K_i} |\delta(\vk')|^2},
-\end{equation}
-%
-where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
-factor that suppresses the high $k$ contribution in Eulerian fields such
-as the density field (with $w=1$ for the Lagrangian displacement field).
-This is because a PM solver gradually loses small scale information step
-by step, and without $w$ setting the priority, the loss function could
-encourage the network to fit the small scales at the cost of the
-intermediate scales.
-Note that $\cL_\vs$ is on the displacement field $\vs(\vq)$ in the
-Lagrangian space, while the density and momentum fields are in the
-Eulerian space:
-%
-\begin{align}
-1 + \delta(\vx) &= \int\!\d^3\vq\,
-  \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
-(1 + \delta)\vv(\vx) &= \int\!\d^3\vq\,
-  \vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
-(1 + \delta)\vv\vv(\vx) &= \int\!\d^3\vq\,
-  \vv\vv(\vq) \deltaD[ \vx - \vq - \vs(\vq)], \nonumber\\
-\end{align}
-%
-which we compute by scattering particles to the choice of grid for loss.
-
-Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
-and isotropic, we can show
-%
-\begin{equation}
-\cL_\delta \propto \int \!\d\ln\!k\, \ln
-\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
-  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
-\end{equation}
-%
-replacing
-%
-\begin{equation}
-\cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
-\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
-  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
-\end{equation}
-%
-that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
-when $\hat\delta$ is identical to $\delta$.
-Above we choose the binned version instead for the robustness against
-the noisy modewise $|\delta(\vk)|^2$ in the denominator.
-The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
-loss, while our modification is important to balance among all scales so
-that the neural network does not focus too much on either high or low
-$k$'s.
-
-We combine Lagrangian and Eulerian fields to form the total loss for a
-single snapshot:
-%
-\begin{equation}
-\cJ = \cL_\vs + \cL_\delta.
-\end{equation}
-%
-
-\YZ{We further sum the loss of all snapshots across a single forward
-simulation at different times, i.e.
-\begin{equation}
-\cJ = \sum_{a_i} \cJ |_{a_i}
-\end{equation}
-}
+We introduce a loss function that sums up its MSE relative to the target
+field power in logarithmic bins:
+%
+\begin{equation}
+  \cL_f = \frac1{N_K} \sum_{i} \ln \biggl[
+  \frac{\sum_{\vk \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
+       {\sum_{\vk' \in K_i} | f(\vk')|^2} + \epsilon \biggr],
+\label{loss_power_ln}
+\end{equation}
+%
+where $N_K$ is the number of $k$-bins, and $\epsilon$ sets the relative
+threshold tolerance, below which we consider $\hat f$ in each $K$
+accurate enough and the gradients from that bin less important.
+% Under the mild assumptions that $T$ \eqref{T} and $r$ \eqref{r} are smooth
+% and isotropic, we can show
+% %
+% \begin{equation}
+% \cL_\delta \propto \int \!\d\ln\!k\, \ln
+% \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+%   + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty,
+% \end{equation}
+% %
+% replacing
+% %
+% \begin{equation}
+% \cL_\delta \propto \int \!\d\ln\!k\, w(\vk)
+% \Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
+%   + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
+% \end{equation}
+% %
+% that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
+% when $\hat\delta$ is identical to $\delta$.
+% Above we choose the binned version instead for the robustness against
+% the noisy modewise $|\delta(\vk)|^2$ in the denominator.
+% The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
+% loss, while our modification is important to balance among all scales so
+% that the neural network does not focus too much on either high or low
+% $k$'s.
+
+In this work, we use the loss function in \eqref{loss_power_ln} for
+both $\vs(\vq)$ and $\delta(\vx)$.
+We combine the two fields to form the total loss for a single snapshot,
+and further sum over all snapshots in a single forward simulation
+%
+\begin{equation}
+\cJ = \sum_{a_i} \left[\cL_{\vs(a_i)} + \cL_{\delta(a_i)}\right].
+\end{equation}
+%
 

### 2024-02-28 docs/papers/sto/so.tex
WIP draft training NNs

@@
 
-In this work, we use the loss function in \eqref{loss_power_ln} for
-both $\vs(\vq)$ and $\delta(\vx)$.
+\YZ{In this work, we use the loss function in \eqref{loss_power_ln} for
+both $\vs(\vq)$ and $\delta(\vx)$.}
 We combine the two fields to form the total loss for a single snapshot,
@@
 We perform distributed data-parallel training using GPU devices hosted
-on multiple CPU nodes.
-With each process being binded to a single GPU device, the training data
+on multiple computing nodes.
+Each process is binded to a single GPU device, and the training data
 samples are uniformly distributed to the processes.
-In total, we use [...] \GPU\ GPUs, each accompanied with [...] CPU cores
-and [...] GB CPU memory mainly for preloading the training data.
+In total, we use XXX \GPU\ GPUs, each accompanied with XXX CPU cores
+and XXX GB CPU memory mainly for preloading the training data.
+The order of data samples within each process is shuffled across
+training epochs.
 In each training step, a data sample is fed into the GPU from the CPU
-memory where the preloading takes place to save the reading time from
-the disk.
+memory where the preloading took place to save the repeated reading time
+from the disk.
 
@@
 over all snapshots.
-Our batch size is [...], i.e. the number of GPU devices, since one
+Our batch size is XXX, i.e. the number of GPU devices, since one
 simulation is run on a single GPU and compared to the corresponding
@@
 Then the loss and gradients are collected and averaged across all GPU
-devices, with which the optimizer is also updated.
-The SO parameters are initialized to be the same for all devices, and
-hence remain the same for all processes given the same averaged
-gradients.
+devices, with which the optimizer is updated.
 We use the Adam \citep{Adam} optimizer.
 
-The SO NN parameters are initialized in a way such that the output
-is very close to one, i.e. no SO.
-We achieve this by setting the kernel of the output layer to be very
-small random values and the bias to be zero, which is around one after
-the exponential function.
+We initialize the parameters of NNs in a way such that the output is
+very close to one, i.e. no SO.
+This is achieved by setting the kernel of the output layer to be very
+small random values and the bias to be zero, whose return is then around
+one after the exponential function.
+The parameters are also initialized to be the same for all devices, and
+hence remain the same during training given the same averaged gradients.
+
+% We use Optuna \citep{optuna_2019} to search for the optimal
+% hyperparameters.
+
+% We empoly a learning rate scheduler that reduces the learning rate on
+% plateau where the loss function stops decreasing.
+% The initial learning rate is set at [...] with a reduction factor of
+% [...] after [...] epochs of no decreasing on loss.
 
 Besides the mesh for the force evaluation, we also need meshes when
-computing the power spectra of the density field for the loss function
-as well as for visualization.
+estimating the density field for the loss function as well as for
+visualization.
 
-Hyperparameters:
-\begin{itemize}
-  \item Model: number of hidden layers and nodes of f and g nets,
-  activation function
-  \item Optimization: learning rate scheduler, momentum, weight decay
-  and dropout rate etc
-  \item Loss: fields to include and loss function forms
-\end{itemize}
-
-We use Optuna \citep{optuna_2019} to search for the optimal
-hyperparameters.
-
-We empoly a learning rate scheduler that reduces the learning rate on
-plateau where the loss function stops decreasing.
-The initial learning rate is set at [...] with a reduction factor of
-[...] after [...] epochs of no decreasing on loss.
-
-We implement the neural networks with \texttt{Flax}
-\citep{flax2020github}, and use the optimizers in \texttt{Optax}, part
-of the DeepMind JAX Ecosystem
-\citep{deepmind2020jax}.
+Same as the \pmwd\ library, this work is also implemented using JAX
+\citep{JAX}.
+We implement the NNs with \texttt{Flax} \citep{flax2020github}, and use
+the optimizers provided in \texttt{Optax}, part of the DeepMind JAX
+Ecosystem \citep{deepmind2020jax}.
 

### 2024-02-28 docs/papers/sto/so.tex
WIP draft abstract

@@
 We optimize the spatial resolution of cosmological particle-mesh (PM)
-simulations by sharpening the PM force with first neural networks and
-then symbolic regression, taking symmetry and dimensional analysis in
-consideration.
-
-We combine differentiable simulation, deep learning, [decision tree
-(feature selection)], and symbolic regression to extract the final
-kernel applicable to all other cosmological simulation code.
+simulations by sharpening the PM force with first neural networks (NNs)
+and then symbolic regression (SR), taking symmetry and dimensional
+analysis in consideration.
+We embed deep NNs in the differentiable cosmological $N$-body
+simulation, and training them to learn for the force kernel.
+We fit the trained NNs with SR to extract an analytic kernel for better
+interpretability and physical insights, which is also applicable to all
+other cosmological PM simulation code.
 

### 2024-02-29 docs/papers/sto/so.tex
FIXUP draft update T and r plots

@@
   \centering
-  \includegraphics[width=.48\columnwidth]{tf_snap120_3091776_e3000.pdf}
-  \includegraphics[width=.48\columnwidth]{cc_snap120_3091776_e3000.pdf}
-  \caption{The TF and CC of PM and PM-SONN simulated density fields with
-  respect to that of the \GADGET. The solid line and shaded region
-  indicate the median and the [16, 84] percentiles of the 64 Sobol
-  simulation setups. last snapshot $a\simeq 1$, mesh shape=1}
+  \includegraphics[width=.48\columnwidth]{tf_snap120_3177874_e3000_mid-per.pdf}
+  \includegraphics[width=.48\columnwidth]{cc_snap120_3177874_e3000_mid-per.pdf} \\
+  \includegraphics[width=.48\columnwidth]{tf_test_snap120_3177874_e3000_mid-per.pdf}
+  \includegraphics[width=.48\columnwidth]{cc_test_snap120_3177874_e3000_mid-per.pdf}
+  \caption{The TF and CC of original PM and optimized PM-SONN simulated
+  density fields with respect to that of the \GADGET.
+  The solid line and shaded region indicate the median and the [16, 84]
+  percentiles of different simulation configurations.
+  The results are shown for the last snapshot sampled in each
+  configuration, which are all around time $a\simeq 1$.
+  The x-axis is the wavenumber $k$ normalized by the Nyquist wavenumber
+  $k_{\rm Nyquist}$ of the particle grid, which is therefore the same
+  for different box sizes.
+  \textit{Upper}: results for the training dataset.
+  \textit{Lower}: results for the test dataset.
+  }
   \label{fig:tfcc}
@@
 therefore we expect $T$ and $r$ to be very close to being unity both
-before and after the force sharpening optimization.
-To show the $T$ and $r$ for different box sizes, we normalize the
-wavenumber by the Nyquist wavenumber $k_{\rm Nyquist} = \pi / l_p$ of
-the particle grid, where $l_p$ is the particle spacing.
-\YZ{We show T and r separately for different mesh shapes?}
+before and after the SO with force sharpening.
+
+In \autoref{fig:tfcc}, we show the results of $T$ and $r$ for both
+training and test data samples.
+To show different box sizes, we normalize the wavenumber by the Nyquist
+wavenumber $k_{\rm Nyquist} = \pi / l_p$ of the particle grid, where
+$l_p$ is the particle spacing.
 
@@
 \vspace{1em}
-\subsection{Configurations and Training Data}
+\subsection{Configurations and Data Samples}
 
@@
 discrepancy (a uniformity measure) proposed in \citet{Zhou2013MD}.
+
+In addition, we sample 9 extra data points for the interpolation test of
+the trained model to different configurations and cosmological
+parameters.
 
@@
 
-Besides the mesh for the force evaluation, we also need meshes when
-estimating the density field for the loss function as well as for
-visualization.
-
 Same as the \pmwd\ library, this work is also implemented using JAX
@@
 \begin{itemize}
-\item
-
-\item (paper) Moving some of the following ongoing and completed tasks
-  to the main text and add preliminary figures.
 \item (data; YL) Incorporate 3lpt branch; testing; find a\_start for
@@
   std acc over the std vel) of each realization.
-\item (model; YL) Change nbody for loop by lax scan and/or while loops.
-  while\_loop can help to save compilation when trained on different
-  snapshots. Performance difference between scan and while?\newline
-  \YZ{It seems that lax.while\_loop only supports fwd grad}
-  \href{https://jax.readthedocs.io/en/latest/notebooks/Common_Gotchas_in_JAX.html#summary}{link}
-\item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
-    steps) results and make sure that they agree perfectly on the large
-    scales. For this purpose the box size should be large enough (in
-    linear regime), say 1 Gpc/$h$.\newline
-    \YZ{\textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are
-    visually consistent at the field level. For power spectrum: For
-    odd mesh shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent
-    on large scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers
-    look slightly different. To figure out this weird issue.}}
-\item (model; YL) Probably necessary to first incorporate List \& Hahn's
-  recent improvement on FastPM integrator arXiv:2301:09655.
-\item (train, test) For $\vs(\vq)$, $\vv(\vq)$, $\delta(\vx)$,
-  $(1+\delta)\vv(\vx)$, $(1+\delta)\vv\vv(\vx)$, show $T - 1$ in symlog
-  with 0.01 linthresh, and $1-r^2$ in log at 3 different $a$'s; Show $f$
-  and $g$ (along axis and face diagonal) at 3 different $a$'s
 \item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add
@@
   structured scheme, e.g., alternating between 2 interlocking grids
-\item (train) Explore if we can use a finer loss grid, from 1x to say
-  1.5x and even 2x.
 \item (train, test) Non-integer multiple mesh shapes: for uniform
@@
   deconvolution, with CIC deconvolution, with SO.
-\item (train) find the best loss function for production training
-\item (train) For generalizability and robustness, consider testing
-  weight decay and dropout. Also they're worth adding if they give lower
-  losses at convergence. Note that 512 realization are not too many, and
-  we have features that can be very correlated.\newline
-  \YZ{Dropout with Flax has been coded. We can use Optax AdamW for weight decay.}
-\item (train) optuna hyperparam search on a subset of training data
-  before production training
-  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna
-  example}).
-\item (train) Reduce learning rate on (mean epoch) loss plateau in
-  production\newline
-  \YZ{A simple scheduler has been coded, to be fine-tuned for production training.}
 \item (test) Interpolation test: compare SO DM and halo fields and
@@
   compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
-\item (paper) Think about what figures and tables to show in the paper. Up
-to 6 items could be included. Some tables could be included in the context as
-list if we have too many figures to show?\newline
-  \YZ{figures:
-  1. $f$ and $g$ functions, both NN and Symbolic.
-  2. Density field (and halo?) with zoom in for small scale sharpening.
-  3. Transfer function and correlation coefficient.
-  4. Halo related stats.}
 \item Fix cross references and other journal quirks as a patch

### 2024-03-01 docs/papers/sto/so.tex
WIP draft: figure on density slab

@@
   functions. The solid line and shaded region indicate the median and
-  the [16, 84] percentiles of the 64 Sobol simulation setups. $a=1$,
+  the [16, 84] percentiles of the 64 sampled configurations and
+  cosmological parameters. $a=1$,
   mesh shape=1}
@@
 
-For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
-fully connected) structures to learn the nonlinear functions.
+For both the $f$ and $g$ spatial optimization NNs (SONNs), we adopt
+multilayer perceptron (MLP, i.e. fully connected) structures to learn
+the nonlinear functions.
 Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
@@
 To compare two fields $\hat f$ and $f$, we introduce for each mode the
-square root ratio of the auto power spectrum
+transfer function (TF), which is defined as the square root ratio of the
+auto power spectrum:
 %
@@
 %
-which is referred to as the transfer function (TF), and the correlation
-coefficient (CC) defined with the cross power spectrum
+and the correlation coefficient (CC):
 %
@@
 \frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
-     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}}.
+     {\sqrt{|\hat f(\vk)|^2 |f(\vk)|^2}},
 \label{r}
@@
 %
+where the numerator is the real part of the cross power spectrum.
 The two fields are identical when both $T$ and $r$ are unity.
@@
   \centering
-  \includegraphics[width=.8\columnwidth]{denslab_0_120_e300.pdf}
-  \caption{Projected dark matter density fields of \GADGET, \pmwd\ without and
-  with force sharpening.
-  Sobol 0 snap 120}
+  \includegraphics[width=.98\columnwidth]{slab_s0_snap120_3177874_e3000.pdf}
+  \caption{Projected dark matter density fields of \GADGET, original PM,
+  and optimized PM-SONN.
+  The examples here are shown for the 0th data point in our samples of
+  configurations and cosmological parameters.
+  This corresponds to an overall simulation box size of 79 Mpc$/h$.
+  We show a zoom-in view with window length XXX Mpc$/h$, and a projected
+  slab thickness of XXX Mpc$/h$.}
   \label{fig:denslab}

### 2024-03-04 docs/papers/sto/so.tex
FIXUP draft: minor

@@
 
-We fit the trained NNs with symbolic regression for analytic
+To make the SO more interpretable and applicable to other simulations,
+we fit the trained NNs with symbolic regression for analytic
 expressions of the $f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3;
@@
 
-In \autoref{fig:tfcc}, we show the results of $T$ and $r$ for both
+In \autoref{fig:tfcc}, we show the results of TF and CC for both
 training and test data samples.
@@
 $l_p$ is the particle spacing.
-
-comparison of PM, PM with SO NN (PM-SONN), PM with SO SR (PM-SOSR)
+Compared to original PM, the TF of PM-SONN is improved to be consistent
+with \GADGET\ within 10\% up to the particle grid Nyquist wavenumber.
 
@@
 
-% density field with zoom in (with halos painted as circles?)
 The effect of force sharpening can also be visually shown by looking
-directly at the density field.
+directly at the higher resolution of the density field
+(\autoref{fig:denslab}) on small scales.
+Compared to orginal PM, PM-SONN is more similar to \GADGET\ with sharper
+big halos and more small-scale structures.
 
@@
 \subsection{Benchmarks}
-We compare the computational cost of \GADGET, \pmwd\ with NNs, and
-\pmwd\ with SR.
+We compare the computational cost of \GADGET, original PM, PM-SONN, and
+PM-SOSR.
 
-We time the computation of force sharpening for one step using NNs and
-SR respectively.
+The forward evaluation of the NNs is more computation and memory
+expensive compared to the analytic expressions.
+We time the computation of SO for one step using NNs and SR
+respectively.
 

### 2024-03-18 pmwd/sto/hypars.py
FIXUP minor

@@ -26,7 +26,7 @@ optimizer = optax.MultiSteps(optimizer, 1)
 # so neural nets
 so_type = 'NN'
 soft_i = 'soft_v1'
-n_input = [soft_len('g'), soft_len('f')]
+n_input = [soft_len(soft_i, 'g'), soft_len(soft_i, 'f')]
 
 so_nodes = [[3*n] * 5 + [1] for n in n_input]
 

### 2024-03-18 docs/papers/sto/so.tex
FIXUP update figures of f and g nets

@@
 \begin{equation}
-- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\left[- \vnabla \varphi\right]_\vk = - i \vk \varphi(\vk) =
+                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\label{force}
 \end{equation}
 %
-we then introduce SO force sharpening through
+which then gives the force field in configuration space after an inverse
+Fourier transform.
+
+In the PM simulation, the density field $\delta(\vx)$ is estimated by
+scattering the particles onto the mesh, and then the force on each
+particle is estimated by gathering the force field from the mesh.
+Therefore, the accuracy of the gravitational force estimated for each
+particle depends on the resolution of the mesh.
+To improve the accuracy of the PM solver, we introduce SO force
+sharpening by modifying the force equation \eqref{force} through
 %
@@
   \centering
-  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
-  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{net_f_a1_j3177874_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{net_g_a1_j3177874_e3000.pdf}
   \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
@@
 PM-SOSR.
+In the generation of our training data, on average each \GADGET\
+simulation takes about 2K CPU hours.
 
@@
 We time the computation of SO for one step using NNs and SR
-respectively.
+respectively, both after the just-in-time (JIT) compilation function
+transformation in JAX.
 
@@
 on multiple computing nodes.
-Each process is binded to a single GPU device, and the training data
-samples are uniformly distributed to the processes.
+With data parallelism, the training data samples are uniformly
+distributed to the processes, and each process is binded to a single GPU
+device.
 In total, we use XXX \GPU\ GPUs, each accompanied with XXX CPU cores
@@
 We initialize the parameters of NNs in a way such that the output is
-very close to one, i.e. no SO.
+around one with slight deviations, i.e. roughly the original PM without
+SO.
 This is achieved by setting the kernel of the output layer to be very
@@
 one after the exponential function.
-The parameters are also initialized to be the same for all devices, and
-hence remain the same during training given the same averaged gradients.
+The parameters are initialized to be the same for all devices, and hence
+remain the same during training given the same averaged gradients.
 

### 2024-03-27 docs/papers/sto/so.tex
FIXUP draft minor

@@
 
-\begin{table*}
-  \centering
-  \caption{Ranges of \GADGET\ and \pmwd\ configurations and cosmological
-  parameters.
-  Note that the grid ratio need next fast len to determine the mesh shape
-  for fast FFT.
-  Given the box size, the mesh shape determines the cell size.
-  The softening parameter gives the ratio of the comoving softening length
-  to the mean particle spacing.
-  The curvature $\OmegaK$ is related to the separate universe simulation.
-  We sample parameters applicable to \GADGET\ during data generation, and
-  those applicable to \pmwd\ during training.
-  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
-  Now it seems more natural to me to have the ptcl spacing sampled
-  uniformly instead...)}
-  }
-  \label{tab:param}
-  \begin{tabular}{lcccr}
-  \toprule
-  applicability & parameter & distribution & range & sampling \\
-  \midrule
-  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
-  \cmidrule(lr){2-5}
-  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
-  \cmidrule(lr){2-4}
-  & snapshot offset & uniform & (0, 1] \\
-  \cmidrule(lr){2-4}
-  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
-  \cmidrule(lr){2-4}
-  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
-  \cmidrule(lr){2-4}
-  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
-  \cmidrule(lr){2-4}
-  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
-  \cmidrule(lr){2-4}
-  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
-  \cmidrule(lr){2-4}
-  & $h$ & log-uniform & [0.5, 1) \\
-  \cmidrule(lr){1-4}
-  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
-  \cmidrule(lr){1-5}
-  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
-  \cmidrule(lr){2-4}
-  \cmidrule(lr){1-4}
-  training & loss grid offset & uniform & $[0, 1)^3$ \\
-  \bottomrule
-  \end{tabular}
-  \end{table*}
-
-\begin{table}
-  \centering
-  \caption{Possible features that can affect force sharpening.
-  All scales are then multiplied by the wavenumber $k$ to form dimensionless
-  factors.}
-  \label{tab:feat}
-  \begin{tabular}{rl}
-  \toprule
-  type & feature \\
-  \midrule
-  & $R_\mathrm{d}$: RMS linear displacement \\
-  \cmidrule(lr){2-2}
-  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
-  \cmidrule(lr){2-2}
-  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
-  \cmidrule(lr){2-2}
-  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
-  \cmidrule(lr){2-2}
-  & $\d R / \d\lna$: time derivatives of the above \\
-  \midrule
-  & particle spacing \\
-  \cmidrule(lr){2-2}
-  other scales & $l$: PM force mesh cell size \\
-  \cmidrule(lr){2-2}
-  & softening length \\
-  \midrule
-  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
-  \cmidrule(lr){2-2}
-  & $\d\ln G_m / \d\lna$ \\
-  \cmidrule(lr){2-2}
-  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
-  \cmidrule(lr){2-2}
-  & $\d\ln\!H / \d\lna$ \\
-  \cmidrule(lr){2-2}
-  & $\Delta\lna$: time step size in $\lna$ \\
-  & \YL{Now this can be an important feature with varying $a_\ic$} \\
-  \bottomrule
-  \end{tabular}
-  \end{table}
-
 Cosmological $N$-body simulations evolve dark matter particles under the
@@
 
+\subsection{Target Simulation and Data}
+
+\begin{table*}
+  \centering
+  \caption{Ranges of \GADGET\ and \pmwd\ configurations and cosmological
+  parameters.
+  Note that the grid ratio need next fast len to determine the mesh shape
+  for fast FFT.
+  Given the box size, the mesh shape determines the cell size.
+  The softening parameter gives the ratio of the comoving softening length
+  to the mean particle spacing.
+  The curvature $\OmegaK$ is related to the separate universe simulation.
+  We sample parameters applicable to \GADGET\ during data generation, and
+  those applicable to \pmwd\ during training.
+  \YL{(I forgot why I wanted to sample mesh shape and mesh cell size...
+  Now it seems more natural to me to have the ptcl spacing sampled
+  uniformly instead...)}
+  }
+  \label{tab:param}
+  \begin{tabular}{lcccr}
+  \toprule
+  applicability & parameter & distribution & range & sampling \\
+  \midrule
+  & $128^3$ white noise & $\mathcal{N}(\vzero, \vI)$ & 512 realizations & MC \\
+  \cmidrule(lr){2-5}
+  & box size in Mpc$/h$ & \YL{log-uniform} & [25.6, 2560) \\
+  \cmidrule(lr){2-4}
+  & snapshot offset & uniform & (0, 1] \\
+  \cmidrule(lr){2-4}
+  \GADGET\ & $\As \times 10^9$ & log-uniform & [1, 4) \\
+  \cmidrule(lr){2-4}
+  \& \pmwd\ & $\ns$ & log-uniform & [0.75, 1.25) & RQMC \\
+  \cmidrule(lr){2-4}
+  & $\Omegam$ & log-uniform & [1/5, 1/2) & see \autoref{fig:sobol} \\
+  \cmidrule(lr){2-4}
+  & $\Omegab / \Omegam$ & log-uniform & [1/8, 1/4) \\
+  \cmidrule(lr){2-4}
+  & $\OmegaK / (1 - \OmegaK)$ & uniform & [-1/3, 1/3) \\
+  \cmidrule(lr){2-4}
+  & $h$ & log-uniform & [0.5, 1) \\
+  \cmidrule(lr){1-4}
+  \GADGET\ & softening ratio & log-uniform & [1/50, 1/20) \\
+  \cmidrule(lr){1-5}
+  \pmwd\ & mesh shape & log-uniform & [128, 512] & MC \\
+  \cmidrule(lr){2-4}
+  \cmidrule(lr){1-4}
+  training & loss grid offset & uniform & $[0, 1)^3$ \\
+  \bottomrule
+  \end{tabular}
+  \end{table*}
+
 Our target data samples are generated with the \GADGET\ simulation code
@@
 For both the $f$ and $g$ spatial optimization NNs (SONNs), we adopt
-multilayer perceptron (MLP, i.e. fully connected) structures to learn
-the nonlinear functions.
+multilayer perceptron (MLP, i.e. NN with fully connected layers)
+structures to learn the nonlinear functions.
 Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
@@
 
-With discrete Fourier transform (DFT) on a regular grid of length $N_c$
-and cell size $l_c$, the range of sampled 1D discrete wavenumbers is
-limited by the fundamental and Nyquist values, i.e. $k_{\rm fund} =
-2\pi/(l_c N_c)$ and $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is
-simply the box size.
-Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
-$[2/N_c, 1]$ for different box sizes.
-For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum
-normalized value is then $\sqrt{3}$.
-
 For $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally symmetric
@@
 To make the SO more interpretable and applicable to other simulations,
-we fit the trained NNs with symbolic regression for analytic
-expressions of the $f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3;
-\vvartheta)$ functions.
+we fit the trained NNs with SR for analytic expressions of the $f(k_i;
+\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ functions.
+
+The data for SR fitting is generated from the trained NNs, while the
+input features for SR could be different from those of the NNs.
+Based on the NN input features, we could further design other features
+for SR fitting.
 
@@
 $l_p$ is the particle spacing.
+With discrete Fourier transform (DFT) on a regular grid of length $N_c$
+and cell size $l_c$, the range of sampled 1D discrete wavenumbers is
+limited by the fundamental and Nyquist values, i.e. $k_{\rm fund} =
+2\pi/(l_c N_c)$ and $k_{\rm Nyquist} = \pi/l_c$, where $l_c N_c$ is
+simply the box size.
+Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
+$[2/N_c, 1]$ for different box sizes.
+For the 3D wavenumber $k=\sqrt{\sum_{i=1}^{3}k_i^2}$, the maximum
+normalized value is then $\sqrt{3}$.
 Compared to original PM, the TF of PM-SONN is improved to be consistent
@@
 \vspace{1em}
-\subsection{Input Features}
+\subsection{Physical Features}
 \label{sec:features}
+
+\begin{table}
+  \centering
+  \caption{Possible features that can affect force sharpening.
+  All scales are then multiplied by the wavenumber $k$ to form dimensionless
+  factors.}
+  \label{tab:feat}
+  \begin{tabular}{rl}
+  \toprule
+  type & feature \\
+  \midrule
+  & $R_\mathrm{d}$: RMS linear displacement \\
+  \cmidrule(lr){2-2}
+  & $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  & $R_\gauss$: $\sigma_\gauss^2(R_\gauss) \triangleq 1$ \\
+  \cmidrule(lr){2-2}
+  & $\d R / \d\lna$: time derivatives of the above \\
+  \midrule
+  & particle spacing \\
+  \cmidrule(lr){2-2}
+  other scales & $l$: PM force mesh cell size \\
+  \cmidrule(lr){2-2}
+  & softening length \\
+  \midrule
+  & $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
+  \cmidrule(lr){2-2}
+  & $\d\ln G_m / \d\lna$ \\
+  \cmidrule(lr){2-2}
+  dimensionless factors & $\Omegam(a), \YL{\OmegaK(a)}$ \\
+  \cmidrule(lr){2-2}
+  & $\d\ln\!H / \d\lna$ \\
+  \cmidrule(lr){2-2}
+  & $\Delta\lna$: time step size in $\lna$ \\
+  & \YL{Now this can be an important feature with varying $a_\ic$} \\
+  \bottomrule
+  \end{tabular}
+  \end{table}
 

### 2024-04-17 docs/papers/sto/so.tex
Draft minor
Update paper draft

@@
 Rapid advances in artificial intelligence (AI) and more specifically
-deep learning (DL) have provided powerful tools that are revolutionizing
+deep learning have provided powerful tools that are revolutionizing
 scientific research across disciplines.
-AI for science consists of not only the state-of-art deep neural
-networks, but also more fundamental computational tools and hardwares
-like automatic differentiation (AD) and GPUs.
-One particular application is the numerical simulation of the Universe.
+AI for science benefits from not only the state-of-art architectures of
+deep neural networks (NNs), but also more fundamental computational
+tools and hardwares like automatic differentiation (AD) and GPUs.
+One of the very promising areas of AI in Cosmology is the numerical
+simulation of the Universe.
+
+% combination of differentiable simulations, NN, and SR
+NNs are flexible and efficient universal function approximators
+\citep{Hornik1989}, which can be used to learn physical processes
+when embedded in numerical simulations.
+NNs are typically trained with gradient based optimizers, which requires
+the simulations to be differentiable.
 
 % a brief review of (differentiable) cosmological simulations
-BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
-\texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
+% BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
+% \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
 
 % introduce PM and its limitation on small scales
-The particle-mesh (PM) method solves the Poisson equation by scattering
-the particles onto a mesh, on which FFT is performed and the Laplacian
-operator reduces to simple multiplication of wavenumbers in Fourier
-space.
-Although being computationally efficient, the PM method is mainly
-limited by its resolution on small scales, i.e. scales around the mesh
+The particle-mesh (PM) method is a fundamental technique for
+cosmological $N$-body simulations.
+PM solves the Poisson equation by scattering the particles onto an
+auxiliary mesh, on which the fast Fourier transform (FFT) is performed
+and the Laplacian operator reduces to a simple multiplication of
+wavenumbers in Fourier space.
+The cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} has been
+widely used to compute the mass fractions of a particle at $\vx'$
+scattered to a nearby mesh grid point at $\vx$
+%
+\begin{equation}
+  W(\vx, \vx') = \prod_{i=1}^3
+    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
+  \label{cic}
+  \end{equation}
+%
+where $l_c$ is the mesh cell size.
+The gravitational forces on the mesh points are then grathered to form
+the interpolated force for each particle, with the same weights as in
+the scattering process.
+Other scattering schemes include the nearest grid point (NGP) and
+triangle shaped cloud (TSC) etc.
+Although being computationally efficient, the PM method is limited by
+its resolution on small scales, i.e. scales around and below the mesh
 cell size.
-This is due to the loss of information during the scattering process of
-continuously distributed particles onto the discrete mesh.
-We use the cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} to
-compute the mass fractions of a particle scattered to nearby grid
-points.
-Other schemes include the nearest grid point (NGP) and triangle shaped
-cloud (TSC) etc.
+This is due to the loss of information during the scattering and
+gathering process between the continuously distributed particles and the
+discrete mesh.
 
@@
 Within this fully differentiable framework, we sharpen the gravitational
-force by embedding neural networks (NNs) in the Poisson equation and
-train them for the spatial optimization (SO) of the simulation
-resolution on small scales.
-For better interpretability, we further fit the trained NNs using
-symbolic regression (SR).
+force by embedding NNs in the Poisson equation and train them for the
+spatial optimization (SO) of the simulation resolution on small scales.
+We further interpret the trained NNs using symbolic regression (SR).
 
@@
 gravitational potential and matter density fluctuation respectively.
-In Fourier space, the gravitational force field $- \vnabla \varphi$
+In Fourier space, the gravitational force field $\vF = - \vnabla \varphi$
 follows
@@
 \begin{equation}
-\left[- \vnabla \varphi\right]_\vk = - i \vk \varphi(\vk) =
+\vF(\vk) = - i \vk \varphi(\vk) =
                   i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
@@
 In the PM simulation, the density field $\delta(\vx)$ is estimated by
-scattering the particles onto the mesh, and then the force on each
-particle is estimated by gathering the force field from the mesh.
+scattering the particles onto the mesh with e.g. the CIC scheme in
+\eqref{cic}.
+Then the force on each particle is estimated by gathering the force
+field from the mesh.
 Therefore, the accuracy of the gravitational force estimated for each
@@
 To improve the accuracy of the PM solver, we introduce SO force
-sharpening by modifying the force equation \eqref{force} through
-%
-\begin{equation}
-\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
-  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
-\end{equation}
-%
-where $f$ and $g$ are NNs, each approximating a nonlinear function of
-the wavevector components and all possible relevant features
-$\vvartheta$.
-The subscript $i\in\{1,2,3\}$ denotes the 3 principal axes of the
-simulation box.
+sharpening by modifying the force equation \eqref{force} to
+%
+\begin{equation}
+F(\vk)_i = i \frac32 \Omegam H_0^2 \,\frac{k_i}{k^2}\,
+                    f(k_i; \vvartheta) g(\vk; \vvartheta) \delta(\vk),
+\end{equation}
+%
+where $f$ and $g$ are nonlinear functions of the wavevector components
+and all possible relevant features $\vvartheta$, which are approximated
+with NNs.
+$k_i$ is a component of $\vk$, with $i\in\{1,2,3\}$ denoting the 3
+principal axes of the simulation box.
 
@@
 
-For both the $f$ and $g$ spatial optimization NNs (SONNs), we adopt
-multilayer perceptron (MLP, i.e. NN with fully connected layers)
-structures to learn the nonlinear functions.
+For both the $f$ and $g$ spatial optimization NNs (SONNs), the input
+features consist of tabular data, therefore we adopt multilayer
+perceptron (MLP, i.e. NN with fully connected layers) structures to
+learn the nonlinear functions.
 Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm

### 2024-04-18 docs/papers/sto/so.tex
Update paper draft

@@
 % Main text: up to 3500 words: introduction + results + discussion
-% so we should keep introduction and discussion brief
 \vspace{1em}
@@
 
-% combination of differentiable simulations, NN, and SR
+% a brief review of (differentiable) cosmological simulations
+% BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
+% \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
+
+Cosmological $N$-body simulations evolve dark matter particles under the
+gravitational interaction.
+The dynamics of the dark matter is governed by the Poisson equation
+%
+\begin{equation}
+  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
+\end{equation}
+%
+where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
+constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
+gravitational potential and matter density fluctuation respectively.
+In Fourier space, the gravitational force field $\vF = - \vnabla
+\varphi$ follows
+%
+\begin{equation}
+\vF(\vk) = - i \vk \varphi(\vk) =
+                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\label{force}
+\end{equation}
+%
+where the Laplacian operator reduces to simple multiplications.
+The force field in configuration space $\vF(\vx)$ can then be recovered
+with an inverse Fourier transform.
+The particle-mesh (PM) method is a fundamental and efficient approach to
+solving the Poisson equation given numerous particles.
+In order to use the fast Fourier transform (FFT), PM estimates the
+matter density field on a regular spatial grid by scattering the
+particles onto an auxiliary mesh.
+The cloud-in-cell (CIC), or trilinear, interpolation scheme
+\citep{HockneyEastwood1988} has been widely used to compute the mass
+fractions of a particle at $\vx'$ scattered to a nearby mesh point
+at $\vx$
+%
+\begin{equation}
+  W(\vx, \vx') = \prod_{i=1}^3
+    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
+  \label{cic}
+  \end{equation}
+%
+where $l_c$ is the mesh cell size.
+Other scattering schemes include the nearest grid point (NGP) and
+triangle shaped cloud (TSC) etc.
+The gravitational forces on the mesh points are computed in the Fourier
+space and transformed back to the configuration space, which are then
+grathered to form the interpolated force for each particle, with the
+same weights as in the scattering process.
+Although being computationally efficient, the accuracy of the
+gravitational force estimated for each particle depends on the
+resolution of the mesh.
+The PM method is mainly limited by its resolution on small scales, i.e.
+scales around and below the mesh cell size.
+This is due to the loss of information during the scattering and
+gathering process between the continuously distributed particles and the
+regular discrete mesh.
+
+In this work, we sharpen the gravitational force by embedding NNs in the
+force equation and train them for the spatial optimization (SO) of the
+PM simulation.
 NNs are flexible and efficient universal function approximators
@@
 the simulations to be differentiable.
-
-% a brief review of (differentiable) cosmological simulations
-% BORG, ELUCID, and BORG-PM \citep{BORG, ELUCID, BORG-PM}
-% \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM, vmad, SeljakEtAl2017, FlowPM}
-
-% introduce PM and its limitation on small scales
-The particle-mesh (PM) method is a fundamental technique for
-cosmological $N$-body simulations.
-PM solves the Poisson equation by scattering the particles onto an
-auxiliary mesh, on which the fast Fourier transform (FFT) is performed
-and the Laplacian operator reduces to a simple multiplication of
-wavenumbers in Fourier space.
-The cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} has been
-widely used to compute the mass fractions of a particle at $\vx'$
-scattered to a nearby mesh grid point at $\vx$
-%
-\begin{equation}
-  W(\vx, \vx') = \prod_{i=1}^3
-    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
-  \label{cic}
-  \end{equation}
-%
-where $l_c$ is the mesh cell size.
-The gravitational forces on the mesh points are then grathered to form
-the interpolated force for each particle, with the same weights as in
-the scattering process.
-Other scattering schemes include the nearest grid point (NGP) and
-triangle shaped cloud (TSC) etc.
-Although being computationally efficient, the PM method is limited by
-its resolution on small scales, i.e. scales around and below the mesh
-cell size.
-This is due to the loss of information during the scattering and
-gathering process between the continuously distributed particles and the
-discrete mesh.
-
-% motivation of this work
+The differentiable cosmological simulation with adjoint method
+\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
+\citep{Li2022b} lay the foundation of the gradient-based optimization
+discussed in this work.
+We further interpret the trained NNs using symbolic regression (SR).
 Optimizing the accuracy of fast and differentiable simulations is
@@
 
-% introduction of this work
-The differentiable cosmological simulation with adjoint method
-\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
-\citep{Li2022b} lay the foundation of the gradient-based optimization
-discussed in this work.
-Within this fully differentiable framework, we sharpen the gravitational
-force by embedding NNs in the Poisson equation and train them for the
-spatial optimization (SO) of the simulation resolution on small scales.
-We further interpret the trained NNs using symbolic regression (SR).
-
 \vspace{1em}
@@
 
-Cosmological $N$-body simulations evolve dark matter particles under the
-gravitational force.
-The dynamics of the particles is governed by the Poisson equation
-%
-\begin{equation}
-  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
-\end{equation}
-%
-where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
-constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
-gravitational potential and matter density fluctuation respectively.
-In Fourier space, the gravitational force field $\vF = - \vnabla \varphi$
-follows
-%
-\begin{equation}
-\vF(\vk) = - i \vk \varphi(\vk) =
-                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
-\label{force}
-\end{equation}
-%
-which then gives the force field in configuration space after an inverse
-Fourier transform.
-
-In the PM simulation, the density field $\delta(\vx)$ is estimated by
-scattering the particles onto the mesh with e.g. the CIC scheme in
-\eqref{cic}.
-Then the force on each particle is estimated by gathering the force
-field from the mesh.
-Therefore, the accuracy of the gravitational force estimated for each
-particle depends on the resolution of the mesh.
 To improve the accuracy of the PM solver, we introduce SO force
-sharpening by modifying the force equation \eqref{force} to
+sharpening by modifying the gravitational force equation \eqref{force}
+to
 %

### 2024-04-18 docs/examples/sto/train/run_train.py
Remove reduce_on_plateau func, optax now has one

@@ -1,10 +1,13 @@
 """Multi-process SO training using jax pmap, one process contains one gpu."""
 import os
-os.environ['CUDA_DEVICE_ORDER'] = 'PCI_BUS_ID'
+
 n_procs = int(os.getenv('SLURM_NTASKS'))
 procid = int(os.getenv('SLURM_PROCID'))
 n_tasks_per_node = int(os.getenv('SLURM_NTASKS_PER_NODE'))
 slurm_job_id = os.getenv('SLURM_JOB_ID')
+
+# setup the CUDA device binded to the proc
+os.environ['CUDA_DEVICE_ORDER'] = 'PCI_BUS_ID'
 os.environ['CUDA_VISIBLE_DEVICES'] = str(procid % n_tasks_per_node)
 
 os.environ['XLA_PYTHON_CLIENT_MEM_FRACTION'] = '.95'
@@ -19,14 +22,12 @@ from torch.utils.tensorboard import SummaryWriter
 from datetime import datetime
 import time
 import pickle
-import optax
 
 from pmwd.sto.train import train_epoch, loss_epoch
 from pmwd.sto.vis import track_figs
 from pmwd.sto.data import read_gsdata
 from pmwd.sto.post import pmwd_fwd
 from pmwd.sto.util import pv2ptcl, global_mean
-from pmwd.sto.hypars import lr_scheduler
 
 
 def printinfo(s, procid=procid, flush=False):
@@ -116,11 +117,10 @@ def run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch, learning_rat
 
     # RNGs with fixed seeds
     # pmwd MC sampling
-    np_rng = np.random.default_rng(0)
-    # shuffle of data samples across epoch
-    np_rng_shuffle = np.random.default_rng(procid)
+    np_rng = np.random.default_rng(42)
 
-    skd_state = None
+    # rng for shuffling data samples across epoch
+    np_rng_shuffle = np.random.default_rng(42+procid)
 
     jax_device_sync()
     if procid == 0:
@@ -135,25 +135,21 @@ def run_train(n_epochs, sobol_ids, gsdata, snap_ids, shuffle_epoch, learning_rat
     loss_epoch_all = []  # to collect the loss of all epochs
     sobol_ids_epoch = sobol_ids.copy()
 
+    # training loop over epochs
     for epoch in range(0, n_epochs+1):
 
         # shuffle the data samples across epoch
[…]

### 2024-04-18 pmwd/sto/hypars.py

@@ -1,35 +1,32 @@
 import numpy as np
-import jax.numpy as jnp
 import optax
-import pickle
 
 from pmwd.sto.so import soft_len
 from pmwd.sto.mlp import init_mlp_params
 
 n_epochs = 5000
 
-# data
+###  data  ###
 sobol_ids_global = np.arange(0, 64)
 snap_ids = np.arange(0, 121, 2)
 shuffle_epoch = True  # shuffle the order of sobols across epochs
 
-# loss
+###  loss  ###
 loss_pars = {
     'log_eps': 0,
 }
 
-# optimizer
+###  optimizer  ###
 learning_rate = 1e-5
 optimizer = optax.adam(learning_rate)
 optimizer = optax.MultiSteps(optimizer, 1)
 
-# so neural nets
+###  model  ###
 so_type = 'NN'
 soft_i = 'soft_v1'
 n_input = [soft_len(soft_i, 'g'), soft_len(soft_i, 'f')]
 
 so_nodes = [[3*n] * 5 + [1] for n in n_input]
-
 
 # start a new training
 so_params = init_mlp_params(n_input, so_nodes, scheme='last_ws')
@@ -41,33 +38,3 @@ opt_state = optimizer.init(so_params)
 #     so_params = dic['so_params']
 #     # opt_state = dic['opt_state']
 #     opt_state = optimizer.init(so_params)
-
-
-
-def get_optimizer(lr):
-    return optax.adam(lr)
-    # return optax.adamw(lr, weight_decay=1e-3)
-
-
-def lr_scheduler(lr, skd_state, loss, patience=15, factor=0.5, threshold=1e-4, lr_min=1e-7):
-    """Reduce the learning rate on plateau, a mini implementation."""
-    if skd_state is None:  # initialize
-        return lr, (0, loss)
-
-    counter, loss_low = skd_state
-
-    if loss_low - loss < abs(loss_low) * threshold:  # not better
-        counter += 1
-        if counter == patience:  # reduce LR
-            lr = max(lr * factor, lr_min)
-            counter = 0
-    else:  # is better
-        counter = 0
-        loss_low = loss
-
-    skd_state = (counter, loss_low)
-
-    return lr, skd_state
-
-# mannually turn off nets by setting the corresponding so_nodes to None
-# for i in [0, 2]: so_nodes[i] = None

### 2024-04-18 docs/examples/sto/train/job.sh
Remove reduce_on_plateau func, optax now has one

@@ -19,7 +19,6 @@ export PYTHONPATH="./pymods"
 source /mnt/home/yzhang/miniforge3/bin/activate jax0411
 
 srun python run_train.py
-# srun python optune.py
 
 
 # Notes:

### 2024-04-18 docs/examples/sto/g4gen/setup_g4sims.py
Split sto.data into data, sample, ccic for clarity

@@ -5,7 +5,8 @@ import jax
 
 import numpy as np
 
-from pmwd.sto.data import scale_Sobol, gen_cc, gen_ic
+from pmwd.sto.sample import scale_Sobol
+from pmwd.sto.ccic import gen_cc, gen_ic
 from pmwd.io_util import write_gadget_hdf5
 
 

### 2024-04-18 docs/examples/sto/sr/sonn_data.py

@@ -6,7 +6,8 @@ from tqdm import tqdm
 from scipy.stats.qmc import Sobol
 
 from pmwd.sto.util import load_soparams
-from pmwd.sto.data import scale_Sobol, gen_cc
+from pmwd.sto.sample import scale_Sobol
+from pmwd.sto.ccic import gen_cc
 from pmwd.sto.mlp import MLP
 from pmwd.sto.so import sotheta, soft_k, soft_kvec, soft_names, soft_names_tex
 

### 2024-04-18 pmwd/sto/post.py

@@ -4,7 +4,8 @@ import matplotlib.pyplot as plt
 
 from pmwd.nbody import nbody
 from pmwd.sto.train import pmodel, init_pmwd
-from pmwd.sto.data import G4snapDataset, gen_cc, gen_ic
+from pmwd.sto.data import G4snapDataset
+from pmwd.sto.ccic import gen_cc, gen_ic
 from pmwd.sto.mlp import mlp_size
 from pmwd.sto.util import power_tfcc, scatter_dens, pv2ptcl, load_soparams, tree_unstack
 

### 2024-04-18 pmwd/sto/sample.py

@@ -1,0 +1,74 @@
+import numpy as np
+
+
+def gen_sobol(filename=None, d=9, m=9, extra=9, seed=55868, seed_max=65536):
+    from scipy.stats.qmc import Sobol, discrepancy
+
+    nicer_seed = seed
+    if seed is None:
+        disc_min = np.inf
+        for s in range(seed_max):
+            sampler = Sobol(d, scramble=True, seed=s)  # d is the dimensionality
+            sample = sampler.random_base2(m)  # m is the log2 of the number of samples
+            disc = discrepancy(sample, method='MD')
+            if disc < disc_min:
+                nicer_seed = s
+                disc_min = disc
+        print(f'0 <= seed = {nicer_seed} < {seed_max}, minimizes mixture discrepancy = '
+                f'{disc_min}')
+        # nicer_seed = 55868, mixture discrepancy = 0.016109347957680598
+
+    sampler = Sobol(d, scramble=True, seed=nicer_seed)
+    sample = sampler.random(n=2**m + extra)  # extra is the additional testing samples
+    if filename is None:
+        return sample
+    else:
+        np.savetxt(filename, sample)
+
+
+def scale_Sobol(sobol=None, fn='sobol.txt', ind=slice(None)):
+    """Scale the Sobol sequence samples, refer to the Table in the paper."""
+    if sobol is None:
+        sobol = np.loadtxt(fn)[ind].T
+    # functions mapping uniform random samples in [0, 1] to a desired one
+    f_uni = lambda x, a, b : a + x * (b - a)
+    f_log_uni = lambda x, a, b : np.exp(f_uni(x, np.log(a), np.log(b)))
+    def f_log_trap(x, a, b, c):
+        # a, b, c are the locations of the first 3 points of the symmetric trapezoid
+        h = 1 / (c - a)
+        x1 = (b - a) * h / 2
+        x2 = (2*c - b - a) * h / 2
+        y = np.zeros_like(x)
+        m = (x < x1)
+        y[m] = a + np.sqrt(2 * (b - a) * x[m] / h)
+        m = (x1 <= x) & (x < x2)
+        y[m] = x[m] / h + (a + b) / 2
+        m = (x2 <= x)
[…]

### 2024-04-18 docs/examples/sto/g4gen/scaled_sobol_csv.py
Split sto.data into data, sample, ccic for clarity

@@ -1,6 +1,6 @@
 """Generate the csv file of scaled Sobol parameters for convenience."""
 import numpy as np
-from pmwd.sto.data import scale_Sobol
+from pmwd.sto.sample import scale_Sobol
 
 sobol = scale_Sobol()
 

### 2024-04-18 docs/examples/sto/g4gen/sobol.py

@@ -3,7 +3,7 @@ import os
 import numpy as np
 import matplotlib.pyplot as plt
 
-from pmwd.sto.data import scale_Sobol, gen_sobol
+from pmwd.sto.sample import scale_Sobol, gen_sobol
 
 
 def plt_proj(filename, max_rows=None, max_cols=None):

### 2024-04-18 docs/examples/sto/train/run_train.py

@@ -27,7 +27,7 @@ from pmwd.sto.train import train_epoch, loss_epoch
 from pmwd.sto.vis import track_figs
 from pmwd.sto.data import read_gsdata
 from pmwd.sto.post import pmwd_fwd
-from pmwd.sto.util import pv2ptcl, global_mean
+from pmwd.sto.util import pv2ptcl, tree_global_mean
 
 
 def printinfo(s, procid=procid, flush=False):
@@ -37,10 +37,10 @@ def printinfo(s, procid=procid, flush=False):
 
 def jax_device_sync(verbose=False):
     """Nothing but to sync all devices, with dummy global mean."""
-    x = global_mean(jnp.array(procid))
+    x = tree_global_mean(jnp.array(procid))
     assert round(2 * x + 1) == n_procs, 'something wrong with global mean'
     if verbose:
-        printinfo(f'# global devices: {len(jax.devices())}, device sync successful',
+        printinfo(f'# global devices: {len(jax.devices())}, sync successful',
                   flush=True)
 
 
@@ -67,33 +67,34 @@ def track(writer, epoch, scalars, check_sobols, check_snaps,
             writer.add_scalar(k, v, epoch)
 
     # check a few training sobols and snaps
-    for sidx in check_sobols:
-        # get target snap and sobol
-        tgts, a_snaps, sobol, snap_ids = (gsdata[sidx][k] for k in (
-                                          'pv', 'a_snaps', 'sobol', 'snap_ids'))
-        a_snaps = tuple(a_snaps[i] for i in check_snaps)
-        tgts = tuple(t[check_snaps] for t in tgts)
-        snap_ids = snap_ids[check_snaps]
+    # for sidx in check_sobols:
+    #     # get target snap and sobol
+    #     tgts, a_snaps, sobol, snap_ids = (gsdata[sidx][k] for k in (
+    #                                       'pv', 'a_snaps', 'sobol', 'snap_ids'))
+    #     a_snaps = tuple(a_snaps[i] for i in check_snaps)
+    #     tgts = tuple(t[check_snaps] for t in tgts)
+    #     snap_ids = snap_ids[check_snaps]
 
-        # run pmwd
-        obsvbl, cosmo, conf = pmwd_fwd(so_params, sidx, sobol, a_snaps,
-                                       mesh_shape, n_steps, so_type, so_nodes,
-                                       soft_i)
[…]

### 2024-04-18 pmwd/sto/ccic.py

@@ -1,0 +1,51 @@
+import jax.numpy as jnp
+
+from pmwd.configuration import Configuration
+from pmwd.cosmology import Cosmology
+from pmwd.boltzmann import boltzmann
+from pmwd.modes import white_noise, linear_modes
+from pmwd.lpt import lpt
+
+
+def gen_cc(sobol, mesh_shape=1, a_snapshots=(1,), a_nbody_num=61,
+           so_type=None, so_nodes=None, soft_i=None, a_start=1/16, a_stop=1+1/128,
+           cal_boltz=True):
+    """Setup conf and cosmo given a sobol."""
+    conf = Configuration(
+        ptcl_spacing = sobol[0] / 128,
+        ptcl_grid_shape = (128,) * 3,
+        a_start = a_start,
+        a_stop = a_stop,
+        float_dtype = jnp.float64,
+        mesh_shape = mesh_shape,
+        a_snapshots = a_snapshots,
+        a_nbody_num = a_nbody_num,
+        so_type = so_type,
+        so_nodes = so_nodes,
+        soft_i = soft_i,
+        softening_length = sobol[8],
+    )
+
+    cosmo = Cosmology(
+        conf = conf,
+        A_s_1e9 = sobol[2],
+        n_s = sobol[3],
+        Omega_m = sobol[4],
+        Omega_b = sobol[5],
+        Omega_k_ = sobol[6],
+        h = sobol[7],
+    )
+    if cal_boltz:
+        cosmo = boltzmann(cosmo, conf)
+
+    return conf, cosmo
+
+
+def gen_ic(seed, conf, cosmo):
+    """Generate the initial condition with lpt for nbody."""
+    modes = white_noise(seed, conf)
+
+    modes = linear_modes(modes, cosmo, conf)
+    ptcl, obsvbl = lpt(modes, cosmo, conf)
+
+    return ptcl

### 2024-04-18 pmwd/sto/data.py

@@ -5,129 +5,8 @@ from torch.utils.data import Dataset
 from joblib import Parallel, delayed
 import h5py
 
-from pmwd.configuration import Configuration
-from pmwd.cosmology import Cosmology
-from pmwd.boltzmann import boltzmann
-from pmwd.modes import white_noise, linear_modes
-from pmwd.lpt import lpt
 from pmwd.io_util import read_gadget_hdf5
-
-
-def gen_sobol(filename=None, d=9, m=9, extra=9, seed=55868, seed_max=65536):
-    from scipy.stats.qmc import Sobol, discrepancy
-
-    nicer_seed = seed
-    if seed is None:
-        disc_min = np.inf
-        for s in range(seed_max):
-            sampler = Sobol(d, scramble=True, seed=s)  # d is the dimensionality
-            sample = sampler.random_base2(m)  # m is the log2 of the number of samples
-            disc = discrepancy(sample, method='MD')
-            if disc < disc_min:
-                nicer_seed = s
-                disc_min = disc
-        print(f'0 <= seed = {nicer_seed} < {seed_max}, minimizes mixture discrepancy = '
-                f'{disc_min}')
-        # nicer_seed = 55868, mixture discrepancy = 0.016109347957680598
-
-    sampler = Sobol(d, scramble=True, seed=nicer_seed)
-    sample = sampler.random(n=2**m + extra)  # extra is the additional testing samples
-    if filename is None:
-        return sample
-    else:
-        np.savetxt(filename, sample)
-
-
-def scale_Sobol(sobol=None, fn='sobol.txt', ind=slice(None)):
-    """Scale the Sobol sequence samples, refer to the Table in the paper."""
-    if sobol is None:
-        sobol = np.loadtxt(fn)[ind].T
-    # functions mapping uniform random samples in [0, 1] to a desired one
-    f_uni = lambda x, a, b : a + x * (b - a)
-    f_log_uni = lambda x, a, b : np.exp(f_uni(x, np.log(a), np.log(b)))
-    def f_log_trap(x, a, b, c):
-        # a, b, c are the locations of the first 3 points of the symmetric trapezoid
-        h = 1 / (c - a)
[…]

### 2024-04-18 pmwd/sto/train.py

@@ -5,9 +5,9 @@ from functools import partial
 import time
 
 from pmwd.nbody import nbody
-from pmwd.sto.data import gen_cc, gen_ic
+from pmwd.sto.ccic import gen_cc, gen_ic
 from pmwd.sto.loss import loss_func
-from pmwd.sto.util import global_mean
+from pmwd.sto.util import tree_global_mean
 
 
 def pmodel(ptcl_ic, so_params, cosmo, conf):
@@ -44,7 +44,7 @@ def train_step(tgts, so_params, pmwd_params, opt_params, loss_pars, loss_mesh_sh
                              loss_mesh_shape)
 
     # average over global devices
-    loss, grad = global_mean((loss, grad))
+    loss, grad = tree_global_mean((loss, grad))
 
     # optimize
     optimizer, opt_state = opt_params
@@ -52,7 +52,7 @@ def train_step(tgts, so_params, pmwd_params, opt_params, loss_pars, loss_mesh_sh
     so_params = optax.apply_updates(so_params, updates)
 
     # not necessary, but no harm
-    so_params = global_mean(so_params)
+    so_params = tree_global_mean(so_params)
 
     return so_params, loss, opt_state
 
@@ -61,7 +61,8 @@ def train_epoch(procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, soft_
                 so_params, opt_state, optimizer, loss_pars, verbose):
     loss_epoch = 0.  # the sum of loss of the whole epoch
 
-    tic = time.perf_counter()
+    if procid == 0:
+        tic = time.perf_counter()
     for step, sidx in enumerate(sobol_ids_epoch):
         tgts, a_snaps, sobol = (gsdata[sidx][k] for k in ('pv', 'a_snaps', 'sobol'))
         tgts = jax.device_put(tgts)  # could be asynchronous
@@ -69,7 +70,7 @@ def train_epoch(procid, epoch, gsdata, sobol_ids_epoch, so_type, so_nodes, soft_
         # mesh shape, [1, 2, 3, 4]
         # mesh_shape = np_rng.integers(1, 5)
         mesh_shape = 1
-        loss_mesh_shape = 1
+        loss_mesh_shape = 3
         loss_pars['grid_offset'] = 0
 
         # number of nbodytime steps
[…]

### 2024-04-18 pmwd/sto/util.py

@@ -2,11 +2,11 @@ import jax
 import jax.numpy as jnp
 from jax.tree_util import tree_map, tree_flatten
 import pickle
-from pmwd.sto.mlp import mlp_size
 
 from pmwd.scatter import scatter
 from pmwd.spec_util import powspec
 from pmwd.particles import Particles
+from pmwd.sto.mlp import mlp_size
 
 
 def tree_stack(trees):
@@ -20,7 +20,7 @@ def tree_unstack(tree):
     return [treedef.unflatten(leaf) for leaf in zip(*leaves, strict=True)]
 
 
-def global_mean(tree):
+def tree_global_mean(tree):
     """Global average of a pytree x, i.e. for all leaves."""
     def global_mean_arr(x):
         x = jnp.expand_dims(x, axis=0)  # leading axis for pmap

### 2024-04-25 docs/papers/sto/so.tex
Update paper draft

@@
 \author[0000-0002-9300-2632]{\normalsize Yucheng Zhang}
-\affiliation{Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
+\affiliation{Peng Cheng Laboratory, Shenzhen, Guangdong 518000, China}
 \affiliation{Center for Cosmology and Particle Physics, New York
@@
 New York, New York 10010, USA}
+
+% \author{\normalsize Heung-Yeung Shum}
+% \affiliation{International Digital Economy Academy, Shenzhen, Guangdong
+% 518000, China}
+% \affiliation{Tsinghua University, Beijing 100084, China}
+% \affiliation{The Hong Kong University of Science and Technology,
+% Hong Kong 999077, China}
 
@@
 \section{Introduction}
+\label{sec:intro}
 % without heading
@@
 tools and hardwares like automatic differentiation (AD) and GPUs.
-One of the very promising areas of AI in Cosmology is the numerical
+One of the very promising areas of AI in cosmology is the numerical
 simulation of the Universe.
@@
 Cosmological $N$-body simulations evolve dark matter particles under the
-gravitational interaction.
-The dynamics of the dark matter is governed by the Poisson equation
+gravitational interaction, which at the field level is described by the
+Poisson equation
 %
@@
 \begin{equation}
-\vF(\vk) = - i \vk \varphi(\vk) =
-                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
+\vF(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
 \label{force}
@@
 %
-where the Laplacian operator reduces to simple multiplications.
+where the Laplacian operator reduces to simple multiplications of the
+wavevectors.
 The force field in configuration space $\vF(\vx)$ can then be recovered
-with an inverse Fourier transform.
+with an inverse Fourier transform of $\vF(\vk)$.
 The particle-mesh (PM) method is a fundamental and efficient approach to
-solving the Poisson equation given numerous particles.
-In order to use the fast Fourier transform (FFT), PM estimates the
-matter density field on a regular spatial grid by scattering the
-particles onto an auxiliary mesh.
+solving the Poisson equation in the $N$-body simulation.
+In order to perform the transformation between configuration and Fourier
+spaces efficiently using the fast Fourier transform (FFT), PM estimates
+the matter density field on a regular grid by scattering the particles
+onto an auxiliary mesh.
 The cloud-in-cell (CIC), or trilinear, interpolation scheme
@@
 triangle shaped cloud (TSC) etc.
-The gravitational forces on the mesh points are computed in the Fourier
-space and transformed back to the configuration space, which are then
-grathered to form the interpolated force for each particle, with the
-same weights as in the scattering process.
+The gravitational force field is evaluated in the Fourier space and
+transformed back to the configuration space on the mesh points, which
+are then grathered to form the interpolated force for each particle,
+with the same weights as in the scattering process.
 Although being computationally efficient, the accuracy of the
@@
 resolution of the mesh.
-The PM method is mainly limited by its resolution on small scales, i.e.
-scales around and below the mesh cell size.
-This is due to the loss of information during the scattering and
-gathering process between the continuously distributed particles and the
-regular discrete mesh.
-
-In this work, we sharpen the gravitational force by embedding NNs in the
-force equation and train them for the spatial optimization (SO) of the
-PM simulation.
+The PM method is mainly limited by its accuracy on small scales around
+and below the mesh cell size, due to the loss of information during the
+scattering and gathering process between the continuously distributed
+particles and the regular discrete mesh.
+
+In this work, we perform the spatial optimization (SO) of the PM
+simulation by sharpening the gravitational force.
+We embed NNs in the force equation \eqref{force} and train them towards
+the targets that are more accurate on small scales.
 NNs are flexible and efficient universal function approximators
-\citep{Hornik1989}, which can be used to learn physical processes
-when embedded in numerical simulations.
-NNs are typically trained with gradient based optimizers, which requires
-the simulations to be differentiable.
-The differentiable cosmological simulation with adjoint method
+\citep{Hornik1989}, which can be used to learn physical processes in
+numerical simulations.
+It is necessary for the simulations to be differentiable in order to
+train the NNs with gradient based optimizers \citep[e.g.][]{Adam}.
+We develop our training pipeline of the SO NNs in the framework of
+the differentiable cosmological simulation with adjoint method
 \citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
-\citep{Li2022b} lay the foundation of the gradient-based optimization
-discussed in this work.
-We further interpret the trained NNs using symbolic regression (SR).
-Optimizing the accuracy of fast and differentiable simulations is
-crucial for their applications in, e.g. generation of mocks for
-cosmological surveys, simulation-based inference or field-level
+\citep{Li2022b}.
+To gain further physical insights into the PM method, we fit the trained
+NNs using symbolic regression (SR).
+Optimizing the accuracy of fast and differentiable simulations like
+\pmwd\ is crucial for their applications in, e.g. generation of mocks
+for cosmological surveys, simulation-based inference or field-level
 inference.
@@
 principal axes of the simulation box.
-
 Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
@@
 the 3 wavevector compontents to achieve the permutation equivariance.
+
+For both the $f$ and $g$ spatial optimization NNs (SONNs), their input
+features consist of tabular data, therefore we adopt multilayer
+perceptrons (MLPs, i.e. NNs with fully connected layers) to learn the
+nonlinear functions.
+Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times
+n_{\rm input}$ nodes for each layer, where $n_{\rm input}$ is the size
+of the input layer, which is [...] and [...] for $f$ and $g$
+respectively.
+The rectified linear unit (ReLU) activation function is used for hidden
+layers.
+We further apply the exponential function to the output as a regulator
+for positive output.
+Being embedded in the differentiable simulation library \pmwd, the NNs
+can be trained using gradient-based optimization methods as in usual
+deep learning tasks.
+The parameters of the NNs, which enter the force evaluation in every
+time integration step, can be treated just the same as other
+cosmological parameters.
+The objective gradients are accumulated, based on which the parameters
+are then optimized.
 
@@
 Our target data samples are generated with the \GADGET\ simulation code
-\citep{GADGET-4}.
+\citep{GADGET-4}, configured with high force accuracy settings.
 We run 512 \GADGET\ simulations that cover a wide range of
@@
 \citep{LiEtAl2014, WagnerEtAl2015}.
-\YZ{We refer to this training dataset of 512 \GADGET\ simulations
-sampled with Sobol sequence as \GSDATA.}
+We refer to this training dataset of 512 \GADGET\ simulations
+sampled with Sobol sequence as \GSDATA.
 On average each \GADGET\ simulation takes roughly 2K CPU hours to run,
@@
 
-\subsection{Spatial Optimization Neural Networks}
+\subsection{Optimized Dark Matter Density Fields}
 
@@
   \centering
-  \includegraphics[width=.48\columnwidth]{net_f_a1_j3177874_e3000.pdf}
-  \includegraphics[width=.48\columnwidth]{net_g_a1_j3177874_e3000.pdf}
-  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
-  functions. The solid line and shaded region indicate the median and
-  the [16, 84] percentiles of the 64 sampled configurations and
-  cosmological parameters. $a=1$,
-  mesh shape=1}
-  \label{fig:func}
+  \includegraphics[width=.98\columnwidth]{slab_s0_snap120_3177874_e3000.pdf}
+  \caption{Projected dark matter density fields of \GADGET, original PM,
+  and optimized PM-SONN.
+  The examples here are shown for the 0th data point in our samples of
+  configurations and cosmological parameters.
+  This corresponds to an overall simulation box size of 79 Mpc$/h$.
+  We show a zoom-in view with window length XXX Mpc$/h$, and a projected
+  slab thickness of XXX Mpc$/h$.}
+  \label{fig:denslab}
 \end{figure*}
-
-For both the $f$ and $g$ spatial optimization NNs (SONNs), the input
-features consist of tabular data, therefore we adopt multilayer
-perceptron (MLP, i.e. NN with fully connected layers) structures to
-learn the nonlinear functions.
-Both $f$ and $g$ networks have $XXX$ hidden layers with $XXX\times n_{\rm
-input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
-input layer, which is [...] and [...] for $f$ and $g$ respectively.
-The rectified linear unit (ReLU) activation function is used for hidden
-layers.
-We further apply the exponential function to the output as a regulator
-for positive output.
-Being embedded in the differentiable simulation library \pmwd, the NNs
-can be trained using gradient-based optimization methods as in usual
-deep learning tasks.
-The parameters of the NNs, which enter the force evaluation in every
-time integration step, can be treated just the same as other
-cosmological parameters.
-The objective gradients are accumulated, based on which the parameters
-are then optimized.
-
-For $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally symmetric
-about $(k_1, k_2, k_3)$, we show it as a function of the wavevector
-along axis, face, and body diagonal directions.
-
-\subsection{Symbolic Expressions}
-
-To make the SO more interpretable and applicable to other simulations,
-we fit the trained NNs with SR for analytic expressions of the $f(k_i;
-\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ functions.
-
-The data for SR fitting is generated from the trained NNs, while the
-input features for SR could be different from those of the NNs.
-Based on the NN input features, we could further design other features
-for SR fitting.
-
-\subsection{Dark Matter Density Fields}
 
@@
 
-Improvements of SO on the dark matter density fields are the first
-things to check.
-We construct these fields by scattering the particles onto a mesh.
+As discussed in \autoref{sec:intro}, our main motivation on performing
+SO of the PM method is to improve its accuracy on simulating dark matter
+distribution on small scales.
+
+The effect of SO can be visually shown by looking directly at the
+density field (\autoref{fig:denslab}) on small scales.
+We construct the dark matter density fields by scattering the particles
+onto another mesh.
+Compared to orginal PM, optimized PM with SONN is more similar to
+\GADGET\ with sharper big halos and more small-scale structures.
+
 To compare two fields $\hat f$ and $f$, we introduce for each mode the
@@
 The two fields are identical when both $T$ and $r$ are unity.
-The PM method is accurate on scales much larger than the mesh cell size,
-therefore we expect $T$ and $r$ to be very close to being unity both
-before and after the SO with force sharpening.
 
@@
 training and test data samples.
-To show different box sizes, we normalize the wavenumber by the Nyquist
-wavenumber $k_{\rm Nyquist} = \pi / l_p$ of the particle grid, where
-$l_p$ is the particle spacing.
 With discrete Fourier transform (DFT) on a regular grid of length $N_c$
@@
 simply the box size.
+To show different box sizes, we normalize the wavenumber by the Nyquist
+value $k_{\rm Nyquist} = \pi / l_c$ of the force mesh grid, where
+$l_c$ is the cell size.
 Then the normzalized wavenumber $k/k_{\rm Nyquist}$ gives the same range
@@
 normalized value is then $\sqrt{3}$.
+The original PM method is accurate on scales much larger than the mesh
+cell size, therefore $T$ and $r$ are already very close to one without
+the application of SONN.
 Compared to original PM, the TF of PM-SONN is improved to be consistent
@@
 
+\subsection{Optimized Dark Matter Halo Distributions}
+
+\subsection{Neural Networks and Symbolic Expressions}
+
 \begin{figure*}
   \centering
-  \includegraphics[width=.98\columnwidth]{slab_s0_snap120_3177874_e3000.pdf}
-  \caption{Projected dark matter density fields of \GADGET, original PM,
-  and optimized PM-SONN.
-  The examples here are shown for the 0th data point in our samples of
-  configurations and cosmological parameters.
-  This corresponds to an overall simulation box size of 79 Mpc$/h$.
-  We show a zoom-in view with window length XXX Mpc$/h$, and a projected
-  slab thickness of XXX Mpc$/h$.}
-  \label{fig:denslab}
+  \includegraphics[width=.48\columnwidth]{net_f_a1_j3177874_e3000.pdf}
+  \includegraphics[width=.48\columnwidth]{net_g_a1_j3177874_e3000.pdf}
+  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
+  functions. The solid line and shaded region indicate the median and
+  the [16, 84] percentiles of the 64 sampled configurations and
+  cosmological parameters. $a=1$,
+  mesh shape=1}
+  \label{fig:func}
 \end{figure*}
 
-The effect of force sharpening can also be visually shown by looking
-directly at the higher resolution of the density field
-(\autoref{fig:denslab}) on small scales.
-Compared to orginal PM, PM-SONN is more similar to \GADGET\ with sharper
-big halos and more small-scale structures.
-
-\subsection{Dark Matter Halos}
+For $g(k_1,k_2,k_3;\vvartheta)$, which is permutationally symmetric
+about $(k_1, k_2, k_3)$, we show it as a function of the wavevector
+along axis, face, and body diagonal directions.
+
+To make the SO more interpretable and applicable to other simulations,
+we fit the trained NNs with SR for analytic expressions of the $f(k_i;
+\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ functions.
+
+The data for SR fitting is generated from the trained NNs, while the
+input features for SR could be different from those of the NNs.
+Based on the NN input features, we could further design other features
+for SR fitting.
 
@@
 In the generation of our training data, on average each \GADGET\
-simulation takes about 2K CPU hours.
+simulation takes about 2,000 CPU hours.
 
@@
 \vspace{1em}
-\subsection{Configurations and Data Samples}
+\subsection{Training Data and Configurations}
 
@@
 \textit{\large Data Availability:}
-We make the \GSDATA\ simulation data set public at ...
+We make the \GSDATA\ simulation dataset public at ...
 

### 2024-05-09 docs/papers/sto/so.tex
Updating paper draft

@@
   \includegraphics[width=.98\columnwidth]{slab_s0_snap120_3177874_e3000.pdf}
-  \caption{Projected dark matter density fields of \GADGET, original PM,
-  and optimized PM-SONN.
+  \caption{Projected dark matter density fields of \GADGET\
+  (\textit{middle}), original PM (\textit{left}), and optimized PM-SONN
+  (\textit{right}).
   The examples here are shown for the 0th data point in our samples of
@@
 for SR fitting.
+
+We design the input features of the NNs and SR in a way such that they
+could properly fit and interpolate given our wide coverage of the
+configurations and cosmological parameters.
+Given the difference between NNs and SR, we could have different input
+features for the two.
+When generating the data for SR with the trained NNs, we could reduce or
+add more features based on the existing features, with the output
+unchanged.
 
@@
   \centering
-  \caption{Possible features that can affect force sharpening.
-  All scales are then multiplied by the wavenumber $k$ to form dimensionless
-  factors.}
+  \caption{Possible features that can affect force sharpening, detailed
+  in \autoref{sec:features}. Scales could be multiplied by the
+  wavenumber $k$ to form dimensionless factors.}
   \label{tab:feat}
@@
   \midrule
-  & particle spacing \\
+  & $l_p$: particle spacing \\
   \cmidrule(lr){2-2}
-  other scales & $l$: PM force mesh cell size \\
+  other scales & $l_c$: PM force mesh cell size \\
   \cmidrule(lr){2-2}
-  & softening length \\
+  & $l_s$: softening length \\
   \midrule
@@
 
-We design the input features of the NNs in a way such that they could
-properly fit and interpolate given our wide coverage of the
-configurations and cosmological parameters.
-
+In this section, we introduce a list of physical quantities that could
+be of possible importance in SO.
+These quantities are used in the input features for either the NNs or
+SR.
 Multiple scales and dimensionless factors could possibly affect force
-sharpening, which are denoted as $\vvartheta$ and enter the NNs of $f(k_i;
-\vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$ as input features.
-A list of these features is summarized in \autoref{tab:feat}, and here we
-present relevant definations and more details.
-
-Below are some definations related to the typical scales.
+sharpening.
+A list of these features is summarized in \autoref{tab:feat}, and here
+we present relevant definitions and more details.
+
+Below are some definitions related to the typical scales.
 \begin{itemize}
@@
 Fourier transform of the top-hat window.
-Then we have $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$.
+Then we have $R_\tophat$ defined through $\sigma_\tophat^2(R_\tophat)
+\triangleq 1$.
 \item Likewise $\sigma_\gauss^2(R)$ is defined with a Gaussian window
-$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$: $\sigma_\gauss^2(R_\gauss)
-\triangleq 1$.
+$W_\gauss(kR) = e^{-(kR)^2/2}$, and $R_\gauss$ with
+$\sigma_\gauss^2(R_\gauss) \triangleq 1$.
 \end{itemize}
 These nonlinear scales are functions of time since the linear power
-spectrum evolves with time through the growth factor $D(a)$, we also
-take the time derivatives of these scales as input features.
-Other relevant scales include the particle spacing in Lagrangian space,
-the cell size of the chosen mesh grid for force evaluation, and the
-softening length in the target \GADGET\ simulations.
+spectrum evolves with time through the growth factor $D(a)$, we could
+also take the time derivatives of these scales as possible features.
+Other relevant scales include the particle spacing $l_p$ in Lagrangian
+space, the cell size $l_c$ of the chosen mesh grid for force evaluation,
+and the softening length $l_s$ in the target \GADGET\ simulations.
 
@@
 
-Dimensionless combinations are formed by multiplying the wavevector
-components with the scales, i.e.,
-%
-\begin{equation}
-f(k_i; \vvartheta) = f(k_iR_d, k_iR_P, \cdots; G_m, \Omegam, \cdots),
-\end{equation}
-%
-\begin{equation}
-g(k_1, k_2, k_3; \vvartheta) = g(\{k_jR_d, k_jR_P, \cdots\}_{j=1,2,3};
-                                 G_m, \Omegam, \cdots).
-\end{equation}
-%
+Scales could be multiplied with the wavenumber $k$ to form dimensionless
+quantities.
+For example, the product of $k$ with the mesh cell size $l_c$ results in
+a quantity $kl_c = \pi k/k_{\rm Nyquist}$ that is invariant to different
+box sizes, which should be an important feature.
 
